### 2025-01-20 Thesis/Latex/Abstract.tex
Organizing repo and adding thesis template

\documentclass[12pt,a4paper]{report}
\usepackage[utf8]{inputenc}
\usepackage[arabic, english, french]{babel}
\begin{document}
	\newpage
	\thispagestyle{empty}
	\begin{center} \end{center}
	\vspace{0.01cm}

	\selectlanguage{arabic}
	\begin{center} {\LARGE ملخص  } \end{center}
	\par ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص ملخص .\\
	\textbf{\textit{كلمات مفتاحية:}} \textit{ملخص, ملخص, ملخص, ملخص, ملخص.}

	\vspace{2cm}

	\selectlanguage{english}
	\begin{center} {\LARGE Abstract} \end{center}
	\par Abstract Abstract Abstract  Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract.\\
	\textbf{\textit{Key words:}} \textit{Abstract, Abstract, Abstract, Abstract, Abstract.}

	\thispagestyle{empty}
\end{document}

### 2025-01-20 Thesis/Latex/Acknowledgements.tex
Organizing repo and adding thesis template

\chapter*{Acknowledgements}
\thispagestyle{empty}
%\addcontentsline{toc}{chapter}{Acknowledgments}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2025-01-20 Thesis/Latex/Chapter_2.tex
Organizing repo and adding thesis template

\chapter{Arabic Language and Machine Translation}
\pagestyle{fancy}\lhead{\textbf \footnotesize\it{Arabic Machine Translation}}
\pagestyle{fancy}\chead{} \pagestyle{fancy}\rhead{}
\pagestyle{fancy}\lfoot{\textbf {\small\it{Univ-Mascara/Computer Science: 2024}}} 
\pagestyle{fancy}\cfoot{} \pagestyle{fancy}\rfoot{\thepage}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Overview} \label{start2}
Translation has become difficult due to the intricate variations across languages \cite{dorr99}. For instance, certain words may have different meanings based on the context, or other words may not have equivalent translations in other languages. Additionally, translating idiomatic expressions calls for a thorough understanding of both the source and target languages. Further, structural variations like word order disparities between languages complicate translation.

Another difficulty is that a good translation needs to be faithful and fluent. A faithful translation accurately conveys the sense of the original text, whereas a fluent translation is easy to understand and sounds natural. A literal, faithful translation could result in an unpleasant and unnatural translation in the target language. For example, fluency rather than faithfulness is more critical when translating literary works. We might have to alter some of the meaning to keep the text flowing smoothly. Readers should feel as if it was written in their native language. The faithfulness of the translation, however, is prioritized when translating a technical manual or a legal document. Even if the translation is not fluent, it must be faithful and convey the same meaning. To produce accurate translations that balance faithfulness and fluency, human translators primarily rely on their experience, knowledge, and reasoning abilities. Due to these issues, various human translators will translate the same text in different ways.

Despite the complex linguistic distinctions, recent decades have seen significant improvements in machine translation. It is even applied in practical, real-world applications. For instance, we employ machine translation for cross-language information retrieval. It allows people to interact and obtain information in other languages.
Machine translation is also used to assist human translators. By creating a draft translation that human translators will edit, it expedites a time-consuming translation task (Plitt and Masselot, 2010). In addition, we can employ machine translation for translations that are speech- and image-centric. Speech-centric translation involves translating a text from a speech recognition system into another language before the text is fully formed. As a result, it mimics a live human interpretation.
Image-centric machine translation uses an optical character recognition system to translate the text included in images, such as billboard advertisements or street signs.

There are various methods for automating the challenging task of translation. Data-driven methods later supplanted the initial rule-based approaches. The most popular data-driven methods are Statistical Machine Translation (SMT) and Neural Machine Translation (NMT). Nevertheless, due to its remarkable successes, NMT has become state-of-the-art.

Language differences and the various machine translation approaches are briefly discussed in the sections that follow, Sections 2.2 and 2.2. Neural networks and NMT are explained in Sections 2.3 and 2.4. The evaluation of machine translation is covered in the final section, Section 2.5.

\section{Arabic Language}
Arabic is considered as one of the six official languages of the United Nation. It is the official language in 22 countries and spoken by more than 350 million people in 24 countries around the world \cite{almansor18}. The Arabic language is morphologically rich and complex, however, it is considered a low-resource language due to the lack of enough parallel dataset.

Arabic is well known for its complex morphology. It has different possibilities of word order that express the same sentence. According to word orders, Arabic sentences can be classified into 4 types: SVO\footnote{SVO: Subject-Verb-Object}, VSO, VOS and SOV \cite{alqudsi14}.

Translating the Arabic language into other languages engenders multiple linguistic problems, as no two languages can match, either in the meaning given to the conforming symbols or in the ways in which such symbols are arranged in sentences. Lexical, syntactic and semantic problems arise when translating the meaning of Arabic words into English and vice versa. Machine translation into morphologically rich languages (MRL) poses many challenges, from handling a complex and rich vocabulary that can reach hundreds of thousands or even millions, to designing adequate MT metrics that take morphology into consideration.

Fehri\cite{fehri93}, Chalabi\cite{chalabi00} and Daimi\cite{daimi01} enumerated major issues involving Arabic in the following points:
\begin{itemize}
	\item Arabic is written from right to left.
	\item There are no capital letters in Arabic.
	\item Gender is used for all nouns (there are no neutral).
	\item Some letters have different shapes depending on their location within a word. e.g. The shape of letter (\RL{`}) in the start of a word is (\RL{`-}) like in \RL{`lb}, in the middle (\RL{-`-}) like in \RL{l`b} and at the end (\RL{`} or \RL{-`}) like in \RL{bl`} or \RL{wd`}.
	\item Arabic words can be partially, fully or not vocalized. Unvocalized words may generate ambiguities for MT. See Table~\ref{tab_voc}.
	\begin{table}[!htb]
		\centering
		\begin{tabular}{lll}
			\hline\\[-1.5ex]
			\multicolumn{1}{l}{Word} & \multicolumn{1}{l}{Transliteration} & \multicolumn{1}{l}{English meaning} \\[0.5ex]
			\hline\\[-1.5ex]
			\RL{`amara} & Eamara & build / live / was populated\\[0.5ex]
			\RL{`ammara} & Eam\~~ara & Live a long time\\[0.5ex]
			\RL{`omira} & Eomira & became populous\\[0.5ex]
			\RL{`ommira} & Eom\~~ira & Given a long life\\[0.5ex]
			\RL{`omaro} & Eumaru & Omar (noun) / plural of Umrah\\[0.5ex]
			\RL{`am"ro} & Eamoru & Age of\\[0.5ex]
			\RL{`om"ro} & Eumoru & Age of\\[0.5ex]
			\RL{`omoro} & Eumuru & Age of\\[0.5ex]
			\RL{`am"ruN} & EamorN & Age\\[0.5ex]
			\RL{`om"ruN} & EumorN & Age\\[0.5ex]
			\RL{`omoruN} & EumurN & Age\\[0.5ex]
			\RL{`amaruN} & EamarN & Head cover for women\\[0.5ex]
			\RL{`amarra} & Eamar\~~a & Strong tough man /\\[0.5ex]
			&  & Longest of everything /\\[0.5ex]
			&  & Shrewd malicious guy /\\[0.5ex]
			&  & Ferocious man\\[0.5ex]
			\hline
		\end{tabular}
		\caption{Possible meanings of the unvocalized Arabic word "\RL{`mr}" (Emr)}
		\label{tab_voc}
	\end{table}

	\item Arabic has a relatively free order of words. See Table~\ref{tab0}.
	\begin{table}[!htb]
		\centering
		\begin{tabular}{lll}
			\hline\\[-1.5ex]
			\multicolumn{1}{c}{Order} & \multicolumn{1}{c}{Sentence} & \multicolumn{1}{c}{English translation} \\[0.5ex]
			\hline\\[-1.5ex]
			VSO & \RL{^srba `mro almaA'a} & Drank Omar the water\\[0.3ex]
			OVS & \RL{almA'a ^srba `mro} & The water drank Omar\\[0.3ex]
			SVO & \RL{`mro ^srba almaA'a} & Omar Drank the water\\[0.5ex]
			VOS & \RL{^srba almaA'a `mro} & Drank the water Omar\\[0.5ex]
			\hline
		\end{tabular}
		\caption{Arabic free word order}
		\label{tab0}
	\end{table}

	\item Some Arabic vocalized words may have multiple senses (polysemy) depending on its context. See Table~\ref{tab_polysem}. 
	\begin{table}[!htb]
		\centering
		\begin{tabular}{lll}
			\hline\\[-1.5ex]
			\multicolumn{1}{l}{Word} & \multicolumn{1}{l}{Transliteration} & \multicolumn{1}{l}{English meanings} \\[0.5ex]
			\hline\\[-1.5ex]
			\RL{sA'il} & sA\}il & liquid, beggar, questioner \\[0.5ex]
			\RL{`ayn} & Eayn & eye, water source, gold \\[0.5ex]
			\hline
		\end{tabular}
		\caption{Examples of vocalized Arabic words polysemy}
		\label{tab_polysem}
	\end{table}

	\item The subject can be omitted. e.g. \RL{y^srb almaA'} (He drinks the water)
	\item Some words hold the meaning of a whole sentence. e.g. \RL{fa'AsqaynAkomUh} (and We gave it to you to drink).
	\item Copula verbs "to be" and "to have" do not exist in Arabic.
	\item The three letters root system can often engender ambiguous words.
	\item Feminine nouns are often derived from masculine nouns, e.g. \RL{mhnds} (Engineer male) \RL{mhndsT} (Engineer female). In some cases they are totally different, e.g. \RL{wld} (Boy) \RL{bnt} (Girl).
	\item In English the number system moves from singular to plural form directly, however Arabic language includes dual form by suffixing morpheme (\RL{An}) or (\RL{yn}) to the singular form. e.g. \RL{mhnds} (Engineer male) \RL{mhndsAn} or \RL{mhndsayn} (Two Engineers male).
	\item The plural form of Arabic masculine nouns is the result of suffixing morpheme (\RL{wn}) or (\RL{yn}) to the singular form.
	e.g. \RL{mhnds} (Engineer male) \RL{mhndsUn} or \RL{mhndsiyn} (Engineers male).
	\item  The plural form of Arabic feminine nouns is the result of suffixing morpheme (\RL{At}) to the singular nouns. e.g. \RL{mhndsT} (Engineer female) \RL{mhndsAt} (Engineers female). 
	\item Some words have no fixed rule for their plural form. e.g. \RL{.tbyb} (Doctor) \RL{'a.tibbaA'} (Doctors).
\end{itemize}
In addition to the preceding specific challenges to MT, common standard issues also are present in Arabic language, such as:
\begin{itemize}
	\item Multi word expressions (MWE) where the meaning of words collocation varies between partially to completely not derivable from its single constituents \cite{Kordoni14}. e.g. \RL{qA`dT `skryT}, (military base) is a phrase that is highly compositional. \RL{mdynT al-mlAhy}, (amusement park), lit. "city of amusements" is an expression that shows a degree of idiomaticity. In extreme cases the meaning of the expression as a whole is utterly unrelated to the component words, such as, \RL{frs Alnby}, (grasshopper), lit. "Horse of the Prophet".

	\item Idioms and idiomatic expressions are frequently used by Arabic speakers. They are a special challenge for MT systems, because their translation mainly does not outcome literally, but logically \cite{Zaninello20}. e.g. \RL{'a.zlam mn timsA.h} , lit. "More oppressive than a crocodile", has the idiomatic meaning of (crocodile tears).

	\item Named entities (NE) refer to abstract entities in the real world such as people (PER) such as \RL{al'amIr `bd-al-qAdr} (Emir Abdelkader), places (LOC) such as \RL{mkT} (Mecca), companies, and organizations that have an appropriate name (ORG) such as \RL{yAhU} (Yahoo). It also refers to expressions of date, time, space and quantity (MISC) such as \RL{sbtmbr} 8 (September 8th), \RL{dj}100 (100DZD) or \RL{k.g}25 (25Kgs).

	Due to the rich lexical variations and the absence of capitalization in Arabic, the task of named entities recognition (NER) is more difficult. Handling NE in Arabic MT is done very carefully based on :
	\begin{enumerate}
		\item meaning translation: e.g. \RL{al-'omam al-motta.hdaT} (The United Nations).
		\item phoneme transliteration: e.g. \RL{jUjl} (Google).
		\item mixture of meaning translation and phoneme transliteration: \RL{kArUlAynA al-^smAlyaT} (North Carolina), where "North" is translated and "Carolina" is transliterated.
	\end{enumerate}   
\end{itemize}

\section{Social Media}
Social media involves online spaces where individuals can post and share information with others. They have become crucial in contemporary communication, fundamentally reshaping how individuals connect, information disseminates, and communities form. These online environments, characterized by user-generated content and interactive features \cite{boyd08},  allow for the creation of virtual spaces fostering real-time interaction and knowledge sharing \cite{yaqub23}.

\textbf{Social Networking Sites (SNS):} Platforms like Facebook prioritize building and maintaining online social connections. Users can share personal updates, engage in discussions through comments and reactions, and participate in groups centered around shared interests \cite{segesten22}. Research suggests that SNS use can foster social capital by strengthening existing relationships and enabling the formation of new ones \cite{yoon14}.

\textbf{Media-Sharing Platforms:} Platforms like Instagram and YouTube focus heavily on the creation and dissemination of multimedia content. Users can share and discover photos, videos, and live streams, with features designed to enhance content visibility through hashtags and algorithmic recommendations \cite{rogers21}. Studies highlight the increasing influence of visual communication and the emergence of new social norms surrounding content creation and consumption on these platforms \cite{lu24}.

\textbf{Messaging Applications:} Platforms like WhatsApp and WeChat enable asynchronous and real-time communication between individuals and groups. These applications offer features like text messaging, voice calls, and video conferencing, facilitating private and semi-private communication, often replacing traditional communication methods \cite{lenhart15}. Research suggests that messaging applications can strengthen interpersonal bonds and serve as crucial tools for information sharing and community mobilization \cite{pang20}.

\textbf{Professional Networking Sites:} LinkedIn serves a distinct purpose within the social media landscape, focusing on professional networking and career development. Users create profiles highlighting their skills and experience, connect with colleagues and potential employers, and discover job opportunities \cite{ellison07}. Studies suggest that LinkedIn usage can positively impact career prospects by facilitating professional connections and knowledge exchange \cite{davis20}.

Social media platforms are a goldmine for data collection. 
However, in the case of Arabic, several issues arise due to the way language is used on these platforms. 
Users often tend to write in colloquial Arabic or a mix of colloquial Arabic and English characters (often referred to as Arabizi). 
This deviates from the standard Arabic used in formal writing. 
This informality poses challenges for data collection and analysis, as algorithms need to be able to understand these diverse and non-standard language patterns.
Another important aspect of Arabic language on social media is the use of dialects.
Arabic is a highly diverse language, with many different dialects and variations.
Users may choose to write in their local dialect, which can vary significantly from Modern Standard Arabic, the standardized form of the language.
With the spread of social media, new challenges arise to the Arabic MT such as:   
\begin{itemize}
	\item  \textbf{Non-standard speech}: which encompasses dialectal languages or the various colloquial forms of standard language. This text frequently contains slang, MWE and unreasonable abbreviations \cite{Moussallem19} like as "idk\footnote{I don’t know}" and "brb\footnote{Be right back}" in English or "hmd\footnote{\RL{al.hmd lillh} (Thank God)}" and "slm\footnote{\RL{salAm} (Peace)}" in Arabic. For best translation results, the MT system is required
	to identify such argot and try to map it to the target language.

	\item \textbf{Arabizi} (sometimes known as Arabic chat alphabet, Franco-Arabic, Arabish, Araby and	Mu’arrab): which is defined as writing informal Arabic dialects in Latin characters and Arabic numerals. This new style of writing doesn’t follow any type of rules, which leads to big variations in writing nearly all Arabic words. The following words present some examples of the Arabizi issue that machine translation systems have to solve \cite{Darwish15}:
	\begin{enumerate}
		\item The Modern Standard Arabic word \RL{t.hrIr}\footnote{Liberation} has the following popular Arabizi equivalents: ta7rir, tahrir, t7rir, t7reer, ta7reer, tahreer, etc.
		\item The dialectal spellings of the MSA words \RL{lA-yl`b}\footnote{He does not play} could be \RL{mAyl`b^s}, \RL{mAbyl`b^s}, \RL{myl`b^s}, \RL{myl`b^sI}, \RL{mAyl`b^sI} etc, and the resultant Arabizi could be: mayel3absh, mayel3abch, mabyelaabsh, mabyel3absh, mayel3abshi, mayel3abchi, etc.
	\end{enumerate}
\end{itemize}

\section{Natural Language Processing}
Natural Language Processing (NLP) is a branch of computer science dedicated to facilitating communication between computers and human languages. Its methodologies find application in social media for the analysis and comprehension of the extensive textual data produced by users (Huang, 2021). Below are several applications of NLP in social media:
\begin{enumerate}
	\item \textbf{Machine Translation:} NLP techniques are utilized to translate text from one language to another, facilitating cross-lingual communication and information access. Machine translation involves converting text from a source language into a target language while preserving meaning, context, and grammatical structure.

	\item \textbf{Text Classification:} NLP techniques involve categorizing text data into predefined categories or labels based on its content. This involves tasks such as sentiment analysis, spam detection, topic classification, and intent recognition in chatbots. Text classification is essential for organizing and analyzing large volumes of text data efficiently.

	\item \textbf{Named Entity Recognition (NER):} NLP algorithms aim to identify and extract named entities from unstructured text, such as names of people, organizations, locations, dates, numerical expressions, and other specific entities. NER is crucial for various applications, including information retrieval, entity linking, and knowledge graph construction.

	\item \textbf{Speech Recognition:} NLP algorithms transcribe spoken language into text, enabling voice-controlled applications, virtual assistants, and speech-to-text transcription services. Speech recognition involves converting audio signals into textual representations, which can then be further processed or analyzed.

	\item \textbf{Information Extraction:} NLP methodologies focus on extracting structured information from unstructured text sources, enabling the retrieval of specific data points or facts from documents, web pages, or other text-based sources. Information extraction techniques involve identifying relevant entities, relationships, and events mentioned in the text.

	\item \textbf{Text Generation:} NLP models are employed to generate human-like text based on given prompts or input. This includes tasks such as automated content creation, dialogue generation, language translation, and summarization. Text generation techniques leverage deep learning architectures such as recurrent neural networks (RNNs) and transformer models.

	\item \textbf{Document Summarization:} NLP techniques condense lengthy documents or articles into shorter summaries while preserving the main points and key information. Document summarization aims to provide a concise overview of the text, facilitating easier comprehension and information retrieval.

	\item \textbf{Question Answering:} NLP systems process natural language questions and retrieve relevant answers from large text corpora or knowledge bases. Question answering involves understanding the semantics and context of the question, searching for relevant information, and generating accurate responses.

	\item \textbf{Language Understanding:} NLP models aim to understand the semantics, context, and intent behind natural language expressions. Language understanding involves tasks such as text comprehension, semantic analysis, and context modeling, enabling machines to interpret human communication more effectively.

	\item \textbf{Dialog Systems:} NLP-powered conversational agents interact with users in natural language, providing assistance, answering queries, and engaging in dialogue across various domains. Dialog systems leverage techniques such as natural language understanding (NLU) and natural language generation (NLG) to enable seamless communication between humans and machines.
\end{enumerate}
Natural Language Processing (NLP) emerges as a transformative force, enriching our lives by bridging the gap between human language and computing systems. This remarkable technology enables machines to comprehend, interpret, and generate human language, revolutionizing how we interact with technology. Through NLP, machines can analyze vast volumes of textual data from diverse sources such as social media, news articles, and customer reviews, extracting meaningful insights that shape various aspects of our lives. From facilitating accurate language translation and sentiment analysis to powering virtual assistants and chatbots, NLP permeates numerous applications, enhancing efficiency and convenience in our daily activities. Its significance lies in its ability to decipher the intricacies of human communication, empowering individuals and businesses alike with deeper understanding and actionable intelligence.

\section{Machine Translation}
Machine Translation is a procedure that uses computer pieces of software to express text from one natural language NL (SL i.e. source language) in another NL (TL i.e. target language). In any human or automated translation process, the meaning of the source sentences must be fully reproduced into the target translated sentences, which is only simple on the surface.

\begin{figure}[htbp]
	\centerline{\includegraphics[scale=.7]{Figures/fig1.4.png}}
	\caption{The Vauquois triangle, illustrating the foundations of machine translation.}
	\label{fig1.4.png}
\end{figure}

The different approaches to MT fall into three categories: methods that depend on rules and knowledge (linguistic-based). Approaches that are empirical and data-driven (corpus-based); and finally, hybrid methods.

\subsection{Linguistic Approaches}
These MT approaches attempt to formalize all the necessary knowledge required for translation, using expert methods. The "Vauquois triangle" presented in Figure~\ref{fig1.4.png} is a generic representation of these techniques.

\subsubsection{\textbf{Direct Approach}}
or Direct MT (DMT) is, the simplest MT approach. It operates at the word level, i.e. the words' translation is done word by word, just, as a dictionary does, and generally without much correspondence of their meaning \cite{okpor14}.

\subsubsection{\textbf{Rule-based MT}}
(RBMT) uses linguistic knowledge of source and target languages fundamentally collected from (bilingual) dictionaries and grammars encompassing the principal morphological, syntactic and/or semantic rules of each language
respectively \cite{okpor14}. 
RBMT approach suffers from the impossibility of writing all the rules of all the languages, because this task requires large and important linguistic knowledge. %\cite{Charoenpornsawat02}.

\subsubsection{\textbf{Interlingual MT}}
The term "Interlingua" refers to a language that serves as a bridge between two languages. In this method, SL is turned into an assistant/mediator language (representation) which is independent of the languages concerned by the translation. This auxiliary form is then used to specify the TL's translated verse. This approach focuses on a single representation for different languages \cite{Ashraf15}.

\subsubsection{\textbf{Transfer-based MT}}
(TBMT) is similar to Interlingual-MT in that it generates a translation from an intermediate structure that mimics the original sentence's meaning. The source text is translated into a less language-specific intermediate representation. This form is then translated into a target language structure with a comparable structure, and the text is generated in the target language. The source and target languages' morphological, syntactic, and/or semantic information is used in the transfer process. As a result, TBMT can make use of knowledge of both the source and target languages. \cite{DO11}.

\subsection{Corpus Approaches}
Corpus techniques use empirical methods to ensure that all linguistic knowledge is learned empirically and automatically from corpora, which are collections of parallel datasets of source and target phrases that are translated to each other.

\subsubsection{\textbf{Example-based MT}}
The main idea behind (EBMT) is analogy \cite{Nagao84}. The primary concept is to build new translations on top of current examples. Bilingual parallel corpora containing sentence pairs are used to train EBMT systems. It's used to translate similar-sounding sentences by looking for the closest source example to the source word or phrase in parallel corpora. Nagao has appropriately classified this procedure into three steps \cite{Nagao84}:
\begin{itemize}
	\item Fragments are matched against a database of real examples.
	\item Identifying the translation fragments that correlate (Alignment)
	\item Putting these together to create the target text
\end{itemize}

\subsubsection{\textbf{Statistical MT}}
(SMT) generates translation hypotheses in a target language $t$ based on a sentence in a source language $s$ with the highest conditional probability $P(t|s)$ \cite{brown88,brown93}. The translation direction will be inverted to a translation model (TM) $P(s|t)$ and a language model (LM) $P(t)$ will be included by applying the Bayes rule. The following equation~\eqref{eq:eq1} is used to optimize the likelihood of the best translation:
\begin{equation}
	\label{eq:eq1}
	t_{best}=arg_{t}max(P(t|s))=arg_{t}max(P(s|t) \times P(t))
\end{equation}
where $P(s|t)$ is the TM and $P(t)$ is the LM.

\begin{figure}[htbp]
	\centerline{\includegraphics[scale=0.3]{Figures/fig1.13.png}}
	\caption{Statistical Machine Translation approach}
	\label{fig1.13.png}
\end{figure}

SMT requires a language model, a translation model, and a decoding method in general. The TM, on the one hand, assures that the target hypothesis created matches the source sentence. The LM, on the other hand, ensures that the output is grammatically correct  (Figure~\ref{fig1.13.png}).

\section{Evaluation}
As an integral part of design science research, evaluation is crucial to assessing the quality of machine translation outputs and comparing different machine translation methods. We need to measure quality to track progress, ideally with a single score.
However, devising such a score is still an open research question (Koehn, 2020).
Nonetheless, some best practices have already been established, and in general,
there is broad consensus on how to track quality gains.
Human (Section 2.5.1) and automatic (Section 2.5.2) evaluation methods exist.
Human evaluation seems more accurate than automatic evaluation because the
translation is, after all, intended for humans. However, running a human evaluation can be time-consuming and expensive.
In practice, it can be used to compare a small number of variant systems. Therefore, automated metrics are prevalent because they can rapidly evaluate system improvements; they are also used as a loss function for training models.

\subsection{Human Evaluation}
Human evaluation can be considered the most common method of judging and measuring translation quality. The linguistics and translation experts can judge the quality of a translation system output from two corners: 
\begin{itemize}
	\item \textbf{Fluency}: The level of smoothness and coherence of the translated text to target language norms,  such as grammaticality, intelligibility and clarity.  When annotating fluency, the evaluators (fluent only in the target language) have access to only the translation being evaluated and not the source text which is not relevant to the fluency assesment. 
	\item \textbf{Adequacy}: Also known as accuracy. It stands for the correspondence of the target text to the source text, including the expressive means in translation, and how well the target text represents the informational content of the source text. In this case, the bilingual evaluators (in both the source and target languages) have access to the source text and translations being evaluated and habitually, they take into consideration the context of the sentence.   
\end{itemize}
The fluency and adequacy are usually measured on a 5-point scale, as presented in the following table \ref{tab:table1.1} \cite{mauvcec19}.

\begin{table}[ht]% <--- it say "let figure be here"
	\centering
	\resizebox{\textwidth}{!}{%
		\begin{tabular}{@{}lccccc@{}}
			\toprule
			\multirow{2}{*}{\textbf{Adequacy}} & \textbf{1}           & \textbf{2}           & \textbf{3}           & \textbf{4}           & \textbf{5}           \\ \cmidrule(l){2-6} 
			& None                 & Little meaning       & Much meaning         & Most meaning         & All meaning          \\ \midrule
			& \multicolumn{1}{l}{} & \multicolumn{1}{l}{} & \multicolumn{1}{l}{} & \multicolumn{1}{l}{} & \multicolumn{1}{l}{} \\ \midrule
			\multirow{2}{*}{\textbf{Fluency}}  & \textbf{1}           & \textbf{2}           & \textbf{3}           & \textbf{4}           & \textbf{5}           \\ \cmidrule(l){2-6} 
			& Incomprehensible     & Disfluent language   & Non-native language  & Good language        & Flawless language    \\ \bottomrule
		\end{tabular}%
	}
	\caption{Numeric scale for adequacy and fluency evaluation.}
	\label{tab:table1.1}
\end{table}

Human evaluation is intrinsically subjective and expensive in time and money. To reduce the problem of subjectivity, more experts are usually invited to evaluate the same translations in the ES, and their assessments are, eventually, justified statistically.

\subsection{Automatic Evaluation} 
Generally, translation evaluation methods are based on counting word- and/or sentence-based errors likely to be identified automatically.
Correlation with human evaluation is the measure of evaluation for metrics.
Different metrics are used in MT evaluation: BLEU \cite{papineni02}, NIST \cite{doddington02}, METEOR \cite{banerjee05}, TER \cite{snover06}, LEPOR \cite{han12} and many others. 
All of these metrics require reference translations because they confront the MT output sentences with reference translations and produce comparison scores.\\
\textbf{BLEU} metric is one of the first and most used metrics to return high correlation with human evaluation of quality. 
It measures the overlap of single words (unigrams) and \textit{n}-grams between MT output and reference translations. 
BLEU operates by not only counting matching words between the translation and reference but also accounting for n-gram alignments. 
This approach values proper word sequencing, enhancing the chances of aligning word pairs (bigrams) or longer sequences like trigrams or 4-grams. 
Additionally, multiple reference translations can be employed to assess the presence of $n$-gram matches across variations.
The BLEU score of a machine-translated output relies on the adjusted $n$-gram precision along with a brevity penalty. 
Essentially, precision measures the proportion of $n$-grams in the machine translation output that align with the reference translation.
BLEU scores are calculated across an entire test set, typically with one or more reference translations. 
However, it's uncommon in practice to utilize multiple reference translations.
\begin{equation}
	BLEU = BP * exp \sum_{n=1}^N log \frac{matching\_i\_grams}{total\_i\_grams}
%	BLEU=BP*\left(\sum_{n=1}^N w_n log\left(precision_n\right)\right)
	\label{eqn:BLEU}
\end{equation}
The brevity penalty (BP) in Equation \ref{eqn:BLEU}, which penalizes shorter output, is expressed as:
\begin{equation*}
	BP = min(1,\frac{output\_length}{reference\_length}).
\end{equation*}
%\[
%BP=\begin{cases} 
%	1 & if c > r\\ 
%	e^{(1-r/c)} & if c \leq r
%\end{cases}
%\]
\\BLEU suffers from notable drawbacks, one of which is its stringent nature. 
For instance, it incorporates trigram or 4-gram precision in its calculation. 
However, a translated sentence might lack any trigram or 4-gram matches with the reference translation, leading to a BLEU score of zero.
\section{Related Work}
The main focus of AMT (Arabic Machine Translation) research initially was on translating from Modern Standard Arabic (MSA) to English, with significantly less emphasis on the reverse direction, from English to Arabic, and even fewer efforts dedicated to translations between Arabic and other languages. 
The first direct-based machine translation system from English to Arabic was developed by Weidner Communication Inc. In 1990, Apptek introduced ArabTrans MT, a tool for translating from English to Arabic. 
Additionally, products like Al-Mutarjim Al-Araby, Al-Alamiyah, and Al-Nakheel were capable of translating between French and Arabic as well as English and Arabic. 
A critical challenge in using the Interlingua approach for AMT is the construction of representations that resolve ambiguity and accurately reflect the semantic structure of the language. 
There have been limited studies that have developed and assessed models using this method.
Table~\ref{tab_linguistic} summarizes the surveyed AMT research studies.
\begin{small}
	\begin{longtable}{|l|l|l|l|l|}
		\hline
		Year & Research & SL-TL$^{\mathrm{a}}$ & Method & Score\\
		\hline
		\endhead
		%	\hline
		%	\endfoot

		\multicolumn{5}{c}{\textbf{\textcolor{blue}{Linguistic-based AMT researches}}}\\
		%	\hline
		\multicolumn{5}{l}{\textbf{Direct-based AMT}}\\
		\hline
		2005 & Al-Taani \& Hailat \cite{altaani05}					& En$\rightarrow$Ar &-& 57.3\%\\
		2007 & Ittycheriah \& Roukos \cite{ittycheriah-roukos-07}	& Ar$\rightarrow$En &Word alignment& Bl 51.27\\
		\hline
		\multicolumn{5}{l}{\textbf{Rule-based AMT}}\\
		\hline
		1995 & Mankai \& Mili \cite{mankai95}				& Ar$\rightarrow$En/Fr &-&\\
		2008 & Salem et al. \cite{salem08} 					& Ar$\rightarrow$En &-&\\
		2008 & Nguyen \& Vogel \cite{nguyen08}				& Ar$\rightarrow$En &-& 56.03\\
		2008 & Samy	\& & Ar-Sp-En &-&\\
		 & Gonz{\'a}lez-Ledesma \cite{samy08}	&  &-&\\
		2008 & AbuShuqier \& Sembok \cite{abushquier08} 	& En$\rightarrow$Ar &-& 96.1\%\\
		2009 & Elming \& Habash \cite{elming09}				& En$\rightarrow$Ar &-&\\
		2009 & Besançon et al. \cite{besanccon09} 			& En/Fr$\rightarrow$Ar &-&\\
		2012 & Salloum \& Habash \cite{salloum12}			& DA$\rightarrow$MSA &-&\\
		2020 & Sghaier \& Zrigui \cite{sghaier20}			& DA$\rightarrow$MSA &-& Bl 55.22\\
		\hline
		\multicolumn{5}{l}{\textbf{Interlingua-based AMT}}\\
		\hline
		2002 & Soudi et al. \cite{soudi02}					& En$\rightarrow$Ar &-&\\
		2006 & Shaalan et al. \cite{shaalan06}				& En$\rightarrow$Ar &-&\\
		2008 & Bouillon et al. \cite{bouillon08}			& Jp$\leftrightarrow$Ar &-&\\
		2014 & Al Ansary \cite{AlAnsary14}					& En$\rightarrow$Ar &Universal Networking Language&\\
		\hline
		\multicolumn{5}{l}{\textbf{Transfer-based AMT}}\\
		\hline
		2002 & Attia \cite{attia02}							& En$\rightarrow$Ar &Agreement features&\\
		2004 & Shaalan et al. \cite{shaalan04}				& En$\rightarrow$Ar &-& 92\%\\
		2010 & Shirko et al. \cite{shirko10}				& Ar$\rightarrow$En &-& 94.6\%\\
		2010 & Shaalan et al. \cite{shaalan10a}				& En$\leftrightarrow$Ar &-& 0.450%4
		/0.458%1
		\\
		2011 & Hatem et al. \cite{hatem11}					& En$\rightarrow$Ar &Morphological analysis&\\
		\hline
		\multicolumn{5}{c}{\textbf{\textcolor{blue}{Corpus-based AMT researches}}}\\
		%	\hline
		\multicolumn{5}{l}{\textbf{Example-based AMT}}\\
		\hline
		2002 & Guidere \cite{guidere02}					& Fr-Ar &-&\\
		2011 & Bar \& Dershowitz \cite{Bar11}			& Ar$\rightarrow$En &Verb paraphrases& 23.98\\
		2012 & Bar \& Dershowitz \cite{Bar12}			& Ar$\rightarrow$En &Semantic equivalents&\\
		2012 & Cavalli-Sforza \& Phillips \cite{Cavalli-Sforza12}	& Ar$\rightarrow$En &Morphological analysis&\\
		2014 & El-Shishtawy 		& En$\rightarrow$Ar &Template-based syntactic &\\
		&\& El-Sammak \cite{elshishtawy14}		& & matching&\\
		\hline
		\multicolumn{5}{l}{\textbf{Statistical-based AMT}}\\
		\hline
		2006 & Hasan et al. \cite{hasan06}				& Ar$\rightarrow$Fr &Pre-/Post-processing& 40.8\%\\
		2007 & Diab et al. \cite{diab07}				& Ar$\rightarrow$En &Pre-/Post-processing& 45.38\%\\
		2007 & Sarikaya \& Deng \cite{sarikaya07}		& En$\rightarrow$Ar &POS tag$^{\mathrm{b}}$/CDW$^{\mathrm{d}}$&+0.3\\
		2009 & Badr et al. \cite{badr09}				& En$\rightarrow$Ar &Syntactic phrase reordering&Bl 32.46\\
		%s		2009 & Mostefa et al. \cite{mostefa09}					& Ar-Fr-En\\[0.3ex]
		2009 & Habash \& Hu \cite{habash-hu09}		& Ar$\rightarrow$En$\rightarrow$Cn &MT Evaluation&Bl +1.1\\
		2010 & Carpuat et al. \cite{carpuat10}			& Ar$\rightarrow$En &Word reordering&Bl 51.70\\
		2010 & Ghurab et al. \cite{ghurab10}			& Ar$\leftrightarrow$Cn &-&0.805%3
		/0.696%5
		\\
		2010 & Bisazza and Federico \cite{bisazza10}	& Ar$\rightarrow$En &Word reordering& 48.96\\
		2017 & Durrani et al. \cite{durrani17}			& Ar$\leftrightarrow$En &MT Evaluation& Bl +4\\	
		2017 & Mallek et al. \cite{mallek17}			& Ar$\leftrightarrow$En &Pre- processing& Bl 10.98\\
		2017 & Ebrahim et al. \cite{ebrahim17}			& En$\rightarrow$Ar &MWE Detection& 19.31/19.22\\
		2019 & Aqlan et al. \cite{aqlan19a}				& Ar$\rightarrow$Cn &Morpho/Vocab/POS tag& Bl 19.40\\
		\hline
		\multicolumn{5}{c}{\textbf{\textcolor{blue}{Hybrid AMT}}}\\
		\hline
		2004 & Alsharaf et al. \cite{alsharaf04}	& Fr$\rightarrow$Ar& Direct+Transfer+Pivot+SMT &\\
		2008 & Toutanova et al. \cite{toutanova08}	& En$\rightarrow$Ar& Inflection prediction models &Bl 36.92\\
	 & 	& &  + SMT &\\
		2008 & Hatem \& Nassar \cite{hatem08}		& En$\rightarrow$Ar& Rule-based + Example-based & 68\%\\
		2009 & Matusov et al. \cite{matusov09} 		& Ar$\rightarrow$En& Multi-engine (5 MT systems) &\\
		2009 & Habash et al. \cite{habash09}		& Ar$\rightarrow$En& SMT + Rule-based &Bl 0.4162\\
		2010 & Al Dam \& Guessoum \cite{aldam10}   	& En$\rightarrow$Ar& Transfer-based + ANN &56\%\\
		2010 & Sawaf \cite{sawaf10}					& DA$\rightarrow$MSA& Rule-based+SMT & 42.1\%\\
		2011 & Alawneh \& Sembok \cite{alawneh11}	& En$\rightarrow$Ar& Rule-based + Example-based &\\
		2012 & Shaalan \& Hany \cite{shaalan12}		& Ar$\rightarrow$En& Rule-based + Example-based & WER 88\%\\
		2014 & Akeel \& Mishra \cite{akeel14}   	& En$\rightarrow$Ar& Rule-based + ANN & 0.6\\
		2015 & Mohamed \& Sadat	\cite{mohamed15}	& Ar$\rightarrow$Fr& Morphological rule + SMT & 34.7\%\\
		2015 & Zantout \& Guessoum \cite{zantout15}	& Ar$\rightarrow$En& Transfer-based + ANN &64.50\%\\
		\hline
		\multicolumn{3}{l}{$^{\mathrm{a}}$Source language-Target language}&
		\multicolumn{2}{l}{$^{\mathrm{b}}$Part Of Speech Tagging}\\
		\multicolumn{3}{l}{$^{\mathrm{c}}$Context Dependent Words}\\
		\caption{Linguistic-, Corpus-based and Hybrid AMT researches}
		\label{tab_linguistic}\\
	\end{longtable}
\end{small}

\section{Summary}
In conclusion, this chapter has delved into the multifaceted challenges and intricacies surrounding Arabic language processing, with a specific focus on machine translation (MT).
We have explored the intricate aspects of Arabic, ranging from its morphology and syntax to its diverse array of dialects, all of which present formidable hurdles for natural language processing (NLP) and MT systems alike. 
Additionally, we have scrutinized the influence of social media on Arabic NLP and MT, taking into account the unique linguistic characteristics and informal language usage prevalent on online platforms.

Our examination has encompassed both linguistic and corpus-based approaches to Arabic MT, underscoring the necessity of tailored resources and datasets to capture the language's nuances effectively. 
Moreover, we have delved into various MT evaluation methods, encompassing both human and automatic assessment techniques.

Lastly, we have surveyed the related work in the field, providing insights into recent advancements, methodologies, and the challenges encountered in Arabic language MT research, thus offering valuable perspectives on the current landscape and future trajectories of the field.

### 2025-01-20 Thesis/Latex/Chapter_3.tex
Organizing repo and adding thesis template

\chapter{Neural Networks and Machine Translation}
\pagestyle{fancy}\lhead{\textbf \footnotesize\it{Neural Networks and Machine Translation}}
\pagestyle{fancy}\chead{} \pagestyle{fancy}\rhead{}
\pagestyle{fancy}\lfoot{\textbf {\small\it{Univ-Mascara/Computer Science: 2024}}} 
\pagestyle{fancy}\cfoot{} \pagestyle{fancy}\rfoot{\thepage}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Overview}\label{start3}
In this chapter, we embark on an in-depth exploration of neural networks, pivotal to understanding the intricate workings of machine translation. 
Beginning with a thorough examination of the foundational feed-forward networks (FFNs), recurrent neural networks (RNNs), long short-term memory networks (LSTMs), and gated recurrent units (GRUs), we lay the groundwork for comprehending the neural machine translation (NMT) process.

We delve into the nuances of training NMT models, and the decoding phase, where trained models generate translations from input sequences. 
Finally, we conducted a comprehensive survey of recent advancements and methodologies in neural machine translation research, with a focus on exploring related works specifically dedicated to the Arabic language context.

\section{Neural Networks}
Neural networks are the foundational computational mechanisms driving Neural Machine Translation, employing complex algorithms to mimic human language understanding and generation. 
At the core of these networks is the node, or processing unit, which functions similarly to neurons in the human brain. 
Each node is designed to receive inputs—typically a vector of real-valued numbers—process these inputs through a series of mathematical operations, including weighted sums and activation functions, and then produce an output that can be passed on to subsequent layers or nodes in the network.
The following subsections will delve into the different types of neural networks, beginning with the simplest form: the feed-forward neural network.

\subsection{Feed-Forward Neural Networks}
A feed-forward neural network is a basic type of neural network where information moves in one direction from the input to the output layer without any loops or cycles back to previous layers. 
Outputs from each node are forwarded upwards to the subsequent layer without any feedback to lower levels. 
An illustration, referred to as Figure~\ref{fig:ffn}, shows a FFN consisting of two layers. 
These networks are composed of three distinct types of nodes: input, hidden, and output. 
The input nodes receive individual scalar values, and notably, among these input nodes is a constant bias node, labeled $x_0$, which is always set to the value of 1. 
In the hidden layer, nodes calculate a weighted sum of their inputs and then apply a nonlinear function to this sum to determine their output.
\begin{figure}[h]
	\centerline{\includegraphics[width=0.7\linewidth]{Figures/FFN}}
	\caption{A feed-forward network consisting of two layers includes an input layer $x$, a hidden layer $h$, and an output layer $y$.}
	\label{fig:ffn}
\end{figure}
In a standard neural network structure, every layer is comprehensively interconnected, allowing nodes to receive inputs from all nodes in the preceding layer. 
These networks are characterized by their depth, with numerous layers contributing to their complexity. 
Hidden nodes within these layers act as automatic feature detectors, eliminating the need for manual feature identification by learning to recognize relevant patterns in the input data through training. 
Each hidden node is defined by its parameters, including a weight vector and a bias term. 
The parameters for the entire hidden layer are aggregated into a weight matrix $U$, where each weight $u_{jk}$ in $U$ corresponds to the connection strength from the $k^{th}$ input node $x_k$ to the $j^{th}$ hidden node $h_j$, and a bias vector that applies to the entire layer.

A notable feature of neural networks is the incorporation of non-linear activation functions, with the rectified linear unit ($ReLU$) standing out for its simplicity and widespread usage. 
$ReLU$ is particularly efficient to compute, as depicted in Figure~\ref{fig:relu} (a). 
This function returns the value of $z$ when $z$ is greater than 0 and returns $0$ when $z$ is less or equal to 0.
\begin{figure}[h]
	\centering
	\includegraphics[width=0.7\linewidth]{Figures/Relu}
	\caption{Common activation functions utilized in neural networks.}
	\label{fig:relu}
\end{figure}
The $sigmoid$ (or logistic) function, another frequently utilized activation function, has been illustrated in Figure~\ref{fig:relu} (b).
An activation function similar to the sigmoid but more widely employed is the hyperbolic tangent ($tanh$), depicted in Figure~\ref{fig:relu} (c).

The computation of the hidden layer in the basic feed-forward network can be performed efficiently using straightforward matrix operations, involving three main steps:
\begin{itemize}
	\item The multiplication of the weight matrix by the input vector $x$,
	\item Addition of the bias vector, and
	\item Application of the activation function $f$, such as $ReLU$, $sigmoid$, or $tanh$.
\end{itemize}
Hence, the neural network depicted in Figure~\ref{fig:ffn} can be expressed mathematically as follows: 
\begin{itemize}
	\item A set of input nodes represented by the vector $x = (x_1, x_2, x_3, . . . , x_n)^T$;
	\item A set of hidden nodes represented by the vector $h = (h_1, h_2, h_3, . . . , h_l)^T$;
	\item A set of output nodes represented by the vector $y = (y_1, y_2, y_3, . . . , y_m)^T$;
	\item A weight matrix connecting input nodes to hidden nodes denoted by $U = u_{jk}$;
	\item A weight matrix connecting hidden nodes to output nodes denoted by $W = w_{ij}$.
\end{itemize}
The computation of the hidden layer output, represented by the vector $h$, is performed according to Equation~\ref{eqn:hj}, incorporating the activation function $f$.
\begin{equation}
	h_j = f(\sum_{k}u_{jk}x_k)
	\label{eqn:hj}
\end{equation}
The value $h$ obtained serves as a representation of the input. 
Subsequently, the output layer's function is to utilize this revised representation $h$ and calculate a conclusive output, as described in Equation~\ref{eqn:yi}.
\begin{equation}
	y_i = \sum_{j}w_{ij}h_j
	\label{eqn:yi}
\end{equation}
In numerous instances, this output may initially be a real-valued number; however, it is often transformed into a probability distribution using a $softmax$ function. For a vector $y$ with a dimensionality of $d$, the $softmax$ function is defined as shown in Equation~\ref{eqn:softmax_yi}, where $1 \leq i \leq d$.
\begin{equation}
	softmax(y_i) =\frac{e^{y_i}}{\sum_{j}e^{y_j}}
	\label{eqn:softmax_yi}
\end{equation}
The $softmax$ function operates on a vector $y = [y_1, y_2, y_3, . . . , y_m]$ of arbitrary values, transforming them into a probability distribution where each value falls within the range of (0, 1) and the sum of all values equals one.

\subsection{Recurrent Neural Networks}
A neural network containing a cycle inside its network connections is called a recurrent neural network (RNN). 
Its preceding outputs directly or indirectly influence the value of a node. 
Figure~\ref{fig:rnn} illustrates the structure of a simple RNN based on \cite{elman90}. 
Similar to conventional FFNs, the values for a layer of hidden nodes are calculated by multiplying an input vector representing the current input, $x$, by a weight matrix and then passing the result through a non-linear activation function. 
The associated output, $y$, is then determined using the hidden layer, which comprises the hidden nodes. 
The context layer is where it differs from a FFN the most. 
It keeps the previous values and sends them to the appropriate nodes in the hidden layer. 
This layer uses the hidden layer’s value from the previous time step as input to the computation at the hidden layer.
\begin{figure}[h]
	\centering
	\includegraphics[width=0.7\linewidth]{Figures/RNN}
	\caption{A basic recurrent neural network.}
	\label{fig:rnn}
\end{figure}

\subsubsection{Stacked Recurrent Neural Networks} 
Stacked recurrent neural networks (RNNs) consist of multiple layers, where the output of one layer serves as the input to the next layer. 
Information flows from lower layers to higher layers, and the final output is generated by the top layer. 
Stacked RNNs have shown improved performance in neural MT compared to single-layer networks \cite{bahdanau15}. 
This improvement can be attributed to the network's ability to create representations at different levels of abstraction across the layers. 
However, the optimal number of stacked RNNs depends on the availability of training data. 
While abundant data can lead to better generalization, limited data may result in overfitting \cite{lample18}. 
Moreover, increasing the number of stacked layers escalates the training costs.

\subsubsection{Bidirectional Recurrent Neural Networks}
In recurrent networks, the hidden state at any given time encapsulates all information about the sequence up to that point, serving as the left context for the current input. 
In neural machine translation (NMT), having access to the entire input sequence simultaneously suggests the importance of utilizing the context to the right of the input. 
Training an RNN in reverse on the input sequence is one approach to retain this information. 
Combining the forward and backward networks results in a bidirectional RNN, where two independent RNNs process input from start to finish and finish to start, respectively \cite{schuster97}. 
The outputs of both networks are then merged to create a unified representation incorporating both left and right input contexts at each time step. 
Methods such as concatenation, element-wise operations, or averaging can be employed to combine the outputs of the forward and backward passes. 
This ensures that the information on both sides of the current input is captured in the output at each time step.
Figure~\ref{fig:rnn-brnn} illustrates the conventional RNN structure alongside the BRNN architecture over a timeline. In the RNN, only a forward hidden layer is present, whereas in the BRNN, both forward and backward hidden layers are depicted.

\begin{figure}[h]
	\centering
	\includegraphics[width=0.9\linewidth]{Figures/RNN-BRNN}
	\caption{Architecture of RNN and BRNN shown over a period of time.}
	\label{fig:rnn-brnn}
\end{figure}

\subsection{Long Short-Term Memory}
One of the main limitations of RNNs in machine translation is their difficulty in handling long-distance dependencies. Distant words play a crucial role in translation tasks, as demonstrated in the following example:
\begin{quotation}
	The lengthy \textit{sentence} composed of numerous words spanning multiple lines still \textit{poses} significant challenges for MT.
\end{quotation}
In this example, the verb \textit{poses} depends on the subject \textit{sentence}, which is separated by a long subordinate clause.

While RNNs can access preceding sequences, the knowledge retained in hidden states tends to be relatively localized. 
Consequently, more complex network structures have been developed to address the difficulty of maintaining context over extended periods. 
It is essential for the network to discard irrelevant information while retaining data crucial for future decision-making. 
Long Short-Term Memory (LSTM) and Gated Recurrent Units (GRUs) are the predominant approaches employed to achieve this objective.

LSTM networks, introduced by \cite{hochreiter97}, address the challenge of managing context by dividing it into two sub-problems: removing unnecessary information and incorporating data crucial for future decisions. 
Unlike incorporating a fixed strategy into the architecture, the key lies in learning to handle context dynamically. 
LSTMs accomplish this by expanding the architecture with an explicit context layer and employing specialized neural units with gates to control the flow of information within the network layers. 
These gates utilize additional weights to sequentially process input, previous hidden layers, and previous context layers.

\subsection{Gated Recurrent Units}
GRU, akin to LSTM but with fewer parameters, offers the advantage of reduced training costs by requiring fewer parameters. 
This reduction in parameters is achieved by eliminating the need for a separate context vector and reducing the number of gates, thereby alleviating the computational burden compared to LSTM. 
Both GRU and LSTM employ sigmoid functions in their gates to either allow information with values close to one to pass through or block information with values close to zero.

In contrast to simple FFNs, LSTMs and GRUs feature more complex neural nodes, as depicted in Figure~\ref{fig:lstm-vs-gru}, with inputs and outputs connected to each node type. 
The additional complexity within LSTM and GRU nodes is primarily contained within the nodes themselves. 
The only external complexity introduced by LSTM, beyond the basic recurrent node, is the availability of the other context vector as an input and output. 
Conversely, GRU nodes exhibit similar input and output architecture to the simple recurrent node.
\begin{figure}[h]
	\centering
	\includegraphics[width=0.7\linewidth]{Figures/LSTM-vs-GRU}
	\caption{Diagram illustrating the network structure of LSTM in the first panel (a) and GRU in the second panel (b)}
	\label{fig:lstm-vs-gru}
\end{figure}

\section{Neural Machine Translation}
The standard architecture for Neural Machine Translation (NMT) is the encoder-decoder network. 
This architecture operates on the fundamental principle of employing an encoder network to convert a sequence of words from the source language sentence into a contextual representation. 
Subsequently, a decoder utilizes this representation to generate an output sequence, essentially providing a plausible translation into the target language. 
Figure~\ref{fig:encoder-decoder} illustrates the encoder-decoder architecture at its most abstract level, comprising three key components: an encoder, context, and decoder. 
The encoder takes in a source language sentence as an input sequence of words, $x_1, x_2, ..., x_m$, and produces a corresponding sequence of contextualized representations known as a context vector. 
This context vector captures the essence of the input and serves as input to the decoder. 
Finally, the decoder utilizes the context vector to generate the most probable translation as a sequence of words, $y_1, y_2, ..., y_n$. 
Encoders and decoders can be implemented using either RNNs or Transformers.
\begin{figure}[h]
	\centering
	\includegraphics[width=0.7\linewidth]{Figures/Encoder-Decoder}
	\caption{Basic Encoder-Decoder architecture}
	\label{fig:encoder-decoder}
\end{figure}

Figure 2.7 illustrates a simplified version of the RNN-based encoder-decoder architecture. 
The encoder processes the input sequence $x$ with the objective of generating a representation of the input. 
This representation is captured in the final hidden state of the encoder, represented as $h^e_n$. 
Subsequently, this context representation, denoted as $c$, is transferred to the decoder. 
Upon receiving this state, the decoder utilizes it to initialize its initial hidden state. 
The first decoder cell employs $c$ as its initial hidden state, $h^d_0$, and proceeds to generate a sequence of outputs iteratively, one element at a time, until an end-of-word marker, </s>, is produced. 
Consequently, each hidden state is dependent on the preceding hidden state and the output generated in the preceding state. 
The embedding layer is composed of word embeddings, which adhere to the notion that semantically related words in similar contexts should possess similar representations. 
Ultimately, the output $y$ at each time step involves a $softmax$ computation over the vocabulary, $V$. The most probable output at each time step can be determined by computing the $argmax$ over the $softmax$ output as per Equation~\ref{eqn:yt}. 
Although Figure~\ref{fig:rnn-encoder-decoder} illustrates a single network layer, stacked and bi-directional networks are typically employed for both the encoder and decoder in practical implementations.
\begin{equation}
	\hat{y}_t = argmax_{w \in V}P(w|x, y_1, . . . , y_{t-1})
	\label{eqn:yt}
\end{equation}
\begin{figure}[]
	\centering
	\includegraphics[width=1.0\linewidth]{Figures/RNN-Encoder-Decoder}
	\caption{The basic RNN-based encoder-decoder architecture.}
	\label{fig:rnn-encoder-decoder}
\end{figure}

A significant drawback of this architecture lies in its inability to evenly represent information from the beginning of a sentence, particularly evident in lengthy sentences. 
To address this issue, the attention mechanism emerges as a solution. 
It enables the decoder to access information from all the hidden states of the encoder, rather than solely relying on the last hidden state. 
The concept behind the attention mechanism is to create a singular context vector, denoted as $c$, by computing a weighted sum of all the encoder's hidden states. 
These weights concentrate on a specific segment of the source text pertinent to the token generated by the decoder. 
Notably, the context vector produced by the attention mechanism is dynamic, varying for each decoded token. 
By introducing the attention mechanism, the static context vector is replaced with $c_i$ a dynamically derived counterpart from the encoder's hidden states at each decoding step $i$, ensuring comprehensive consideration of all encoder states.

\subsection{Training Neural Machine Translation Models}
Neural Machine Translation (NMT) operates within a supervised machine learning framework, where the correct output, denoted as $y$, is known for each input observation, $x$. 
The system's output, represented as $\hat{y}$, serves as an estimate of the actual $y$. 
During the training phase, the objective is to adjust the parameters within each layer so that $\hat{y}$ closely resembles $y$ for every training instance.
To accomplish this, a loss function is employed to quantify the disparity between the system's output and the actual output. 
The optimization process, typically facilitated by the gradient descent algorithm, seeks to minimize this loss function by adjusting the parameters. 
Gradient descent utilizes the gradient of the loss function, computed as the partial derivative of the loss function with respect to each parameter, to iteratively update the model's parameters. 
However, in the realm of neural networks, characterized by numerous layers and millions of parameters, computing the partial derivative of a weight in one layer concerning the loss associated with another layer necessitates employing techniques such as error back-propagation or reverse differentiation. 
These methodologies are crucial for efficiently navigating the complex landscape of neural network optimization.

Moreover, NMT models undergo end-to-end training, where each training instance comprises a pair of sentences, one in the source language and the other in the target language. These sentence pairs are concatenated with a designated separator token, <s>, to form the training data. Typically, this training data is sourced from established datasets containing aligned pairs of sentences, known as parallel corpora. Optimization within the NMT framework is characterized by a non-convex problem landscape.
Nevertheless, there exist several best practices for effectively training NMT models. For instance, it is advisable to initialize the model weights with small random numbers and to employ random seeds for reproducibility. Additionally, normalizing the input values to have zero mean and unit variance can be beneficial in enhancing training stability and convergence.

NMT model training involves multiple epochs, representing complete iterations over the training data. Typically, as training progresses, the error on the training set steadily decreases. However, a key challenge arises when the model starts to overfit, meaning it memorizes the training data excessively without generalizing well to unseen examples. To diagnose overfitting, a separate set of examples known as the development (or validation) set is used, which is not involved in the training process. Monitoring the error on this development set over the course of training reveals a point where the error begins to increase, indicating overfitting.
In theory, training should halt when the minimum error on the development set is reached. However, in practice, this is more nuanced for NMT models due to the non-deterministic nature of training and the lack of clear convergence or overfitting points, particularly with sizable datasets. While some studies suggest stopping criteria based on approximate training durations or a fixed number of epochs, most NMT research does not specify precise stopping conditions. To mitigate overfitting, various regularization techniques are employed, such as dropout, which randomly disables some nodes and connections during training.

In addition to regularization, hyperparameter tuning plays a crucial role in NMT architecture design. These hyperparameters, which include the learning rate, mini-batch size, number of layers, hidden nodes per layer, and activation functions, are chosen by the architect to optimize model performance.
\subsection{Decoding}
The decoding (inference) algorithm employed to generate translations (as illustrated in the Figure~\ref{fig:rnn-encoder-decoder}) faces a challenge. 
Opting for the single most probable word at each step implies a 1-best greedy search, where a greedy algorithm makes locally optimal decisions. 
However, there are instances where following a sequence of words leads to the realization that an earlier mistake was made. 
In such cases, the best sequence may initially comprise less probable words that are refined by subsequent words in the context of the entire output. 
For instance, when translating an idiomatic expression, the initial words chosen may seem unusual (e.g., "\textit{take your hat off}" for "\textit{respect, admire, or congratulate someone}").

In NMT decoding, the predominant method employed is known as beam search. 
Unlike selecting the best word at each step, beam search maintains $k$ possible words, where $k$ denotes the beam size or beam width. 
Initially, a $softmax$ is computed over the entire vocabulary to assign probabilities to each word. 
From this $softmax$ output, the k-best candidates are selected, forming hypotheses. 
Each hypothesis represents an initial output sequence along with its probability. 
These top $k$ hypotheses are progressively expanded by passing them through different decoders in subsequent steps. 
At each decoder, a $softmax$ is generated over the entire vocabulary to determine the next word for each hypothesis. 
Subsequently, each of the $k * V$ potential sequences is scored based on the product of the current word's probability and the path leading to it. 
These hypotheses are then pruned down to the $k$ best ones to ensure there are never more than $k$ hypotheses at any given time. 
This iterative process continues until a complete translation candidate marked by "</s>" is generated, signifying the end of a hypothesis. 
The finished hypothesis is then removed, and the beam size is reduced by one. 
This search persists until the beam size diminishes to zero, resulting in $k$ hypotheses. 
The best translation is determined from the complete hypothesis with the highest score. 
Figure~\ref{fig:beam} illustrates this process using a beam size of six.
When evaluating the top paths, we assess each based on the product of their word prediction probabilities. 
In practical applications, we tend to achieve improved outcomes by normalizing the score according to the length of the translation produced, which involves dividing the score by the number of words. 
This normalization process occurs after the search is finalized.
\begin{figure}
	\centering
	\includegraphics[width=0.7\linewidth]{Figures/beam}
	\caption{Beam search decoding with a beam size of six.}
	\label{fig:beam}
\end{figure}

\section{Related Work}
In the recent years, Research in AMT has predominantly centered on Neural MT, targeting translation from Arabic to English and various other languages. 
Neural machine translation, in particular, has become a compelling substitute for phrase-based Statistical Machine Translation (SMT), especially for the Arabic language.
Numerous studies are currently focusing on the translation between Dialectal Arabic and Modern Standard Arabic (MSA) in low-resource settings, both from Dialectal Arabic to MSA and vice versa.
Table~\ref{tab_ANMT} summarizes the surveyed Neural-based AMT research studies.
\begin{small}
	\begin{longtable}{|l|l|l|l|l|}
		\caption{Neural-based AMT researches}
		\label{tab_ANMT}\\
		\hline
		Year & Research & SL-TL$^{\mathrm{a}}$ & Method & Score\\
		\hline
		\endhead
		%	\hline
		%	\endfoot

%		\multicolumn{5}{c}{\textbf{Corpus-based AMT researches}}\\
		%	\hline
		\multicolumn{5}{l}{\textbf{Neural-based AMT}}\\
		\hline
		2016 & Almahairi et al. \cite{almahairi16}		& Ar$\leftrightarrow$En &MT Evaluation& 49.7/33.62\\
		2016 & Guzm{\'a}n et al. \cite{guzman16}		& En$\rightarrow$Ar &Morpho-syntactic analysis& +75\%\\
		2017 & Belinkov et al. \cite{belinkov17}		& Ar$\leftrightarrow$En &Morpho/Vocab/Factored NMT&Bl 28.42\\
		2017 & Durrani et al. \cite{durrani17}			& Ar$\leftrightarrow$En &MT Evaluation& Bl +4\\	
		2017 & Choi et al. \cite{choi17}				& Kor$\rightarrow$Ar &Corpus extension&Bl 27.07\\
		2018 & Almansor \& Al-Ani \cite{almansor18}		& Ar$\rightarrow$En &Low-resource& Bl +10\\
		2018 & Shapiro \& Duh \cite{shapiro18}			& Ar$\rightarrow$En &Morpho/Vocab/Factored NMT&Bl 29.10\\
		2018 & Alrajeh \cite{alrajeh18}					& Ar$\rightarrow$En &MT Evaluation& Bl +13\\
		2018 & Alkhatib \& Shaalan \cite{alkhatib18}	& Ar$\leftrightarrow$En &Paraphrasing model& 86.9\%/94.1\%\\
		2018 & Baniata et al. \cite{baniata18}			& DA$\rightarrow$MSA &Multitask Learning& 0.41/0.30\\
		2019 & Aqlan et al. \cite{aqlan19}				& Ar$\rightarrow$Cn &Pre-/Post-processing& Bl 24.66\\
		2019 & Oudah et al. \cite{oudah19}				& Ar$\rightarrow$En &Pre-/Post-processing& 55.64/53.54\\
		2019 & Hadj Ameur et al. \cite{hadjameur19}		& Ar$\leftrightarrow$En &Pre-/Post-processing&29.76/20.41\\	
		2019 & Gashaw \& Shashirekha \cite{gashaw19}	& Amh$\rightarrow$Ar &MT Evaluation& Bl 12\\	
		2020 & Ji et al. \cite{ji20}					& $\rightarrow$Ar &Multilingual/Low-resource& 25.49\\
		2021 & Bensalah et al. \cite{bensalah21}		&  Ar$\rightarrow$En &Pre-processing/MT Evaluation& 41.87\%\\
		2021 & Berrichi \& Mazroui \cite{berrichi21}	& Ar$\leftrightarrow$En & Morpho/Vocab/Factored NMT&Bl 33.02\\	
		2021 & Moukafih et al. \cite{moukafih21}		& DA$\leftrightarrow$Ar &Multi-Task learning/MT Eval&Bl 35.06\\	
		2021 & Nagoudi et al. \cite{nagoudi21}			& DA$\rightarrow$En &Pre-processing/MT Evaluation& Bl 25.72\\	
		2021 & Stergiadis et al. \cite{stergiadis21}	& Ar$\rightarrow$En &Multidimensional Tagging&Bl 50.84\\	
		2022 & Bensalah et al. \cite{bensalah22}		& Ar$\rightarrow$En &MT Evaluation&Bl 0.575\\	
		2022 & Slim et al. \cite{slim22}				& DA$\rightarrow$MSA &Transductive transfer learning&Bl 35.87\\
		2022 & Gaser et al. \cite{gaser22}				& DA$\rightarrow$En &Segmentation& chrF2 51.3\\
		2022 & Nagoudi et al. \cite{nagoudi22}			& 20 lang$\rightarrow$Ar &Multi-lingual model& Bl 32.07\\
		2022 & Hameed et al. \cite{hameed22}			& Ru$\rightarrow$Ar & MT Evaluation& 51.57\% \\
		2022 & Baniata et al. \cite{baniata22}			& DA$\rightarrow$Ar & RPE$^{\mathrm{b}}$+sub-word & Bl 66.87\\
\hline
\multicolumn{3}{l}{$^{\mathrm{a}}$Source language-Target language}&
\multicolumn{2}{l}{$^{\mathrm{b}}$Reverse Positional Encoding}\\
\end{longtable}
\end{small}

\section{Summary}
In this chapter, we embarked on a comprehensive exploration of neural networks, particularly focusing on feedforward networks (FFN), recurrent neural networks (RNN), long short-term memory networks (LSTM), and gated recurrent units (GRU). 
These architectures lay the foundation for understanding the mechanisms behind neural machine translation (NMT).

We delved into the intricacies of training NMT modelsand the decoding phase, where the trained models generate translations from input sequences. 
Finally, we delved into related works in the field, dedicating special attention to advancements, methodologies, and challenges specific to neural machine translation in the context of the Arabic language.

### 2025-01-20 Thesis/Latex/Chapter_4.tex
Organizing repo and adding thesis template

\chapter{Large Language Models}
\pagestyle{fancy}\lhead{\textbf \footnotesize\it{Large Language Models}}
\pagestyle{fancy}\chead{} \pagestyle{fancy}\rhead{}
\pagestyle{fancy}\lfoot{\textbf {\small\it{Univ-Mascara/Computer Science: 2024}}} 
\pagestyle{fancy}\cfoot{} \pagestyle{fancy}\rfoot{\thepage}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Overview}\label{start4}
This chapter provides an in-depth exploration into the transformative landscape of large language models (LLMs) and their pivotal role in revolutionizing natural language processing (NLP) and machine translation (MT).

We begin by delving into the foundational architecture of LLMs, notably Transformers, BERT, and GPT, which have reshaped the understanding of linguistic patterns and relationships. 
Within this context, we examine how LLMs, particularly BERT and GPT, have evolved to address bidirectional context understanding and generative text tasks, respectively.

Furthermore, we explore the application of LLMs in MT, showcasing variants like BERT, T5, and ChatGPT, which demonstrate promising potential in overcoming language barriers and enhancing translation accuracy.

Following this exploration, we present a comprehensive section on related work, offering insights into the latest research and advancements in the field of LLM-based MT.

\section{Transformers}\label{sec:transformers}
While LSTMs and GRUs alleviate the issue of losing distant information inherent in simple RNNs, they are unable to leverage parallel computing resources due to their sequential nature. 
This limitation is addressed by Transformers, as introduced by \cite{vaswani17}.
Transformers revolutionize sequence processing by completely replacing RNNs. 
They operate by mapping sequences of input vectors $(x_1, x_2, ..., x_m)$ to sequences of output vectors $(y_1, y_2, ..., y_n)$ through stacks of network layers featuring customized connections of basic feed-forward networks and self-attentions. 
Unlike RNNs, Transformers enable the extraction and utilization of information from broader contexts through self-attention mechanisms, without the need for recurrent intermediary connections.
\begin{figure}
	\centering
	\includegraphics[width=0.5\linewidth]{Figures/Transformer}
	\caption{Transformer architecture adapted from \cite{vaswani17}}
	\label{fig:transformer}
\end{figure}

The Transformer, as depicted in Figure~\ref{fig:transformer}, is a sequence-to-sequence (Seq2Seq) network that captures sequential information through stacked self-attention and cross-attention layers. 
The output $O$ of each attention sub-layer is calculated using scaled multiplicative formulations, as defined by:
\begin{equation}
	A = (QW^Q)(KW^K)^T/\sqrt{d};\;\;\;   
	Att(Q, K, V ) = softmax(A)(V W^V )
\end{equation}
\begin{equation}
	O = Att(Q, K, V )W^O
\end{equation}
Where, $Q = (q_1, ..., q_{l_q}) \in \mathbb{R}^{l_q \times d}$,$K = (k_1, ..., k_{l_k}) \in \mathbb{R}^{l_k \times d}$, $V = (v_1, ..., v_{l_k}) \in \mathbb{R}^{l_k \times d}$ represent matrices of query, key, and value vectors respectively. Additionally, $W^Q$, $W^K$, $W^V$, and $W^O \in \mathbb{R}^{d \times d}$ denote the associated trainable weight matrices. $A$ signifies the affinity scores (or attention scores) between queries and keys, while $Att(Q, K, V)$ represents the attention vectors.
In practical applications, multi-head attention is utilized rather than single-head attention. 
Here, the hidden dimension $d$ is divided into $h$ segments, each processed independently through an attention layer before being amalgamated back together. 
Subsequently, the final output of a Transformer layer is determined as follows:
\begin{equation}
	\phi(A, Q) = LN(FFN(LN(O + Q)) + LN(O + Q))
\end{equation}
Here, $\phi$ denotes the standard sequential operations of a Transformer layer incorporating layer normalization (LN) and feed-forward (FFN) layers. The feed-forward layers essentially consist of sequences of fully connected layers interspersed with $ReLU$ activation, which are applied to each token's latent vector separately.

Since attention layers are insensitive to order, implying that the arrangement of key sequences doesn't influence the outcome of a specific query, it's crucial to introduce some form of ordering information for the model to grasp temporal characteristics of the input. 
To address this, the Transformer integrates positional encoding into the word embeddings immediately after the embedding layer and before the initial attention layer. 
This positional encoding relies on sine (Equation~\ref{eqn:E_sin}) and cosine (Equation~\ref{eqn:E_cos}) functions and is defined as follows:
\begin{equation}
	E(pos, 2i) = \sin(pos/10000^{2i/d})
	\label{eqn:E_sin}
\end{equation}
\begin{equation}
	E(pos, 2i+1) = \cos(pos/10000^{2i/d})
	\label{eqn:E_cos}
\end{equation}
Here, $pos$ denotes the position of the word within the sequence, while $i$ is the dimension.

\section{Large Language Models}
At the forefront of Natural Language Processing (NLP) stands a novel class of models termed Large Language Models (LLMs). 
These intricate artificial neural networks undergo training on massive volumes of textual data. Diverging from conventional NLP models tailored to specific tasks, LLMs acquire a broader comprehension of linguistic structures and correlations. 
This versatility enables them to undertake a diverse array of tasks, encompassing sentiment analysis, question answering, text summarization, and even creative writing. 
Notably, LLMs have significantly propelled the field of machine translation, a critical conduit between languages. 
Through extensive analysis of translated text datasets, these models can grasp the nuances inherent in different languages, yielding translations that are notably more precise and natural compared to traditional rule-based methodologies. 
However, the effectiveness of LLMs crucially depends on the caliber and extent of their training data. 
Greater diversity and comprehensiveness in the dataset enhance the model's understanding of the intricacies of human language. 
This has led to the exploration of specific LLM architectures such as BERT, mBERT, and GPT, among others, each contributing to the ongoing evolution of NLP and machine translation through their distinctive training methods and capabilities.

\subsection{BERT}
BERT, or Bidirectional Encoder Representations from Transformers, emerged as a groundbreaking pre-training method in the realm of Natural Language Processing (NLP). 
Unveiled by \cite{devlin18}, BERT reshaped the landscape by achieving top-tier performance across a diverse array of NLP tasks. 
Its standout feature lies in its adeptness at leveraging extensive, unlabeled text corpora during pre-training. 
Unlike earlier approaches that relied on task-specific labeled data, BERT follows a two-phase strategy. 
Initially, it undergoes pre-training on tasks like masked language modeling and next sentence prediction, exposing it to various linguistic structures and relationships. 
This equips the model with a robust grasp of word semantics and contextual nuances, forming a strong foundation for fine-tuning on specific tasks. 
BERT produces embeddings as its output, rather than predicting the next words in a sequence. 
To utilize these embeddings, additional layers must be added on top, such as those for text classification or question answering tasks.

BERT's architectural backbone relies on the Transformer encoder (see Figure~\ref{fig:transformer}), a potent neural network architecture introduced by \cite{vaswani17}. 
Unlike traditional sequential models, the Transformer enables parallel processing of entire input sentences, facilitating the capture of long-range dependencies essential for tasks such as sentiment analysis and question answering (see Figure~\ref{fig:bert}). 
Notably, BERT employs a pre-trained encoder model and foregoes the decoder component typically found in sequence-to-sequence models like those used in machine translation. 
This design choice underscores BERT's focus on comprehending text rather than generating new sequences. A key contributor to BERT's efficacy is the scale of its pre-training data, drawn from vast text corpora like BookCorpus and English Wikipedia. 
\begin{figure}
	\centering
	\includegraphics[width=0.7\linewidth]{Figures/BERT}
	\caption{BERT base architecture with twelve encoder blocks, adapted from \cite{khalid21}.}
	\label{fig:bert}
\end{figure}

While BERT's original training was centered on English, subsequent research has explored its applicability to multilingual NLP tasks. 
mBERT, short for multilingual BERT, extends the capabilities of BERT to encompass tasks spanning multiple languages. It inherits BERT's architecture but undergoes training on an extensive dataset containing text samples from a staggering 104 languages.
Efforts such as mBERT have demonstrated the feasibility of fine-tuning pre-trained models across multiple languages, showcasing BERT's potential for cross-lingual NLP endeavors\cite{pires19}.

\subsection{GPT}
In contrast to BERT's emphasis on pre-training encoders, Generative Pre-trained Transformer (GPT) models offer a compelling alternative for tasks involving text generation. 
Introduced by OpenAI\footnote{https://openai.com/}, GPT utilizes a decoder-based architecture derived from the Transformer family. 
This design enables GPT to excel in generating various forms of creative text, including poems, code, scripts, musical compositions, and even realistic chat dialogues \cite{radford18}.

Unlike BERT, GPT is predominantly trained on a single, extensive dataset consisting of text and code. 
The initial GPT model, GPT-1, was trained on a dataset compiled from internet sources, while subsequent versions like GPT-2 and GPT-3 have utilized progressively larger and more diverse datasets. 
Focusing on a single language, typically English, allows GPT to grasp statistical patterns and sequential dependencies within the text, thereby generating text of human-like quality that can often be mistaken for content written by humans.

The fundamental architecture of GPT relies on a Transformer decoder (see Figure~\ref{fig:transformer}). 
Unlike encoders that process the entire input sequence simultaneously, decoders handle the input sequence one element (word) at a time. 
This iterative approach enables GPT to predict the next word in a sequence based on previously generated words and the overall context, making it proficient in tasks like text summarization, machine translation, and creative writing, where generating coherent text is crucial.
\begin{figure}
	\centering
	\includegraphics[width=0.7\linewidth]{Figures/GPT2}
	\caption[]{GPT-2 model architecture adapted from \cite{yang23}. }
	\label{fig:gpt2}
\end{figure}

In language scenarios, decoders play a crucial role in generating subsequent words, such as in text translation or story generation, where the outputs are words with probabilities.
Decoders integrate attention mechanisms, employing them twice in their operation. Initially, during model training, they utilize Masked Multi-Head Attention, where only the initial words of the target sentence are revealed to prevent the model from cheating during learning. This approach resembles the MASK concept introduced in BERT.
Subsequently, decoders employ Multi-Head Attention \cite{liu21}, similar to how it is utilized in the encoder. Transformer-based models incorporating both encoders and decoders employ a technique for enhanced efficiency. The output of the encoders serves as input to the decoders, specifically as keys and values. Decoders can then make queries to locate the most relevant keys. This facilitates tasks such as comprehending the original sentence's meaning and translating it into other languages, even when the resulting word count and order vary.
Figure~\ref{fig:gpt2} illustrates the structure of GPT-2, consisting of N Transformer decoder blocks. Each block is equipped with multi-head masked attention, multi-layer perceptrons, normalization, and dropout layers. Residual connections link these blocks, enhancing the model's ability to learn from earlier inputs. The multi-head masked attention mechanism employs Q, K, and V vectors for calculating attention scores, effectively capturing and encoding the sequential relationships within the data.
%While GPT models are primarily trained on a single language, ongoing research is exploring their adaptability to multilingual tasks. 
%Techniques such as multilingual fine-tuning allow GPT models to be adjusted for generating text in multiple languages. 
%However, due to their single-language focus during pre-training, GPT models generally exhibit lower performance on multilingual tasks compared to models like mBERT, which are explicitly designed for handling multiple languages.

\section{LLM-based Machine Translation}
Unlike conventional rule-based methodologies that depend on predetermined linguistic principles, Large Language Models (LLMs) utilize extensive real-world translation datasets to grasp the intricacies and statistical regularities of language \cite{brants07}. This enables them to generate translations that are more natural-sounding and precise compared to traditional techniques.

A significant advantage of LLM-based MT lies in its capacity to comprehend context. By analyzing extensive sets of translated texts, LLMs can discern the connections between words, phrases, and the overall context of a sentence. This contextual comprehension empowers them to produce translations that not only adhere to grammatical rules but also capture the intended meaning and nuances of the source language.

Numerous research studies have investigated the efficacy of LLMs in MT. For example, the study conducted by  showcases how LLMs can achieve superior performance on diverse language pairs, sometimes outperforming traditional statistical MT approaches.

Nevertheless, LLM-based MT encounters certain challenges. One concern is the potential presence of bias in the training data, which may manifest in the translated output \cite{wang24}. 
Translating content into English may enhance multilingual NLP tasks for English-centric LLMs, yet it's not always the best approach, particularly for tasks requiring deep cultural and linguistic comprehension. 
Using the native language directly often yields better results by capturing cultural nuances more effectively \cite{liu24}.
Additionally, LLMs can pose computational challenges due to their extensive size and complexity, making training and deployment resource-intensive \cite{yang24}.
Despite these hurdles, LLM-based MT offers a promising avenue for seamless cross-lingual communication. As research advances and training datasets expand, we anticipate further progress in this dynamic field.

\subsection{BART}
BART (Bidirectional and Autoregressive Transformer) arises as a potent Large Language Model, differentiating itself from GPT's focus on decoders and BERT's emphasis on encoders. 
It achieves this by integrating both encoder and decoder elements, enabling it to tackle tasks related to both Natural Language Generation (NLG) and Natural Language Understanding (NLU). 
This dual structure synergizes the comprehensive understanding capabilities of the encoder with the fluent generative abilities of the decoder.

As highlighted in \cite{lewis19}, BART's architecture enables it to deliver superior performance in MT by thoroughly grasping the source language context with its encoder and producing coherent translations with its decoder. 
The ability to pre-train on extensive multilingual datasets further enhances BART's proficiency in handling a wide array of language pairs, positioning it as an effective solution for overcoming language barriers.
Figure~\ref{fig:bart} illustrates the BART model adapted for MT. In this scenario, a supplementary encoder is introduced to substitute word embeddings within BART. This new encoder has the flexibility to utilize a separate vocabulary.
\begin{figure}[h]
	\centering
	\includegraphics[width=0.7\linewidth]{Figures/Bart}
	\caption{Fine tuning BART for machine translation.}
	\label{fig:bart}
\end{figure}

\subsection{T5}
T5, or Text-to-Text Transfer Transformer, emerges as a formidable player in the landscape of LLM-based MT. 
Unlike models such as BERT, which concentrate on pre-training encoders, or GPT, which prioritize decoders, T5 takes a unified approach. 
It employs a single Transformer-based architecture that can be fine-tuned for various NLP tasks, including machine translation (see Figure~\ref{fig:t5}).

%This design, outlined in the paper "Exploring the Limits of Language Learning with a Unified Text-to-Text Transformer" by Raffel et al. (2020), grants T5 exceptional adaptability in addressing MT challenges. 
The model can be trained on extensive multilingual text and code datasets, enabling it to comprehend the nuances of different languages and perform effective translations between them. 
Moreover, T5's capability to adjust to diverse NLP tasks equips it to handle a wide range of translation scenarios, from straightforward sentence translations to intricate document summarization with translation components. 
This versatility positions T5 as a valuable resource for applications necessitating seamless and nuanced language transfer across various domains.
\begin{figure}
	\centering
	\includegraphics[width=0.7\linewidth]{Figures/T5}
	\caption{T5 model architecture adapted from \cite{wang23}.}
	\label{fig:t5}
\end{figure}

\subsection{ChatGPT}
While ChatGPT demonstrates remarkable proficiency in generating diverse text formats creatively, its utility in MT is circumscribed by certain constraints \cite{lyu23}. 
Unlike specialized models like BART or T5 tailored explicitly for managing both source and target languages, ChatGPT's fundamental capability resides in its decoder-centric architecture. 
This design excels in text generation based on provided prompts or contexts, rendering it suitable for tasks such as creative writing or text summarization.
owever, for achieving precise and natural-sounding translations, comprehending the context of the source language is imperative. 
ChatGPT's primary emphasis on text generation within a single language, typically English (GPT data proportion is only 7\% non-English \cite{pourkamali24}), constrains its capacity to fully comprehend the subtleties and intricacies involved in translating between diverse languages. 
Therefore, while it can generate text in various languages based on the input it receives, its proficiency in languages other than English may vary depending on factors such as the availability and diversity of training data in those languages.

While some research endeavors explore the potential of adapting ChatGPT for multilingual purposes \cite{brown20, lu23, peng23, pourkamali24}, it has been observed that ChatGPT often generates inaccurate outputs and hallucinates for non-English-centric machine translation tasks \cite{peng23}. 
Its performance generally lags behind models like mBERT or BART, specifically designed for multilingual understanding and translation. 
Additionally, studies have shown that while ChatGPT can compete with commercial translation products such as Google Translate and Microsoft Translator for high-resource languages, it exhibits limited capabilities for low-resource and distant languages \cite{peng23, hendy23, jiao23, pourkamali24}.
\begin{figure}
	\centering
	\includegraphics[width=1.0\linewidth]{Figures/ChatGPT-model}
	\caption{Transformer architecture of ChatGPT}
	\label{fig:chatgpt-model}
\end{figure}

\section{Related Work}
\hl{.....................................................................}

\cite{brown20}
\cite{lu23}
\cite{peng23}
\cite{hendy23}
\cite{jiao23}
\cite{pourkamali24}
\cite{jiang24}
\\
\hl{.....................................................................}

\section{Summary}
In this chapter, we delved into the transformative impact of large language models (LLMs), notably focusing on the advancements brought forth by models such as Transformers, BERT, and GPT within the realm of natural language processing (NLP). 
These models, built upon Transformer architectures, have revolutionized NLP tasks by learning intricate linguistic patterns and relationships. 
Particularly, BERT and GPT represent pioneering models, with BERT excelling in understanding bidirectional context and GPT specializing in generative tasks.

Leveraging LLMs for machine translation (MT), variants like BERT, T5, and ChatGPT demonstrate promising capabilities in bridging language barriers and enhancing translation accuracy. 
However, challenges persist, including the need for diverse and extensive training data and mitigating biases inherent in these models. 
Nevertheless, ongoing research endeavors continue to explore advancements and refine techniques in leveraging LLMs for MT, underscoring the dynamic evolution of NLP and MT technologies.

### 2025-01-20 Thesis/Latex/Chapter_6.tex
Organizing repo and adding thesis template

--- 2025-01-20 Thesis/Latex/Chapter_4.tex
@@
-\chapter{Large Language Models}
-\pagestyle{fancy}\lhead{\textbf \footnotesize\it{Large Language Models}}
+\chapter{Seq2Seq Neural Machine Translation}
+\pagestyle{fancy}\lhead{\textbf \footnotesize\it{Seq2Seq Neural Machine Translation}}
 \pagestyle{fancy}\chead{} \pagestyle{fancy}\rhead{}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Overview}\label{start4}
-This chapter provides an in-depth exploration into the transformative landscape of large language models (LLMs) and their pivotal role in revolutionizing natural language processing (NLP) and machine translation (MT).
+\section{Overview}\label{start6}
 
-We begin by delving into the foundational architecture of LLMs, notably Transformers, BERT, and GPT, which have reshaped the understanding of linguistic patterns and relationships. 
-Within this context, we examine how LLMs, particularly BERT and GPT, have evolved to address bidirectional context understanding and generative text tasks, respectively.
-
-Furthermore, we explore the application of LLMs in MT, showcasing variants like BERT, T5, and ChatGPT, which demonstrate promising potential in overcoming language barriers and enhancing translation accuracy.
-
-Following this exploration, we present a comprehensive section on related work, offering insights into the latest research and advancements in the field of LLM-based MT.
-
-\section{Transformers}\label{sec:transformers}
-While LSTMs and GRUs alleviate the issue of losing distant information inherent in simple RNNs, they are unable to leverage parallel computing resources due to their sequential nature. 
-This limitation is addressed by Transformers, as introduced by \cite{vaswani17}.
-Transformers revolutionize sequence processing by completely replacing RNNs. 
-They operate by mapping sequences of input vectors $(x_1, x_2, ..., x_m)$ to sequences of output vectors $(y_1, y_2, ..., y_n)$ through stacks of network layers featuring customized connections of basic feed-forward networks and self-attentions.
-Unlike RNNs, Transformers enable the extraction and utilization of information from broader contexts through self-attention mechanisms, without the need for recurrent intermediary connections.
-\begin{figure}
-	\centering
-	\includegraphics[width=0.5\linewidth]{Figures/Transformer}
-	\caption{Transformer architecture adapted from \cite{vaswani17}}
-	\label{fig:transformer}
-\end{figure}
-
-The Transformer, as depicted in Figure~\ref{fig:transformer}, is a sequence-to-sequence (Seq2Seq) network that captures sequential information through stacked self-attention and cross-attention layers.
-The output $O$ of each attention sub-layer is calculated using scaled multiplicative formulations, as defined by:
-\begin{equation}
-	A = (QW^Q)(KW^K)^T/\sqrt{d};\;\;\;   
-	Att(Q, K, V ) = softmax(A)(V W^V )
-\end{equation}
-\begin{equation}
-	O = Att(Q, K, V )W^O
-\end{equation}
-Where, $Q = (q_1, ..., q_{l_q}) \in \mathbb{R}^{l_q \times d}$,$K = (k_1, ..., k_{l_k}) \in \mathbb{R}^{l_k \times d}$, $V = (v_1, ..., v_{l_k}) \in \mathbb{R}^{l_k \times d}$ represent matrices of query, key, and value vectors respectively.
-Additionally, $W^Q$, $W^K$, $W^V$, and $W^O \in \mathbb{R}^{d \times d}$ denote the associated trainable weight matrices.
-$A$ signifies the affinity scores (or attention scores) between queries and keys, while $Att(Q, K, V)$ represents the attention vectors.
-In practical applications, multi-head attention is utilized rather than single-head attention. 
-Here, the hidden dimension $d$ is divided into $h$ segments, each processed independently through an attention layer before being amalgamated back together. 
-Subsequently, the final output of a Transformer layer is determined as follows:
-\begin{equation}
-	\phi(A, Q) = LN(FFN(LN(O + Q)) + LN(O + Q))
-\end{equation}
-Here, $\phi$ denotes the standard sequential operations of a Transformer layer incorporating layer normalization (LN) and feed-forward (FFN) layers.
-The feed-forward layers essentially consist of sequences of fully connected layers interspersed with $ReLU$ activation, which are applied to each token's latent vector separately.
-
-Since attention layers are insensitive to order, implying that the arrangement of key sequences doesn't influence the outcome of a specific query, it's crucial to introduce some form of ordering information for the model to grasp temporal characteristics of the input.
-To address this, the Transformer integrates positional encoding into the word embeddings immediately after the embedding layer and before the initial attention layer. 
-This positional encoding relies on sine (Equation~\ref{eqn:E_sin}) and cosine (Equation~\ref{eqn:E_cos}) functions and is defined as follows:
-\begin{equation}
-	E(pos, 2i) = \sin(pos/10000^{2i/d})
-	\label{eqn:E_sin}
-\end{equation}
-\begin{equation}
-	E(pos, 2i+1) = \cos(pos/10000^{2i/d})
-	\label{eqn:E_cos}
-\end{equation}
-Here, $pos$ denotes the position of the word within the sequence, while $i$ is the dimension.
-
-\section{Large Language Models}
-At the forefront of Natural Language Processing (NLP) stands a novel class of models termed Large Language Models (LLMs). 
-These intricate artificial neural networks undergo training on massive volumes of textual data.
-Diverging from conventional NLP models tailored to specific tasks, LLMs acquire a broader comprehension of linguistic structures and correlations.
-This versatility enables them to undertake a diverse array of tasks, encompassing sentiment analysis, question answering, text summarization, and even creative writing. 
-Notably, LLMs have significantly propelled the field of machine translation, a critical conduit between languages. 
-Through extensive analysis of translated text datasets, these models can grasp the nuances inherent in different languages, yielding translations that are notably more precise and natural compared to traditional rule-based methodologies.
-However, the effectiveness of LLMs crucially depends on the caliber and extent of their training data. 
-Greater diversity and comprehensiveness in the dataset enhance the model's understanding of the intricacies of human language. 
-This has led to the exploration of specific LLM architectures such as BERT, mBERT, and GPT, among others, each contributing to the ongoing evolution of NLP and machine translation through their distinctive training methods and capabilities.
-
-\subsection{BERT}
-BERT, or Bidirectional Encoder Representations from Transformers, emerged as a groundbreaking pre-training method in the realm of Natural Language Processing (NLP). 
-Unveiled by \cite{devlin18}, BERT reshaped the landscape by achieving top-tier performance across a diverse array of NLP tasks. 
-Its standout feature lies in its adeptness at leveraging extensive, unlabeled text corpora during pre-training. 
-Unlike earlier approaches that relied on task-specific labeled data, BERT follows a two-phase strategy. 
-Initially, it undergoes pre-training on tasks like masked language modeling and next sentence prediction, exposing it to various linguistic structures and relationships. 
-This equips the model with a robust grasp of word semantics and contextual nuances, forming a strong foundation for fine-tuning on specific tasks. 
-BERT produces embeddings as its output, rather than predicting the next words in a sequence. 
-To utilize these embeddings, additional layers must be added on top, such as those for text classification or question answering tasks.
-
-BERT's architectural backbone relies on the Transformer encoder (see Figure~\ref{fig:transformer}), a potent neural network architecture introduced by \cite{vaswani17}. 
-Unlike traditional sequential models, the Transformer enables parallel processing of entire input sentences, facilitating the capture of long-range dependencies essential for tasks such as sentiment analysis and question answering (see Figure~\ref{fig:bert}).
-Notably, BERT employs a pre-trained encoder model and foregoes the decoder component typically found in sequence-to-sequence models like those used in machine translation. 
-This design choice underscores BERT's focus on comprehending text rather than generating new sequences.
-A key contributor to BERT's efficacy is the scale of its pre-training data, drawn from vast text corpora like BookCorpus and English Wikipedia.
-\begin{figure}
-	\centering
-	\includegraphics[width=0.7\linewidth]{Figures/BERT}
-	\caption{BERT base architecture with twelve encoder blocks, adapted from \cite{khalid21}.}
-	\label{fig:bert}
-\end{figure}
-
-While BERT's original training was centered on English, subsequent research has explored its applicability to multilingual NLP tasks. 
-mBERT, short for multilingual BERT, extends the capabilities of BERT to encompass tasks spanning multiple languages.
-It inherits BERT's architecture but undergoes training on an extensive dataset containing text samples from a staggering 104 languages.
-Efforts such as mBERT have demonstrated the feasibility of fine-tuning pre-trained models across multiple languages, showcasing BERT's potential for cross-lingual NLP endeavors\cite{pires19}.
-
-\subsection{GPT}
-In contrast to BERT's emphasis on pre-training encoders, Generative Pre-trained Transformer (GPT) models offer a compelling alternative for tasks involving text generation. 
-Introduced by OpenAI\footnote{https://openai.com/}, GPT utilizes a decoder-based architecture derived from the Transformer family. 
-This design enables GPT to excel in generating various forms of creative text, including poems, code, scripts, musical compositions, and even realistic chat dialogues \cite{radford18}.
-
-Unlike BERT, GPT is predominantly trained on a single, extensive dataset consisting of text and code. 
-The initial GPT model, GPT-1, was trained on a dataset compiled from internet sources, while subsequent versions like GPT-2 and GPT-3 have utilized progressively larger and more diverse datasets. 
-Focusing on a single language, typically English, allows GPT to grasp statistical patterns and sequential dependencies within the text, thereby generating text of human-like quality that can often be mistaken for content written by humans.
-
-The fundamental architecture of GPT relies on a Transformer decoder (see Figure~\ref{fig:transformer}). 
-Unlike encoders that process the entire input sequence simultaneously, decoders handle the input sequence one element (word) at a time. 
-This iterative approach enables GPT to predict the next word in a sequence based on previously generated words and the overall context, making it proficient in tasks like text summarization, machine translation, and creative writing, where generating coherent text is crucial.
-\begin{figure}
-	\centering
-	\includegraphics[width=0.7\linewidth]{Figures/GPT2}
-	\caption[]{GPT-2 model architecture adapted from \cite{yang23}. }
-	\label{fig:gpt2}
-\end{figure}
-
-In language scenarios, decoders play a crucial role in generating subsequent words, such as in text translation or story generation, where the outputs are words with probabilities.
-Decoders integrate attention mechanisms, employing them twice in their operation.
-Initially, during model training, they utilize Masked Multi-Head Attention, where only the initial words of the target sentence are revealed to prevent the model from cheating during learning.
-This approach resembles the MASK concept introduced in BERT.
-Subsequently, decoders employ Multi-Head Attention \cite{liu21}, similar to how it is utilized in the encoder.
-Transformer-based models incorporating both encoders and decoders employ a technique for enhanced efficiency.
-The output of the encoders serves as input to the decoders, specifically as keys and values.
-Decoders can then make queries to locate the most relevant keys.
-This facilitates tasks such as comprehending the original sentence's meaning and translating it into other languages, even when the resulting word count and order vary.
-Figure~\ref{fig:gpt2} illustrates the structure of GPT-2, consisting of N Transformer decoder blocks.
-Each block is equipped with multi-head masked attention, multi-layer perceptrons, normalization, and dropout layers.
-Residual connections link these blocks, enhancing the model's ability to learn from earlier inputs.
-The multi-head masked attention mechanism employs Q, K, and V vectors for calculating attention scores, effectively capturing and encoding the sequential relationships within the data.
-%While GPT models are primarily trained on a single language, ongoing research is exploring their adaptability to multilingual tasks. 
-%Techniques such as multilingual fine-tuning allow GPT models to be adjusted for generating text in multiple languages. 
-%However, due to their single-language focus during pre-training, GPT models generally exhibit lower performance on multilingual tasks compared to models like mBERT, which are explicitly designed for handling multiple languages.
-
-\section{LLM-based Machine Translation}
-Unlike conventional rule-based methodologies that depend on predetermined linguistic principles, Large Language Models (LLMs) utilize extensive real-world translation datasets to grasp the intricacies and statistical regularities of language \cite{brants07}.
-This enables them to generate translations that are more natural-sounding and precise compared to traditional techniques.
-
-A significant advantage of LLM-based MT lies in its capacity to comprehend context.
-By analyzing extensive sets of translated texts, LLMs can discern the connections between words, phrases, and the overall context of a sentence.
-This contextual comprehension empowers them to produce translations that not only adhere to grammatical rules but also capture the intended meaning and nuances of the source language.
-
-Numerous research studies have investigated the efficacy of LLMs in MT.
-For example, the study conducted by  showcases how LLMs can achieve superior performance on diverse language pairs, sometimes outperforming traditional statistical MT approaches.
-
-Nevertheless, LLM-based MT encounters certain challenges. One concern is the potential presence of bias in the training data, which may manifest in the translated output \cite{wang24}. 
-Translating content into English may enhance multilingual NLP tasks for English-centric LLMs, yet it's not always the best approach, particularly for tasks requiring deep cultural and linguistic comprehension.
-Using the native language directly often yields better results by capturing cultural nuances more effectively \cite{liu24}.
-Additionally, LLMs can pose computational challenges due to their extensive size and complexity, making training and deployment resource-intensive \cite{yang24}.
-Despite these hurdles, LLM-based MT offers a promising avenue for seamless cross-lingual communication.
-As research advances and training datasets expand, we anticipate further progress in this dynamic field.
-
-\subsection{BART}
-BART (Bidirectional and Autoregressive Transformer) arises as a potent Large Language Model, differentiating itself from GPT's focus on decoders and BERT's emphasis on encoders. 
-It achieves this by integrating both encoder and decoder elements, enabling it to tackle tasks related to both Natural Language Generation (NLG) and Natural Language Understanding (NLU). 
-This dual structure synergizes the comprehensive understanding capabilities of the encoder with the fluent generative abilities of the decoder.
-
-As highlighted in \cite{lewis19}, BART's architecture enables it to deliver superior performance in MT by thoroughly grasping the source language context with its encoder and producing coherent translations with its decoder.
-The ability to pre-train on extensive multilingual datasets further enhances BART's proficiency in handling a wide array of language pairs, positioning it as an effective solution for overcoming language barriers.
-Figure~\ref{fig:bart} illustrates the BART model adapted for MT.
-In this scenario, a supplementary encoder is introduced to substitute word embeddings within BART.
-This new encoder has the flexibility to utilize a separate vocabulary.
-\begin{figure}[h]
-	\centering
-	\includegraphics[width=0.7\linewidth]{Figures/Bart}
-	\caption{Fine tuning BART for machine translation.}
-	\label{fig:bart}
-\end{figure}
-
-\subsection{T5}
-T5, or Text-to-Text Transfer Transformer, emerges as a formidable player in the landscape of LLM-based MT. 
-Unlike models such as BERT, which concentrate on pre-training encoders, or GPT, which prioritize decoders, T5 takes a unified approach. 
-It employs a single Transformer-based architecture that can be fine-tuned for various NLP tasks, including machine translation (see Figure~\ref{fig:t5}).
-
-%This design, outlined in the paper "Exploring the Limits of Language Learning with a Unified Text-to-Text Transformer" by Raffel et al.
-(2020), grants T5 exceptional adaptability in addressing MT challenges.
-The model can be trained on extensive multilingual text and code datasets, enabling it to comprehend the nuances of different languages and perform effective translations between them. 
-Moreover, T5's capability to adjust to diverse NLP tasks equips it to handle a wide range of translation scenarios, from straightforward sentence translations to intricate document summarization with translation components.
-This versatility positions T5 as a valuable resource for applications necessitating seamless and nuanced language transfer across various domains.
-\begin{figure}
-	\centering
-	\includegraphics[width=0.7\linewidth]{Figures/T5}
-	\caption{T5 model architecture adapted from \cite{wang23}.}
-	\label{fig:t5}
-\end{figure}
-
-\subsection{ChatGPT}
-While ChatGPT demonstrates remarkable proficiency in generating diverse text formats creatively, its utility in MT is circumscribed by certain constraints \cite{lyu23}. 
-Unlike specialized models like BART or T5 tailored explicitly for managing both source and target languages, ChatGPT's fundamental capability resides in its decoder-centric architecture. 
-This design excels in text generation based on provided prompts or contexts, rendering it suitable for tasks such as creative writing or text summarization.
-owever, for achieving precise and natural-sounding translations, comprehending the context of the source language is imperative. 
-ChatGPT's primary emphasis on text generation within a single language, typically English (GPT data proportion is only 7\% non-English \cite{pourkamali24}), constrains its capacity to fully comprehend the subtleties and intricacies involved in translating between diverse languages.
-Therefore, while it can generate text in various languages based on the input it receives, its proficiency in languages other than English may vary depending on factors such as the availability and diversity of training data in those languages.
-
-While some research endeavors explore the potential of adapting ChatGPT for multilingual purposes \cite{brown20, lu23, peng23, pourkamali24}, it has been observed that ChatGPT often generates inaccurate outputs and hallucinates for non-English-centric machine translation tasks \cite{peng23}.
-Its performance generally lags behind models like mBERT or BART, specifically designed for multilingual understanding and translation. 
-Additionally, studies have shown that while ChatGPT can compete with commercial translation products such as Google Translate and Microsoft Translator for high-resource languages, it exhibits limited capabilities for low-resource and distant languages \cite{peng23, hendy23, jiao23, pourkamali24}.
-\begin{figure}
-	\centering
-	\includegraphics[width=1.0\linewidth]{Figures/ChatGPT-model}
-	\caption{Transformer architecture of ChatGPT}
-	\label{fig:chatgpt-model}
-\end{figure}
-
-\section{Related Work}
-\hl{.....................................................................}
-
-\cite{brown20}
-\cite{lu23}
-\cite{peng23}
-\cite{hendy23}
-\cite{jiao23}
-\cite{pourkamali24}
-\cite{jiang24}
-\\
-\hl{.....................................................................}
+\section{System Architecture}
+\section{Baseline System}
+\section{Experiments and Evaluation}
+\section{Results and Discussions}
 
 \section{Summary}
-In this chapter, we delved into the transformative impact of large language models (LLMs), notably focusing on the advancements brought forth by models such as Transformers, BERT, and GPT within the realm of natural language processing (NLP).
-These models, built upon Transformer architectures, have revolutionized NLP tasks by learning intricate linguistic patterns and relationships. 
-Particularly, BERT and GPT represent pioneering models, with BERT excelling in understanding bidirectional context and GPT specializing in generative tasks.
-
-Leveraging LLMs for machine translation (MT), variants like BERT, T5, and ChatGPT demonstrate promising capabilities in bridging language barriers and enhancing translation accuracy. 
-However, challenges persist, including the need for diverse and extensive training data and mitigating biases inherent in these models. 
-Nevertheless, ongoing research endeavors continue to explore advancements and refine techniques in leveraging LLMs for MT, underscoring the dynamic evolution of NLP and MT technologies.

### 2025-01-20 Thesis/Latex/Cover_page.tex
Organizing repo and adding thesis template

\thispagestyle{empty}
\begin{center}
	\begin{title}
		\textbf{\uppercase{ People's Democratic Republic of Algeria }} \\[0.2cm]
		\textbf{\uppercase{ MINISTRY OF HIGHER EDUCATION AND SCIENTIFIC RESEARCH }} \\[0.5cm]
		\textbf{\uppercase{ UNIVERSITY MUSTAPHA STAMBOULI OF MASCARA}}
	\end{title}
	\begin{figure}[H]
		\centering
		\label{logo_UD}\includegraphics[width=0.2\textwidth]{Figures/Logo_UM.jpg}
	\end{figure}
	{\large Faculty of Exact Sciences} \\[0.2cm]
	{\large Department of Computer Science} \\[0.2cm]
	{\large Dissertation } \\[0.2cm]
	{\ Submitted in partial fulfilment of the requirements for Doctorate degree in Computer Science} \\[0.2cm]
	{\large Option: Artificial Intelligence}\\[0.5cm]

    {\large Theme} \\[0.3cm]
%\hline	
%    \\[0.7cm]
	{\LARGE {\textbf{Neural Machine Translation for\\ Arabic Media Content}}}
	\\[0.7cm]
%\hline
%	 \\[0.5cm]

	{\large Presented by \textbf{Baligh BABAALI} }\\[0.5cm]

\end{center}
Jury:\\
\begin{tabular}{llll}
	President    &  Aaaaa AAAAA & Professor & University of Mascara\\
	Director     &  Mohammed SALEM & Professor & University of Mascara\\
%	Co-Director  &  Ccccc CCCCC & Professor & University of Mascara\\
	Examiner     &  Ddddd DDDDD & Professor	& University of Mascara\\
	Examiner	 &  Eeeee EEEEE & Associate Prof.  & University of Mascara\\
	Examiner	 &  Fffff FFFFF & Associate Prof.  & University of Mascara\\
\end{tabular}
\vfill
\begin{center}

	\it{Month  2024}
\end{center}

### 2025-01-20 Thesis/Latex/Dedication.tex
Organizing repo and adding thesis template

\thispagestyle{empty}
\begin{tikzpicture}[remember picture, overlay]
	\node at (current page){
		\includegraphics[width=\paperwidth,height=\paperheight]{Figures/bgg4.png}} ;
\end{tikzpicture}
\vskip15mm
\begin{center}
	\resizebox{6cm}{1.1cm}{\calligra Dedication} 
\end{center}
%

\thispagestyle{empty}

### 2025-01-20 Thesis/Latex/Images_sources.tex
Organizing repo and adding thesis template

\begin{landscape}
	\thispagestyle{empty}
	\begin{center}
		\Huge{\textbf{Sources of Images}} \\
		\normalsize{\textit{for Copyright purposes}}
	\end{center}
	Figure 2.1 from \url{www.site1.com} \\
	Figure 2.3 from \url{www.site2.fr} \\
	Figure 2.5 from \url{www.site3.com} \\
	Figure 2.9 from \url{www.site4.com} \\
	Figure 3.5 from \url{www.site5.com} \\
	Figure 2.3 from \url{www.site2.fr} \\
	Figure 2.5 from \url{www.site3.com} \\
	Figure 2.9 from \url{www.site4.com} \\
	Figure 3.5 from \url{www.site5.com} \\
\end{landscape}

### 2025-01-21 Paper/latex/paper_RAG.tex
Paper template added

%Version 3 October 2023
% See section 11 of the User Manual for version history
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                                                 %%
%% Please do not use \input{...} to include other tex files.       %%
%% Submit your LaTeX manuscript as one .tex document.              %%
%%                                                                 %%
%% All additional figures and files should be attached             %%
%% separately and not embedded in the \TeX\ document itself.       %%
%%                                                                 %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%\documentclass[referee,sn-basic]{sn-jnl}% referee option is meant for double line spacing

%%=======================================================%%
%% to print line numbers in the margin use lineno option %%
%%=======================================================%%

%%\documentclass[lineno,sn-basic]{sn-jnl}% Basic Springer Nature Reference Style/Chemistry Reference Style

%%======================================================%%
%% to compile with pdflatex/xelatex use pdflatex option %%
%%======================================================%%

%%\documentclass[pdflatex,sn-basic]{sn-jnl}% Basic Springer Nature Reference Style/Chemistry Reference Style

%%Note: the following reference styles support Namedate and Numbered referencing. By default the style follows the most common style. To switch between the options you can add or remove “Numbered” in the optional parenthesis. 
%%The option is available for: sn-basic.bst, sn-vancouver.bst, sn-chicago.bst%

%%\documentclass[sn-nature]{sn-jnl}% Style for submissions to Nature Portfolio journals
%%\documentclass[sn-basic]{sn-jnl}% Basic Springer Nature Reference Style/Chemistry Reference Style
\documentclass[sn-mathphys-num]{sn-jnl}% Math and Physical Sciences Numbered Reference Style 
%%\documentclass[sn-mathphys-ay]{sn-jnl}% Math and Physical Sciences Author Year Reference Style
%%\documentclass[sn-aps]{sn-jnl}% American Physical Society (APS) Reference Style
%%\documentclass[sn-vancouver,Numbered]{sn-jnl}% Vancouver Reference Style
%%\documentclass[sn-apa]{sn-jnl}% APA Reference Style 
%%\documentclass[sn-chicago]{sn-jnl}% Chicago-based Humanities Reference Style

%%%% Standard Packages
%%<additional latex packages if required can be included here>

\usepackage{graphicx}%
\usepackage{multirow}%
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{amsthm}%
\usepackage{mathrsfs}%
\usepackage[title]{appendix}%
\usepackage{xcolor}%
\usepackage{textcomp}%
\usepackage{manyfoot}%
\usepackage{booktabs}%
\usepackage{algorithm}%
\usepackage{algorithmicx}%
\usepackage{algpseudocode}%
\usepackage{listings}%
%%%%

%%%%%=============================================================================%%%%
%%%%  Remarks: This template is provided to aid authors with the preparation
%%%%  of original research articles intended for submission to journals published 
%%%%  by Springer Nature. The guidance has been prepared in partnership with 
%%%%  production teams to conform to Springer Nature technical requirements. 
%%%%  Editorial and presentation requirements differ among journal portfolios and 
%%%%  research disciplines. You may find sections in this template are irrelevant 
%%%%  to your work and are empowered to omit any such section if allowed by the 
%%%%  journal you intend to submit to. The submission guidelines and policies 
%%%%  of the journal take precedence. A detailed User Manual is available in the 
%%%%  template package for technical guidance.
%%%%%=============================================================================%%%%

%% as per the requirement new theorem styles can be included as shown below
\theoremstyle{thmstyleone}%
\newtheorem{theorem}{Theorem}%  meant for continuous numbers
%%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
%% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
\newtheorem{proposition}[theorem]{Proposition}% 
%%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.

\theoremstyle{thmstyletwo}%
\newtheorem{example}{Example}%
\newtheorem{remark}{Remark}%

\theoremstyle{thmstylethree}%
\newtheorem{definition}{Definition}%

\raggedbottom
%%\unnumbered% uncomment this for unnumbered level heads

\begin{document}

\title[Article Title]{ttttttttttttttttttttt}

%%=============================================================%%
%% GivenName	-> \fnm{Joergen W.}
%% Particle	-> \spfx{van der} -> surname prefix
%% FamilyName	-> \sur{Ploeg}
%% Suffix	-> \sfx{IV}
%% \author*[1,2]{\fnm{Joergen W.} \spfx{van der} \sur{Ploeg} 
%%  \sfx{IV}}\email{iauthor@gmail.com}
%%=============================================================%%

\author*[1]{\fnm{Asmaa} \sur{Boudjenane }}\email{}

\author[2]{\fnm{Mohammed} \sur{Salem}}\email{salem@univ-mascara.dz}
\equalcont{These authors contributed equally to this work.}

%\author[1,2]{\fnm{Third} \sur{Author}}\email{iiiauthor@gmail.com}
%\equalcont{These authors contributed equally to this work.}

\affil*[1]{\orgdiv{Computer Science Department}, \orgname{University of Mascara}, \orgaddress{\street{Street}, \city{Mascara}, \postcode{29000}, \state{}, \country{Algeria}}}

\affil*[2]{\orgdiv{LISYS Lab}, \orgname{University of Mascara}, \orgaddress{\street{Street}, \city{City}, \postcode{100190}, \state{State}, \country{Algeria}}}

%\affil[3]{\orgdiv{Department}, \orgname{Organization}, \orgaddress{\street{Street}, \city{City}, \postcode{610101}, \state{State}, \country{Country}}}

%%==================================%%
%% Sample for unstructured abstract %%
%%==================================%%

\abstract{rrrrrrrrrrrrrrrr}

%%\pacs[JEL Classification]{D8, H51}

%%\pacs[MSC Classification]{35A01, 65L10, 65L12, 65L20, 65L70}

\maketitle

\section{Introduction}\label{sec1}

\section{Related works}\label{sec2}

\section{section1 }\label{sec3}

\subsection{section1}\label{subsec2}

\section{section1}\label{sec3}
\subsection{section1}\label{subsec2}
\subsection{section1}\label{subsec2}

\section{section1n}\label{sec3}

\section{Experimental Results}\label{sec3}
\subsection{Experiments settings}\label{subsec2}

Equations in \LaTeX\ can either be inline or on-a-line by itself (``display equations''). For
inline equations use the \verb+$...$+ commands. E.g.: The equation
$H\psi = E \psi$ is written via the command \verb+$H \psi = E \psi$+.

For display equations (with auto generated equation numbers)
one can use the equation or align environments:
% KAN Archecture
\begin{equation}
f(x_1, x_2, \cdots, x_n)=\sum_{q=1}^{2n+1}\Phi_q \left( \sum_{p=1}^{n}\phi_{q, p}(x_p)\right)
\end{equation}

\begin{equation}
\Phi = \{\phi_{p, q}\}, \qquad p=1,2,\cdots, n_{in},\qquad q=1,2,\cdots, n_{out}
\end{equation}

\begin{equation}
x_{l+1,j}=\sum_{i=1}^{n_l}\phi_{l, j, i}(x_{l,i}), \qquad j=1,2,\cdots n_{l+1}
\end{equation}

\begin{equation}
\mathbf{x}_{l+1} = \begin{pmatrix}
\phi_{l,1,1}(\cdot)& \phi_{l,1,2}(\cdot) &\cdots &\phi_{l,1,n_l}(\cdot)  \\
\phi_{l,2,1}(\cdot)&\phi_{l,2,2}(\cdot)  & \cdots &  \phi_{l,2,n_l}(\cdot)\\
\vdots&\vdots  &\ddots  &\vdots  \\
\phi_{l,n_{l+1},1}(\cdot)& \phi_{l,n_{l+1},2}(\cdot) &\cdots  & \phi_{l,n_{l+1},n_l}(\cdot)
\end{pmatrix}\mathbf{x}_{l} =\mathbf{ \Phi}_{l}\mathbf{x}_{l}
\end{equation}
\begin{equation}
\mathbf{KAN(x)}=(\mathbf{\Phi}_{L-1}\circ \mathbf{\Phi}_{L-2}\circ \cdots \circ \mathbf{\Phi}_{1}\circ \mathbf{\Phi}_{0})\mathbf{x}
\end{equation}
% models equations
% system 2
\begin{equation}
y_p(k+1) = f\big[y_p(k), y_p(k-1)\big] + u(k)
\end{equation}
where
\begin{equation*}
f\big[y_p(k), y_p(k-1)\big]=\dfrac{y_p(k)y_p(k-1)[y_p(k)+2.5]}{1+y_p^2(k)+y_p^2(k-1)}
\end{equation*}

% system 3
\begin{equation}
y_p(k+1) = \dfrac{y_p(k)}{1+y_p^2(k)}+ u^3(k)
\end{equation}
% system 4
\begin{equation}
y_p(k+1) = f\big[y_p(k), y_p(k-1), y_p(k-2), u(k), u(k-1)\big] 
\end{equation}
where
\begin{equation*}
f\big[x_1, x_2, x_3,x_4, x_5\big]=\dfrac{x_1x_2x_3x_5(x_3-1)+x_4}{1+x_2^2+x_3^2}
\end{equation*}

\section{Tables}\label{sec5}

Tables can be inserted via the normal table and tabular environment. To put
footnotes inside tables you should use \verb+\footnotetext[]{...}+ tag.
The footnote appears just below the table itself (refer Tables~\ref{tab1} and \ref{tab2}). 
For the corresponding footnotemark use \verb+\footnotemark[...]+

\begin{table}[h]
\caption{Caption text}\label{tab1}%
\begin{tabular}{@{}llll@{}}
\toprule
Column 1 & Column 2  & Column 3 & Column 4\\
\midrule
row 1    & data 1   & data 2  & data 3  \\
row 2    & data 4   & data 5\footnotemark[1]  & data 6  \\
row 3    & data 7   & data 8  & data 9\footnotemark[2]  \\
\botrule
\end{tabular}
\footnotetext{Source: This is an example of table footnote. This is an example of table footnote.}
\footnotetext[1]{Example for a first table footnote. This is an example of table footnote.}
\footnotetext[2]{Example for a second table footnote. This is an example of table footnote.}
\end{table}

\begin{table}[h]
\caption{Example of a lengthy table which is set to full textwidth}\label{tab2}
\begin{tabular*}{\textwidth}{@{\extracolsep\fill}lcccccc}
\toprule%
& \multicolumn{3}{@{}c@{}}{Element 1\footnotemark[1]} & \multicolumn{3}{@{}c@{}}{Element 2\footnotemark[2]} \\\cmidrule{2-4}\cmidrule{5-7}%
Project & Energy & $\sigma_{calc}$ & $\sigma_{expt}$ & Energy & $\sigma_{calc}$ & $\sigma_{expt}$ \\
\midrule
Element 3  & 990 A & 1168 & $1547\pm12$ & 780 A & 1166 & $1239\pm100$\\
Element 4  & 500 A & 961  & $922\pm10$  & 900 A & 1268 & $1092\pm40$\\
\botrule
\end{tabular*}
\footnotetext{Note: This is an example of table footnote. This is an example of table footnote this is an example of table footnote this is an example of~table footnote this is an example of table footnote.}
\footnotetext[1]{Example for a first table footnote.}
\footnotetext[2]{Example for a second table footnote.}
\end{table}

In case of double column layout, tables which do not fit in single column width should be set to full text width. For this, you need to use \verb+\begin{table*}+ \verb+...+ \verb+\end{table*}+ instead of \verb+\begin{table}+ \verb+...+ \verb+\end{table}+ environment. Lengthy tables which do not fit in textwidth should be set as rotated table. For this, you need to use \verb+\begin{sidewaystable}+ \verb+...+ \verb+\end{sidewaystable}+ instead of \verb+\begin{table*}+ \verb+...+ \verb+\end{table*}+ environment. This environment puts tables rotated to single column width. For tables rotated to double column width, use \verb+\begin{sidewaystable*}+ \verb+...+ \verb+\end{sidewaystable*}+.

\begin{sidewaystable}
\caption{Tables which are too long to fit, should be written using the ``sidewaystable'' environment as shown here}\label{tab3}
\begin{tabular*}{\textheight}{@{\extracolsep\fill}lcccccc}
\toprule%
& \multicolumn{3}{@{}c@{}}{Element 1\footnotemark[1]}& \multicolumn{3}{@{}c@{}}{Element\footnotemark[2]} \\\cmidrule{2-4}\cmidrule{5-7}%
Projectile & Energy	& $\sigma_{calc}$ & $\sigma_{expt}$ & Energy & $\sigma_{calc}$ & $\sigma_{expt}$ \\
\midrule
Element 3 & 990 A & 1168 & $1547\pm12$ & 780 A & 1166 & $1239\pm100$ \\
Element 4 & 500 A & 961  & $922\pm10$  & 900 A & 1268 & $1092\pm40$ \\
Element 5 & 990 A & 1168 & $1547\pm12$ & 780 A & 1166 & $1239\pm100$ \\
Element 6 & 500 A & 961  & $922\pm10$  & 900 A & 1268 & $1092\pm40$ \\
\botrule
\end{tabular*}
\footnotetext{Note: This is an example of table footnote this is an example of table footnote this is an example of table footnote this is an example of~table footnote this is an example of table footnote.}
\footnotetext[1]{This is an example of table footnote.}
\end{sidewaystable}

\section{Figures}\label{sec6}

\begin{figure}[h]
\centering
\includegraphics[width=0.9\textwidth]{fig.eps}
\caption{This is a widefig. This is an example of long caption this is an example of long caption  this is an example of long caption this is an example of long caption}\label{fig1}
\end{figure}

In case of double column layout, the above format puts figure captions/images to single column width. To get spanned images, we need to provide \verb+\begin{figure*}+ \verb+...+ \verb+\end{figure*}+.

For sample purpose, we have included the width of images in the optional argument of \verb+\includegraphics+ tag. Please ignore this.

\section{Algorithms, Program codes and Listings}\label{sec7}

You may refer above listed package documentations for more details before setting \verb+algorithm+ environment. For program codes, the ``verbatim'' package is required and the command to be used is \verb+\begin{verbatim}+ \verb+...+ \verb+\end{verbatim}+.

Similarly, for \verb+listings+, use the \verb+listings+ package. \verb+\begin{lstlisting}+ \verb+...+ \verb+\end{lstlisting}+ is used to set environments similar to \verb+verbatim+ environment. Refer to the \verb+lstlisting+ package documentation for more details.

A fast exponentiation procedure:

\lstset{texcl=true,basicstyle=\small\sf,commentstyle=\small\rm,mathescape=true,escapeinside={(*}{*)}}
\begin{lstlisting}
begin
  for $i:=1$ to $10$ step $1$ do
      expt($2,i$);  
      newline() od                (*\textrm{Comments will be set flush to the right margin}*)
where
proc expt($x,n$) $\equiv$
  $z:=1$;
  do if $n=0$ then exit fi;
     do if odd($n$) then exit fi;                 
        comment: (*\textrm{This is a comment statement;}*)
        $n:=n/2$; $x:=x*x$ od;
     { $n>0$ };
     $n:=n-1$; $z:=z*x$ od;
  print($z$). 
end
\end{lstlisting}

\begin{algorithm}
\caption{Calculate $y = x^n$}\label{algo1}
\begin{algorithmic}[1]
\Require $n \geq 0 \vee x \neq 0$
\Ensure $y = x^n$ 
\State $y \Leftarrow 1$
\If{$n < 0$}\label{algln2}
        \State $X \Leftarrow 1 / x$
        \State $N \Leftarrow -n$
\Else
        \State $X \Leftarrow x$
        \State $N \Leftarrow n$
\EndIf
\While{$N \neq 0$}
        \If{$N$ is even}
            \State $X \Leftarrow X \times X$
            \State $N \Leftarrow N / 2$
        \Else[$N$ is odd]
            \State $y \Leftarrow y \times X$
            \State $N \Leftarrow N - 1$
        \EndIf
\EndWhile
\end{algorithmic}
\end{algorithm}

%%=============================================%%
%% For presentation purpose, we have included  %%
%% \bigskip command. Please ignore this.       %%
%%=============================================%%
\bigskip
\begin{minipage}{\hsize}%
\lstset{frame=single,framexleftmargin=-1pt,framexrightmargin=-17pt,framesep=12pt,linewidth=0.98\textwidth,language=pascal}% Set your language (you can change the language for each code-block optionally)
%%% Start your code-block
\begin{lstlisting}
for i:=maxint to 0 do
begin
{ do nothing }
end;
Write('Case insensitive ');
Write('Pascal keywords.');
\end{lstlisting}
\end{minipage}

\section{Cross referencing}\label{sec8}

Environments such as figure, table, equation and align can have a label
declared via the \verb+\label{#label}+ command. For figures and table
environments use the \verb+\label{}+ command inside or just
below the \verb+\caption{}+ command. You can then use the
\verb+\ref{#label}+ command to cross-reference them. As an example, consider
the label declared for Figure~\ref{fig1} which is
\verb+\label{fig1}+. To cross-reference it, use the command 
\verb+Figure \ref{fig1}+, for which it comes up as
``Figure~\ref{fig1}''.

To reference line numbers in an algorithm, consider the label declared for the line number 2 of Algorithm~\ref{algo1} is \verb+\label{algln2}+. To cross-reference it, use the command \verb+\ref{algln2}+ for which it comes up as line~\ref{algln2} of Algorithm~\ref{algo1}.

\subsection{Details on reference citations}\label{subsec7}

Standard \LaTeX\ permits only numerical citations. To support both numerical and author-year citations this template uses \verb+natbib+ \LaTeX\ package. For style guidance please refer to the template user manual.

Here is an example for \verb+\cite{...}+: \cite{bib1}. Another example for \verb+\citep{...}+: \citep{bib2}. For author-year citation mode, \verb+\cite{...}+ prints Jones et al. (1990) and \verb+\citep{...}+ prints (Jones et al., 1990).

All cited bib entries are printed at the end of this article: \cite{bib3}, \cite{bib4}, \cite{bib5}, \cite{bib6}, \cite{bib7}, \cite{bib8}, \cite{bib9}, \cite{bib10}, \cite{bib11}, \cite{bib12} and \cite{bib13}.

\section{Examples for theorem like environments}\label{sec10}

For theorem like environments, we require \verb+amsthm+ package. There are three types of predefined theorem styles exists---\verb+thmstyleone+, \verb+thmstyletwo+ and \verb+thmstylethree+

%%=============================================%%
%% For presentation purpose, we have included  %%
%% \bigskip command. Please ignore this.       %%
%%=============================================%%
\bigskip
\begin{tabular}{|l|p{19pc}|}
\hline
\verb+thmstyleone+ & Numbered, theorem head in bold font and theorem text in italic style \\\hline
\verb+thmstyletwo+ & Numbered, theorem head in roman font and theorem text in italic style \\\hline
\verb+thmstylethree+ & Numbered, theorem head in bold font and theorem text in roman style \\\hline
\end{tabular}

\section{Discussion}\label{sec12}

Discussions should be brief and focused. In some disciplines use of Discussion or `Conclusion' is interchangeable. It is not mandatory to use both. Some journals prefer a section `Results and Discussion' followed by a section `Conclusion'. Please refer to Journal-level guidance for any specific requirements.

\section{Conclusion}\label{sec13}

Conclusions may be used to restate your hypothesis or research question, restate your major findings, explain the relevance and the added value of your work, highlight any limitations of your study, describe future directions for research and recommendations.

In some disciplines use of Discussion or 'Conclusion' is interchangeable. It is not mandatory to use both. Please refer to Journal-level guidance for any specific requirements.

\backmatter

\bmhead{Supplementary information}

If your article has accompanying supplementary file/s please state so here.

Authors reporting data from electrophoretic gels and blots should supply the full unprocessed scans for key as part of their Supplementary information. This may be requested by the editorial team/s if it is missing.

Please refer to Journal-level guidance for any specific requirements.

\bmhead{Acknowledgements}

Acknowledgements are not compulsory. Where included they should be brief. Grant or contribution numbers may be acknowledged.

Please refer to Journal-level guidance for any specific requirements.

\section*{Declarations}

Some journals require declarations to be submitted in a standardised format. Please check the Instructions for Authors of the journal to which you are submitting to see if you need to complete this section. If yes, your manuscript must contain the following sections under the heading `Declarations':

\begin{itemize}
\item Funding
\item Conflict of interest/Competing interests (check journal-specific guidelines for which heading to use)
\item Ethics approval and consent to participate
\item Consent for publication
\item Data availability 
\item Materials availability
\item Code availability 
\item Author contribution
\end{itemize}

%%===================================================%%
%% For presentation purpose, we have included        %%
%% \bigskip command. Please ignore this.             %%
%%===================================================%%

%%===========================================================================================%%
%% If you are submitting to one of the Nature Portfolio journals, using the eJP submission   %%
%% system, please include the references within the manuscript file itself. You may do this  %%
%% by copying the reference list from your .bbl file, paste it into the main manuscript .tex %%
%% file, and delete the associated \verb+\bibliography+ commands.                            %%
%%===========================================================================================%%

\bibliography{sn-bibliography}% common bib file
%% if required, the content of .bbl file can be included here once bbl is generated
%%\input sn-article.bbl

\end{document}

### 2025-01-22 Thesis/Latex/Chapter_2.tex
Updated thesis template with content for chapters

@@
-\chapter{Arabic Language and Machine Translation}
-\pagestyle{fancy}\lhead{\textbf \footnotesize\it{Arabic Machine Translation}}
+\chapter{Large Language Models (LLMs)}
+\pagestyle{fancy}\lhead{\textbf \footnotesize\it{Enhancing Legal Information Access with Retrieval Augmented LLMs for Juridical Data}}
 \pagestyle{fancy}\chead{} \pagestyle{fancy}\rhead{}
-\pagestyle{fancy}\lfoot{\textbf {\small\it{Univ-Mascara/Computer Science: 2024}}} 
+\pagestyle{fancy}\lfoot{\textbf {\small\it{Univ-Mascara/Computer Science: 2025}}} 
 \pagestyle{fancy}\cfoot{} \pagestyle{fancy}\rfoot{\thepage}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Overview} \label{start2}
-Translation has become difficult due to the intricate variations across languages \cite{dorr99}.
-For instance, certain words may have different meanings based on the context, or other words may not have equivalent translations in other languages.
-Additionally, translating idiomatic expressions calls for a thorough understanding of both the source and target languages.
-Further, structural variations like word order disparities between languages complicate translation.
+\section{Introduction} \label{start1}
+Large Language Models (LLMs) represent a transformative leap in Natural Language Processing (NLP), allowing machines to perform complex language tasks with remarkable accuracy.
+From generating coherent text to answering nuanced questions, LLMs have pushed the boundaries of what AI can achieve.
+This chapter delves into the architecture and principles behind LLMs, their development through scaling and training techniques, and the models that have defined this field.
+It also addresses the challenges these models face and their expanding role in various real-world applications.
 
-Another difficulty is that a good translation needs to be faithful and fluent.
-A faithful translation accurately conveys the sense of the original text, whereas a fluent translation is easy to understand and sounds natural.
-A literal, faithful translation could result in an unpleasant and unnatural translation in the target language.
-For example, fluency rather than faithfulness is more critical when translating literary works.
-We might have to alter some of the meaning to keep the text flowing smoothly.
-Readers should feel as if it was written in their native language.
-The faithfulness of the translation, however, is prioritized when translating a technical manual or a legal document.
-Even if the translation is not fluent, it must be faithful and convey the same meaning.
-To produce accurate translations that balance faithfulness and fluency, human translators primarily rely on their experience, knowledge, and reasoning abilities.
-Due to these issues, various human translators will translate the same text in different ways.
+\section{Natural language processing (NLP)}
+Natural Language Processing (NLP) is a field within artificial intelligence (AI) that concentrates on the interaction between computers and human language.
+It involves the development of algorithms and models  that allow machines to comprehend, interpret, and generate human language in a meaningful context.NLP is crucial for enabling computers to process and respond to human language effectively, as demonstrated by features like Google's predictive text in keyboards and language translation systems that manage multiple languages efficiently \cite{murugan2024nlp}
 
-Despite the complex linguistic distinctions, recent decades have seen significant improvements in machine translation.
-It is even applied in practical, real-world applications.
-For instance, we employ machine translation for cross-language information retrieval.
-It allows people to interact and obtain information in other languages.
-Machine translation is also used to assist human translators.
-By creating a draft translation that human translators will edit, it expedites a time-consuming translation task (Plitt and Masselot, 2010).
-In addition, we can employ machine translation for translations that are speech- and image-centric.
-Speech-centric translation involves translating a text from a speech recognition system into another language before the text is fully formed.
-As a result, it mimics a live human interpretation.
-Image-centric machine translation uses an optical character recognition system to translate the text included in images, such as billboard advertisements or street signs.
-
-There are various methods for automating the challenging task of translation.
-Data-driven methods later supplanted the initial rule-based approaches.
-The most popular data-driven methods are Statistical Machine Translation (SMT) and Neural Machine Translation (NMT).
-Nevertheless, due to its remarkable successes, NMT has become state-of-the-art.
-
-Language differences and the various machine translation approaches are briefly discussed in the sections that follow, Sections 2.2 and 2.2.
-Neural networks and NMT are explained in Sections 2.3 and 2.4.
-The evaluation of machine translation is covered in the final section, Section 2.5.
-
-\section{Arabic Language}
-Arabic is considered as one of the six official languages of the United Nation.
-It is the official language in 22 countries and spoken by more than 350 million people in 24 countries around the world \cite{almansor18}.
-The Arabic language is morphologically rich and complex, however, it is considered a low-resource language due to the lack of enough parallel dataset.
-
-Arabic is well known for its complex morphology.
-It has different possibilities of word order that express the same sentence.
-According to word orders, Arabic sentences can be classified into 4 types: SVO\footnote{SVO: Subject-Verb-Object}, VSO, VOS and SOV \cite{alqudsi14}.
-
-Translating the Arabic language into other languages engenders multiple linguistic problems, as no two languages can match, either in the meaning given to the conforming symbols or in the ways in which such symbols are arranged in sentences.
-Lexical, syntactic and semantic problems arise when translating the meaning of Arabic words into English and vice versa.
-Machine translation into morphologically rich languages (MRL) poses many challenges, from handling a complex and rich vocabulary that can reach hundreds of thousands or even millions, to designing adequate MT metrics that take morphology into consideration.
-
-Fehri\cite{fehri93}, Chalabi\cite{chalabi00} and Daimi\cite{daimi01} enumerated major issues involving Arabic in the following points:
-\begin{itemize}
-	\item Arabic is written from right to left.
-	\item There are no capital letters in Arabic.
-	\item Gender is used for all nouns (there are no neutral).
-	\item Some letters have different shapes depending on their location within a word.
-e.g.
-The shape of letter (\RL{`}) in the start of a word is (\RL{`-}) like in \RL{`lb}, in the middle (\RL{-`-}) like in \RL{l`b} and at the end (\RL{`} or \RL{-`}) like in \RL{bl`} or \RL{wd`}.
-	\item Arabic words can be partially, fully or not vocalized. Unvocalized words may generate ambiguities for MT. See Table~\ref{tab_voc}.
-	\begin{table}[!htb]
-		\centering
-		\begin{tabular}{lll}
-			\hline\\[-1.5ex]
-			\multicolumn{1}{l}{Word} & \multicolumn{1}{l}{Transliteration} & \multicolumn{1}{l}{English meaning} \\[0.5ex]
-			\hline\\[-1.5ex]
-			\RL{`amara} & Eamara & build / live / was populated\\[0.5ex]
-			\RL{`ammara} & Eam\~~ara & Live a long time\\[0.5ex]
-			\RL{`omira} & Eomira & became populous\\[0.5ex]
-			\RL{`ommira} & Eom\~~ira & Given a long life\\[0.5ex]
-			\RL{`omaro} & Eumaru & Omar (noun) / plural of Umrah\\[0.5ex]
-			\RL{`am"ro} & Eamoru & Age of\\[0.5ex]
-			\RL{`om"ro} & Eumoru & Age of\\[0.5ex]
-			\RL{`omoro} & Eumuru & Age of\\[0.5ex]
-			\RL{`am"ruN} & EamorN & Age\\[0.5ex]
-			\RL{`om"ruN} & EumorN & Age\\[0.5ex]
-			\RL{`omoruN} & EumurN & Age\\[0.5ex]
-			\RL{`amaruN} & EamarN & Head cover for women\\[0.5ex]
-			\RL{`amarra} & Eamar\~~a & Strong tough man /\\[0.5ex]
-			&  & Longest of everything /\\[0.5ex]
-			&  & Shrewd malicious guy /\\[0.5ex]
-			&  & Ferocious man\\[0.5ex]
-			\hline
-		\end{tabular}
-		\caption{Possible meanings of the unvocalized Arabic word "\RL{`mr}" (Emr)}
-		\label{tab_voc}
-	\end{table}
-
-	\item Arabic has a relatively free order of words. See Table~\ref{tab0}.
-	\begin{table}[!htb]
-		\centering
-		\begin{tabular}{lll}
-			\hline\\[-1.5ex]
-			\multicolumn{1}{c}{Order} & \multicolumn{1}{c}{Sentence} & \multicolumn{1}{c}{English translation} \\[0.5ex]
-			\hline\\[-1.5ex]
-			VSO & \RL{^srba `mro almaA'a} & Drank Omar the water\\[0.3ex]
-			OVS & \RL{almA'a ^srba `mro} & The water drank Omar\\[0.3ex]
-			SVO & \RL{`mro ^srba almaA'a} & Omar Drank the water\\[0.5ex]
-			VOS & \RL{^srba almaA'a `mro} & Drank the water Omar\\[0.5ex]
-			\hline
-		\end{tabular}
-		\caption{Arabic free word order}
-		\label{tab0}
-	\end{table}
-
-	\item Some Arabic vocalized words may have multiple senses (polysemy) depending on its context. See Table~\ref{tab_polysem}. 
-	\begin{table}[!htb]
-		\centering
-		\begin{tabular}{lll}
-			\hline\\[-1.5ex]
-			\multicolumn{1}{l}{Word} & \multicolumn{1}{l}{Transliteration} & \multicolumn{1}{l}{English meanings} \\[0.5ex]
-			\hline\\[-1.5ex]
-			\RL{sA'il} & sA\}il & liquid, beggar, questioner \\[0.5ex]
-			\RL{`ayn} & Eayn & eye, water source, gold \\[0.5ex]
-			\hline
-		\end{tabular}
-		\caption{Examples of vocalized Arabic words polysemy}
-		\label{tab_polysem}
-	\end{table}
-
-	\item The subject can be omitted. e.g. \RL{y^srb almaA'} (He drinks the water)
-	\item Some words hold the meaning of a whole sentence. e.g. \RL{fa'AsqaynAkomUh} (and We gave it to you to drink).
-	\item Copula verbs "to be" and "to have" do not exist in Arabic.
-	\item The three letters root system can often engender ambiguous words.
-	\item Feminine nouns are often derived from masculine nouns, e.g.
-\RL{mhnds} (Engineer male) \RL{mhndsT} (Engineer female).
-In some cases they are totally different, e.g.
-\RL{wld} (Boy) \RL{bnt} (Girl).
-	\item In English the number system moves from singular to plural form directly, however Arabic language includes dual form by suffixing morpheme (\RL{An}) or (\RL{yn}) to the singular form.
-e.g.
-\RL{mhnds} (Engineer male) \RL{mhndsAn} or \RL{mhndsayn} (Two Engineers male).
-	\item The plural form of Arabic masculine nouns is the result of suffixing morpheme (\RL{wn}) or (\RL{yn}) to the singular form.
-	e.g. \RL{mhnds} (Engineer male) \RL{mhndsUn} or \RL{mhndsiyn} (Engineers male).
-	\item  The plural form of Arabic feminine nouns is the result of suffixing morpheme (\RL{At}) to the singular nouns. e.g. \RL{mhndsT} (Engineer female) \RL{mhndsAt} (Engineers female). 
-	\item Some words have no fixed rule for their plural form. e.g. \RL{.tbyb} (Doctor) \RL{'a.tibbaA'} (Doctors).
-\end{itemize}
-In addition to the preceding specific challenges to MT, common standard issues also are present in Arabic language, such as:
-\begin{itemize}
-	\item Multi word expressions (MWE) where the meaning of words collocation varies between partially to completely not derivable from its single constituents \cite{Kordoni14}.
-e.g.
-\RL{qA`dT `skryT}, (military base) is a phrase that is highly compositional.
-\RL{mdynT al-mlAhy}, (amusement park), lit.
-"city of amusements" is an expression that shows a degree of idiomaticity.
-In extreme cases the meaning of the expression as a whole is utterly unrelated to the component words, such as, \RL{frs Alnby}, (grasshopper), lit.
-"Horse of the Prophet".
-
-	\item Idioms and idiomatic expressions are frequently used by Arabic speakers.
-They are a special challenge for MT systems, because their translation mainly does not outcome literally, but logically \cite{Zaninello20}.
-e.g.
-\RL{'a.zlam mn timsA.h} , lit.
-"More oppressive than a crocodile", has the idiomatic meaning of (crocodile tears).
-
-	\item Named entities (NE) refer to abstract entities in the real world such as people (PER) such as \RL{al'amIr `bd-al-qAdr} (Emir Abdelkader), places (LOC) such as \RL{mkT} (Mecca), companies, and organizations that have an appropriate name (ORG) such as \RL{yAhU} (Yahoo).
-It also refers to expressions of date, time, space and quantity (MISC) such as \RL{sbtmbr} 8 (September 8th), \RL{dj}100 (100DZD) or \RL{k.g}25 (25Kgs).
-
-	Due to the rich lexical variations and the absence of capitalization in Arabic, the task of named entities recognition (NER) is more difficult.
-Handling NE in Arabic MT is done very carefully based on :
-	\begin{enumerate}
-		\item meaning translation: e.g. \RL{al-'omam al-motta.hdaT} (The United Nations).
-		\item phoneme transliteration: e.g. \RL{jUjl} (Google).
-		\item mixture of meaning translation and phoneme transliteration: \RL{kArUlAynA al-^smAlyaT} (North Carolina), where "North" is translated and "Carolina" is transliterated.
-	\end{enumerate}   
-\end{itemize}
-
-\section{Social Media}
-Social media involves online spaces where individuals can post and share information with others.
-They have become crucial in contemporary communication, fundamentally reshaping how individuals connect, information disseminates, and communities form.
-These online environments, characterized by user-generated content and interactive features \cite{boyd08},  allow for the creation of virtual spaces fostering real-time interaction and knowledge sharing \cite{yaqub23}.
-
-\textbf{Social Networking Sites (SNS):} Platforms like Facebook prioritize building and maintaining online social connections.
-Users can share personal updates, engage in discussions through comments and reactions, and participate in groups centered around shared interests \cite{segesten22}.
-Research suggests that SNS use can foster social capital by strengthening existing relationships and enabling the formation of new ones \cite{yoon14}.
-
-\textbf{Media-Sharing Platforms:} Platforms like Instagram and YouTube focus heavily on the creation and dissemination of multimedia content.
-Users can share and discover photos, videos, and live streams, with features designed to enhance content visibility through hashtags and algorithmic recommendations \cite{rogers21}.
-Studies highlight the increasing influence of visual communication and the emergence of new social norms surrounding content creation and consumption on these platforms \cite{lu24}.
-
-\textbf{Messaging Applications:} Platforms like WhatsApp and WeChat enable asynchronous and real-time communication between individuals and groups.
-These applications offer features like text messaging, voice calls, and video conferencing, facilitating private and semi-private communication, often replacing traditional communication methods \cite{lenhart15}.
-Research suggests that messaging applications can strengthen interpersonal bonds and serve as crucial tools for information sharing and community mobilization \cite{pang20}.
-
-\textbf{Professional Networking Sites:} LinkedIn serves a distinct purpose within the social media landscape, focusing on professional networking and career development.
-Users create profiles highlighting their skills and experience, connect with colleagues and potential employers, and discover job opportunities \cite{ellison07}.
-Studies suggest that LinkedIn usage can positively impact career prospects by facilitating professional connections and knowledge exchange \cite{davis20}.
-
-Social media platforms are a goldmine for data collection. 
-However, in the case of Arabic, several issues arise due to the way language is used on these platforms. 
-Users often tend to write in colloquial Arabic or a mix of colloquial Arabic and English characters (often referred to as Arabizi). 
-This deviates from the standard Arabic used in formal writing. 
-This informality poses challenges for data collection and analysis, as algorithms need to be able to understand these diverse and non-standard language patterns.
-Another important aspect of Arabic language on social media is the use of dialects.
-Arabic is a highly diverse language, with many different dialects and variations.
-Users may choose to write in their local dialect, which can vary significantly from Modern Standard Arabic, the standardized form of the language.
-With the spread of social media, new challenges arise to the Arabic MT such as:   
-\begin{itemize}
-	\item  \textbf{Non-standard speech}: which encompasses dialectal languages or the various colloquial forms of standard language.
-This text frequently contains slang, MWE and unreasonable abbreviations \cite{Moussallem19} like as "idk\footnote{I don’t know}" and "brb\footnote{Be right back}" in English or "hmd\footnote{\RL{al.hmd lillh} (Thank God)}" and "slm\footnote{\RL{salAm} (Peace)}" in Arabic.
-For best translation results, the MT system is required
-	to identify such argot and try to map it to the target language.
-
-	\item \textbf{Arabizi} (sometimes known as Arabic chat alphabet, Franco-Arabic, Arabish, Araby and	Mu’arrab): which is defined as writing informal Arabic dialects in Latin characters and Arabic numerals.
-This new style of writing doesn’t follow any type of rules, which leads to big variations in writing nearly all Arabic words.
-The following words present some examples of the Arabizi issue that machine translation systems have to solve \cite{Darwish15}:
-	\begin{enumerate}
-		\item The Modern Standard Arabic word \RL{t.hrIr}\footnote{Liberation} has the following popular Arabizi equivalents: ta7rir, tahrir, t7rir, t7reer, ta7reer, tahreer, etc.
-		\item The dialectal spellings of the MSA words \RL{lA-yl`b}\footnote{He does not play} could be \RL{mAyl`b^s}, \RL{mAbyl`b^s}, \RL{myl`b^s}, \RL{myl`b^sI}, \RL{mAyl`b^sI} etc, and the resultant Arabizi could be: mayel3absh, mayel3abch, mabyelaabsh, mabyel3absh, mayel3abshi, mayel3abchi, etc.
-	\end{enumerate}
-\end{itemize}
-
-\section{Natural Language Processing}
-Natural Language Processing (NLP) is a branch of computer science dedicated to facilitating communication between computers and human languages.
-Its methodologies find application in social media for the analysis and comprehension of the extensive textual data produced by users (Huang, 2021).
-Below are several applications of NLP in social media:
-\begin{enumerate}
-	\item \textbf{Machine Translation:} NLP techniques are utilized to translate text from one language to another, facilitating cross-lingual communication and information access.
-Machine translation involves converting text from a source language into a target language while preserving meaning, context, and grammatical structure.
-
-	\item \textbf{Text Classification:} NLP techniques involve categorizing text data into predefined categories or labels based on its content.
-This involves tasks such as sentiment analysis, spam detection, topic classification, and intent recognition in chatbots.
-Text classification is essential for organizing and analyzing large volumes of text data efficiently.
-
-	\item \textbf{Named Entity Recognition (NER):} NLP algorithms aim to identify and extract named entities from unstructured text, such as names of people, organizations, locations, dates, numerical expressions, and other specific entities.
-NER is crucial for various applications, including information retrieval, entity linking, and knowledge graph construction.
-
-	\item \textbf{Speech Recognition:} NLP algorithms transcribe spoken language into text, enabling voice-controlled applications, virtual assistants, and speech-to-text transcription services.
-Speech recognition involves converting audio signals into textual representations, which can then be further processed or analyzed.
-
-	\item \textbf{Information Extraction:} NLP methodologies focus on extracting structured information from unstructured text sources, enabling the retrieval of specific data points or facts from documents, web pages, or other text-based sources.
-Information extraction techniques involve identifying relevant entities, relationships, and events mentioned in the text.
-
-	\item \textbf{Text Generation:} NLP models are employed to generate human-like text based on given prompts or input.
-This includes tasks such as automated content creation, dialogue generation, language translation, and summarization.
-Text generation techniques leverage deep learning architectures such as recurrent neural networks (RNNs) and transformer models.
-
-	\item \textbf{Document Summarization:} NLP techniques condense lengthy documents or articles into shorter summaries while preserving the main points and key information.
-Document summarization aims to provide a concise overview of the text, facilitating easier comprehension and information retrieval.
-
-	\item \textbf{Question Answering:} NLP systems process natural language questions and retrieve relevant answers from large text corpora or knowledge bases.
-Question answering involves understanding the semantics and context of the question, searching for relevant information, and generating accurate responses.
-
-	\item \textbf{Language Understanding:} NLP models aim to understand the semantics, context, and intent behind natural language expressions.
-Language understanding involves tasks such as text comprehension, semantic analysis, and context modeling, enabling machines to interpret human communication more effectively.
-
-	\item \textbf{Dialog Systems:} NLP-powered conversational agents interact with users in natural language, providing assistance, answering queries, and engaging in dialogue across various domains.
-Dialog systems leverage techniques such as natural language understanding (NLU) and natural language generation (NLG) to enable seamless communication between humans and machines.
-\end{enumerate}
-Natural Language Processing (NLP) emerges as a transformative force, enriching our lives by bridging the gap between human language and computing systems.
-This remarkable technology enables machines to comprehend, interpret, and generate human language, revolutionizing how we interact with technology.
-Through NLP, machines can analyze vast volumes of textual data from diverse sources such as social media, news articles, and customer reviews, extracting meaningful insights that shape various aspects of our lives.
-From facilitating accurate language translation and sentiment analysis to powering virtual assistants and chatbots, NLP permeates numerous applications, enhancing efficiency and convenience in our daily activities.
-Its significance lies in its ability to decipher the intricacies of human communication, empowering individuals and businesses alike with deeper understanding and actionable intelligence.
-
-\section{Machine Translation}
-Machine Translation is a procedure that uses computer pieces of software to express text from one natural language NL (SL i.e.
-source language) in another NL (TL i.e.
-target language).
-In any human or automated translation process, the meaning of the source sentences must be fully reproduced into the target translated sentences, which is only simple on the surface.
-
+\subsection{Historical development of NLP}
+The figure below presents a timeline outlining the key developments in Natural Language Processing (NLP)  beginning with rule-based approaches in the 1950s and progressing through the rise of statistical methods and early neural networks in the late 1980s.
+It then highlights the impact of deep learning from the 2000s onward, leading to the development of pre-trained models like BERT and GPT.
+The timeline concludes with the emergence of large language models (LLMs) from 2019 to the present, marking a significant shift in NLP research and applications.
 \begin{figure}[htbp]
-	\centerline{\includegraphics[scale=.7]{Figures/fig1.4.png}}
+	\centerline{\includegraphics[scale=.7]{Figures/the_historyOf_nlp.png}}
 	\caption{The Vauquois triangle, illustrating the foundations of machine translation.}
-	\label{fig1.4.png}
+	\label{the_historyOf_nlp.png}
 \end{figure}
 
-The different approaches to MT fall into three categories: methods that depend on rules and knowledge (linguistic-based).
-Approaches that are empirical and data-driven (corpus-based); and finally, hybrid methods.
+\subsection{Importance of natural language processing}
 
-\subsection{Linguistic Approaches}
-These MT approaches attempt to formalize all the necessary knowledge required for translation, using expert methods.
-The "Vauquois triangle" presented in Figure~\ref{fig1.4.png} is a generic representation of these techniques.
+\begin{enumerate}
+	\item \textbf{Efficient Data Processing:} NLP enables businesses to process and analyze large volumes of unstructured, text-heavy data that would be challenging to handle otherwise..
 
-\subsubsection{\textbf{Direct Approach}}
-or Direct MT (DMT) is, the simplest MT approach.
-It operates at the word level, i.e.
-the words' translation is done word by word, just, as a dictionary does, and generally without much correspondence of their meaning \cite{okpor14}.
+	\item \textbf{Understanding Human Language:}NLP can interpret complex language elements, such as acronyms, abbreviations, and context, improving the accuracy of machine learning models..
 
-\subsubsection{\textbf{Rule-based MT}}
-(RBMT) uses linguistic knowledge of source and target languages fundamentally collected from (bilingual) dictionaries and grammars encompassing the principal morphological, syntactic and/or semantic rules of each language
-respectively \cite{okpor14}. 
-RBMT approach suffers from the impossibility of writing all the rules of all the languages, because this task requires large and important linguistic knowledge. %\cite{Charoenpornsawat02}.
+	\item \textbf{Advancements in Technology:}Improvements in deep learning and machine learning have made NLP more effective, expanding the types of data that can be analyzed.
 
-\subsubsection{\textbf{Interlingual MT}}
-The term "Interlingua" refers to a language that serves as a bridge between two languages.
-In this method, SL is turned into an assistant/mediator language (representation) which is independent of the languages concerned by the translation.
-This auxiliary form is then used to specify the TL's translated verse.
-This approach focuses on a single representation for different languages \cite{Ashraf15}.
+	\item \textbf{Natural Interactions:} NLP allows users to interact naturally with AI chatbots and voice assistants, like Siri, without needing to use specific,predefined\cite{techtarget_nlp}.
 
-\subsubsection{\textbf{Transfer-based MT}}
-(TBMT) is similar to Interlingual-MT in that it generates a translation from an intermediate structure that mimics the original sentence's meaning.
-The source text is translated into a less language-specific intermediate representation.
-This form is then translated into a target language structure with a comparable structure, and the text is generated in the target language.
-The source and target languages' morphological, syntactic, and/or semantic information is used in the transfer process.
-As a result, TBMT can make use of knowledge of both the source and target languages.
-\cite{DO11}.
+\end{enumerate}
 
-\subsection{Corpus Approaches}
-Corpus techniques use empirical methods to ensure that all linguistic knowledge is learned empirically and automatically from corpora, which are collections of parallel datasets of source and target phrases that are translated to each other.
+\section{Neural Networks in NLP}
+A neural network, or artificial neural network, is a machine learning algorithm inspired by the human brain.
+It is a key component of deep learning, a branch of machine learning effective in solving complex problems like image recognition and language processing.
+Unlike traditional computer programs that use a step-by-step algorithmic approach, neural networks learn from examples, mimicking the way neurons in the human brain operate.
+They consist of interconnected nodes (processing elements) that work together in parallel to solve specific problems.
+\\
+A neural network has three basic sections, or parts, each composed of "nodes." \\ The input layer is the first part, receiving raw data, with each node (or neuron) representing a feature of this data.
+The hidden layers, which form the intermediate section, perform various transformations and computations, enabling the network to learn complex patterns and relationships.
+Finally, the output layer, which is the last part, produces the network's output, with the number of nodes corresponding to the desired output classes or regression values \cite{murugan2024nlp}.
+\begin{figure}[htbp]
 
-\subsubsection{\textbf{Example-based MT}}
-The main idea behind (EBMT) is analogy \cite{Nagao84}.
-The primary concept is to build new translations on top of current examples.
-Bilingual parallel corpora containing sentence pairs are used to train EBMT systems.
-It's used to translate similar-sounding sentences by looking for the closest source example to the source word or phrase in parallel corpora.
-Nagao has appropriately classified this procedure into three steps \cite{Nagao84}:
-\begin{itemize}
-	\item Fragments are matched against a database of real examples.
-	\item Identifying the translation fragments that correlate (Alignment)
-	\item Putting these together to create the target text
-\end{itemize}
-
-\subsubsection{\textbf{Statistical MT}}
-(SMT) generates translation hypotheses in a target language $t$ based on a sentence in a source language $s$ with the highest conditional probability $P(t|s)$ \cite{brown88,brown93}.
-The translation direction will be inverted to a translation model (TM) $P(s|t)$ and a language model (LM) $P(t)$ will be included by applying the Bayes rule.
-The following equation~\eqref{eq:eq1} is used to optimize the likelihood of the best translation:
-\begin{equation}
-	\label{eq:eq1}
-	t_{best}=arg_{t}max(P(t|s))=arg_{t}max(P(s|t) \times P(t))
-\end{equation}
-where $P(s|t)$ is the TM and $P(t)$ is the LM.
-
-\begin{figure}[htbp]
-	\centerline{\includegraphics[scale=0.3]{Figures/fig1.13.png}}
-	\caption{Statistical Machine Translation approach}
-	\label{fig1.13.png}
+	\centerline{\includegraphics[width=.7\linewidth]{Figures/ANN.png}}
+	\caption{Artificial Neural Network (ANN).}
+	\label{ANN.png}
 \end{figure}
 
-SMT requires a language model, a translation model, and a decoding method in general.
-The TM, on the one hand, assures that the target hypothesis created matches the source sentence.
-The LM, on the other hand, ensures that the output is grammatically correct  (Figure~\ref{fig1.13.png}).
+\subsection{Basics Concepts in Neural Networks}
+\textbf{Weights} \\
+Variables on the edges between nodes that are multiplied with node outputs to form the input for the next layer.Weights are crucial for training and tuning a neural network and are often initialized within a range of -1 to 1.
+\\ \\
+\textbf{Biases} \\
+Additional nodes in hidden and output layers that connect to every node within their respective layers but not to the previous layer.
+Biases add a constant value (typically 1 or -1) to the input of a layer, helping shift the activation function and aiding in effective learning.
+\textbf{Activation Functions} \\
+These functions introduce non-linearity to the network, enabling it to learn and model complex data patterns.
+Common activation functions include the sigmoid, hyperbolic tangent (tanh), and rectified linear unit (ReLU).
+\cite{taylor2017neural}.
 
-\section{Evaluation}
-As an integral part of design science research, evaluation is crucial to assessing the quality of machine translation outputs and comparing different machine translation methods.
-We need to measure quality to track progress, ideally with a single score.
-However, devising such a score is still an open research question (Koehn, 2020).
-Nonetheless, some best practices have already been established, and in general,
-there is broad consensus on how to track quality gains.
-Human (Section 2.5.1) and automatic (Section 2.5.2) evaluation methods exist.
-Human evaluation seems more accurate than automatic evaluation because the
-translation is, after all, intended for humans. However, running a human evaluation can be time-consuming and expensive.
-In practice, it can be used to compare a small number of variant systems.
-Therefore, automated metrics are prevalent because they can rapidly evaluate system improvements; they are also used as a loss function for training models.
+\subsection{ Recurrent Neural Networks (RNNs)}
+Recurrent neural networks (RNNs) are a type of neural network designed for processing sequential data, such as text or time series, by effectively handling variable-length inputs.
+An RNN consists of a hidden state h and an optical output y, which operate on a variable  sequence x=($x_1$....$x_t$)At each time step t, the hidden state $h^t$
+of the RNN is updated according to the function $h^{(t)} = f(h^{(t-1)}, x_t)$
+where f is a non-linear activation function, such as a logistic sigmoid or a more complex long short-term memory (LSTM) unit.
+This design allows RNNs to remember past inputs and incorporate them into future outputs, enabling them to recognize patterns that occur at multiple positions within a sequence.
+Additionally, RNNs utilize parameter sharing across time steps, which helps in generalizing across sequences of varying lengths and learning complex dependencies over time.
+Despite challenges with long-term dependencies, RNNs are particularly effective for tasks requiring context or sequential understanding, making them valuable for various applications, including time series analysis and natural language processing\cite{cho2014learning}
+\begin{figure}[htbp]
 
-\subsection{Human Evaluation}
-Human evaluation can be considered the most common method of judging and measuring translation quality.
-The linguistics and translation experts can judge the quality of a translation system output from two corners:
+	\centerline{\includegraphics[width=1\linewidth]{Figures/RNN.png}}
+	\caption{RNN Architecture.}
+	\label{RNN.png}
+\end{figure}
+\subsection{ Long-Short Term Memory (LSTM)}
+Long Short-Term Memory (LSTM) networks, designed by Hochreiter and Schmidhuber, are an improved version of recurrent neural networks (RNNs) designed to learn long-term dependencies in sequential data, making them suitable for tasks like time series forecasting and language translation
+They address the limitations of traditional RNNs by introducing a memory cell that maintains information over extended periods.
+This memory cell is regulated by three gates—input, forget, and output gates—which control the flow of information in and out of the cell as shown in the fig \cite{geeksforgeeks_lstm}
+\begin{figure}[htbp]
+	\centering\includegraphics[width=0.9\linewidth]{Figures/LSTM.png}
+	\caption{LSTM Architecture.}
+	\label{/LSTM.png}
+\end{figure}
+Forget Gate: Decides what information to discard from the cell state
+\[
+f_t = \sigma \left( W_f \cdot [h_{t-1}, x_t] + b_f \right)
+\]
+
+Input Gate: Determines what new information to add to the cell state
+\[
+i_t = \sigma \left( W_i \cdot [h_{t-1}, x_t] + b_i \right)
+\]
+\[
+\tilde{C}_t = \tanh \left( W_c \cdot [h_{t-1}, x_t] + b_c \right)
+\]
+
+Output Gate: Controls what information to output from the cell state
+\[
+o_t = \sigma \left( W_o \cdot [h_{t-1}, x_t] + b_o \right)
+\]
+\[
+h_t = o_t \odot \tanh(C_t)
+\]
+
+It is worth mentioning that Bidirectional LSTM networks extend the traditional LSTM architecture by processing data in both forward and backward directions.
+This approach enables the network to capture dependencies from both past and future contexts, improving its ability to resolve temporal dependencies.
+Bidirectional LSTMs are particularly effective at handling multidimensional problems, encapsulating spatially and temporally distributed information, and dealing with incomplete data through flexible connection mechanisms \cite{hochreiter1997long}.
+\section{ Transformer Architecture}
+The Transformer architecture is a deep learning model introduced in June 2017 by Vaswani et al.
+from Google Brain.
+Their paper, titled "Attention Is All You Need," presented a groundbreaking approach to processing sequential data through the use of a self-attention mechanism.
+This innovative method allows the model to assign different levels of importance to various parts of the input, enabling it to capture long-range dependencies much more effectively than earlier models like RNNs and LSTMs.The original Transformer model is structured as a stack of six layers, where the output of each layer i serves as the input to the subsequent layer i+1, continuing this process until the final prediction is reached.It features a six-layer encoder on the left and a corresponding six-layer decoder on the right,  both of which work together to transform input sequences into meaningful outputs.
+Each encoder and decoder consists of six identical layers that allow the model to process and generate language efficiently \cite{rothman2021transformers}.
+\begin{figure}[htbp]
+
+	\centerline{\includegraphics[width=.6\linewidth]{
+			Figures/trasformers.png}}
+	\caption{The Transformer - model architecture..}
+	\label{trasformers.png}
+
+\end{figure}
+\subsection{Encoder and Decoder Stacks}
+\textbf{Encoder:} Each layer in the encoder consists of two sub-layers: \\
+Multi-head self-attention mechanism: This allows the model to focus on different parts of the input sequence. \\
+Position-wise feed-forward network: A fully connected network applied to each position separately and identically. \\
+Residual Connections: Residual connections are applied around each sub-layer, followed by layer normalization. The output of each sub-layer is computed as Layer Norm(x + Sublayer(x)). \\
+Output Dimension: All sub-layers and embedding layers output vectors of dimension d-model = 512.\\
+\textbf{Decoder:} Similar to the encoder, the decoder also has 6 identical layers, each containing the same two sub-layers:
+Multi-head self-attention mechanism ,
+Position-wise feed-forward network. \\
+Additional Sub-layer: The decoder includes a third sub-layer that performs multi-head attention over the output of the encoder stack. \\
+Residual Connections and Normalization: Like the encoder, residual connections are applied around each sub-layer, followed by layer normalization. \\
+Masked Self-Attention: The self-attention mechanism in the decoder is modified to prevent positions from attending to subsequent positions, ensuring that predictions for position i depend only on the known outputs at positions before i \cite{vaswani2017attention}.
+\begin{figure}[htbp]
+
+	\centerline{\includegraphics[width=.8\linewidth]{
+			Figures/incoderDecoder.png}}
+	\caption{Encoder and Decoder Stacks}
+	\label{incoderDecoder.png}
+
+\end{figure}
+\subsection{Self-Attention Mechanism}
+Self-attention is a mechanism that allows the model to weigh the importance of different words in a sequence relative to each other.
+This is crucial for understanding context and relationships between words \cite{vaswani2017attention}.
+\subsection{Query, Key, and Value Vectors:}
+\textbf{Query Vectors (Q):} Generated from the input word's embedding and represent the word's role in finding relevant information in other words.\\
+\textbf{Key Vectors (K):} Derived similarly from embedding and serve as a comparison standard for each word.\\
+\textbf{Value Vectors (V):} Contain the actual information to be passed on and are weighted according to relevance determined by the Query and Key comparison \cite{vaswani2017attention}.. 
+\subsection{Self-Attention Calculation:}
+The score is computed by taking the dot product of the Query vector of a word with the Key vectors of all other words.
+This score determines how much focus should be placed on each word when encoding the current word.
+The scores are scaled and passed through a softmax function, producing a distribution of attention weights. \\
+Each Value vector is weighted by these attention scores and summed to produce the final output of the self-attention layer\cite{vaswani2017attention}.
+\subsection{Role of Multi-Head Attention}
+Multi-head attention enhances the model’s ability to focus on different aspects of the input by creating multiple sets of Query, Key, and Value vectors.
+Each set is processed independently and then combined.
+\\
+This approach allows the model to attend to information from different representation subspaces at different positions, improving its ability to capture complex relationships\cite{vaswani2017attention}.
+\subsection{Positional Encoding:}
+Since self-attention mechanisms do not inherently capture the order of words, positional encoding is added to the word embeddings to provide information about the relative position of each word in the sequence.
+These encodings are added directly to the input embeddings, allowing the model to consider both the position and the content of 
+each word during processing\cite{alammar2018illustratedtransformer}.
+\subsection{Transformers in NLP}
+Transformers have revolutionized the field of Natural Language Processing (NLP) by introducing a powerful and efficient way to handle textual data.
+This advancement has led to the creation of highly effective models like BERT and GPT, with its deep bidirectional context, excels at understanding and improving performance across various NLP tasks.
+GPT, on the other hand, is renowned for generating coherent and contextually relevant text, significantly advancing applications such as text generation and translation\cite{devlin2019bert}.\\
+\textbf{Scalability:} Transformers efficiently handle large datasets and long sequences, overcoming the limitations of RNNs.
+This scalability allows for training with billions of parameters, enhancing model capabilities.\\
+\textbf{Rich Contextual Understanding: }The self-attention mechanism in transformers captures relationships between words across the entire sequence, enabling deep contextual understanding and more accurate language processing.\\
+\textbf{Model Efficiency:} Transformers enable parallel processing, which speeds up training and makes them more efficient than sequential models like RNNs.
+This efficiency supports the rapid development and deployment of advanced language models\cite{vaswani2017attention}.
+\section{ Emergence of Large Language Models (LLMs)}
+
+Large Language Models (LLMs) are advanced artificial intelligence systems designed to process and generate human-like text.
+They are typically based on transformer architectures and are characterized by their enormous scale, with billions of parameters.
+LLMs are pre-trained on vast amounts of text data in a self-supervised manner, enabling them to develop a broad understanding of language.
+They are capable of performing a wide range of tasks with minimal task-specific fine-tuning, often achieving significant performance improvements through few-shot or zero-shot learning\cite{brown2020language}.
+\subsection{Scaling in Large Language Models (LLMs)}
+Scaling is crucial in the evolution of large language models.
+The history of scaling shows that increasing both model size and dataset size leads to significant improvements in performance across various NLP tasks.
+For instance, early work by Brants et al.
+(2007) demonstrated the benefits of using language models trained on vast datasets, such as 2 trillion tokens, which led to significant advancements in machine translation quality.
+This was followed by efforts like those of Heafield et al.
+(2013), who scaled traditional models to Web-scale data, and Jozefowicz et al.
+(2016), who scaled LSTMs to 1 billion parameters, achieving state-of-the-art results on large benchmarks.The advent of transformer-based models marked a significant shift.Models like BERT, GPT-2, and GPT-3, with their enormous parameter counts—up to 175 billion for GPT-3—demonstrated that scaling up not only the model but also the dataset size yields substantial gains in performance.
+Researchers like Kaplan et al.
+(2020) and Hoffmann et al.
+(2022) studied how scaling affects model performance, proposing power laws that show a predictable relationship between model size, dataset size, and performance.
+These studies emphasized the importance of scaling for the continued progress of LLMs\cite{touvron2023llama}.
+\subsection{Pre-training}
+Pre-training is a crucial stage in developing Large Language Models (LLMs), where the model learns from extensive unlabeled datasets through a process called self-supervision.
+This stage allows the model to recognize and internalize a wide range of linguistic patterns, laying the groundwork for fine-tuning on specific tasks.\\
+Several pre-training objectives have been implemented to maximize the effectiveness of this learning process, each offering distinct benefits to the model's performance
+\textbf{Full Language Modeling:} Used since GPT-2, this approach trains decoder-only models to predict the next token in a sequence based on previous tokens.
+This autoregressive method enables models like GPT-3 to generate coherent and contextually relevant text.\\
+\textbf{Prefix Language Modeling: }Employed in encoder-decoder and non-causal decoder-only models, this technique uses a non-causal (considering both past and future tokens) prefix for predicting subsequent tokens, offering more flexibility and enhancing the model's adaptability across various language tasks.\\
+\textbf{Masked Language Modeling: }Popularized by BERT, this method involves masking certain tokens in the input text and training the model to predict them, helping the model understand word context.
+An extension, span corruption, masks entire text spans for prediction, further improving contextual comprehension\cite{wang2023language}.
+\begin{figure}[htbp]
+
+	\includegraphics[width=1\linewidth]{Figures/pretraining.png}
+	\caption{Training Tokenization in Full, Prefix, and Masked Language Modeling.}
+\end{figure}
+\subsection{Fine-tuning }
+Fine-tuning has been the most common approach for adapting LLMs to specific tasks.
+After pre-training on large datasets, these models are further trained on smaller, supervised datasets tailored to the desired task.
+Fine-tuning updates the weights of the pre-trained model, allowing it to excel at specific tasks like sentiment analysis, machine translation, or question answering.
+The main advantage of fine-tuning is that it often results in strong performance on task-specific benchmarks.
+However, it comes with challenges, such as the need for a new large dataset for every task, the risk of overfitting to narrow distributions, and potential poor generalization to out-of-distribution examples\cite{touvron2023llama}.
+\subsection{Few-Shot, One-Shot, and Zero-Shot Learning}
+Few-shot, one-shot, and zero-shot learning represent different paradigms of using LLMs without extensive fine-tuning.
+These methods involve using pre-trained models to perform tasks with minimal task-specific data.\\
+\textbf{Few-Shot Learning} involves giving the model a few examples (typically 10-100) of the task during inference.
+This approach reduces the need for large, task-specific datasets and limits the risk of overfitting to narrow data distributions.
+However, few-shot results are generally not as strong as those from fully fine-tuned models.\\
+\textbf{One-Shot Learning} is similar to few-shot learning but uses only a single example alongside a natural language description of the task.
+This method mirrors how humans are often instructed for tasks, making it an interesting approach for tasks where providing multiple examples is impractical.\\
+\textbf{Zero-Shot Learning} requires the model to perform a task based solely on a natural language description, with no examples provided.
+While this method offers maximum flexibility and robustness, it is also the most challenging for models.
+Zero-shot performance is often weaker than few-shot or one-shot, but it represents a significant step towards task-agnostic AI, where models can generalize across a wide range of tasks with minimal human intervention\cite{brown2020language}.
+\subsection{Evaluation Datasets and Tasks}
+Evaluating Large Language Models (LLMs) is essential for determining their effectiveness and limitations in comprehending and generating human language.
+This evaluation typically falls into two primary categories:
 \begin{itemize}
-	\item \textbf{Fluency}: The level of smoothness and coherence of the translated text to target language norms,  such as grammaticality, intelligibility and clarity.
-When annotating fluency, the evaluators (fluent only in the target language) have access to only the translation being evaluated and not the source text which is not relevant to the fluency assesment.
-	\item \textbf{Adequacy}: Also known as accuracy.
-It stands for the correspondence of the target text to the source text, including the expressive means in translation, and how well the target text represents the informational content of the source text.
-In this case, the bilingual evaluators (in both the source and target languages) have access to the source text and translations being evaluated and habitually, they take into consideration the context of the sentence.
+	\item Natural Language Understanding (NLU):This measures the model's proficiency in comprehending language, encompassing tasks such as sentiment analysis, text classification, natural language inference (NLI), question answering (QA), commonsense reasoning (CR), mathematical reasoning (MR), and reading comprehension (RC).
+	\item Natural Language Generation (NLG):This assesses the model's capability to produce text based on given context.
+It includes tasks like summarization, sentence completion, machine translation (MT), and dialogue generation.
 \end{itemize}
-The fluency and adequacy are usually measured on a 5-point scale, as presented in the following table \ref{tab:table1.1} \cite{mauvcec19}.
+\textbf{Benchmarks} play a critical role in evaluating LLMs, providing standardized tests to measure their performance across various tasks:
+\begin{itemize}
+	\item MMLU: Measures model knowledge from pretraining and evaluates performance in zero-shot and few-shot scenarios across 57 subjects, testing world knowledge and problem-solving abilities.
+	\item SuperGLUE: An advanced benchmark that builds on GLUE, assessing tasks like question answering and natural language inference.
+It is designed to test deeper aspects of language understanding and requires significant advancements in various learning methodologies.
+	\item BIG-bench: A large-scale benchmark for evaluating LLMs across diverse tasks including reasoning, creativity, ethics, and domain-specific knowledge.
+	\item GLUE: A foundational benchmark for evaluating and analyzing natural language understanding, offering a range of resources for model assessment\cite{Naveed2024}.
+\end{itemize}
+\section{Popular Models and Datasets}
+Large Language Models (LLMs) like GPT-3, GPT-4, and BERT have revolutionized NLP by leveraging vast datasets such as Common Crawl and WebText.
+These datasets provide a diverse linguistic foundation, enabling models to perform a wide range of tasks with remarkable accuracy and contextual understanding.
+\subsection{GPT-N Models}
+GPT models are advanced autoregressive language models that generate substantial and complex machine-produced text from minimal input.
+They leverage deep learning techniques to mimic human text generation by predicting the current value based on preceding values.
+NLP models initially struggled with tasks outside their training sets due to data restrictions. OpenAI addressed this with GPT-1,introduced in 2018.
+\begin{itemize}
+	\item GPT-1, trained on the BooksCorpus dataset, utilized a 12-layer transformer decoder with self-attention.
+Its pre-training allowed for zero-shot performance on various tasks, demonstrating the potential of generative language models.
+	\item In 2019, GPT-2 improved upon GPT-1 by using a larger dataset and 1.5 billion parameters (compared to GPT-1’s 117 million).
+It excelled in tasks like translation and summarization, enhancing accuracy in recognizing long-distance relationships.
+	\item GPT-3, released later, featured around 175 billion parameters and was trained on the Common Crawl dataset.
+It could generate human-like text, perform basic math, and write code.
+Despite its capabilities, its size and cost made it challenging to implement.
+	\item GPT-4, launched in March 2023, advanced further with multimodal capabilities and context windows of up to 32,768 tokens.
+It incorporates reinforcement learning for better alignment with human input and policy\cite{yenduri2023gpt}
+\end{itemize}
+\subsection{BERT}
+BERT (Bidirectional Encoder Representations from Transformers) is a groundbreaking language representation model that pretrains deep bidirectional representations from unlabeled text by jointly conditioning on both left and right contexts in all layers.
+This bidirectional approach enables BERT to be fine-tuned with just a single additional output layer for various tasks, such as question answering and language inference, without the need for extensive architectural modifications.
+BERT has achieved remarkable performance, setting new state-of-the-art results across eleven NLP tasks, including a GLUE score of 80.5\%, MultiNLI accuracy of 86.7\%, SQuAD v1.1 F1 score of 93.2\%, and SQuAD v2.0 F1 score of 83.1\%.
+Its combination of conceptual simplicity and empirical effectiveness makes BERT a powerful and versatile tool for a wide range of natural language processing applications\cite{devlin2019bert}.
+\subsection{T5}
+The T5 (Text-To-Text Transfer Transformer) model represents a unified framework for natural language processing (NLP), designed to address various text processing tasks by treating them as "text-to-text" problems.
+This approach involves converting all tasks into a format where both input and output are text, allowing the same model, objective, and training procedure to be applied across different tasks.
+T5 leverages extensive pre-training on a large, unlabeled dataset, enabling it to develop general-purpose knowledge that enhances its performance on a range of NLP tasks, such as question answering, document summarization, and sentiment classification\cite{raffel2023exploring}.
+\subsection{LLAMA2}
+\subsection{WebText }
+WebText is a dataset developed by OpenAI to encapsulate a wide variety of high-quality web content.
+It comprises text sourced from outbound links on popular websites like Reddit, chosen based on their popularity and the quality of their content.
+Unlike Common Crawl, which indiscriminately includes a broad range of web pages, WebText is more selective, focusing on high-quality, engaging material.
+Although smaller in size compared to Common Crawl, WebText is curated to ensure the text reflects diverse, well-written content, making it particularly valuable for training language models to produce coherent and contextually appropriate text\cite{radford2019language}.
+\subsection{RuleTaker}
+RuleTaker is a collection of datasets designed to assess the deductive reasoning capabilities of language models.
+Each dataset includes a set of facts, rules, and a boolean question that the model must use logical reasoning to answer.
+The datasets feature synthetic subsets requiring various levels of reasoning complexity, with different numbers of deduction steps necessary to reach an answer.
+Among the datasets are the Bird dataset, which illustrates McCarthy’s abnormality problem , the Electricity dataset, which models appliance functions, and the ParaRules corpus, where sentences like “Bob is cold” are paraphrased to “In the snow sits Bob, crying from being cold.” The collection comprises 580,000 training examples, 84,000 validation examples, and 173,000 testing examples\cite{helwe2024}.
+\begin{figure}
+	\centering
+	\includegraphics[width=0.7\linewidth]{Figures/dataset-exmple.png}
+	\caption{exammple of RuleTaker dataset }
+\end{figure}
+\section{Challenges and Limitations}
+LLMs have greatly advanced NLP, but they also present challenges such as high computational costs, issues with adversarial robustness, and difficulties in interpretability.
+As these models scale, they encounter new challenges in scalability, privacy, and real-time processing.
+Ongoing research is exploring multi-modality, transfer learning, and continuous learning, which bring additional complexities and highlight the evolving impact of LLMs in practical applications\cite{Naveed2024}.
+\subsection{ Computational Cost}
+Training large language models (LLMs) demands significant computational resources, leading to higher production costs and environmental concerns due to the substantial energy used in large-scale training.
+Although enhancing computational resources can boost performance, the gains diminish over time when the size of the model and dataset stay constant, adhering to the power law of diminishing returns.
+\subsection{Overfitting}
+Despite their advanced learning abilities, they can overfit to noisy or unusual patterns in their large training datasets, which may result in generating nonsensical responses.
+The ongoing discussion about Memorization versus Generalization in LLMs revolves around finding an optimal balance.
+Memorization helps the model retain specific details from its training, allowing it to answer precise questions accurately.
+On the other hand, generalization enables the model to make predictions and generate responses for new, unseen inputs, which is crucial for handling diverse real-world tasks.
+The challenge is to find the right balance: excessive memorization can lead to overfitting, reducing the model's flexibility and ability to handle novel inputs.
+\subsection{ Interpretability and Explainability}
+The "black-box" nature of LLMs makes it difficult to understand how they make decisions, which is vital for gaining broader acceptance and trust, particularly in sensitive areas.
+Although these models have advanced capabilities, the lack of transparency into their workings limits their effectiveness and reliability.
+Efforts are underway to enhance the explainability of LLMs to build user trust and promote responsible use of AI.
+Understanding how LLMs generate their responses is crucial for ensuring they align with human values and legal standards.
+\subsection{ Hallucinations}
+LLMs sometimes produce "hallucinations," or responses that, despite appearing plausible, are incorrect or do not match the provided information.
+These hallucinations can be classified into three types:
 
-\begin{table}[ht]% <--- it say "let figure be here"
-	\centering
-	\resizebox{\textwidth}{!}{%
-		\begin{tabular}{@{}lccccc@{}}
-			\toprule
-			\multirow{2}{*}{\textbf{Adequacy}} & \textbf{1}           & \textbf{2}           & \textbf{3}           & \textbf{4}           & \textbf{5}           \\ \cmidrule(l){2-6} 
-			& None                 & Little meaning       & Much meaning         & Most meaning         & All meaning          \\ \midrule
-			& \multicolumn{1}{l}{} & \multicolumn{1}{l}{} & \multicolumn{1}{l}{} & \multicolumn{1}{l}{} & \multicolumn{1}{l}{} \\ \midrule
-			\multirow{2}{*}{\textbf{Fluency}}  & \textbf{1}           & \textbf{2}           & \textbf{3}           & \textbf{4}           & \textbf{5}           \\ \cmidrule(l){2-6} 
-			& Incomprehensible     & Disfluent language   & Non-native language  & Good language        & Flawless language    \\ \bottomrule
-		\end{tabular}%
-	}
-	\caption{Numeric scale for adequacy and fluency evaluation.}
-	\label{tab:table1.1}
-\end{table}
+\begin{itemize}
+	\item Input-conflicting hallucination: When the model generates responses that do not align with the user's input.
+	\item Context-conflicting hallucination: When the model produces content that contradicts information it has previously generated.
+	\item Fact-conflicting hallucination: When the model creates responses that conflict with established knowledge.
+\end{itemize}
+\subsection{Privacy Concerns}
+As Large Language Models (LLMs) have become more complex and sizable, privacy concerns have intensified, particularly regarding data sharing and potential misuse.
+Risks include the creation of harmful content, evasion of filters, and issues related to data privacy, especially in areas like e-commerce where safeguarding customer information is vital.
+When LLMs are trained on private data and then made publicly available, additional privacy risks arise.
+Since LLMs can memorize phrases from their training data, there's a danger that these phrases could be exploited by malicious actors to retrieve sensitive information, thus threatening personal privacy.
+\subsection{Real-Time Processing}
+Real-time processing in Large Language Models (LLMs) is crucial for many applications, particularly as mobile AI solutions become more popular and concerns about information security and privacy grow.
+However, LLMs typically consist of hundreds of layers and millions of parameters, which create significant challenges for real-time processing due to their high computational demands and the limited storage capacity of hardware platforms, especially in edge computing environments.
+Although efforts such as MobileBERT attempt to lessen memory usage, they still encounter considerable execution overhead because of the extensive number of model layers, resulting in high inference latency.
+\section{Domains of Application}
+The use of Large Language Models (LLMs) in various downstream tasks has become increasingly prevalent in both AI research and industry, with new applications being identified and explored regularly.
+These models, which excel at understanding and generating human-like text, are finding valuable applications across diverse fields.
+\subsection{Law} 
+In the legal field, Large Language Models (LLMs) can support the analysis of legal documents by assisting with tasks such as generating initial coding for datasets, identifying key themes, and classifying information accordingly.
+This synergy between legal professionals and LLMs has been effective in examining legal texts, such as court opinions on theft, enhancing both the efficiency and quality of legal research.
+Additionally, LLMs have been tested for their capability to generate explanations of legal terms, with a focus on improving factual accuracy by integrating sentences from relevant case law.
+By incorporating pertinent case law, these enhanced models can produce more accurate and relevant explanations with fewer factual errors.
+Furthermore, LLMs can be specialized with domain-specific knowledge to tackle legal reasoning tasks and address legal queries effectively \cite{Naveed2024}.
+\subsection{Cybersecurity}
+large Language Models (LLMs) have garnered significant attention in the field of cybersecurity.
+Recent research has highlighted their potential in addressing software bugs created by human developers and identifying cybersecurity threats.
+For example, Arora et al.
+have proposed methods for utilizing LLMs to evaluate cyber threats on social media through sentiment analysis.
+LLMs are also employed to detect cybersecurity-related information in Open Source Intelligence (OSINT), aiding in the identification of potential cyber threats.
+Additionally, LLMs have shown promise in detecting scams, such as phishing.
+Initial tests with models like GPT-3.5 and GPT-4 have demonstrated their ability to recognize common phishing indicators in emails.
+While LLMs exhibit considerable potential in cybersecurity, improving their reasoning abilities could enhance their effectiveness further, such as in uncovering zero-day vulnerabilities in open-source software by analyzing logic and source code\cite{helwe2024}.
+\subsection{Medicine}
+The integration of Large Language Models (LLMs) into medicine is transforming both healthcare delivery and research.
+In clinical settings, LLMs are increasingly utilized in decision support systems to offer evidence-based treatment recommendations.
+By analyzing patient data and medical literature, these models can assist in diagnosing conditions, suggesting relevant tests, and proposing effective treatment options.
+Additionally, LLMs improve patient interactions through applications like chatbots that answer questions about symptoms and medications, schedule appointments, and provide health advice.\\
+In medical research, LLMs help sift through vast amounts of literature to extract, filter, and summarize relevant information, identify key studies, and predict future research directions.
+They also play a role in medical education by generating training materials, creating exam questions, explaining complex topics, and offering personalized feedback.
+Furthermore, LLMs simulate patient interactions, aiding students in honing their clinical skills \cite{Naveed2024}.
 
-Human evaluation is intrinsically subjective and expensive in time and money.
-To reduce the problem of subjectivity, more experts are usually invited to evaluate the same translations in the ES, and their assessments are, eventually, justified statistically.
+\subsection{journalism}
+Large Language Models (LLMs) offer valuable support to journalists, especially in fact-checking and news verification.
+They can process and cross-reference large volumes of data with established knowledge bases.
+Research has demonstrated that LLMs, such as GPT-3.5, can be used to detect fake news by providing rationales that enhance other models like BERT, which can then be fine-tuned for this purpose .
+In addition to fact-checking, LLMs are useful for analyzing political debates, helping journalists to identify key themes, monitor how discussions evolve, and evaluate sentiments.
+They can also assist in detecting logical fallacies and underlying motives in political discourse and propaganda .
+By enhancing their reasoning capabilities, LLMs can uncover deeper insights into propaganda and misinformation, making them a powerful tool for modern journalism\cite{helwe2024}.
 
-\subsection{Automatic Evaluation} 
-Generally, translation evaluation methods are based on counting word- and/or sentence-based errors likely to be identified automatically.
-Correlation with human evaluation is the measure of evaluation for metrics.
-Different metrics are used in MT evaluation: BLEU \cite{papineni02}, NIST \cite{doddington02}, METEOR \cite{banerjee05}, TER \cite{snover06}, LEPOR \cite{han12} and many others. 
-All of these metrics require reference translations because they confront the MT output sentences with reference translations and produce comparison scores.\\
-\textbf{BLEU} metric is one of the first and most used metrics to return high correlation with human evaluation of quality. 
-It measures the overlap of single words (unigrams) and \textit{n}-grams between MT output and reference translations. 
-BLEU operates by not only counting matching words between the translation and reference but also accounting for n-gram alignments. 
-This approach values proper word sequencing, enhancing the chances of aligning word pairs (bigrams) or longer sequences like trigrams or 4-grams. 
-Additionally, multiple reference translations can be employed to assess the presence of $n$-gram matches across variations.
-The BLEU score of a machine-translated output relies on the adjusted $n$-gram precision along with a brevity penalty. 
-Essentially, precision measures the proportion of $n$-grams in the machine translation output that align with the reference translation.
-BLEU scores are calculated across an entire test set, typically with one or more reference translations. 
-However, it's uncommon in practice to utilize multiple reference translations.
-\begin{equation}
-	BLEU = BP * exp \sum_{n=1}^N log \frac{matching\_i\_grams}{total\_i\_grams}
-%	BLEU=BP*\left(\sum_{n=1}^N w_n log\left(precision_n\right)\right)
-	\label{eqn:BLEU}
-\end{equation}
-The brevity penalty (BP) in Equation \ref{eqn:BLEU}, which penalizes shorter output, is expressed as:
-\begin{equation*}
-	BP = min(1,\frac{output\_length}{reference\_length}).
-\end{equation*}
-%\[
-%BP=\begin{cases} 
-%	1 & if c > r\\ 
-%	e^{(1-r/c)} & if c \leq r
-%\end{cases}
-%\]
-\\BLEU suffers from notable drawbacks, one of which is its stringent nature. 
-For instance, it incorporates trigram or 4-gram precision in its calculation. 
-However, a translated sentence might lack any trigram or 4-gram matches with the reference translation, leading to a BLEU score of zero.
-\section{Related Work}
-The main focus of AMT (Arabic Machine Translation) research initially was on translating from Modern Standard Arabic (MSA) to English, with significantly less emphasis on the reverse direction, from English to Arabic, and even fewer efforts dedicated to translations between Arabic and other languages.
-The first direct-based machine translation system from English to Arabic was developed by Weidner Communication Inc.
-In 1990, Apptek introduced ArabTrans MT, a tool for translating from English to Arabic.
-Additionally, products like Al-Mutarjim Al-Araby, Al-Alamiyah, and Al-Nakheel were capable of translating between French and Arabic as well as English and Arabic. 
-A critical challenge in using the Interlingua approach for AMT is the construction of representations that resolve ambiguity and accurately reflect the semantic structure of the language. 
-There have been limited studies that have developed and assessed models using this method.
-Table~\ref{tab_linguistic} summarizes the surveyed AMT research studies.
-\begin{small}
-	\begin{longtable}{|l|l|l|l|l|}
-		\hline
-		Year & Research & SL-TL$^{\mathrm{a}}$ & Method & Score\\
-		\hline
-		\endhead
-		%	\hline
-		%	\endfoot
-
-		\multicolumn{5}{c}{\textbf{\textcolor{blue}{Linguistic-based AMT researches}}}\\
-		%	\hline
-		\multicolumn{5}{l}{\textbf{Direct-based AMT}}\\
-		\hline
-		2005 & Al-Taani \& Hailat \cite{altaani05}					& En$\rightarrow$Ar &-& 57.3\%\\
-		2007 & Ittycheriah \& Roukos \cite{ittycheriah-roukos-07}	& Ar$\rightarrow$En &Word alignment& Bl 51.27\\
-		\hline
-		\multicolumn{5}{l}{\textbf{Rule-based AMT}}\\
-		\hline
-		1995 & Mankai \& Mili \cite{mankai95}				& Ar$\rightarrow$En/Fr &-&\\
-		2008 & Salem et al. \cite{salem08} 					& Ar$\rightarrow$En &-&\\
-		2008 & Nguyen \& Vogel \cite{nguyen08}				& Ar$\rightarrow$En &-& 56.03\\
-		2008 & Samy	\& & Ar-Sp-En &-&\\
-		 & Gonz{\'a}lez-Ledesma \cite{samy08}	&  &-&\\
-		2008 & AbuShuqier \& Sembok \cite{abushquier08} 	& En$\rightarrow$Ar &-& 96.1\%\\
-		2009 & Elming \& Habash \cite{elming09}				& En$\rightarrow$Ar &-&\\
-		2009 & Besançon et al. \cite{besanccon09} 			& En/Fr$\rightarrow$Ar &-&\\
-		2012 & Salloum \& Habash \cite{salloum12}			& DA$\rightarrow$MSA &-&\\
-		2020 & Sghaier \& Zrigui \cite{sghaier20}			& DA$\rightarrow$MSA &-& Bl 55.22\\
-		\hline
-		\multicolumn{5}{l}{\textbf{Interlingua-based AMT}}\\
-		\hline
-		2002 & Soudi et al. \cite{soudi02}					& En$\rightarrow$Ar &-&\\
-		2006 & Shaalan et al. \cite{shaalan06}				& En$\rightarrow$Ar &-&\\
-		2008 & Bouillon et al. \cite{bouillon08}			& Jp$\leftrightarrow$Ar &-&\\
-		2014 & Al Ansary \cite{AlAnsary14}					& En$\rightarrow$Ar &Universal Networking Language&\\
-		\hline
-		\multicolumn{5}{l}{\textbf{Transfer-based AMT}}\\
-		\hline
-		2002 & Attia \cite{attia02}							& En$\rightarrow$Ar &Agreement features&\\
-		2004 & Shaalan et al. \cite{shaalan04}				& En$\rightarrow$Ar &-& 92\%\\
-		2010 & Shirko et al. \cite{shirko10}				& Ar$\rightarrow$En &-& 94.6\%\\
-		2010 & Shaalan et al. \cite{shaalan10a}				& En$\leftrightarrow$Ar &-& 0.450%4
-		/0.458%1
-		\\
-		2011 & Hatem et al. \cite{hatem11}					& En$\rightarrow$Ar &Morphological analysis&\\
-		\hline
-		\multicolumn{5}{c}{\textbf{\textcolor{blue}{Corpus-based AMT researches}}}\\
-		%	\hline
-		\multicolumn{5}{l}{\textbf{Example-based AMT}}\\
-		\hline
-		2002 & Guidere \cite{guidere02}					& Fr-Ar &-&\\
-		2011 & Bar \& Dershowitz \cite{Bar11}			& Ar$\rightarrow$En &Verb paraphrases& 23.98\\
-		2012 & Bar \& Dershowitz \cite{Bar12}			& Ar$\rightarrow$En &Semantic equivalents&\\
-		2012 & Cavalli-Sforza \& Phillips \cite{Cavalli-Sforza12}	& Ar$\rightarrow$En &Morphological analysis&\\
-		2014 & El-Shishtawy 		& En$\rightarrow$Ar &Template-based syntactic &\\
-		&\& El-Sammak \cite{elshishtawy14}		& & matching&\\
-		\hline
-		\multicolumn{5}{l}{\textbf{Statistical-based AMT}}\\
-		\hline
-		2006 & Hasan et al. \cite{hasan06}				& Ar$\rightarrow$Fr &Pre-/Post-processing& 40.8\%\\
-		2007 & Diab et al. \cite{diab07}				& Ar$\rightarrow$En &Pre-/Post-processing& 45.38\%\\
-		2007 & Sarikaya \& Deng \cite{sarikaya07}		& En$\rightarrow$Ar &POS tag$^{\mathrm{b}}$/CDW$^{\mathrm{d}}$&+0.3\\
-		2009 & Badr et al. \cite{badr09}				& En$\rightarrow$Ar &Syntactic phrase reordering&Bl 32.46\\
-		%s		2009 & Mostefa et al. \cite{mostefa09}					& Ar-Fr-En\\[0.3ex]
-		2009 & Habash \& Hu \cite{habash-hu09}		& Ar$\rightarrow$En$\rightarrow$Cn &MT Evaluation&Bl +1.1\\
-		2010 & Carpuat et al. \cite{carpuat10}			& Ar$\rightarrow$En &Word reordering&Bl 51.70\\
-		2010 & Ghurab et al. \cite{ghurab10}			& Ar$\leftrightarrow$Cn &-&0.805%3
-		/0.696%5
-		\\
-		2010 & Bisazza and Federico \cite{bisazza10}	& Ar$\rightarrow$En &Word reordering& 48.96\\
-		2017 & Durrani et al. \cite{durrani17}			& Ar$\leftrightarrow$En &MT Evaluation& Bl +4\\	
-		2017 & Mallek et al. \cite{mallek17}			& Ar$\leftrightarrow$En &Pre- processing& Bl 10.98\\
-		2017 & Ebrahim et al. \cite{ebrahim17}			& En$\rightarrow$Ar &MWE Detection& 19.31/19.22\\
-		2019 & Aqlan et al. \cite{aqlan19a}				& Ar$\rightarrow$Cn &Morpho/Vocab/POS tag& Bl 19.40\\
-		\hline
-		\multicolumn{5}{c}{\textbf{\textcolor{blue}{Hybrid AMT}}}\\
-		\hline
-		2004 & Alsharaf et al. \cite{alsharaf04}	& Fr$\rightarrow$Ar& Direct+Transfer+Pivot+SMT &\\
-		2008 & Toutanova et al. \cite{toutanova08}	& En$\rightarrow$Ar& Inflection prediction models &Bl 36.92\\
-	 & 	& &  + SMT &\\
-		2008 & Hatem \& Nassar \cite{hatem08}		& En$\rightarrow$Ar& Rule-based + Example-based & 68\%\\
-		2009 & Matusov et al. \cite{matusov09} 		& Ar$\rightarrow$En& Multi-engine (5 MT systems) &\\
-		2009 & Habash et al. \cite{habash09}		& Ar$\rightarrow$En& SMT + Rule-based &Bl 0.4162\\
-		2010 & Al Dam \& Guessoum \cite{aldam10}   	& En$\rightarrow$Ar& Transfer-based + ANN &56\%\\
-		2010 & Sawaf \cite{sawaf10}					& DA$\rightarrow$MSA& Rule-based+SMT & 42.1\%\\
-		2011 & Alawneh \& Sembok \cite{alawneh11}	& En$\rightarrow$Ar& Rule-based + Example-based &\\
-		2012 & Shaalan \& Hany \cite{shaalan12}		& Ar$\rightarrow$En& Rule-based + Example-based & WER 88\%\\
-		2014 & Akeel \& Mishra \cite{akeel14}   	& En$\rightarrow$Ar& Rule-based + ANN & 0.6\\
-		2015 & Mohamed \& Sadat	\cite{mohamed15}	& Ar$\rightarrow$Fr& Morphological rule + SMT & 34.7\%\\
-		2015 & Zantout \& Guessoum \cite{zantout15}	& Ar$\rightarrow$En& Transfer-based + ANN &64.50\%\\
-		\hline
-		\multicolumn{3}{l}{$^{\mathrm{a}}$Source language-Target language}&
-		\multicolumn{2}{l}{$^{\mathrm{b}}$Part Of Speech Tagging}\\
-		\multicolumn{3}{l}{$^{\mathrm{c}}$Context Dependent Words}\\
-		\caption{Linguistic-, Corpus-based and Hybrid AMT researches}
-		\label{tab_linguistic}\\
-	\end{longtable}
-\end{small}
-
-\section{Summary}
-In conclusion, this chapter has delved into the multifaceted challenges and intricacies surrounding Arabic language processing, with a specific focus on machine translation (MT).
-We have explored the intricate aspects of Arabic, ranging from its morphology and syntax to its diverse array of dialects, all of which present formidable hurdles for natural language processing (NLP) and MT systems alike.
-Additionally, we have scrutinized the influence of social media on Arabic NLP and MT, taking into account the unique linguistic characteristics and informal language usage prevalent on online platforms.
-
-Our examination has encompassed both linguistic and corpus-based approaches to Arabic MT, underscoring the necessity of tailored resources and datasets to capture the language's nuances effectively. 
-Moreover, we have delved into various MT evaluation methods, encompassing both human and automatic assessment techniques.
-
-Lastly, we have surveyed the related work in the field, providing insights into recent advancements, methodologies, and the challenges encountered in Arabic language MT research, thus offering valuable perspectives on the current landscape and future trajectories of the field.
+\section{Conclusion}
+This chapter has outlined the evolution and significance of Natural Language Processing (NLP), from early neural networks to advanced Large Language Models (LLMs).
+We covered key concepts in neural networks and the transformative impact of the Transformer architecture.
+The exploration of LLMs highlighted their emergence, scaling, and the intricacies of pre-training and fine-tuning.
+We also discussed various challenges, including computational costs, interpretability, and privacy concerns.
+Finally, the chapter reviewed the applications of LLMs in fields like law, cybersecurity, medicine, and journalism, showcasing their potential and the ongoing need for further development.

### 2025-01-22 Thesis/Latex/Chapter_3.tex
Updated thesis template with content for chapters

@@
-\chapter{Neural Networks and Machine Translation}
-\pagestyle{fancy}\lhead{\textbf \footnotesize\it{Neural Networks and Machine Translation}}
+\chapter{RAG (Retrieval-Augmented Generation)}
+\pagestyle{fancy}\lhead{\textbf \footnotesize\it{Enhancing Legal Information Access with Retrieval Augmented LLMs for Juridical Data}}
 \pagestyle{fancy}\chead{} \pagestyle{fancy}\rhead{}
-\pagestyle{fancy}\lfoot{\textbf {\small\it{Univ-Mascara/Computer Science: 2024}}} 
+\pagestyle{fancy}\lfoot{\textbf {\small\it{Univ-Mascara/Computer Science: 2025}}} 
 \pagestyle{fancy}\cfoot{} \pagestyle{fancy}\rfoot{\thepage}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Overview}\label{start3}
-In this chapter, we embark on an in-depth exploration of neural networks, pivotal to understanding the intricate workings of machine translation. 
-Beginning with a thorough examination of the foundational feed-forward networks (FFNs), recurrent neural networks (RNNs), long short-term memory networks (LSTMs), and gated recurrent units (GRUs), we lay the groundwork for comprehending the neural machine translation (NMT) process.
+\section{ Introduction}\label{start2}
 
-We delve into the nuances of training NMT models, and the decoding phase, where trained models generate translations from input sequences. 
-Finally, we conducted a comprehensive survey of recent advancements and methodologies in neural machine translation research, with a focus on exploring related works specifically dedicated to the Arabic language context.
+Large language models (LLMs) have seen significant advancements but face challenges, particularly in tasks that demand extensive knowledge or deal with queries beyond their training data.
+These limitations often result in inaccuracies or “hallucinations.” To address this, Retrieval-Augmented Generation (RAG) supplements LLMs by retrieving relevant document chunks from external knowledge sources based on semantic similarity.
+By doing so, RAG helps reduce factual errors, allowing LLMs to produce more accurate content.
+This has led to the widespread adoption of RAG, particularly in chatbots and other real-world applications, making it a crucial technology in advancing the capabilities of LLMs.
+\section{Fundamentals of Retrieval-Augmented Generation}
+RAG represents a significant advancement in the capabilities of large language models (LLMs) by integrating external knowledge retrieval with the generation of text.
+Understanding the individual components of retrieval and generation is essential to appreciate how their synergy improves the overall performance of RAG systems.
+\subsection{Definition of RAG}
+Retrieval-Augmented Generation (RAG) is a technique that leverages the strengths of pre-trained large language models (LLMs) and external data sources.
+By merging the generative abilities of LLMs like GPT-3 or GPT-4 with the accuracy of specialized data search mechanisms, RAG systems can generate more sophisticated and contextually relevant responses\cite{selvaraj2024}.
+\subsection{Historical Development of RAG}
+The foundations of Retrieval-Augmented Generation (RAG) can be traced back to traditional information retrieval (IR) systems,  Early models primarily relied on keyword matching and simple ranking mechanisms to retrieve relevant documents from databases, establishing a basic framework for information access.
+The landscape began to shift with the rise of neural networks in the 2010s.
+Notable innovations like Word2Vec and Transformer-based architectures enabled retrieval methods to incorporate deeper semantic understanding, paving the way for more sophisticated document retrieval processes.
+A significant advancement occurred with the introduction of dense passage retrieval (DPR) in 2020, which utilized bi-encoder architectures to map both queries and documents into dense vector spaces. 
+The integration of these advanced retrieval techniques with large language models (LLMs) became a focal point as models like GPT-3 gained prominence.
+Researchers explored how LLMs could be augmented with retrieval capabilities, leading to the development of RAG systems that leverage the strengths of both retrieval and generative capabilities\cite{gao2024retrieval}.
+\subsection{ Differences Between RAG and Fine-Tuning}
+The enhancement of large language models (LLMs) has gained significant interest due to their increasing use.
+Among the various optimization strategies for LLMs, Retrieval-Augmented Generation (RAG) is often compared to fine-tuning (FT) and prompt engineering.
+Each method possesses unique attributes,
+\begin{enumerate}
+	\item Methodological Characteristics :\\
+	RAG is compared to providing a tailored textbook for information retrieval, making it suitable for precise information retrieval tasks.
+It excels in dynamic environments with real-time knowledge updates.
+\\
+	Fine-tuning is likened to a student internalizing knowledge, making it more static and suitable for replicating specific structures, styles, or formats.
+	\item External Knowledge and Adaptation: \\
+	RAG relies on external knowledge sources and allows for high interpretability but may involve higher latency and ethical considerations regarding data retrieval. \\
+	Fine-tuning requires retraining for updates and involves significant computational resources for dataset preparation and training.
+It enables deep customization of the model’s behavior but may struggle with unfamiliar data.
+	\item Performance: \\ RAG consistently outperforms unsupervised fine-tuning in knowledge-intensive tasks, particularly for both previously encountered and new knowledge.
+	LLMs often struggle to learn new factual information through unsupervised fine-tuning.
+	\item Use Cases and Combination: \\
+	The choice between RAG and fine-tuning depends on specific needs for data dynamics, customization, and computational capabilities.
+They are not mutually exclusive; their combined use may yield optimal performance and often requires multiple iterations to refine the results \cite{gao2024retrieval}.
 
-\section{Neural Networks}
-Neural networks are the foundational computational mechanisms driving Neural Machine Translation, employing complex algorithms to mimic human language understanding and generation. 
-At the core of these networks is the node, or processing unit, which functions similarly to neurons in the human brain. 
-Each node is designed to receive inputs—typically a vector of real-valued numbers—process these inputs through a series of mathematical operations, including weighted sums and activation functions, and then produce an output that can be passed on to subsequent layers or nodes in the network.
-The following subsections will delve into the different types of neural networks, beginning with the simplest form: the feed-forward neural network.
+\end{enumerate}
+\section{Types of RAG Systems}
+The RAG research field is continuously evolving, with three main stages: Naive RAG, Advanced RAG, and Modular RAG.
+\subsection{Naive RAG}
+Naive RAG, a foundational approach in Retrieval-Augmented Generation, operates on a straightforward "Retrieve-Read-Generate" paradigm. This method involves three primary steps:
 
-\subsection{Feed-Forward Neural Networks}
-A feed-forward neural network is a basic type of neural network where information moves in one direction from the input to the output layer without any loops or cycles back to previous layers. 
-Outputs from each node are forwarded upwards to the subsequent layer without any feedback to lower levels. 
-An illustration, referred to as Figure~\ref{fig:ffn}, shows a FFN consisting of two layers. 
-These networks are composed of three distinct types of nodes: input, hidden, and output. 
-The input nodes receive individual scalar values, and notably, among these input nodes is a constant bias node, labeled $x_0$, which is always set to the value of 1. 
-In the hidden layer, nodes calculate a weighted sum of their inputs and then apply a nonlinear function to this sum to determine their output.
-\begin{figure}[h]
-	\centerline{\includegraphics[width=0.7\linewidth]{Figures/FFN}}
-	\caption{A feed-forward network consisting of two layers includes an input layer $x$, a hidden layer $h$, and an output layer $y$.}
-	\label{fig:ffn}
-\end{figure}
-In a standard neural network structure, every layer is comprehensively interconnected, allowing nodes to receive inputs from all nodes in the preceding layer. 
-These networks are characterized by their depth, with numerous layers contributing to their complexity. 
-Hidden nodes within these layers act as automatic feature detectors, eliminating the need for manual feature identification by learning to recognize relevant patterns in the input data through training.
-Each hidden node is defined by its parameters, including a weight vector and a bias term. 
-The parameters for the entire hidden layer are aggregated into a weight matrix $U$, where each weight $u_{jk}$ in $U$ corresponds to the connection strength from the $k^{th}$ input node $x_k$ to the $j^{th}$ hidden node $h_j$, and a bias vector that applies to the entire layer.
+\begin{itemize}
+	\item Indexing: Raw data is cleaned, extracted, and converted into a uniform text format.
+It's then segmented into smaller chunks and encoded into vector representations using an embedding model.
+These vectors are stored in a vector database for efficient similarity searches.
+	\item Retrieval: When a user query is received, it's encoded into a vector and compared to the indexed chunks.
+The top K most similar chunks are retrieved and included in the prompt for the language model.
+	\item Generation: The query and retrieved chunks are combined into a prompt, which is fed to a large language model.
+The model generates a response based on the provided context and its internal knowledge.
+\end{itemize}
+While Naive RAG offers a basic framework, it faces several challenges in 
+\begin{itemize}
+	\item Retrieval Issues: The retrieval process can be imprecise, leading to the selection of irrelevant or missing information.
+	\item Generation Challenges: The language model may hallucinate, generating content not supported by the retrieved context. It may also produce irrelevant, toxic, or biased outputs.
+	\item Augmentation Difficulties: Integrating retrieved information into the generation process can be challenging, leading to disjointed or redundant responses.
+\end{itemize}
+To address these limitations, more sophisticated RAG techniques have emerged, which we will explore in the next section.
+\subsection{Advanced RAG}
+Taking aim at the shortcomings of Naive RAG, Advanced RAG introduces specific improvements to enhance retrieval quality. This approach utilizes pre-retrieval and post-retrieval strategies.
+\subsubsection{Pre-retrieval Strategies:}
+\begin{itemize}
+	\item Enhanced Indexing: Advanced RAG tackles indexing issues through a sliding window approach, finer segmentation of data, and inclusion of metadata.
+Additionally, it optimizes the retrieval process by employing various methods.
+	\item Query Optimization: This stage focuses on refining the user's initial query to make it clearer and more suitable for retrieval.
+Techniques like query rewriting, transformation, and expansion are commonly used.
+\end{itemize}
+\subsubsection{Post-Retrieval Strategies:}
+\begin{itemize}
+	\item Re-ranking Chunks: After relevant information is retrieved, Advanced RAG prioritizes the most relevant content by re-ranking the retrieved chunks and placing them strategically within the prompt.
+	\item Context Compression: To avoid overwhelming the LLM with too much information, post-retrieval efforts focus on selecting the most essential parts of the retrieved context, highlighting critical sections, and compressing the data to be processed.
+\end{itemize}
+By addressing indexing issues and refining the query and retrieved information, Advanced RAG aims to improve the overall accuracy and relevance of the generated response.
+\subsection{Modular RAG} 
+Modular RAG represents the latest evolution in RAG, offering greater adaptability. it introduces specialized modules and innovative patterns to enhance retrieval and processing capabilities.
+\subsubsection{New Modules}
+\begin{itemize}
+	\item Search Module: Adapts to specific scenarios by leveraging LLM-generated code and query languages to search across various data sources.
+	\item RAG Fusion: Employs a multi-query strategy to expand user queries, uncover both explicit and implicit knowledge, and improve retrieval results.
+	\item Memory Module: Leverages the LLM's memory to guide retrieval and create an unbounded memory pool, aligning the text more closely with data distribution.
+	\item Routing Module: Navigates through diverse data sources, selecting the optimal pathway for a query based on its specific needs.
+	\item Predict Module: Reduces redundancy and noise by generating relevant context directly through the LLM.
+	\item Task Adapter Module: Tailors RAG to various downstream tasks by automating prompt retrieval and creating task-specific retrievers.
+\end{itemize}
+\subsubsection{New Patterns} 
+\begin{itemize}
+	\item Flexible Module Arrangement: Modular RAG allows for the substitution and reconfiguration of modules to address specific challenges, surpassing the fixed structures of previous RAG paradigms.
+	\item Innovative Retrieval Strategies: Techniques like Rewrite-Retrieve-Read, Generate-Read, and Recite-Read leverage the LLM's capabilities to refine queries, generate content, and retrieve information from model weights.
+	\item Hybrid Retrieval: Combines keyword, semantic, and vector searches to cater to diverse queries, improving retrieval relevance.
+	\item Dynamic Module Interaction: Frameworks like Demonstrate-Search-Predict and ITERRETGEN demonstrate the dynamic use of module outputs to enhance each other's functionality.
+	\item Adaptive Retrieval: Techniques like FLARE and Self-RAG evaluate the necessity of retrieval based on different scenarios, allowing for a more flexible and efficient approach.
 
-A notable feature of neural networks is the incorporation of non-linear activation functions, with the rectified linear unit ($ReLU$) standing out for its simplicity and widespread usage. 
-$ReLU$ is particularly efficient to compute, as depicted in Figure~\ref{fig:relu} (a). 
-This function returns the value of $z$ when $z$ is greater than 0 and returns $0$ when $z$ is less or equal to 0.
-\begin{figure}[h]
-	\centering
-	\includegraphics[width=0.7\linewidth]{Figures/Relu}
-	\caption{Common activation functions utilized in neural networks.}
-	\label{fig:relu}
-\end{figure}
-The $sigmoid$ (or logistic) function, another frequently utilized activation function, has been illustrated in Figure~\ref{fig:relu} (b).
-An activation function similar to the sigmoid but more widely employed is the hyperbolic tangent ($tanh$), depicted in Figure~\ref{fig:relu} (c).
+\end{itemize}
+\section{Core Components of RAG}
+The core components of Retrieval-Augmented Generation (RAG) systems consist of a retrieval mechanism, a generation process, and augmentation techniques.
+These elements work together to enhance the model's ability to access relevant external information, generate coherent and contextually appropriate responses, and improve overall performance in knowledge-intensive tasks.
+\subsection{ Retrieval Mechanism}
+RAG systems combine parametric memory (a pre-trained language model) with non-parametric memory (a retrieval mechanism).
+The retrieval mechanism allows RAG models to access external information sources (e.g., Wikipedia), and this process is central to improving the model's ability to generate factual and accurate outputs \cite{lewis2020retrieval}.
+\begin{itemize}
+	\item Traditional Techniques: \\
+	Classical retrieval methods include \textbf{TF-IDF and BM25}, which rely on sparse vector representations based on term frequencies.
+These methods use exact keyword matching, making them limited in semantic understanding.
+	\item Modern Retrieval Approaches:\\
+	\textbf{ Dense Passage Retrieval (DPR): }This approach utilizes dense vector representations learned by neural models like BERT to encode both the query and document.
+It allows for a more semantic understanding of the content, making retrieval more effective.
+DPR, for example, computes the similarity between the query and documents using Maximum Inner Product Search (MIPS), which finds the closest passages based on the dense vector space.
+	\item Implementation in RAG:\\ In RAG, dense retrieval methods are often paired with techniques like \textbf{Maximum Inner Product Search} (MIPS), which efficiently matches query and document embeddings.
+This enables the system to return the most relevant documents for subsequent generation \cite{karpukhin2020dense}.
 
-The computation of the hidden layer in the basic feed-forward network can be performed efficiently using straightforward matrix operations, involving three main steps:
+\end{itemize}
+\subsection{Generation Process}
+The generation in RAG occurs through a combination of retrieved passages and the input query.\\
+Large Language Models (LLMs) like BART or T5 are utilized for this generation, leveraging their advanced capabilities in understanding and producing human-like text.
+The retrieval component supplies external context, which the generator conditions on to formulate a response.This context is crucial, as it enriches the LLM's understanding, enabling it to incorporate real-time, relevant information.
+Moreover, the generation process benefits from the LLM's ability to synthesize information, allowing it to create responses that are not only coherent but also informed by the latest data, thus improving accuracy and relevance in knowledge-intensive tasks \cite{lewis2020retrieval}.
+\\
+Two models have been proposed in RAG:
 \begin{itemize}
-	\item The multiplication of the weight matrix by the input vector $x$,
-	\item Addition of the bias vector, and
-	\item Application of the activation function $f$, such as $ReLU$, $sigmoid$, or $tanh$.
+	\item \textbf{RAG-Sequence} uses the same document to generate the entire sequence of output. It marginalizes over the top K retrieved documents to produce a final answer.
+	\item \textbf{RAG-Token} allows different tokens in the output sequence to be generated based on different documents, making it more flexible when combining information from various
 \end{itemize}
-Hence, the neural network depicted in Figure~\ref{fig:ffn} can be expressed mathematically as follows: 
+\subsection{ Augmentation Techniques}
+The augmentation of the retrieval process in Retrieval-Augmented Generation (RAG) systems focuses on improving how queries are refined and how relevant information is retrieved for downstream generation tasks.
+Key augmentation methods include.
 \begin{itemize}
-	\item A set of input nodes represented by the vector $x = (x_1, x_2, x_3, . . . , x_n)^T$;
-	\item A set of hidden nodes represented by the vector $h = (h_1, h_2, h_3, . . . , h_l)^T$;
-	\item A set of output nodes represented by the vector $y = (y_1, y_2, y_3, . . . , y_m)^T$;
-	\item A weight matrix connecting input nodes to hidden nodes denoted by $U = u_{jk}$;
-	\item A weight matrix connecting hidden nodes to output nodes denoted by $W = w_{ij}$.
+	\item \textbf{Query Augmentation:} In traditional retrieval pipelines, user queries are often under-specified or ambiguous, leading to poor retrieval performance.
+Query augmentation involves dynamically rewriting the user's query to better match the documents in the knowledge base.
+This can be done by leveraging large language models (LLMs) to generate tailored queries or synthetic questions and answers (QAs) that better align with the search objective​
+	\item \textbf{Synthetic QA Generation:} Instead of using raw document chunks, retrieval is enhanced by generating and embedding synthetic QA pairs from documents.
+This helps to capture the semantic essence of long texts more effectively, reducing noise and improving retrieval precision.
+These synthetic QAs can be used to rewrite the user query, making it more specific to the task at hand \cite{mombaerts2024meta}.
 \end{itemize}
-The computation of the hidden layer output, represented by the vector $h$, is performed according to Equation~\ref{eqn:hj}, incorporating the activation function $f$.
-\begin{equation}
-	h_j = f(\sum_{k}u_{jk}x_k)
-	\label{eqn:hj}
-\end{equation}
-The value $h$ obtained serves as a representation of the input. 
-Subsequently, the output layer's function is to utilize this revised representation $h$ and calculate a conclusive output, as described in Equation~\ref{eqn:yi}.
-\begin{equation}
-	y_i = \sum_{j}w_{ij}h_j
-	\label{eqn:yi}
-\end{equation}
-In numerous instances, this output may initially be a real-valued number; however, it is often transformed into a probability distribution using a $softmax$ function.
-For a vector $y$ with a dimensionality of $d$, the $softmax$ function is defined as shown in Equation~\ref{eqn:softmax_yi}, where $1 \leq i \leq d$.
-\begin{equation}
-	softmax(y_i) =\frac{e^{y_i}}{\sum_{j}e^{y_j}}
-	\label{eqn:softmax_yi}
-\end{equation}
-The $softmax$ function operates on a vector $y = [y_1, y_2, y_3, .
-.
-.
-, y_m]$ of arbitrary values, transforming them into a probability distribution where each value falls within the range of (0, 1) and the sum of all values equals one.
-
-\subsection{Recurrent Neural Networks}
-A neural network containing a cycle inside its network connections is called a recurrent neural network (RNN). 
-Its preceding outputs directly or indirectly influence the value of a node. 
-Figure~\ref{fig:rnn} illustrates the structure of a simple RNN based on \cite{elman90}. 
-Similar to conventional FFNs, the values for a layer of hidden nodes are calculated by multiplying an input vector representing the current input, $x$, by a weight matrix and then passing the result through a non-linear activation function.
-The associated output, $y$, is then determined using the hidden layer, which comprises the hidden nodes. 
-The context layer is where it differs from a FFN the most. 
-It keeps the previous values and sends them to the appropriate nodes in the hidden layer. 
-This layer uses the hidden layer’s value from the previous time step as input to the computation at the hidden layer.
-\begin{figure}[h]
-	\centering
-	\includegraphics[width=0.7\linewidth]{Figures/RNN}
-	\caption{A basic recurrent neural network.}
-	\label{fig:rnn}
-\end{figure}
-
-\subsubsection{Stacked Recurrent Neural Networks} 
-Stacked recurrent neural networks (RNNs) consist of multiple layers, where the output of one layer serves as the input to the next layer. 
-Information flows from lower layers to higher layers, and the final output is generated by the top layer. 
-Stacked RNNs have shown improved performance in neural MT compared to single-layer networks \cite{bahdanau15}. 
-This improvement can be attributed to the network's ability to create representations at different levels of abstraction across the layers. 
-However, the optimal number of stacked RNNs depends on the availability of training data. 
-While abundant data can lead to better generalization, limited data may result in overfitting \cite{lample18}. 
-Moreover, increasing the number of stacked layers escalates the training costs.
-
-\subsubsection{Bidirectional Recurrent Neural Networks}
-In recurrent networks, the hidden state at any given time encapsulates all information about the sequence up to that point, serving as the left context for the current input. 
-In neural machine translation (NMT), having access to the entire input sequence simultaneously suggests the importance of utilizing the context to the right of the input. 
-Training an RNN in reverse on the input sequence is one approach to retain this information. 
-Combining the forward and backward networks results in a bidirectional RNN, where two independent RNNs process input from start to finish and finish to start, respectively \cite{schuster97}. 
-The outputs of both networks are then merged to create a unified representation incorporating both left and right input contexts at each time step. 
-Methods such as concatenation, element-wise operations, or averaging can be employed to combine the outputs of the forward and backward passes. 
-This ensures that the information on both sides of the current input is captured in the output at each time step.
-Figure~\ref{fig:rnn-brnn} illustrates the conventional RNN structure alongside the BRNN architecture over a timeline.
-In the RNN, only a forward hidden layer is present, whereas in the BRNN, both forward and backward hidden layers are depicted.
-
-\begin{figure}[h]
-	\centering
-	\includegraphics[width=0.9\linewidth]{Figures/RNN-BRNN}
-	\caption{Architecture of RNN and BRNN shown over a period of time.}
-	\label{fig:rnn-brnn}
-\end{figure}
-
-\subsection{Long Short-Term Memory}
-One of the main limitations of RNNs in machine translation is their difficulty in handling long-distance dependencies.
-Distant words play a crucial role in translation tasks, as demonstrated in the following example:
-\begin{quotation}
-	The lengthy \textit{sentence} composed of numerous words spanning multiple lines still \textit{poses} significant challenges for MT.
-\end{quotation}
-In this example, the verb \textit{poses} depends on the subject \textit{sentence}, which is separated by a long subordinate clause.
-
-While RNNs can access preceding sequences, the knowledge retained in hidden states tends to be relatively localized. 
-Consequently, more complex network structures have been developed to address the difficulty of maintaining context over extended periods. 
-It is essential for the network to discard irrelevant information while retaining data crucial for future decision-making. 
-Long Short-Term Memory (LSTM) and Gated Recurrent Units (GRUs) are the predominant approaches employed to achieve this objective.
-
-LSTM networks, introduced by \cite{hochreiter97}, address the challenge of managing context by dividing it into two sub-problems: removing unnecessary information and incorporating data crucial for future decisions.
-Unlike incorporating a fixed strategy into the architecture, the key lies in learning to handle context dynamically. 
-LSTMs accomplish this by expanding the architecture with an explicit context layer and employing specialized neural units with gates to control the flow of information within the network layers. 
-These gates utilize additional weights to sequentially process input, previous hidden layers, and previous context layers.
-
-\subsection{Gated Recurrent Units}
-GRU, akin to LSTM but with fewer parameters, offers the advantage of reduced training costs by requiring fewer parameters. 
-This reduction in parameters is achieved by eliminating the need for a separate context vector and reducing the number of gates, thereby alleviating the computational burden compared to LSTM. 
-Both GRU and LSTM employ sigmoid functions in their gates to either allow information with values close to one to pass through or block information with values close to zero.
-
-In contrast to simple FFNs, LSTMs and GRUs feature more complex neural nodes, as depicted in Figure~\ref{fig:lstm-vs-gru}, with inputs and outputs connected to each node type. 
-The additional complexity within LSTM and GRU nodes is primarily contained within the nodes themselves. 
-The only external complexity introduced by LSTM, beyond the basic recurrent node, is the availability of the other context vector as an input and output. 
-Conversely, GRU nodes exhibit similar input and output architecture to the simple recurrent node.
-\begin{figure}[h]
-	\centering
-	\includegraphics[width=0.7\linewidth]{Figures/LSTM-vs-GRU}
-	\caption{Diagram illustrating the network structure of LSTM in the first panel (a) and GRU in the second panel (b)}
-	\label{fig:lstm-vs-gru}
-\end{figure}
-
-\section{Neural Machine Translation}
-The standard architecture for Neural Machine Translation (NMT) is the encoder-decoder network. 
-This architecture operates on the fundamental principle of employing an encoder network to convert a sequence of words from the source language sentence into a contextual representation. 
-Subsequently, a decoder utilizes this representation to generate an output sequence, essentially providing a plausible translation into the target language. 
-Figure~\ref{fig:encoder-decoder} illustrates the encoder-decoder architecture at its most abstract level, comprising three key components: an encoder, context, and decoder. 
-The encoder takes in a source language sentence as an input sequence of words, $x_1, x_2, ..., x_m$, and produces a corresponding sequence of contextualized representations known as a context vector. 
-This context vector captures the essence of the input and serves as input to the decoder. 
-Finally, the decoder utilizes the context vector to generate the most probable translation as a sequence of words, $y_1, y_2, ..., y_n$. 
-Encoders and decoders can be implemented using either RNNs or Transformers.
-\begin{figure}[h]
-	\centering
-	\includegraphics[width=0.7\linewidth]{Figures/Encoder-Decoder}
-	\caption{Basic Encoder-Decoder architecture}
-	\label{fig:encoder-decoder}
-\end{figure}
-
-Figure 2.7 illustrates a simplified version of the RNN-based encoder-decoder architecture. 
-The encoder processes the input sequence $x$ with the objective of generating a representation of the input. 
-This representation is captured in the final hidden state of the encoder, represented as $h^e_n$. 
-Subsequently, this context representation, denoted as $c$, is transferred to the decoder. 
-Upon receiving this state, the decoder utilizes it to initialize its initial hidden state. 
-The first decoder cell employs $c$ as its initial hidden state, $h^d_0$, and proceeds to generate a sequence of outputs iteratively, one element at a time, until an end-of-word marker, </s>, is produced.
-Consequently, each hidden state is dependent on the preceding hidden state and the output generated in the preceding state. 
-The embedding layer is composed of word embeddings, which adhere to the notion that semantically related words in similar contexts should possess similar representations. 
-Ultimately, the output $y$ at each time step involves a $softmax$ computation over the vocabulary, $V$.
-The most probable output at each time step can be determined by computing the $argmax$ over the $softmax$ output as per Equation~\ref{eqn:yt}.
-Although Figure~\ref{fig:rnn-encoder-decoder} illustrates a single network layer, stacked and bi-directional networks are typically employed for both the encoder and decoder in practical implementations.
-\begin{equation}
-	\hat{y}_t = argmax_{w \in V}P(w|x, y_1, . . . , y_{t-1})
-	\label{eqn:yt}
-\end{equation}
-\begin{figure}[]
-	\centering
-	\includegraphics[width=1.0\linewidth]{Figures/RNN-Encoder-Decoder}
-	\caption{The basic RNN-based encoder-decoder architecture.}
-	\label{fig:rnn-encoder-decoder}
-\end{figure}
-
-A significant drawback of this architecture lies in its inability to evenly represent information from the beginning of a sentence, particularly evident in lengthy sentences. 
-To address this issue, the attention mechanism emerges as a solution. 
-It enables the decoder to access information from all the hidden states of the encoder, rather than solely relying on the last hidden state. 
-The concept behind the attention mechanism is to create a singular context vector, denoted as $c$, by computing a weighted sum of all the encoder's hidden states. 
-These weights concentrate on a specific segment of the source text pertinent to the token generated by the decoder. 
-Notably, the context vector produced by the attention mechanism is dynamic, varying for each decoded token. 
-By introducing the attention mechanism, the static context vector is replaced with $c_i$ a dynamically derived counterpart from the encoder's hidden states at each decoding step $i$, ensuring comprehensive consideration of all encoder states.
-
-\subsection{Training Neural Machine Translation Models}
-Neural Machine Translation (NMT) operates within a supervised machine learning framework, where the correct output, denoted as $y$, is known for each input observation, $x$. 
-The system's output, represented as $\hat{y}$, serves as an estimate of the actual $y$. 
-During the training phase, the objective is to adjust the parameters within each layer so that $\hat{y}$ closely resembles $y$ for every training instance.
-To accomplish this, a loss function is employed to quantify the disparity between the system's output and the actual output. 
-The optimization process, typically facilitated by the gradient descent algorithm, seeks to minimize this loss function by adjusting the parameters. 
-Gradient descent utilizes the gradient of the loss function, computed as the partial derivative of the loss function with respect to each parameter, to iteratively update the model's parameters. 
-However, in the realm of neural networks, characterized by numerous layers and millions of parameters, computing the partial derivative of a weight in one layer concerning the loss associated with another layer necessitates employing techniques such as error back-propagation or reverse differentiation.
-These methodologies are crucial for efficiently navigating the complex landscape of neural network optimization.
-
-Moreover, NMT models undergo end-to-end training, where each training instance comprises a pair of sentences, one in the source language and the other in the target language.
-These sentence pairs are concatenated with a designated separator token, <s>, to form the training data.
-Typically, this training data is sourced from established datasets containing aligned pairs of sentences, known as parallel corpora.
-Optimization within the NMT framework is characterized by a non-convex problem landscape.
-Nevertheless, there exist several best practices for effectively training NMT models.
-For instance, it is advisable to initialize the model weights with small random numbers and to employ random seeds for reproducibility.
-Additionally, normalizing the input values to have zero mean and unit variance can be beneficial in enhancing training stability and convergence.
-
-NMT model training involves multiple epochs, representing complete iterations over the training data.
-Typically, as training progresses, the error on the training set steadily decreases.
-However, a key challenge arises when the model starts to overfit, meaning it memorizes the training data excessively without generalizing well to unseen examples.
-To diagnose overfitting, a separate set of examples known as the development (or validation) set is used, which is not involved in the training process.
-Monitoring the error on this development set over the course of training reveals a point where the error begins to increase, indicating overfitting.
-In theory, training should halt when the minimum error on the development set is reached.
-However, in practice, this is more nuanced for NMT models due to the non-deterministic nature of training and the lack of clear convergence or overfitting points, particularly with sizable datasets.
-While some studies suggest stopping criteria based on approximate training durations or a fixed number of epochs, most NMT research does not specify precise stopping conditions.
-To mitigate overfitting, various regularization techniques are employed, such as dropout, which randomly disables some nodes and connections during training.
-
-In addition to regularization, hyperparameter tuning plays a crucial role in NMT architecture design.
-These hyperparameters, which include the learning rate, mini-batch size, number of layers, hidden nodes per layer, and activation functions, are chosen by the architect to optimize model performance.
-\subsection{Decoding}
-The decoding (inference) algorithm employed to generate translations (as illustrated in the Figure~\ref{fig:rnn-encoder-decoder}) faces a challenge. 
-Opting for the single most probable word at each step implies a 1-best greedy search, where a greedy algorithm makes locally optimal decisions. 
-However, there are instances where following a sequence of words leads to the realization that an earlier mistake was made. 
-In such cases, the best sequence may initially comprise less probable words that are refined by subsequent words in the context of the entire output. 
-For instance, when translating an idiomatic expression, the initial words chosen may seem unusual (e.g., "\textit{take your hat off}" for "\textit{respect, admire, or congratulate someone}").
-
-In NMT decoding, the predominant method employed is known as beam search. 
-Unlike selecting the best word at each step, beam search maintains $k$ possible words, where $k$ denotes the beam size or beam width. 
-Initially, a $softmax$ is computed over the entire vocabulary to assign probabilities to each word. 
-From this $softmax$ output, the k-best candidates are selected, forming hypotheses. 
-Each hypothesis represents an initial output sequence along with its probability. 
-These top $k$ hypotheses are progressively expanded by passing them through different decoders in subsequent steps. 
-At each decoder, a $softmax$ is generated over the entire vocabulary to determine the next word for each hypothesis. 
-Subsequently, each of the $k * V$ potential sequences is scored based on the product of the current word's probability and the path leading to it. 
-These hypotheses are then pruned down to the $k$ best ones to ensure there are never more than $k$ hypotheses at any given time. 
-This iterative process continues until a complete translation candidate marked by "</s>" is generated, signifying the end of a hypothesis. 
-The finished hypothesis is then removed, and the beam size is reduced by one. 
-This search persists until the beam size diminishes to zero, resulting in $k$ hypotheses. 
-The best translation is determined from the complete hypothesis with the highest score. 
-Figure~\ref{fig:beam} illustrates this process using a beam size of six.
-When evaluating the top paths, we assess each based on the product of their word prediction probabilities. 
-In practical applications, we tend to achieve improved outcomes by normalizing the score according to the length of the translation produced, which involves dividing the score by the number of words. 
-This normalization process occurs after the search is finalized.
 \begin{figure}
 	\centering
-	\includegraphics[width=0.7\linewidth]{Figures/beam}
-	\caption{Beam search decoding with a beam size of six.}
-	\label{fig:beam}
+	\includegraphics[width=0.9\linewidth]{Figures/rag_architectur.png}
+	\caption{Rag architecture}
+	\label{rag_architectur.png}
 \end{figure}
+\section{Training and Fine-Tuning RAG Models}
+Training and fine-tuning are essential steps in enhancing the performance of RAG models by optimizing both the retriever and generator components leading to more coherent, relevant, and accurate responses.
+\subsection{Training the Retriever}
+A powerful technique for improving retrieval relevance involves training the retriever model using \textbf{contrastive learning}.
+This approach aims to maximize the similarity between query and relevant document embeddings while minimizing the similarity between query and irrelevant document embeddings.
+By training the model on numerous positive (relevant) and negative (irrelevant) pairs, the retriever learns to effectively distinguish between relevant and irrelevant information\cite{sbert2024}.
+\section{Task and Evaluation}
+The rapid development and increasing use of Retrieval-Augmented Generation (RAG) models in NLP have made their evaluation a critical area of research within the LLM community.
+This section covers the primary downstream tasks associated with RAG, the datasets used, and the methods for evaluating RAG systems \cite{zhou2020trustworthiness}.
+\subsection{Factuality}
+\begin{itemize}
+	\item Fact-Checking Benchmarks: Using datasets like FEVER and SQuAD to evaluate the model's ability to identify and correct factual errors.
+	\item Adversarial Testing: Creating adversarial examples to test the model's robustness against misleading information.
+	\item Contextual Understanding: Assessing the model's ability to understand the context of a query and provide accurate answers.
+\end{itemize}
+\subsection{ Robustness}
+\begin{itemize}
+	\item Noise Injection: Introducing noise into the retrieved documents to test the model's ability to handle imperfect information.
+	\item Adversarial Attacks: Evaluating the model's resilience to attacks that aim to manipulate its outputs.
+	\item Domain Adaptation: Testing the model's ability to adapt to new domains and data distributions.
+\end{itemize}
+\subsection{Fairness}
+\begin{itemize}
 
-\section{Related Work}
-In the recent years, Research in AMT has predominantly centered on Neural MT, targeting translation from Arabic to English and various other languages. 
-Neural machine translation, in particular, has become a compelling substitute for phrase-based Statistical Machine Translation (SMT), especially for the Arabic language.
-Numerous studies are currently focusing on the translation between Dialectal Arabic and Modern Standard Arabic (MSA) in low-resource settings, both from Dialectal Arabic to MSA and vice versa.
-Table~\ref{tab_ANMT} summarizes the surveyed Neural-based AMT research studies.
-\begin{small}
-	\begin{longtable}{|l|l|l|l|l|}
-		\caption{Neural-based AMT researches}
-		\label{tab_ANMT}\\
-		\hline
-		Year & Research & SL-TL$^{\mathrm{a}}$ & Method & Score\\
-		\hline
-		\endhead
-		%	\hline
-		%	\endfoot
+	\item Bias Detection: Identifying and quantifying biases in the training data and model outputs.
+	\item Fairness Metrics: Using metrics like demographic parity and equalized odds to evaluate the model's fairness.
+	\item Mitigation Techniques: Implementing techniques like debiasing and fairness constraints to mitigate biases.
+\end{itemize}
+\subsection{Objective Metrics:}
+\begin{itemize}
+	\item Accuracy: Precision, recall, and F1-score for evaluating the correctness of the generated responses.
+	\item Consistency: Measuring the consistency of the model's outputs across different queries and contexts.
+	\item Coherence: Assessing the coherence and fluency of the generated text.
+\end{itemize}
+\subsection{Subjective Metrics}
+\begin{itemize}
+	\item Human Evaluation: Using human raters to evaluate the quality of the generated responses.
+	\item User Studies: Conducting user studies to gather feedback on the user experience
+\end{itemize}
+\section{Limitations}
+While RAG has gained significant traction across diverse applications, it still faces certain limitations in terms of effectiveness and efficiency\cite{zhao2024retrieval}.
+\subsection{Noisy Retrieval Results}
+\begin{itemize}
+	\item Retrieval Quality: The quality of retrieved information can be affected by factors like indexing techniques, query formulation, and the underlying dataset.
+	\item Hallucinations: Noisy or irrelevant information can lead to the generation of hallucinated or factually incorrect responses.
+	\item Contextual Understanding: The LLM may struggle to understand the context of the retrieved information, especially if it's poorly formatted or contains inconsistencies.
+\end{itemize}
+\subsection{Extra Overhead}
+\begin{itemize}
+	\item Computational Cost: Retrieval and processing additional information can increase computational costs, especially for large-scale models and complex queries.
+	\item Latency: Retrieval and processing can introduce latency, impacting the real-time performance of RAG systems.
+	\item Complexity: Implementing and deploying RAG systems requires careful consideration of various factors, including data preparation, model selection, and system architecture.
+\end{itemize}
+\subsection{Interaction of Retrieval and Generation}
+\begin{itemize}
+	\item Aligning the goals of the retriever and generator is challenging.
+	\item Optimizing the interaction between the two components requires careful design and tuning.
+	\item The impact of various factors, such as metric selection and hyperparameter tuning, on RAG performance is still not fully understood.
+	\item
 
-%		\multicolumn{5}{c}{\textbf{Corpus-based AMT researches}}\\
-		%	\hline
-		\multicolumn{5}{l}{\textbf{Neural-based AMT}}\\
-		\hline
-		2016 & Almahairi et al. \cite{almahairi16}		& Ar$\leftrightarrow$En &MT Evaluation& 49.7/33.62\\
-		2016 & Guzm{\'a}n et al. \cite{guzman16}		& En$\rightarrow$Ar &Morpho-syntactic analysis& +75\%\\
-		2017 & Belinkov et al. \cite{belinkov17}		& Ar$\leftrightarrow$En &Morpho/Vocab/Factored NMT&Bl 28.42\\
-		2017 & Durrani et al. \cite{durrani17}			& Ar$\leftrightarrow$En &MT Evaluation& Bl +4\\	
-		2017 & Choi et al. \cite{choi17}				& Kor$\rightarrow$Ar &Corpus extension&Bl 27.07\\
-		2018 & Almansor \& Al-Ani \cite{almansor18}		& Ar$\rightarrow$En &Low-resource& Bl +10\\
-		2018 & Shapiro \& Duh \cite{shapiro18}			& Ar$\rightarrow$En &Morpho/Vocab/Factored NMT&Bl 29.10\\
-		2018 & Alrajeh \cite{alrajeh18}					& Ar$\rightarrow$En &MT Evaluation& Bl +13\\
-		2018 & Alkhatib \& Shaalan \cite{alkhatib18}	& Ar$\leftrightarrow$En &Paraphrasing model& 86.9\%/94.1\%\\
-		2018 & Baniata et al. \cite{baniata18}			& DA$\rightarrow$MSA &Multitask Learning& 0.41/0.30\\
-		2019 & Aqlan et al. \cite{aqlan19}				& Ar$\rightarrow$Cn &Pre-/Post-processing& Bl 24.66\\
-		2019 & Oudah et al. \cite{oudah19}				& Ar$\rightarrow$En &Pre-/Post-processing& 55.64/53.54\\
-		2019 & Hadj Ameur et al. \cite{hadjameur19}		& Ar$\leftrightarrow$En &Pre-/Post-processing&29.76/20.41\\	
-		2019 & Gashaw \& Shashirekha \cite{gashaw19}	& Amh$\rightarrow$Ar &MT Evaluation& Bl 12\\	
-		2020 & Ji et al. \cite{ji20}					& $\rightarrow$Ar &Multilingual/Low-resource& 25.49\\
-		2021 & Bensalah et al. \cite{bensalah21}		&  Ar$\rightarrow$En &Pre-processing/MT Evaluation& 41.87\%\\
-		2021 & Berrichi \& Mazroui \cite{berrichi21}	& Ar$\leftrightarrow$En & Morpho/Vocab/Factored NMT&Bl 33.02\\	
-		2021 & Moukafih et al. \cite{moukafih21}		& DA$\leftrightarrow$Ar &Multi-Task learning/MT Eval&Bl 35.06\\	
-		2021 & Nagoudi et al. \cite{nagoudi21}			& DA$\rightarrow$En &Pre-processing/MT Evaluation& Bl 25.72\\	
-		2021 & Stergiadis et al. \cite{stergiadis21}	& Ar$\rightarrow$En &Multidimensional Tagging&Bl 50.84\\	
-		2022 & Bensalah et al. \cite{bensalah22}		& Ar$\rightarrow$En &MT Evaluation&Bl 0.575\\	
-		2022 & Slim et al. \cite{slim22}				& DA$\rightarrow$MSA &Transductive transfer learning&Bl 35.87\\
-		2022 & Gaser et al. \cite{gaser22}				& DA$\rightarrow$En &Segmentation& chrF2 51.3\\
-		2022 & Nagoudi et al. \cite{nagoudi22}			& 20 lang$\rightarrow$Ar &Multi-lingual model& Bl 32.07\\
-		2022 & Hameed et al. \cite{hameed22}			& Ru$\rightarrow$Ar & MT Evaluation& 51.57\% \\
-		2022 & Baniata et al. \cite{baniata22}			& DA$\rightarrow$Ar & RPE$^{\mathrm{b}}$+sub-word & Bl 66.87\\
-\hline
-\multicolumn{3}{l}{$^{\mathrm{a}}$Source language-Target language}&
-\multicolumn{2}{l}{$^{\mathrm{b}}$Reverse Positional Encoding}\\
-\end{longtable}
-\end{small}
+\end{itemize}
+\subsection{Long Context Generation:}
 
-\section{Summary}
-In this chapter, we embarked on a comprehensive exploration of neural networks, particularly focusing on feedforward networks (FFN), recurrent neural networks (RNN), long short-term memory networks (LSTM), and gated recurrent units (GRU).
-These architectures lay the foundation for understanding the mechanisms behind neural machine translation (NMT).
-
-We delved into the intricacies of training NMT modelsand the decoding phase, where the trained models generate translations from input sequences. 
-Finally, we delved into related works in the field, dedicating special attention to advancements, methodologies, and challenges specific to neural machine translation in the context of the Arabic language.
+\begin{itemize}
+	\item Context Length Limits: LLMs have limitations on the amount of context they can process at once.
+	\item Information Loss: Long documents may be truncated or summarized, leading to loss of important information.
+	\item Computational Cost: Processing long contexts can be computationally expensive.
+\end{itemize}
+\newpage
+\section{Conclusion}
+This chapter provided an overview of Retrieval-Augmented Generation (RAG), highlighting its significance in natural language processing.
+We discussed the key concepts of retrieval and generation, emphasizing the advantages of their combination in enhancing language model capabilities.
+The historical development of RAG was explored, along with its core components, including the retrieval mechanism, generation process, and augmentation techniques.
+Additionally, we examined various downstream tasks and evaluation targets, illustrating RAG's versatility in applications like open-domain question answering and fact verification.
+Overall, RAG represents a promising advancement in knowledge-intensive tasks, paving the way for further research and innovation in the field.

### 2025-01-22 Thesis/Latex/Cover_page.tex
Updated thesis template with content for chapters

@@
 	{\large Dissertation } \\[0.2cm]
-	{\ Submitted in partial fulfilment of the requirements for Doctorate degree in Computer Science} \\[0.2cm]
+	{\ Submitted in partial fulfilment of the requirements for Master degree in Computer Science} \\[0.2cm]
 	{\large Option: Artificial Intelligence}\\[0.5cm]
@@
 %    \\[0.7cm]
-	{\LARGE {\textbf{Neural Machine Translation for\\ Arabic Media Content}}}
+	{\LARGE {\textbf{Enhancing Legal Information Access and \\ Reasoning  with  Retrieval-Augmented LLMs for Juridical Data}}}
 	\\[0.7cm]
@@
 
-	{\large Presented by \textbf{Baligh BABAALI} }\\[0.5cm]
+	{\large Presented by \textbf{Boudjenane Zoubida Asmaa} }\\[0.5cm]
 
@@
 
-	\it{Month  2024}
+	\it{Month  2025}
 \end{center}

### 2025-01-24 Paper/latex/paper_RAG.tex
Added paper introduction

@@
 \usepackage{listings}%
+
 %%%%
@@
 
-\title[Article Title]{ttttttttttttttttttttt}
+\title[Article Title]{Beyond Static Retrieval: Adaptive Method for Improving RAG Performance}
 
@@
 
-\author*[1]{\fnm{Asmaa} \sur{Boudjenane }}\email{}
+\author*[1]{\fnm{Zoubida Asmaa} \sur{Boudjenane }}\email{asmaa.boudj04@gmail.com}
 
@@
 \section{Introduction}\label{sec1}
+Large Language Models (LLMs) have emerged as state-of-the-art artificial intelligence systems that excel at understanding and generating human-like text,Built on deep learning architectures,particularly transformers \cite{vaswani2017attention}.these models are trained on massive datasets to capture linguistic patterns and semantic relationships showcasing impressive adaptability across diverse tasks such as text generation, summarization, question answering.Notable examples of LLMs include OpenAI GPT series\cite{brown2020language} , Google BERT\cite{devlin2018bert} and T5 \cite{raffel2020exploring} which have set new benchmarks in NLP tasks .\\
+despite their impressive capabilities,LLMs suffer from hallucination where they confidently generate factually incorrect or entirely fabricated information.
+This poses challenges in applications requiring high accuracy, Additionally significant limitation of LLMs is their static nature, as they are trained on fixed datasets and cannot dynamically update their knowledge which makes them prone to generating outdated or inconsistent responses, especially for time-sensitive or evolving data\cite{Mousavi2025}.
+\\
+\\
+researchers have explored various approaches To address this limitation, including transfer learning\cite{zhuang2020comprehensive}a technique that improves pre-trained models performance in new tasks by transferring knowledge from related tasks.and fine-tuning \cite{Howard2018} which adapt pre-trained models to specific tasks or domains.
+While these methods improve performance on targeted tasks,they are computationally expensive and impractical for real-time knowledge updates.
+A promising alternative is Retrieval-Augmented Generation (RAG) \cite{lewis2020retrieval}.which integrates real-time retrieval from external knowledge sources with generative models.
+RAG enables LLMs to access up-to-date information without retraining, making it particularly useful for applications like open-domain question answering and fact-checking.\\
+\\
+Retrieval-Augmented Generation (RAG) faces several limitations, The performance of RAG heavily depends on the quality of the retriever \cite{karpukhin2020dpr}, as irrelevant or noisy documents can degrade the generator output.
+Additionally, retrieving and processing a large number of documents increases computational overhead , making RAG less efficient for real-time applications.
+A critical challenge is determining the optimal number of retrieved passages (K) to feed into the generator\cite{lewis2020retrieval}.
+Selecting too few passages (small K) risks missing crucial context, while too many (large K) can introduce noise, reduce coherence and increase computational overhead.
+This trade-off makes K a key factor in balancing performance and efficiency.
+\\
+\\
+Current approaches, such as static K selection \cite{lewis2020retrieval} involves retrieving a fixed number of top results regardless of the query characteristics ,it struggles in dynamic datasets as it cannot adapt to the complexity of individual queries , dynamic K adjustment attempt to address this trade-off but often introduce complexity or latency ,Hybrid approaches like those proposed yuan et al .
+\cite{yuan2024hybrid},that integrates static retrieval with dynamic retrieval to improve accuracy and robustness.
+Furthermore, the "Blended RAG" approach \cite{blended_rag2024} enhances retrieval accuracy by combining semantic search techniques with hybrid query strategies, resulting in improved performance in information retrieval tasks.
+However, the approach faces significant limitations, including high computational overhead due to the need for multiple retrieval methods and dependency on high-quality metadata for optimal performance.
+These limitations highlight the need for a more robust approach to K selection, capable of balancing efficiency and accuracy across varying query types.\\
+\\
+In this paper, we investigate the problem of selecting K in RAG and propose a novel approach to dynamically adjust K based on relevance thresholds and a Mixture of Logits (MoL)\cite{ding2024efficient} scoring mechanism, ensuring efficient and context-sensitive candidate selection while reducing computational costs.
+By using adaptive thresholds the algorithm improves the relevance of retrieved documents, refining the set to include only the most contextually appropriate matches.
+Unlike traditional approaches, it adapts to the query and dataset characteristics, making it more flexible for diverse contexts.
+Furthermore, its design for handling large datasets ensures scalable and precise retrieval, which is crucial for RAG systems working with extensive corpora.
 
@@
 % KAN Archecture
-\begin{equation}
-f(x_1, x_2, \cdots, x_n)=\sum_{q=1}^{2n+1}\Phi_q \left( \sum_{p=1}^{n}\phi_{q, p}(x_p)\right)
-\end{equation}
+%\begin{equation}
+%f(x_1, x_2, \cdots, x_n)=\sum_{q=1}^{2n+1}\Phi_q %\left( \sum_{p=1}^{n}\phi_{q, p}(x_p)\right)
+%\end{equation}
 

### 2025-01-28 Paper/latex/paper_RAG.tex
add Related works

@@
 \section{Introduction}\label{sec1}
+
 Large Language Models (LLMs) have emerged as state-of-the-art artificial intelligence systems that excel at understanding and generating human-like text,Built on deep learning architectures,particularly transformers \cite{vaswani2017attention}.these models are trained on massive datasets to capture linguistic patterns and semantic relationships showcasing impressive adaptability across diverse tasks such as text generation, summarization, question answering.Notable examples of LLMs include OpenAI GPT series\cite{brown2020language} , Google BERT\cite{devlin2018bert} and T5 \cite{raffel2020exploring} which have set new benchmarks in NLP tasks .\\
-despite their impressive capabilities,LLMs suffer from hallucination where they confidently generate factually incorrect or entirely fabricated information.
+Despite their impressive capabilities,LLMs suffer from hallucination where they confidently generate factually incorrect or entirely fabricated information.
 This poses challenges in applications requiring high accuracy, Additionally significant limitation of LLMs is their static nature, as they are trained on fixed datasets and cannot dynamically update their knowledge which makes them prone to generating outdated or inconsistent responses, especially for time-sensitive or evolving data\cite{Mousavi2025}.
-\\
-\\
-researchers have explored various approaches To address this limitation, including transfer learning\cite{zhuang2020comprehensive}a technique that improves pre-trained models performance in new tasks by transferring knowledge from related tasks.and fine-tuning \cite{Howard2018} which adapt pre-trained models to specific tasks or domains.
+
+Researchers have explored various approaches To address this limitation, including transfer learning\cite{zhuang2020comprehensive}a technique that improves pre-trained models performance in new tasks by transferring knowledge from related tasks.and fine-tuning \cite{Howard2018} which adapt pre-trained models to specific tasks or domains.
 While these methods improve performance on targeted tasks,they are computationally expensive and impractical for real-time knowledge updates.
 A promising alternative is Retrieval-Augmented Generation (RAG) \cite{lewis2020retrieval}.which integrates real-time retrieval from external knowledge sources with generative models.
-RAG enables LLMs to access up-to-date information without retraining, making it particularly useful for applications like open-domain question answering and fact-checking.\\
-\\
+
+RAG enables LLMs to access up-to-date information without retraining, making it particularly useful for applications like open-domain question answering and fact-checking.
 Retrieval-Augmented Generation (RAG) faces several limitations, The performance of RAG heavily depends on the quality of the retriever \cite{karpukhin2020dpr}, as irrelevant or noisy documents can degrade the generator output.
@@
 This trade-off makes K a key factor in balancing performance and efficiency.
-\\
-\\
+
 Current approaches, such as static K selection \cite{lewis2020retrieval} involves retrieving a fixed number of top results regardless of the query characteristics ,it struggles in dynamic datasets as it cannot adapt to the complexity of individual queries , dynamic K adjustment attempt to address this trade-off but often introduce complexity or latency ,Hybrid approaches like those proposed yuan et al .
@@
 However, the approach faces significant limitations, including high computational overhead due to the need for multiple retrieval methods and dependency on high-quality metadata for optimal performance.
-These limitations highlight the need for a more robust approach to K selection, capable of balancing efficiency and accuracy across varying query types.\\
-\\
+These limitations highlight the need for a more robust approach to K selection, capable of balancing efficiency and accuracy across varying query types.
+
 In this paper, we investigate the problem of selecting K in RAG and propose a novel approach to dynamically adjust K based on relevance thresholds and a Mixture of Logits (MoL)\cite{ding2024efficient} scoring mechanism, ensuring efficient and context-sensitive candidate selection while reducing computational costs.
@@
 
+Our contributions in this paper are the following :
+\begin{itemize}
+	\item we review existing approaches to Retrieval-Augmented Generation (RAG), with a specific focus on static , dynamic and Hybrid selection strategies.
+We critically analyze the strengths and limitations of these methods, identifying key gaps in their design and performance.
+These gaps motivate the need for an improved approach that addresses the challenges of both static and dynamic relevance in retrieval-augmented systems.
+	\item Overview of Retrieval-Augmented Generation (RAG) in This section We provide a detailed explanation of the RAG framework, including its components and how it operates.
+Key concepts, such as retrievers, generators, and augmented are discussed to establish the foundational context.
+	\item Proposed Solution: This section introduces our novel hybrid dynamic  selection algorithm.
+We describe its methodology, explain its design, and discuss how it overcomes the limitations of existing static and dynamic approaches.
+The benefits of this approach, particularly in improving retrieval accuracy and computational efficiency, are emphasized.
+	\item Results : We present experimental results to demonstrate the effectiveness of our hybrid dynamic  selection algorithm.
+Comparisons with baseline method highlight its superiority in terms of retrieval performance and adaptability to diverse query complexities.
+\end{itemize}
+
 \section{Related works}\label{sec2}
+The selection of an optimal number of retrieved items (K) in Retrieval-Augmented Generation (RAG) systems plays a crucial role in ensuring accurate and efficient responses.Several methods, including static, dynamic, and hybrid approaches have been proposed to address this challenge While these methods show promise in improving performance,they often face trade-offs between adaptability, efficiency when applied to complex or diverse query patterns.
+In this section, we will explore these methods, highlight their respective strengths and weaknesses, and introduce our hybrid solution as a more robust and adaptable alternative.
+\\
+\textbf{Static Top-K selection in Retrieval}
 
-\section{section1 }\label{sec3}
+In traditional Retrieval-Augmented Generation (RAG) systems, static K-selection is a widely used method where a fixed number of top-k documents are retrieved based on their relevance scores.
+Sparse retrieval methods where the "keys" represent documents to be retrieved, and the "values" are the same documents in this context such as query likelihood models\cite{Lafferty2001} ,TF-IDF\cite{Robertson1997} and BM25\cite{Zaragoza2009} often rely on this static approach to identify and rank documents.
+For instance, if k  =5, the system retrieves the top 5 documents deemed most relevant to a query, regardless of the query's complexity or the variability in relevance distribution among documents.
+A major drawback of this approach is its rigidity.
+Since the number of retrieved documents (k) is fixed, it fails to adapt to the varying complexity of queries.
+For simple queries, retrieving too many documents may introduce noise, while for complex queries, retrieving too few may result in incomplete or insufficient information.
+This one-size-fits-all strategy can lead to suboptimal performance across diverse datasets or tasks.\\
+\textbf{Dynamic top-k Selection in Retrieval}
 
+Dynamic k-selection addresses the limitations of static k by varying the number of retrieved documents based on the specific query or context.
+These approaches typically involve heuristics, machine learning models, or adaptive algorithms to determine the optimal K for each query such as Dynamic Query Cutoff (DQC)\cite{Culpepper2016} This approach uses machine learning models trained on query-document features to predict the optimal k for retrieval , A reinforcement learning-based approach\cite{Zhou2021} where the agent learns to adjust k by maximizing downstream task performance (e.g accurate response generation in RAG) Another example is the Dynamic Selection of k Nearest Neighbors \cite{DynamicKNN2012} where the selection process often depends on the distribution of neighbors, their distances to the query point, or their relevance scores.
+However, these methods come with limitations.They often introduce computational overhead due to the need for dynamic evaluation, which can increase latency in large-scale systems.
+Additionally, their performance heavily depends on the quality of the underlying models or distance metrics, making them sensitive to noisy data or poorly tuned hyperparameters.
+As a result, achieving generalization across diverse datasets or domains remains a significant challenge.
+\\
+\textbf{Hybrid Approaches}
+
+Hybrid approaches combine elements of static and dynamic K-selection to achieve a balance between simplicity and adaptability.These systems might use a static 
+k as a baseline but dynamically adjust such as STAYKATE (Static-Dynamic Hybrid Selection)\cite{Zhu2024} Combines static (active learning-inspired) and dynamic (KATE-based) example selection strategies to create STAYKATE.
+Another example is DR-RAG (Dynamic Relevance for Retrieval-Augmented Generation)\cite{Hei2024} retrieval framework designed to improve the relevance between queries and documents.
+In the first stage, a similarity matching method retrieves an initial set of documents.
+In the second stage, a deeper analysis is performed to identify dynamic-relevant documents, supported by a classifier that filters out irrelevant ones .
+however hybrid methods may face challenges in determining when to apply static or dynamic k-selection effectively, leading to potential trade-offs between retrieval accuracy and system efficiency.
+These limitations highlight the ongoing need for robust mechanisms that dynamically and efficiently adapt k to the unique characteristics of each query and dataset.
+
+Our method leverages adaptive thresholds to dynamically refine the set of retrieved documents, ensuring that only the most contextually relevant matches are included.
+Unlike traditional static or fixed
+K-selection methods, which apply a uniform retrieval strategy across all queries, our approach adapts to the specific characteristics of both the query and the dataset.
+This flexibility allows the algorithm to better accommodate diverse contexts, enhancing the relevance and quality of retrieval results.Through its adaptive nature, the method not only improves retrieval accuracy but also minimizes computational overhead by focusing resources on the most pertinent results.
 \subsection{section1}\label{subsec2}
@@
 
-Environments such as figure, table, equation and align can have a label
-declared via the \verb+\label{#label}+ command. For figures and table
-environments use the \verb+\label{}+ command inside or just
-below the \verb+\caption{}+ command. You can then use the
-\verb+\ref{#label}+ command to cross-reference them. As an example, consider
-the label declared for Figure~\ref{fig1} which is
-\verb+\label{fig1}+. To cross-reference it, use the command 
-\verb+Figure \ref{fig1}+, for which it comes up as
-``Figure~\ref{fig1}''.
+%Environments such as figure, table, equation and align can have a label
+%declared via the \verb+\label{#label}+ command. For figures and table
+%environments use the \verb+\label{}+ command inside or just
+%below the \verb+\caption{}+ command. You can then use the
+%\verb+\ref{#label}+ command to cross-reference them. As an example, consider
+%the label declared for Figure~\ref{fig1} which is
+%\verb+\label{fig1}+. To cross-reference it, use the command 
+%\verb+Figure \ref{fig1}+, for which it comes up as
+%``Figure~\ref{fig1}''.
 
@@
 
-Here is an example for \verb+\cite{...}+: \cite{bib1}.
-Another example for \verb+\citep{...}+: \citep{bib2}.
-For author-year citation mode, \verb+\cite{...}+ prints Jones et al.
+%Here is an example for \verb+\cite{...}+: \cite{bib1}.
+Another example for %\verb+\citep{...}+: \citep{bib2}.
+For author-year citation mode, \verb+\cite{...}+ %prints Jones et al.
 (1990) and \verb+\citep{...}+ prints (Jones et al., 1990).
 
-All cited bib entries are printed at the end of this article: \cite{bib3}, \cite{bib4}, \cite{bib5}, \cite{bib6}, \cite{bib7}, \cite{bib8}, \cite{bib9}, \cite{bib10}, \cite{bib11}, \cite{bib12} and \cite{bib13}.
+%All cited bib entries are printed at the end of this article: \cite{bib3}, %\cite{bib4}, \cite{bib5}, \cite{bib6}, \cite{bib7}, \cite{bib8}, \cite{bib9}, %\cite{bib10}, \cite{bib11}, \cite{bib12} and \cite{bib13}.
 

### 2025-01-30 Paper/latex/paper_RAG.tex
add overview of RAG

@@
 Large Language Models (LLMs) have emerged as state-of-the-art artificial intelligence systems that excel at understanding and generating human-like text,Built on deep learning architectures,particularly transformers \cite{vaswani2017attention}.these models are trained on massive datasets to capture linguistic patterns and semantic relationships showcasing impressive adaptability across diverse tasks such as text generation, summarization, question answering.Notable examples of LLMs include OpenAI GPT series\cite{brown2020language} , Google BERT\cite{devlin2018bert} and T5 \cite{raffel2020exploring} which have set new benchmarks in NLP tasks .\\
-Despite their impressive capabilities,LLMs suffer from hallucination where they confidently generate factually incorrect or entirely fabricated information.
-This poses challenges in applications requiring high accuracy, Additionally significant limitation of LLMs is their static nature, as they are trained on fixed datasets and cannot dynamically update their knowledge which makes them prone to generating outdated or inconsistent responses, especially for time-sensitive or evolving data\cite{Mousavi2025}.
+Despite their impressive capabilities,LLMs suffer from hallucination\cite{ji2023survey} where they confidently generate factually incorrect or entirely fabricated information.
+This poses challenges in applications requiring high accuracy like knowledge-intensive tasks\cite{kandpal2023large} , Additionally significant limitation of LLMs is their static nature, as they are trained on fixed datasets\cite{bender2021dangers} and cannot dynamically update their knowledge which makes them prone to generating outdated or inconsistent responses, especially for time-sensitive or evolving data\cite{Mousavi2025}.
 
@@
 Our contributions in this paper are the following :
-\begin{itemize}
-	\item we review existing approaches to Retrieval-Augmented Generation (RAG), with a specific focus on static , dynamic and Hybrid selection strategies.
+we review existing approaches to Retrieval-Augmented Generation (RAG), with a specific focus on static , dynamic and Hybrid selection strategies.
 We critically analyze the strengths and limitations of these methods, identifying key gaps in their design and performance.
 These gaps motivate the need for an improved approach that addresses the challenges of both static and dynamic relevance in retrieval-augmented systems.
-	\item Overview of Retrieval-Augmented Generation (RAG) in This section We provide a detailed explanation of the RAG framework, including its components and how it operates.
-Key concepts, such as retrievers, generators, and augmented are discussed to establish the foundational context.
-	\item Proposed Solution: This section introduces our novel hybrid dynamic  selection algorithm.
+\\
+Overview of Retrieval-Augmented Generation (RAG) in This section We provide a detailed explanation of the RAG framework, including its components and how it operates.
+Key concepts, such as retrievers, generators, and augmented are discussed to establish the foundational context.\\
+Proposed Solution: This section introduces our novel hybrid dynamic  selection algorithm.
 We describe its methodology, explain its design, and discuss how it overcomes the limitations of existing static and dynamic approaches.
-The benefits of this approach, particularly in improving retrieval accuracy and computational efficiency, are emphasized.
-	\item Results : We present experimental results to demonstrate the effectiveness of our hybrid dynamic  selection algorithm.
+The benefits of this approach, particularly in improving retrieval accuracy and computational efficiency, are emphasized.\\
+Results :in This section We present experimental results to demonstrate the effectiveness of our hybrid dynamic  selection algorithm.
 Comparisons with baseline method highlight its superiority in terms of retrieval performance and adaptability to diverse query complexities.
-\end{itemize}
 
@@
 k as a baseline but dynamically adjust such as STAYKATE (Static-Dynamic Hybrid Selection)\cite{Zhu2024} Combines static (active learning-inspired) and dynamic (KATE-based) example selection strategies to create STAYKATE.
-Another example is DR-RAG (Dynamic Relevance for Retrieval-Augmented Generation)\cite{Hei2024} retrieval framework designed to improve the relevance between queries and documents.
+Another example is DR-RAG (Dynamic Relevance for Retrieval-Augmented Generation)\cite{hei2024drragapplyingdynamicdocument} retrieval framework designed to improve the relevance between queries and documents.
 In the first stage, a similarity matching method retrieves an initial set of documents.
@@
 
-Our method leverages adaptive thresholds to dynamically refine the set of retrieved documents, ensuring that only the most contextually relevant matches are included.
-Unlike traditional static or fixed
-K-selection methods, which apply a uniform retrieval strategy across all queries, our approach adapts to the specific characteristics of both the query and the dataset.
+Our method leverages adaptive thresholds to dynamically refine the list of retrieved documents, to make sure that only the most contextually relevant matches are included.
+In contrast to traditional static or fixed K-selection techniques, which employ a consistent retrieval strategy for every query, our approach adapts to the specific characteristics of both the query and the dataset.
 This flexibility allows the algorithm to better accommodate diverse contexts, enhancing the relevance and quality of retrieval results.Through its adaptive nature, the method not only improves retrieval accuracy but also minimizes computational overhead by focusing resources on the most pertinent results.
-\subsection{section1}\label{subsec2}
 
-\section{section1}\label{sec3}
-\subsection{section1}\label{subsec2}
-\subsection{section1}\label{subsec2}
+\section{Overview of Retrieval-Augmented Generation }\label{sec3}
+Retrieval-Augmented Generation (RAG) is a method designed to enhance the performance of large language models (LLMs) by integrating external, reliable knowledge sources into their response generation process\cite{awsRAG}.
+RAG combines search capabilities with LLM prompting.
+The process involves feeding both the userss query and the relevant information retrieved by a search algorithm into the LLMs prompt\cite{ilin2023advancedrag}, enabling the model to generate answers based on the provided context.
 
-\section{section1n}\label{sec3}
+\begin{figure}[h]
+	\centering
+	\includegraphics[width=0.9\textwidth]{rag.PNG}
+	\caption{RAG system component }\label{fig1}
+\end{figure}
+As illustrated in Figure 1, the RAG system is built around two main components:the retriever and the generator.
+The retriever?s role is to find relevant information from a pre-built data store, while the generator utilizes this retrieved data to produce coherent and contextually appropriate content.
+The overall RAG process operates as follows:
+\subsection{Retrieval in RAG Systems}\label{subsec2}
+
+Retrieval is the process of identifying and gathering information system resources that match a specific information need.
+To break it down, imagine these resources as a key-value store, represented as pairs:
+
+\[
+\{(k_i, v_i)\}_{i=1}^{N}
+\]
+
+where each key \( k_i \) is linked to a corresponding value \( v_i \) (often, the key and value are the same).
+When a query \( q \) is provided, the goal is to find the top-\( k \) keys that are most similar to the query using a similarity function \( s \), and then retrieve their associated values.
+
+Depending on the similarity function used, retrieval methods can be grouped into categories such as sparse retrieval, dense retrieval, and others.
+For the widely used sparse and dense retrieval approaches, the process typically involves two main steps:  \\
+1. Encoding each object into a specific representation.\\
+2. Building an index to organize the data for efficient searching .
+
+This structured approach ensures that the most relevant information is retrieved quickly and accurately\cite{zhao2024retrievalaugmentedgenerationaigeneratedcontent}.
+
+\subsection{Generator  in RAG Systems}\label{subsec2}
+In Retrieval-Augmented Generation (RAG) systems, the generator is a key component responsible for crafting the final output.
+It works by combining the retrieved information with the users input query to produce a well-structured and contextually relevant response \cite{gupta2024comprehensivesurveyretrievalaugmentedgeneration}.
+
+Once the retriever fetches relevant data from external sources,the generator processes and integrates this information into a well-structured and meaningful answer.
+At the core of this process is a Large Language Model (LLM), which ensures that the generated text is not only fluent and accurate but also remains relevant to the original query.
+
+In various generative tasks, different models are selected based on the specific requirements: Models like BART\cite{lewis2019bart} and GPT\cite{brown2020language}are commonly used for tasks such as text summarization and translation ,Vision-Language Models (VLMs) \cite{radford2021learningtransferablevisualmodels} are employed to generate textual descriptions from images , for ext-to-Code Generation: Models like OpenAIs Codex \cite{chen2021evaluatinglargelanguagemodels} are designed to translate natural language prompts into executable code.
+
+\section{proposed Slution}\label{sec3}
 

### 2025-02-01 Paper/latex/paper_RAG.tex
added proposed slution

@@
 	\includegraphics[width=0.9\textwidth]{rag.PNG}
-	\caption{RAG system component }\label{fig1}
+	\caption{RAG system component }\label{rag.PNG}
 \end{figure}
-As illustrated in Figure 1, the RAG system is built around two main components:the retriever and the generator.
+As illustrated in figure \ref{rag.PNG}, the RAG system is built around two main components:the retriever and the generator.
 The retriever?s role is to find relevant information from a pre-built data store, while the generator utilizes this retrieved data to produce coherent and contextually appropriate content.
@@
 where each key \( k_i \) is linked to a corresponding value \( v_i \) (often, the key and value are the same).
-When a query \( q \) is provided, the goal is to find the top-\( k \) keys that are most similar to the query using a similarity function \( s \), and then retrieve their associated values.
+When a query \( q \) is provided, the goal is to find the top-\( k \) keys that are most similar to the query using a similarity function \( s \), and then retrieve their associated values\cite{gupta2024comprehensivesurveyretrievalaugmentedgeneration}.
 
@@
 In Retrieval-Augmented Generation (RAG) systems, the generator is a key component responsible for crafting the final output.
-It works by combining the retrieved information with the users input query to produce a well-structured and contextually relevant response \cite{gupta2024comprehensivesurveyretrievalaugmentedgeneration}.
+It works by combining the retrieved information with the users input query to produce a well-structured and contextually relevant response .
 
@@
 \section{proposed Slution}\label{sec3}
+In traditional candidate selection systems, the reliance on static or fixed top-K retrieval lead to suboptimal retrieval performance, especially in scenarios where the relevance distribution of candidates is highly variable.
+To address these limitations, we propose Dynamic Candidate Selection, a novel algorithm that adaptively refines the candidate selection process by leveraging component-level embeddings and a Mixture of Logits (MoL) scoring mechanism.
 
+The \textbf{Mixture of Logits (MoL)}\cite{zhai2023revisiting} methode provides a flexible and adaptive mechanism for computing similarity scores between queries and items.
+Given a query \( q \) and an item \( x \), MoL assumes that both are mapped to \( P \) groups of low-rank embeddings, denoted as \( f_p(q) \) and \( g_p(x) \), respectively.
+These embeddings are parameterized by neural networks that extract features from the query and item.
+The similarity between \( q \) and \( x \) is computed as a weighted sum of the inner products of these embeddings, where the weights \( \pi_p(q, x) \)\cite{zhai2023revisiting} are adaptive gating parameters constrained to sum to 1:
+\[
+\text{Similarity}(q, x) = \sum_{p=1}^P \pi_p(q, x) \cdot \langle f_p(q), g_p(x) \rangle.
+\]
+This formulation allows MoL to capture complex relationships between queries and items, making it particularly suitable for dynamic candidate selection tasks.
+
+Inspired by the approach in\cite{ding2024efficient}, we introduce an adaptive threshold mechanism (\textbf{Dynamic Candidate Selection}), the MoL framework is employed to refine the candidate retrieval process.
+Specifically:
+\begin{enumerate}
+	\item \textbf{Component-Level Embeddings:}  
+	 generate component-level embeddings for all items in \( X \). These embeddings allow efficient similarity computations during retrieval.
+	\begin{equation*}
+		X_p \leftarrow \{ g_p(x) \mid x \in X \}
+	\end{equation*}
+
+	\item \textbf{Dynamic Threshold Adjustment:}  
+	To refine retrieval quality, we compute Mixture of Logits (MoL) scores for each candidate \( x \in G \).  
+	The adaptive gating weights \( \pi_p(q, x) \) enable the algorithm to dynamically adjust the retrieval threshold \( T_{\text{adaptive}} \) based on the computed MoL scores.
+	\begin{equation*}
+		\varphi(q, x) = \sum_{p=1}^{P} \pi_p(q, x) \cdot \langle f_p(q), g_p(x) \rangle
+	\end{equation*}
+	\begin{equation*}
+		T_{\text{adaptive}} = \min \{ s \mid s \in G \}
+	\end{equation*}
+
+	\item \textbf{Refinement and Top-K Selection:}  
+	Additional relevant candidates are retrieved using the adaptive threshold \( T_{\text{adaptive}} \), expanding the candidate set \( G' \).
+The algorithm then selects the most relevant top-k results by sorting \( G' \) based on MoL scores.
+	\begin{equation*}
+		G' \leftarrow G \cup \{ x \mid s_p \geq T_{\text{adaptive}} \}
+	\end{equation*}
+	Finally, we sort candidates by MoL scores and extract the exact top-k:
+	\begin{equation*}
+		G_{\text{final}} = \text{Top-}k(G', \varphi(q, x))
+	\end{equation*}
+\end{enumerate}
+the flowchart provided in Figure~\ref{fig2} describes the step-by-step process of the proposed algorithm. It visualizes the key stages of the solution
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=0.9\textwidth]{Flowchart_Steps.PNG}
+	\caption{Algorithm workflow. }\label{fig2}
+\end{figure}
+
+As shown in Algorithm~\ref{Algo1}, the retrieval process dynamically adjusts the threshold based on MoL scores.
+\begin{algorithm}
+	\caption{Hybrid Exact Top-k with Threshold-Based k Selection}\label{ALgo1}
+	\begin{algorithmic}[1]
+		\State \textbf{Input:}
+		\begin{itemize}
+			\item Query $q$
+			\item Set of items $X$
+			\item Component-level embeddings: $f_p(q)$, $g_p(x)$ for $p \in P$, $x \in X$
+			\item Initial threshold $T_{\text{init}}$
+		\end{itemize}
+
+		\State \textbf{Output:}
+		\begin{itemize}
+			\item Exact top $k$ items based on dynamic threshold selection, $G_{\text{final}}$
+		\end{itemize}
+
+		\State \textbf{1. Initialize:}
+		\begin{itemize}
+			\item Set $G \gets \emptyset$ \Comment{Initial candidate set}
+		\end{itemize}
+
+		\State \textbf{2. Generate Component-Level Embeddings:}
+		\For{each component $p \in P$}
+		\State $X_p \gets \{g_p(x) \mid x \in X\}$ \Comment{Precompute embeddings}
+		\EndFor
+
+		\State \textbf{3. Initial Candidate Retrieval:}
+		\For{each component $p \in P$}
+		\State Compute dot product scores: 
+		\[
+		S_p = \{ \langle f_p(q), g_p(x) \rangle : x \in X_p \}
+		\]
+		\State Retrieve items with scores $S_p \geq T_{\text{init}}$
+		\State Add these items to $G$
+		\EndFor
+
+		\State \textbf{4. Adjust k Dynamically:}
+		\For{each $x \in G$}
+		\State Compute MoL scores $s = \phi(q, x)$ for each $x \in G$ using:
+		\[
+		\phi(q, x) = \sum_{p=1}^{P} \pi_p(q, x) \cdot \langle f_p(q), g_p(x) \rangle
+		\]
+		\State Set $T_{\text{adaptive}} = \min \{s : s \in G\}$
+		\EndFor
+
+		\State \textbf{5. Refine Candidate Set with Adaptive k:}
+		\For{each component $p \in P$}
+		\State Retrieve items from $X_p$ with scores $S_p \geq T_{\text{adaptive}}$
+		\State Add these items to $G'$
+		\EndFor
+
+		\State \textbf{6. Select Exact Top-k Items:}
+		\State Compute MoL scores for all items in $G'$
+		\State Sort $G'$ by MoL scores in descending order
+		\State Select the top $k$ items from $G'$ where $k$ is the number of items in $G'$ exceeding $T_{\text{adaptive}}$
+
+		\State \textbf{7. Return:} $G_{\text{final}}$ \Comment{Top $k$ items from $G'$}
+	\end{algorithmic}
+\end{algorithm}
 \section{Experimental Results}\label{sec3}
 \subsection{Experiments settings}\label{subsec2}
-
-Equations in \LaTeX\ can either be inline or on-a-line by itself (``display equations''). For
-inline equations use the \verb+$...$+ commands. E.g.: The equation
-$H\psi = E \psi$ is written via the command \verb+$H \psi = E \psi$+.
-
-For display equations (with auto generated equation numbers)
-one can use the equation or align environments:
-% KAN Archecture
-%\begin{equation}
-%f(x_1, x_2, \cdots, x_n)=\sum_{q=1}^{2n+1}\Phi_q %\left( \sum_{p=1}^{n}\phi_{q, p}(x_p)\right)
-%\end{equation}
-
-\begin{equation}
-\Phi = \{\phi_{p, q}\}, \qquad p=1,2,\cdots, n_{in},\qquad q=1,2,\cdots, n_{out}
-\end{equation}
-
-\begin{equation}
-x_{l+1,j}=\sum_{i=1}^{n_l}\phi_{l, j, i}(x_{l,i}), \qquad j=1,2,\cdots n_{l+1}
-\end{equation}
-
-\begin{equation}
-\mathbf{x}_{l+1} = \begin{pmatrix}
-\phi_{l,1,1}(\cdot)& \phi_{l,1,2}(\cdot) &\cdots &\phi_{l,1,n_l}(\cdot)  \\
-\phi_{l,2,1}(\cdot)&\phi_{l,2,2}(\cdot)  & \cdots &  \phi_{l,2,n_l}(\cdot)\\
-\vdots&\vdots  &\ddots  &\vdots  \\
-\phi_{l,n_{l+1},1}(\cdot)& \phi_{l,n_{l+1},2}(\cdot) &\cdots  & \phi_{l,n_{l+1},n_l}(\cdot)
-\end{pmatrix}\mathbf{x}_{l} =\mathbf{ \Phi}_{l}\mathbf{x}_{l}
-\end{equation}
-\begin{equation}
-\mathbf{KAN(x)}=(\mathbf{\Phi}_{L-1}\circ \mathbf{\Phi}_{L-2}\circ \cdots \circ \mathbf{\Phi}_{1}\circ \mathbf{\Phi}_{0})\mathbf{x}
-\end{equation}
-% models equations
-% system 2
-\begin{equation}
-y_p(k+1) = f\big[y_p(k), y_p(k-1)\big] + u(k)
-\end{equation}
-where
-\begin{equation*}
-f\big[y_p(k), y_p(k-1)\big]=\dfrac{y_p(k)y_p(k-1)[y_p(k)+2.5]}{1+y_p^2(k)+y_p^2(k-1)}
-\end{equation*}
-
-% system 3
-\begin{equation}
-y_p(k+1) = \dfrac{y_p(k)}{1+y_p^2(k)}+ u^3(k)
-\end{equation}
-% system 4
-\begin{equation}
-y_p(k+1) = f\big[y_p(k), y_p(k-1), y_p(k-2), u(k), u(k-1)\big] 
-\end{equation}
-where
-\begin{equation*}
-f\big[x_1, x_2, x_3,x_4, x_5\big]=\dfrac{x_1x_2x_3x_5(x_3-1)+x_4}{1+x_2^2+x_3^2}
-\end{equation*}
 
@@
 
-\section{Figures}\label{sec6}
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=0.9\textwidth]{fig.eps}
-\caption{This is a widefig. This is an example of long caption this is an example of long caption  this is an example of long caption this is an example of long caption}\label{fig1}
-\end{figure}
-
-In case of double column layout, the above format puts figure captions/images to single column width. To get spanned images, we need to provide \verb+\begin{figure*}+ \verb+...+ \verb+\end{figure*}+.
-
-For sample purpose, we have included the width of images in the optional argument of \verb+\includegraphics+ tag. Please ignore this.
-
-\section{Algorithms, Program codes and Listings}\label{sec7}
-
-You may refer above listed package documentations for more details before setting \verb+algorithm+ environment.
-For program codes, the ``verbatim'' package is required and the command to be used is \verb+\begin{verbatim}+ \verb+...+ \verb+\end{verbatim}+.
-
-Similarly, for \verb+listings+, use the \verb+listings+ package.
-\verb+\begin{lstlisting}+ \verb+...+ \verb+\end{lstlisting}+ is used to set environments similar to \verb+verbatim+ environment.
-Refer to the \verb+lstlisting+ package documentation for more details.
-
 A fast exponentiation procedure:
@@
 \end{lstlisting}
-
-\begin{algorithm}
-\caption{Calculate $y = x^n$}\label{algo1}
-\begin{algorithmic}[1]
-\Require $n \geq 0 \vee x \neq 0$
-\Ensure $y = x^n$ 
-\State $y \Leftarrow 1$
-\If{$n < 0$}\label{algln2}
-        \State $X \Leftarrow 1 / x$
-        \State $N \Leftarrow -n$
-\Else
-        \State $X \Leftarrow x$
-        \State $N \Leftarrow n$
-\EndIf
-\While{$N \neq 0$}
-        \If{$N$ is even}
-            \State $X \Leftarrow X \times X$
-            \State $N \Leftarrow N / 2$
-        \Else[$N$ is odd]
-            \State $y \Leftarrow y \times X$
-            \State $N \Leftarrow N - 1$
-        \EndIf
-\EndWhile
-\end{algorithmic}
-\end{algorithm}
 
@@
 %``Figure~\ref{fig1}''.
-
-To reference line numbers in an algorithm, consider the label declared for the line number 2 of Algorithm~\ref{algo1} is \verb+\label{algln2}+.
-To cross-reference it, use the command \verb+\ref{algln2}+ for which it comes up as line~\ref{algln2} of Algorithm~\ref{algo1}.
 

### 2025-02-03 Paper/latex/paper_RAG.tex
added Experimental Results section

@@
 \section{Experimental Results}\label{sec3}
-\subsection{Experiments settings}\label{subsec2}
+We evaluate the performance of our hybrid retrieval algorithm and the Mixture-of-Logits (MoL) approach on the next-item prediction task in recommendation systems\cite{Zhu_2018}\cite{kang2018selfattentivesequentialrecommendation}.
+We conducted experiments using two MovieLens datasets: ML-100K and ML-1M \cite{Harper2015}  which are widely used benchmarks for evaluating sequential recommendation models.\\
+\textbf{MovieLens-100K} \\
+100,000 ratings (1-5 scale) from 943 users on 1,682 movies\\
+Each user has rated at least 20 movies \\
+Includes user demographic information (age, gender, occupation, zip code)\\
+\textbf{MovieLens-1M} \\
+1,000,209 ratings from 6,040 users on approximately 3,900 movies \\
+Data collected from users who joined MovieLens in 2000 \\
+Represents a larger-scale recommendation scenario \\
+\subsection{ Setup }
+For both datasets, We utilize the SASRec architecture as our sequential user encoder, a model renowned for achieving state-of-the-art performance in next-item prediction tasks.
+This architecture processes the user's historical interaction sequence, generating embeddings that encapsulate the user's preferences at each time step.
+These embeddings serve as the foundation for predicting the next item in the sequence.\\
 
-\section{Tables}\label{sec5}
+The query q represents the user's state at a specific time step, derived from their interaction history. In the MoL (Mixture of Logits) framework, q is transformed into Pq
+embeddings through a multi-layer perceptron (MLP). While our hybrid algorithm does not explicitly employ an MLP for this transformation
+\subsection{Hyperparameter settings.}
+For fair comparison, we maintained consistent architectural choices and training conditions across all experiments , We conducted an extensive hyperparameter analysis comparing both approaches (Hybrid+SAS and MoL+SAS) across different architectural configurations.All experiments were performed on Google Colab using T4 GPU.We implemented the model using TensorFlow .We discuss detailed
+hyperparameter settings Table\ref{tab:100k_movies},table\ref{tab:1m_movies}
 
-Tables can be inserted via the normal table and tabular environment. To put
-footnotes inside tables you should use \verb+\footnotetext[]{...}+ tag.
-The footnote appears just below the table itself (refer Tables~\ref{tab1} and \ref{tab2}). 
-For the corresponding footnotemark use \verb+\footnotemark[...]+
-
-\begin{table}[h]
-\caption{Caption text}\label{tab1}%
-\begin{tabular}{@{}llll@{}}
+\begin{table}[htht]
+	\centering
+	\resizebox{\textwidth}{!}{ % Scales the table to text width
+		\begin{tabular}{lcccccccccc}
 			\toprule
-Column 1 & Column 2  & Column 3 & Column 4\\
+			Model & Max Seq Len & Embed Dim & Num Heads & FF Dim & Batch Size & Epochs & Val Loss & Val Accuracy \\
 			\midrule
-row 1    & data 1   & data 2  & data 3  \\
-row 2    & data 4   & data 5\footnotemark[1]  & data 6  \\
-row 3    & data 7   & data 8  & data 9\footnotemark[2]  \\
-\botrule
+			Hybrid+SAS & 50  & 128  & 2 & 128  & 128 & 12 & 5.3036 & 0.1584  \\
+			MoL+SAS    & 50  & 128  & 2 & 128  & 128 & 12 & 5.3152 & 0.1582  \\
+			Hybrid+SAS & 128 & 256  & 4 & 256  & 128 & 10 & 3.6364 & 0.4301  \\
+			MoL+SAS    & 128 & 256  & 4 & 256  & 128 & 10 & 3.6374 & 0.4299 \\
+			Hybrid+SAS & 512 & 512  & 4 & 512  & 128 & 10 & 1.2808 & 0.8120 \\
+			MoL+SAS    & 512 & 512  & 4 & 512  & 128 & 10 & 1.3050 & 0.8016 \\
+			\bottomrule
 		\end{tabular}
-\footnotetext{Source: This is an example of table footnote. This is an example of table footnote.}
-\footnotetext[1]{Example for a first table footnote. This is an example of table footnote.}
-\footnotetext[2]{Example for a second table footnote. This is an example of table footnote.}
+	}
+	\caption{Results on 100kMovies dataset}
+	\label{tab:100k_movies}
 \end{table}
@@
 \begin{table}[h]
-\caption{Example of a lengthy table which is set to full textwidth}\label{tab2}
-\begin{tabular*}{\textwidth}{@{\extracolsep\fill}lcccccc}
-\toprule%
-& \multicolumn{3}{@{}c@{}}{Element 1\footnotemark[1]} & \multicolumn{3}{@{}c@{}}{Element 2\footnotemark[2]} \\\cmidrule{2-4}\cmidrule{5-7}%
-Project & Energy & $\sigma_{calc}$ & $\sigma_{expt}$ & Energy & $\sigma_{calc}$ & $\sigma_{expt}$ \\
+	\centering
+	\resizebox{\textwidth}{!}{ 
+		\begin{tabular}{lcccccccccc}
+			\toprule
+			Model & Max Seq Len & Embed Dim & Num Heads & FF Dim & Batch Size & Epochs & Val Loss & Val Accuracy  \\
 			\midrule
-Element 3  & 990 A & 1168 & $1547\pm12$ & 780 A & 1166 & $1239\pm100$\\
-Element 4  & 500 A & 961  & $922\pm10$  & 900 A & 1268 & $1092\pm40$\\
-\botrule
-\end{tabular*}
-\footnotetext{Note: This is an example of table footnote.
-This is an example of table footnote this is an example of table footnote this is an example of~table footnote this is an example of table footnote.}
-\footnotetext[1]{Example for a first table footnote.}
-\footnotetext[2]{Example for a second table footnote.}
+			Hybrid+SAS & 50  & 128  & 2 & 128  & 128 & 10 & 4.5721 & 0.1530  \\
+			MoL+SAS    & 50  & 128  & 2 & 128  & 128 & 10 & 4.5733 & 0.1526  \\
+			Hybrid+SAS & 128 & 256  & 4 & 256  & 128 & 10 & 3.4152 & 0.3680  \\
+			MoL+SAS    & 128 & 256  & 4 & 256  & 128 & 10 & 3.4671 & 0.3627  \\
+			Hybrid+SAS & 512 & 512  & 4 & 512  & 128 & 10 & 1.0275 & 0.8350  \\
+			MoL+SAS    & 512 & 512  & 4 & 512  & 128 & 10 & 1.0350 & 0.8276  \\
+			\bottomrule
+		\end{tabular}
+	}
+	\caption{Results on 1M Movies dataset}
+	\label{tab:1m_movies}
 \end{table}
 
-In case of double column layout, tables which do not fit in single column width should be set to full text width.
-For this, you need to use \verb+\begin{table*}+ \verb+...+ \verb+\end{table*}+ instead of \verb+\begin{table}+ \verb+...+ \verb+\end{table}+ environment.
-Lengthy tables which do not fit in textwidth should be set as rotated table.
-For this, you need to use \verb+\begin{sidewaystable}+ \verb+...+ \verb+\end{sidewaystable}+ instead of \verb+\begin{table*}+ \verb+...+ \verb+\end{table*}+ environment.
-This environment puts tables rotated to single column width.
-For tables rotated to double column width, use \verb+\begin{sidewaystable*}+ \verb+...+ \verb+\end{sidewaystable*}+.
+Both approaches demonstrate significant performance improvements as model capacity increases, with larger configurations consistently delivering better results.
+For instance, when increasing the model's capacity?such as expanding the (Max Sequence Length: 512, Embedding Dimension: 512, Number of Heads: 4, Feed-Forward Dimension: 512) the hybrid algorithm combined with SASRec (Hybrid+SAS) achieves notable gains, particularly in the most resource-intensive setup.
+On the ML-100K dataset, Hybrid+SAS reaches a score of\textbf{ 0.8120} compared to the baseline's 0.8016\textbf{0.8016}, while on ML-1M, it achieves \textbf{0.8350} versus \textbf{0.8276.} The ML-1M dataset generally benefits more from increased capacity, with the performance gap between datasets narrowing as the model scales.
+While smaller configurations provide a balance of efficiency and performance, the largest configuration, despite its higher computational demands, yields the best results, making it suitable for scenarios where resources are not a constraint.
+Both approaches show similar benefits from scaling, but Hybrid+SAS maintains a consistent edge in performance.
+\subsection{metrics }
 
-\begin{sidewaystable}
-\caption{Tables which are too long to fit, should be written using the ``sidewaystable'' environment as shown here}\label{tab3}
-\begin{tabular*}{\textheight}{@{\extracolsep\fill}lcccccc}
-\toprule%
-& \multicolumn{3}{@{}c@{}}{Element 1\footnotemark[1]}& \multicolumn{3}{@{}c@{}}{Element\footnotemark[2]} \\\cmidrule{2-4}\cmidrule{5-7}%
-Projectile & Energy	& $\sigma_{calc}$ & $\sigma_{expt}$ & Energy & $\sigma_{calc}$ & $\sigma_{expt}$ \\
-\midrule
-Element 3 & 990 A & 1168 & $1547\pm12$ & 780 A & 1166 & $1239\pm100$ \\
-Element 4 & 500 A & 961  & $922\pm10$  & 900 A & 1268 & $1092\pm40$ \\
-Element 5 & 990 A & 1168 & $1547\pm12$ & 780 A & 1166 & $1239\pm100$ \\
-Element 6 & 500 A & 961  & $922\pm10$  & 900 A & 1268 & $1092\pm40$ \\
-\botrule
-\end{tabular*}
-\footnotetext{Note: This is an example of table footnote this is an example of table footnote this is an example of table footnote this is an example of~table footnote this is an example of table footnote.}
-\footnotetext[1]{This is an example of table footnote.}
-\end{sidewaystable}
+We employ standard ranking metrics widely used in sequential recommendation:
+\textbf{Hit Rate at k (HR@k)}: Measures the proportion of cases where the target item appears in the top-k recommendations ,
+\textbf{Mean Reciprocal Rank (MRR):} Evaluates the average reciprocal of the rank at which the first relevant item is retrieved and
+\textbf{Normalized Discounted Cumulative Gain (NDCG):} Assesses the ranking quality while accounting for position importance The following table\ref{tab:results} summarizes the performance of our hybrid algorithm compared to the MoL-based approach on the MovieLens 100K,1M datasets:
 
-A fast exponentiation procedure:
+\begin{table}[h]
+	\centering
+	\label{tab:results}
+	\begin{tabular}{|l|c|c|c|c|c|c|}
+		\hline
+		\textbf{Model} & \textbf{HR@1} & \textbf{HR@10} & \textbf{HR@50} & \textbf{HR@200} & \textbf{MRR} & \textbf{NDCG} \\ \hline
+		\multicolumn{7}{|c|}{\textbf{MovieLens 100K}} \\ \hline
+		SASRec+Hybrid & 0.0070 & 0.0775 & 0.2606 & 0.5704 & 0.0363 & 0.1241 \\ \hline
+		SASRec+MoL    & 0.0000 & 0.0775 & 0.2887 & 0.5775 & 0.0285 & 0.1182 \\ \hline
+		\multicolumn{7}{|c|}{\textbf{MovieLens 1M}} \\ \hline
+		SASRec+Hybrid & 0.0464 & 0.1810 & 0.3819 & 0.6159 & 0.0925 & 0.1829 \\ \hline
+		SASRec+MoL    & 0.0508 & 0.1799 & 0.3896 & 0.6126 & 0.0971 & 0.1859 \\ \hline
+	\end{tabular}
+			\caption{Performance Comparison of SASRec+Hybrid and SASRec+MoL on MovieLens 100K and 1M Datasets}
+\end{table}
 
-\lstset{texcl=true,basicstyle=\small\sf,commentstyle=\small\rm,mathescape=true,escapeinside={(*}{*)}}
-\begin{lstlisting}
-begin
-  for $i:=1$ to $10$ step $1$ do
-      expt($2,i$);  
-      newline() od                (*\textrm{Comments will be set flush to the right margin}*)
-where
-proc expt($x,n$) $\equiv$
-  $z:=1$;
-  do if $n=0$ then exit fi;
-     do if odd($n$) then exit fi;                 
-        comment: (*\textrm{This is a comment statement;}*)
-        $n:=n/2$; $x:=x*x$ od;
-     { $n>0$ };
-     $n:=n-1$; $z:=z*x$ od;
-  print($z$). 
-end
-\end{lstlisting}
+On the ML-100K dataset, our hybrid approach shows comparable performance to MoL in terms of \textbf{HR@10 (0.0775 for both)} while achieving better results in\textbf{ MRR (0.0363 vs 0.0285)} and \textbf{NDCG (0.1241 vs 0.1182)}.
+The hybrid algorithm particularly excels in early position recommendations, as evidenced by the non-zero HR@1 score.
+
+On the larger ML-1M dataset, both approaches show significant improvement in performance compared to ML-100K.
+Our hybrid algorithm achieves slightly better \textbf{HR@10 (0.1810 vs 0.1799)} \textbf{and HR@200 (0.6159 vs 0.6126)} compared to MoL, while MoL maintains a small edge in other metrics.
+
+\newpage
 
@@
 %%% Start your code-block
-\begin{lstlisting}
-for i:=maxint to 0 do
-begin
-{ do nothing }
-end;
-Write('Case insensitive ');
-Write('Pascal keywords.');
-\end{lstlisting}
+
 \end{minipage}
-
-\section{Cross referencing}\label{sec8}
-
-%Environments such as figure, table, equation and align can have a label
-%declared via the \verb+\label{#label}+ command. For figures and table
-%environments use the \verb+\label{}+ command inside or just
-%below the \verb+\caption{}+ command. You can then use the
-%\verb+\ref{#label}+ command to cross-reference them. As an example, consider
-%the label declared for Figure~\ref{fig1} which is
-%\verb+\label{fig1}+. To cross-reference it, use the command 
-%\verb+Figure \ref{fig1}+, for which it comes up as
-%``Figure~\ref{fig1}''.
-
-\subsection{Details on reference citations}\label{subsec7}
-
-Standard \LaTeX\ permits only numerical citations.
-To support both numerical and author-year citations this template uses \verb+natbib+ \LaTeX\ package.
-For style guidance please refer to the template user manual.
 
@@
 %All cited bib entries are printed at the end of this article: \cite{bib3}, %\cite{bib4}, \cite{bib5}, \cite{bib6}, \cite{bib7}, \cite{bib8}, \cite{bib9}, %\cite{bib10}, \cite{bib11}, \cite{bib12} and \cite{bib13}.
-
-\section{Examples for theorem like environments}\label{sec10}
-
-For theorem like environments, we require \verb+amsthm+ package. There are three types of predefined theorem styles exists---\verb+thmstyleone+, \verb+thmstyletwo+ and \verb+thmstylethree+
 
@@
 \end{tabular}
-
-\section{Discussion}\label{sec12}
-
-Discussions should be brief and focused.
-In some disciplines use of Discussion or `Conclusion' is interchangeable.
-It is not mandatory to use both.
-Some journals prefer a section `Results and Discussion' followed by a section `Conclusion'.
-Please refer to Journal-level guidance for any specific requirements.
 

### 2025-02-08 Paper/latex/paper_RAG.tex
add conclusion

@@
 \subsection{Hyperparameter settings.}
-For fair comparison, we maintained consistent architectural choices and training conditions across all experiments , We conducted an extensive hyperparameter analysis comparing both approaches (Hybrid+SAS and MoL+SAS) across different architectural configurations.All experiments were performed on Google Colab using T4 GPU.We implemented the model using TensorFlow .We discuss detailed
+For fair comparison, we maintained consistent architectural choices and training conditions across all experiments , We conducted an extensive hyperparameter analysis comparing both approaches (Hybrid+SAS and MoL+SAS) across different architectural configurations.All experiments were pmplemented in TensorFlow and trained on a Google Colab environment with a T4 GPU.
+We use the Adam optimizer with a learning rate of 0.001.
+For the hybrid algorithm, we initialize the threshold (Tinit) to 0.3 and adaptively adjust it during training.
+.We discuss detailed
 hyperparameter settings Table\ref{tab:100k_movies},table\ref{tab:1m_movies}
-
-\begin{table}[htht]
+\begin{table}[h]
 	\centering
@@
 \end{table}
-
+\newpage
 Both approaches demonstrate significant performance improvements as model capacity increases, with larger configurations consistently delivering better results.
 For instance, when increasing the model's capacity?such as expanding the (Max Sequence Length: 512, Embedding Dimension: 512, Number of Heads: 4, Feed-Forward Dimension: 512) the hybrid algorithm combined with SASRec (Hybrid+SAS) achieves notable gains, particularly in the most resource-intensive setup.
-On the ML-100K dataset, Hybrid+SAS reaches a score of\textbf{ 0.8120} compared to the baseline's 0.8016\textbf{0.8016}, while on ML-1M, it achieves \textbf{0.8350} versus \textbf{0.8276.} The ML-1M dataset generally benefits more from increased capacity, with the performance gap between datasets narrowing as the model scales.
+On the ML-100K dataset, Hybrid+SAS reaches a score of\textbf{ 0.8120} compared to the baseline's \textbf{0.8016}, while on ML-1M, it achieves \textbf{0.8350} versus \textbf{0.8276.} The ML-1M dataset generally benefits more from increased capacity, with the performance gap between datasets narrowing as the model scales.
 While smaller configurations provide a balance of efficiency and performance, the largest configuration, despite its higher computational demands, yields the best results, making it suitable for scenarios where resources are not a constraint.
@@
 \textbf{Normalized Discounted Cumulative Gain (NDCG):} Assesses the ranking quality while accounting for position importance The following table\ref{tab:results} summarizes the performance of our hybrid algorithm compared to the MoL-based approach on the MovieLens 100K,1M datasets:
-
 \begin{table}[h]
@@
 \end{table}
-
+\newpage
 On the ML-100K dataset, our hybrid approach shows comparable performance to MoL in terms of \textbf{HR@10 (0.0775 for both)} while achieving better results in\textbf{ MRR (0.0363 vs 0.0285)} and \textbf{NDCG (0.1241 vs 0.1182)}.
@@
 On the larger ML-1M dataset, both approaches show significant improvement in performance compared to ML-100K.
-Our hybrid algorithm achieves slightly better \textbf{HR@10 (0.1810 vs 0.1799)} \textbf{and HR@200 (0.6159 vs 0.6126)} compared to MoL, while MoL maintains a small edge in other metrics.
+Our hybrid algorithm achieves slightly better \textbf{HR@10 (0.1810 vs 0.1799)} \textbf{and HR@200 (0.6159 vs 0.6126)} compared to MoL, while MoL maintains a small edge in other metrics.\\
 
-\newpage
-
+The results demonstrate that our hybrid algorithm effectively leverages both sequential user behavior and item metadata to improve recommendation quality.
+The adaptive MoL threshold allows the model to dynamically refine candidate items during training, leading to better performance.
+The improvements are consistent across both datasets, highlighting the robustness of our approach.
 %%=============================================%%
@@
 %All cited bib entries are printed at the end of this article: \cite{bib3}, %\cite{bib4}, \cite{bib5}, \cite{bib6}, \cite{bib7}, \cite{bib8}, \cite{bib9}, %\cite{bib10}, \cite{bib11}, \cite{bib12} and \cite{bib13}.
-
 %%=============================================%%
@@
 %%=============================================%%
-\bigskip
-\begin{tabular}{|l|p{19pc}|}
-\hline
-\verb+thmstyleone+ & Numbered, theorem head in bold font and theorem text in italic style \\\hline
-\verb+thmstyletwo+ & Numbered, theorem head in roman font and theorem text in italic style \\\hline
-\verb+thmstylethree+ & Numbered, theorem head in bold font and theorem text in roman style \\\hline
-\end{tabular}
 
@@
 
-Conclusions may be used to restate your hypothesis or research question, restate your major findings, explain the relevance and the added value of your work, highlight any limitations of your study, describe future directions for research and recommendations.
+Choosing the right value for K in retrieval-based systems is a significant challenge, as it directly impacts both the efficiency and accuracy of information retrieval.
+While using a fixed K is straightforward, it often fails to adapt to the varying complexities of different queries.
+This can result in either missing important information or retrieving too much irrelevant data, which adds noise to the results.
+On the other hand, dynamically adjusting K offers greater flexibility but comes with its own set of challenges, such as increased computational costs and inconsistencies in determining the optimal retrieval size.\\
+To address these limitations, we introduced a hybrid dynamic K selection strategy that strikes a balance between adaptability and efficiency.
+Our approach adjusts K based on the unique characteristics of each query while keeping computational demands manageable.
+This method has shown promising results, improving retrieval performance by minimizing unnecessary noise and ensuring that critical information is consistently captured.\\
+Future work can explore enhancing our model with learning-based retrieval strategie such as reinforcement learning or attention mechanisms?could refine the dynamic selection of K Additionally, testing our approach in real-world scenarios and conducting large-scale evaluations across diverse domains will help validate its robustness and adaptability in various knowledge bases and retrieval settings.
 
-In some disciplines use of Discussion or 'Conclusion' is interchangeable. It is not mandatory to use both. Please refer to Journal-level guidance for any specific requirements.
-
-\backmatter
-
-\bmhead{Supplementary information}
-
-If your article has accompanying supplementary file/s please state so here.
-
-Authors reporting data from electrophoretic gels and blots should supply the full unprocessed scans for key as part of their Supplementary information.
-This may be requested by the editorial team/s if it is missing.
-
-Please refer to Journal-level guidance for any specific requirements.
-
-\bmhead{Acknowledgements}
-
-Acknowledgements are not compulsory. Where included they should be brief. Grant or contribution numbers may be acknowledged.
-
-Please refer to Journal-level guidance for any specific requirements.
-
-\section*{Declarations}
-
-Some journals require declarations to be submitted in a standardised format.
-Please check the Instructions for Authors of the journal to which you are submitting to see if you need to complete this section.
-If yes, your manuscript must contain the following sections under the heading `Declarations':
-
-\begin{itemize}
-\item Funding
-\item Conflict of interest/Competing interests (check journal-specific guidelines for which heading to use)
-\item Ethics approval and consent to participate
-\item Consent for publication
-\item Data availability 
-\item Materials availability
-\item Code availability 
-\item Author contribution
-\end{itemize}
+\newpage
 

### 2025-02-10 Paper/latex/paper_RAG.tex
add abstract

@@
 
-\abstract{rrrrrrrrrrrrrrrr}
+\abstract{Retrieval-Augmented Generation (RAG) improves the quality of language model responses by incorporating external knowledge.
+However, its performance heavily relies on choosing the right number of documents to retrieve (K).
+Using a fixed number of documents (K) is computationally efficient but fails to adapt to the complexity of different queries.
+On the other hand, dynamically adjusting K offers flexibility but comes with higher computational costs and inconsistency.
+To overcome these challenges, we introduce a hybrid approach to dynamic K selection.
+This method adapts K based on the specific characteristics of each query while keeping computational demands manageable.
+Our strategy enhances retrieval performance by minimizing irrelevant information, ensuring the retrieval of relevant data, and boosting overall system accuracy.
+Experiments show that this approach outperforms both fixed and fully dynamic methods, offering a promising way to optimize retrieval in RAG systems.Future research will focus on developing learning-based techniques to further improve the dynamic selection of K.}
 
@@
 
-Large Language Models (LLMs) have emerged as state-of-the-art artificial intelligence systems that excel at understanding and generating human-like text,Built on deep learning architectures,particularly transformers \cite{vaswani2017attention}.these models are trained on massive datasets to capture linguistic patterns and semantic relationships showcasing impressive adaptability across diverse tasks such as text generation, summarization, question answering.Notable examples of LLMs include OpenAI GPT series\cite{brown2020language} , Google BERT\cite{devlin2018bert} and T5 \cite{raffel2020exploring} which have set new benchmarks in NLP tasks .\\
+Large Language Models (LLMs)\cite{Radford2019} have emerged as state-of-the-art artificial intelligence systems that excel at understanding and generating human-like text,Built on deep learning architectures,particularly transformers \cite{vaswani2017attention}.these models are trained on massive datasets to capture linguistic patterns and semantic relationships showcasing impressive adaptability across diverse tasks such as  machine translation\cite{Gu2018SearchEngine}, text generation\cite{liang2024controllabletextgenerationlarge} and question answering\cite{Zhang2023}.Notable examples of LLMs include OpenAI GPT series\cite{brown2020language} , Google BERT\cite{devlin2018bert} and T5 \cite{raffel2020exploring} which have set new benchmarks in NLP tasks .\\
 Despite their impressive capabilities,LLMs suffer from hallucination\cite{ji2023survey} where they confidently generate factually incorrect or entirely fabricated information.
@@
 A critical challenge is determining the optimal number of retrieved passages (K) to feed into the generator\cite{lewis2020retrieval}.
-Selecting too few passages (small K) risks missing crucial context, while too many (large K) can introduce noise, reduce coherence and increase computational overhead.
+often resulting in either insufficient or excessive information being retrieved, reduce coherence and increase computational overhead.
 This trade-off makes K a key factor in balancing performance and efficiency.
 
-Current approaches, such as static K selection \cite{lewis2020retrieval} involves retrieving a fixed number of top results regardless of the query characteristics ,it struggles in dynamic datasets as it cannot adapt to the complexity of individual queries , dynamic K adjustment attempt to address this trade-off but often introduce complexity or latency ,Hybrid approaches like those proposed yuan et al .
-\cite{yuan2024hybrid},that integrates static retrieval with dynamic retrieval to improve accuracy and robustness.
-Furthermore, the "Blended RAG" approach \cite{blended_rag2024} enhances retrieval accuracy by combining semantic search techniques with hybrid query strategies, resulting in improved performance in information retrieval tasks.
+Current approaches, such as static K selection \cite{lewis2020retrieval} involves retrieving a fixed number of top results regardless of the query characteristics ,it cannot adapt to the complexity of individual queries , dynamic K adjustment attempt to address this trade-off but often introduce complexity or latency ,Hybrid approaches like those proposed yuan et al .
+\cite{yuan2024hybrid},that integrates static retrieval with dynamic retrieval to improve accuracy.
+Furthermore, the "Blended RAG"\cite{sawarkar2024blendedragimprovingrag} approach \cite{blended_rag2024} enhances retrieval accuracy by combining semantic search techniques with hybrid query strategies.
 However, the approach faces significant limitations, including high computational overhead due to the need for multiple retrieval methods and dependency on high-quality metadata for optimal performance.
@@
 
-In this paper, we investigate the problem of selecting K in RAG and propose a novel approach to dynamically adjust K based on relevance thresholds and a Mixture of Logits (MoL)\cite{ding2024efficient} scoring mechanism, ensuring efficient and context-sensitive candidate selection while reducing computational costs.
-By using adaptive thresholds the algorithm improves the relevance of retrieved documents, refining the set to include only the most contextually appropriate matches.
+In this paper, we investigate the problem of selecting K in RAG and propose a novel approach to dynamically adjust K based on  thresholds and a Mixture of Logits (MoL)\cite{ding2024efficient} scoring mechanism, ensuring efficient and context-sensitive candidate selection while reducing computational costs.
+the use of  adaptive thresholds algorithm improves the relevance of retrieved documents, refining the set to include only the most contextually appropriate matches.
 Unlike traditional approaches, it adapts to the query and dataset characteristics, making it more flexible for diverse contexts.
@@
 We critically analyze the strengths and limitations of these methods, identifying key gaps in their design and performance.
-These gaps motivate the need for an improved approach that addresses the challenges of both static and dynamic relevance in retrieval-augmented systems.
+These gaps motivate the need for an improved approach that addresses the challenges of both static and dynamic in retrieval-augmented systems.
 \\
 Overview of Retrieval-Augmented Generation (RAG) in This section We provide a detailed explanation of the RAG framework, including its components and how it operates.
-Key concepts, such as retrievers, generators, and augmented are discussed to establish the foundational context.\\
+Key concepts, such as retrievers, generators are discussed to establish the foundational context.\\
 Proposed Solution: This section introduces our novel hybrid dynamic  selection algorithm.
@@
 This can result in either missing important information or retrieving too much irrelevant data, which adds noise to the results.
-On the other hand, dynamically adjusting K offers greater flexibility but comes with its own set of challenges, such as increased computational costs and inconsistencies in determining the optimal retrieval size.\\
+On the other hand, dynamically adjusting K offers greater flexibility but comes with its own set of challenges, such as increased computational costs and inconsistencies in determining the optimal retrieval size.
 To address these limitations, we introduced a hybrid dynamic K selection strategy that strikes a balance between adaptability and efficiency.
 Our approach adjusts K based on the unique characteristics of each query while keeping computational demands manageable.
-This method has shown promising results, improving retrieval performance by minimizing unnecessary noise and ensuring that critical information is consistently captured.\\
+This method has shown promising results, improving retrieval performance by minimizing unnecessary noise and ensuring that critical information is consistently captured.
 Future work can explore enhancing our model with learning-based retrieval strategie such as reinforcement learning or attention mechanisms?could refine the dynamic selection of K Additionally, testing our approach in real-world scenarios and conducting large-scale evaluations across diverse domains will help validate its robustness and adaptability in various knowledge bases and retrieval settings.

### 2025-02-15 arabic_survey_paper/latex/samplepaper.tex
added a table summarizing studies on the use of Arabic AI in rag

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.21 of 2022/01/12
%
\documentclass[runningheads]{llncs}
%
\usepackage[T1]{fontenc}
% T1 fonts will be used to generate the final print and online PDFs,
% so please use T1 fonts in your manuscript whenever possible.
% Other font encondings may result in incorrect characters.
%
\usepackage{graphicx}

% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following two lines
% to display URLs in blue roman font according to Springer's eBook style:
\usepackage{color}
%\renewcommand\UrlFont{\color{blue}\rmfamily}
%
\begin{document}
%
\title{Contribution Title\thanks{Supported by organization x.}}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{First Author\inst{1}\orcidID{0000-1111-2222-3333} \and
Second Author\inst{2,3}\orcidID{1111-2222-3333-4444} \and
Third Author\inst{3}\orcidID{2222--3333-4444-5555}}
%
\authorrunning{F. Author et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Princeton University, Princeton NJ 08544, USA \and
Springer Heidelberg, Tiergartenstr. 17, 69121 Heidelberg, Germany
\email{lncs@springer.com}\\
\url{http://www.springer.com/gp/computer-science/lncs} \and
ABC Institute, Rupert-Karls-University Heidelberg, Heidelberg, Germany\\
\email{\{abc,lncs\}@uni-heidelberg.de}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
The abstract should briefly summarize the contents of the paper in
150--250 words.

\keywords{First keyword  \and Second keyword \and Another keyword.}
\end{abstract}
%
%
%
\section{First Section}
\begin{table}
	\caption { Arabic AI Models Studies in Retrieval-Augmented Generation (RAG)  }\label{tab:arabic_rag}
	\centering
	\resizebox{\textwidth}{!}{ 
		\begin{tabular}{|p{3.5cm}|p{1.4cm}|p{3.5cm}|p{4.5cm}|p{3cm}|p{3.5cm}|p{3.5cm}|}
		\hline
		\multicolumn{7}{|c|}{\textbf{Studies on Retrieval-Only Models}} \\
		\hline
			\textbf{Paper} & \textbf{year} &
		    \textbf{Retrieval Component Studied} 	& 
		    \textbf{Arabic Model Name} & \textbf{Datasets Used} & \textbf{Evaluation Metrics} & \textbf{Addressed Challenges}  \\
			\hline
			Semantic Embeddings for Arabic Retrieval
			Augmented Generation (ARAG) \cite{ref_arag2023} & 2023 &\textbf{Retrieval} :Semantic   Embeddings &
			Microsoft(e5s,e5b,e5l)\newline DistillBert(hf1,hf2),\newline Openai  Ada embedding \newline  Cohere Multilingual Embedding \newline Meta SONAR  \newline Google LaBSE\newline mpnet-base-v2   & Arabic Reading Comprehension Dataset (ARCD) & Recall@k.&-Embedding Size Constraints \newline      -Need for Language Specific Evaluation Metrics \\
			\hline
			Evaluation of Semantic Search and its Role in Retrieved for Arabic Language\cite{Mahboub2024} & 2024 &\textbf{Retrieval} :Semantic search in Arabic &Encoder 1: MiniLM \newline
			Encoder 2: CMLM\newline
			Encoder 3: MPNet \newline
		    Encoder 4: DistilBERT\newline
		    Encoder 5 : XLM RoBERTa &FAQs: 816 questions with verifiable answers & NDCG@3 \newline MRR@3 \newline mAP@3 & -Embedding size constraints. \newline -Arabic complexity \\
			\hline
			Arabic RAG Leaderboard: A Comprehensive Framework for Evaluating Arabic Language Retrieval Systems\cite{ref_TARL2025} & 2025&
			\textbf{Retrieval}: Semantic Embedding . \newline 
			\textbf{Reranking}:Refine retrieved documents & Retrieval :GATE-AraBERT-v1 \newline
			Reranking : ARA-Reranker-V1\newline
			 &Retrieval:"Web Search Dataset" \newline Reranking:sourced from TyDi QA and MKQA datasets & NDCG \newline MRR \newline mAP \newline Recall@k &  Arabic’s morphological complexity \newline  Dialect diversity\\
			\hline
	        \multicolumn{7}{|c|}{\textbf{Studies on Both Retrieval and Generation}} \\
	        \hline

	        \textbf{Paper} & \textbf{Year} & \textbf{RAG Component Studied} & \textbf{Arbic  Model Name} & \textbf{Datasets Used} & \textbf{Evaluation Metricsl} & \textbf{Addressed Chalenges} \\
	        \hline
			Exploring Retrieval Augmented Generation in Arabic \cite{ref_elbeltagy2024} & 2024 . &  \textbf{Retrieval}: Semantic Embedding \newline \textbf{Generation} : generate Arabic response  & \textbf{Retrieval}: AraVec , AraBERT ,OpenAI ,Cohere, Microsoft’s E5,  Ollama ,JAIS,BGE \newline \textbf{Generator}:GPT3.5, urbo, Mistral 
			7B, Llama 3, Mixtral, and JAIS.
			  & Ar-EduText dataset, \newline ARCD dataset 
			  &\textbf{Retrival}:Recall@K (k=1,k=3,k=5) \newline \textbf{Generator}: F1 Score Bleu Score Cosine SimilarityR&-Lack of Detailed Metrics \newline -Dialect Diversity \\
			\hline
		Evaluating RAG Pipelines for Arabic Lexical Information Retrieval: A Comparative Study of Embedding and Generation Models\cite{ref_alrasheed2025} & 2025 & \textbf{Retrieval}: Word Embeddings , Sentence Embeddings \newline \textbf{Generator} : generatte Arabic lexical information&\textbf{Retrival}:CAMeLBERT, AraBERT-v2, E5-large Arabic-NLI ,AraELECTRA  \newline \textbf{Generator}: GPT-4o, Gemini-1.5-flash, SILMA-9B-Instruct, Aya8B GPT-3.5, AceGPT13B&Riyadh dictionary (88,000+ words)&\textbf{Retrival} Top-k Recall (k = 1, 3, 5) and MRR \newline \textbf{Generator}:F1-score(F1),
		Accuracy (Acc), and CosineSimilarity(Cos). &-handling of dialectal variations in queries and documents. \newline -Disparity in performance between sentence embeddings and word \\
			\hline

		\end{tabular}
	} % End resizebox
\end{table}

%\subsection{A Subsection Sample}

%\subsubsection{Sample Heading} Only two levels of

%\paragraph{Sample Heading (Fourth Level)}

%\begin{figure}
%\includegraphics[width=\textwidth]{fig1.eps}
%\caption{A figure caption is always placed below the illustration.
%Please note that short captions are centered, while long ones are
%justified by the macro package automatically.} \label{fig1}
%\end{figure}%

% the environments 'definition', 'lemma', 'proposition', 'corollary',
% 'remark', and 'example' are defined in the LLNCS documentclass as well.
%

%For citations of references, we prefer the use of square brackets
%and consecutive numbers. Citations using labels or the author/year
%convention are also acceptable. The following bibliography provides
%a sample reference list with entries for journal
%articles~\cite{ref_article1}, an LNCS chapter~\cite{ref_lncs1}, a
%book~\cite{ref_book1}, proceedings without editors~\cite{ref_proc1},
%and a homepage~\cite{ref_url1}. Multiple citations are grouped
%\cite{ref_article1,ref_lncs1,ref_book1},
%\cite{ref_article1,ref_book1,ref_proc1,ref_url1}.

%\subsubsection{Acknowledgements} Please place your acknowledgments at
%the end of the paper, preceded by an unnumbered run-in heading (i.e.
%3rd-level heading).

%
% ---- Bibliography ----
%
% BibTeX users should specify bibliography style 'splncs04'.
% References will then be sorted and formatted in the correct style.
%
% \bibliographystyle{splncs04}
% \bibliography{mybibliography}
%
\newpage
\begin{thebibliography}{8}
\bibitem{ref_arag2023}
Abdelazim, H., Mohamed, A., Tharwat, M.: Semantic Embeddings for Arabic Retrieval Augmented Generation (ARAG). International Journal of Advanced Computer Science and Applications \textbf{14}(11), xx--yy (2023)

\bibitem{Mahboub2024}
Mahboub, A., Za’ter, M. E., Al-Rfooh, B., Estaitia, Y., Jaljuli, A., Hakouz, A.:  
Evaluation of Semantic Search and its Role in Retrieved-Augmented-Generation (RAG) for Arabic Language.  
In: arXiv preprint, arXiv:2403.18350v2, May 2024. Available at: \url{https://arxiv.org/abs/2403.18350}

\bibitem{ref_TARL2025}
Rashad, M.A., Shahid, H.: The Arabic RAG Leaderboard. \textit{Navid-AI} (2025). Available online at: \url{https://huggingface.co/spaces/Navid-AI/The-Arabic-Rag-Leaderboard}.

\bibitem{ref_elbeltagy2024}
El-Beltagy, S.R., Abdallah, M.A.: Exploring Retrieval Augmented Generation in Arabic. Procedia Computer Science, \textbf{00}, 000--000 (2024). Available online at: \url{www.sciencedirect.com}.

\bibitem{ref_alrasheed2025}
Al-Rasheed, R., Al Muaddi, A., Aljasim, H., Al-Matham, R., Alhoshan, M., Al Wazrah, A., AlOsaimy, A.:  
Evaluating RAG Pipelines for Arabic Lexical Information Retrieval: A Comparative Study of Embedding and Generation Models.  
In: El-Haj, M. (ed.), Proc. 1st Workshop on NLP for Languages Using Arabic Script, pp. 155--164.  
Association for Computational Linguistics, Abu Dhabi, UAE (2025).  
Available: \url{https://aclanthology.org/2025.abjadnlp-1.16/}

\bibitem{ref_proc1}
Author, A.-B.: Contribution title. In: 9th International Proceedings
on Proceedings, pp. 1--2. Publisher, Location (2010)

\bibitem{ref_url1}
LNCS Homepage, \url{http://www.springer.com/lncs}. Last accessed 4
Oct 2017
\end{thebibliography}
\end{document}

### 2025-02-19 Paper/latex/paper_RAG.tex
last update

@@
 \abstract{Retrieval-Augmented Generation (RAG) improves the quality of language model responses by incorporating external knowledge.
-However, its performance heavily relies on choosing the right number of documents to retrieve (K).
-Using a fixed number of documents (K) is computationally efficient but fails to adapt to the complexity of different queries.
-On the other hand, dynamically adjusting K offers flexibility but comes with higher computational costs and inconsistency.
+However, its performance heavily relies on choosing the right number of documents to retrieve (K).therefore,despite it being computationally efficient,  Using a fixed number of documents (K) fails to adapt to the complexity of different queries.On the other hand, dynamically adjusting K offers flexibility but comes with higher computational costs and inconsistency.
 To overcome these challenges, we introduce a hybrid approach to dynamic K selection.
 This method adapts K based on the specific characteristics of each query while keeping computational demands manageable.
-Our strategy enhances retrieval performance by minimizing irrelevant information, ensuring the retrieval of relevant data, and boosting overall system accuracy.
+Consequently, our  strategy enhances retrieval performance by minimizing irrelevant information, ensuring the retrieval of relevant data, and boosting overall system accuracy.
 Experiments show that this approach outperforms both fixed and fully dynamic methods, offering a promising way to optimize retrieval in RAG systems.Future research will focus on developing learning-based techniques to further improve the dynamic selection of K.}
@@
 
-Large Language Models (LLMs)\cite{Radford2019} have emerged as state-of-the-art artificial intelligence systems that excel at understanding and generating human-like text,Built on deep learning architectures,particularly transformers \cite{vaswani2017attention}.these models are trained on massive datasets to capture linguistic patterns and semantic relationships showcasing impressive adaptability across diverse tasks such as  machine translation\cite{Gu2018SearchEngine}, text generation\cite{liang2024controllabletextgenerationlarge} and question answering\cite{Zhang2023}.Notable examples of LLMs include OpenAI GPT series\cite{brown2020language} , Google BERT\cite{devlin2018bert} and T5 \cite{raffel2020exploring} which have set new benchmarks in NLP tasks .\\
-Despite their impressive capabilities,LLMs suffer from hallucination\cite{ji2023survey} where they confidently generate factually incorrect or entirely fabricated information.
-This poses challenges in applications requiring high accuracy like knowledge-intensive tasks\cite{kandpal2023large} , Additionally significant limitation of LLMs is their static nature, as they are trained on fixed datasets\cite{bender2021dangers} and cannot dynamically update their knowledge which makes them prone to generating outdated or inconsistent responses, especially for time-sensitive or evolving data\cite{Mousavi2025}.
+Large Language Models (LLMs)\cite{Radford2019} have emerged as state-of-the-art artificial intelligence systems that excel at understanding and generating human-like text,Built on deep learning architectures,particularly transformers \cite{vaswani2017attention}.
+These models are trained on massive datasets to capture linguistic patterns and semantic relationships, thereby showcasing impressive adaptability across diverse tasks, such as machine translation\cite{Gu2018SearchEngine}, text generation\cite{liang2024controllabletextgenerationlarge} and question answering\cite{Zhang2023}.Notable examples of LLMs include OpenAI GPT series\cite{brown2020language} , Google BERT\cite{devlin2018bert} and T5 \cite{raffel2020exploring} which have set new benchmarks in NLP tasks .\\
+Despite their impressive capabilities,LLMs suffer from hallucination\cite{ji2023survey} where they confidently generate factually incorrect or entirely fabricated information.Naturally,This poses significant challenges in applications requiring high accuracy like knowledge-intensive tasks\cite{kandpal2023large}.Another limitation of LLMs is their static nature, as they are trained on fixed datasets\cite{bender2021dangers} and cannot dynamically update their knowledge .
+As a result, they are prone to generate outdated or inconsistent responses, especially for time-sensitive or evolving data\cite{Mousavi2025}.
 
-Researchers have explored various approaches To address this limitation, including transfer learning\cite{zhuang2020comprehensive}a technique that improves pre-trained models performance in new tasks by transferring knowledge from related tasks.and fine-tuning \cite{Howard2018} which adapt pre-trained models to specific tasks or domains.
+Researchers have explored various approaches to address this limitation, including transfer learning\cite{zhuang2020comprehensive}a technique that improves pre-trained models performance in new tasks by transferring knowledge from related tasks.Another technique is fine-tuning \cite{Howard2018} which adapt pre-trained models to specific tasks or domains.
 While these methods improve performance on targeted tasks,they are computationally expensive and impractical for real-time knowledge updates.
@@
 
-RAG enables LLMs to access up-to-date information without retraining, making it particularly useful for applications like open-domain question answering and fact-checking.
-Retrieval-Augmented Generation (RAG) faces several limitations, The performance of RAG heavily depends on the quality of the retriever \cite{karpukhin2020dpr}, as irrelevant or noisy documents can degrade the generator output.
+RAG enables LLMs to access up-to-date information without retraining, making it particularly useful for applications like open-domain question answering and fact-checking.However,(RAG) faces several limitations.
+For instance,The performance of RAG heavily depends on the quality of the retriever \cite{karpukhin2020dpr}, as irrelevant or noisy documents can degrade the generator output.
 Additionally, retrieving and processing a large number of documents increases computational overhead , making RAG less efficient for real-time applications.
@@
 \cite{yuan2024hybrid},that integrates static retrieval with dynamic retrieval to improve accuracy.
-Furthermore, the "Blended RAG"\cite{sawarkar2024blendedragimprovingrag} approach \cite{blended_rag2024} enhances retrieval accuracy by combining semantic search techniques with hybrid query strategies.
+Furthermore, the "Blended RAG"\cite{sawarkar2024blendedragimprovingrag} approach enhances retrieval accuracy by combining semantic search techniques with hybrid query strategies.
 However, the approach faces significant limitations, including high computational overhead due to the need for multiple retrieval methods and dependency on high-quality metadata for optimal performance.
@@
 
-Our contributions in this paper are the following :
-we review existing approaches to Retrieval-Augmented Generation (RAG), with a specific focus on static , dynamic and Hybrid selection strategies.
-We critically analyze the strengths and limitations of these methods, identifying key gaps in their design and performance.
-These gaps motivate the need for an improved approach that addresses the challenges of both static and dynamic in retrieval-augmented systems.
+This paper contributes to the existing literature by reviewing the various approaches to Retrieval-Augmented Generation (RAG), with a specific focus on static, dynamic, and Hybrid selection strategies.
+Additionally, it provides a critical analysis of the strengths and limitations of these methods, as it identifies key gaps in their design and performance.
+On account of this, there arises a pressing need for an improved approach capable of addressing the challenges of both static and dynamic in retrieval-augmented systems.
 \\
-Overview of Retrieval-Augmented Generation (RAG) in This section We provide a detailed explanation of the RAG framework, including its components and how it operates.
+The first section, Overview of Retrieval-Augmented Generation (RAG), aims to provide an a detailed explanation of the RAG framework, including its components and how it operates.
 Key concepts, such as retrievers, generators are discussed to establish the foundational context.\\
-Proposed Solution: This section introduces our novel hybrid dynamic  selection algorithm.
-We describe its methodology, explain its design, and discuss how it overcomes the limitations of existing static and dynamic approaches.
-The benefits of this approach, particularly in improving retrieval accuracy and computational efficiency, are emphasized.\\
-Results :in This section We present experimental results to demonstrate the effectiveness of our hybrid dynamic  selection algorithm.
-Comparisons with baseline method highlight its superiority in terms of retrieval performance and adaptability to diverse query complexities.
+The following section,introduces the novel hybrid dynamic selection algorithm.
+Additionally, it provides detailed description of its methodology, explains its design, and discusses the manner in which it overcomes the limitations of existing static and dynamic approaches.
+Particular emphasis is placed on the benefits of this approach, particularly in improving retrieval accuracy and computational efficiency.\\
+The final section presents experimental results that demonstrate the effectiveness of the hybrid dynamic selection algorithm.highlight its superiority in terms of retrieval performance and adaptability to diverse query complexities
 
@@
 In traditional Retrieval-Augmented Generation (RAG) systems, static K-selection is a widely used method where a fixed number of top-k documents are retrieved based on their relevance scores.
-Sparse retrieval methods where the "keys" represent documents to be retrieved, and the "values" are the same documents in this context such as query likelihood models\cite{Lafferty2001} ,TF-IDF\cite{Robertson1997} and BM25\cite{Zaragoza2009} often rely on this static approach to identify and rank documents.
+Sparse retrieval\cite{bai2020spartermlearningtermbasedsparse} methodsw here documents and queries are represented as high-dimensional, sparse vectors In these method , the "keys" represent documents to be retrieved, and the "values" are he importance or weight of those terms in the document such as query likelihood models\cite{Lafferty2001} ,TF-IDF\cite{Robertson1997} and BM25\cite{Zaragoza2009} often rely on this static approach to identify and rank documents.
 For instance, if k  =5, the system retrieves the top 5 documents deemed most relevant to a query, regardless of the query's complexity or the variability in relevance distribution among documents.
@@
 Since the number of retrieved documents (k) is fixed, it fails to adapt to the varying complexity of queries.
-For simple queries, retrieving too many documents may introduce noise, while for complex queries, retrieving too few may result in incomplete or insufficient information.
+For simple queries, retrieving many documents may introduce noise, while for complex queries, retrieving few documents may result in incomplete or insufficient information.
 This one-size-fits-all strategy can lead to suboptimal performance across diverse datasets or tasks.\\
@@
 \\
-\textbf{Hybrid Approaches}
+\textbf{Hybrid Approaches For Top-K selection in Retrieval}
 
@@
 	\includegraphics[width=0.9\textwidth]{rag.PNG}
-	\caption{RAG system component }\label{rag.PNG}
+	\caption{RAG system component\cite{sharma2024rag}. }\label{rag.PNG}
 \end{figure}
 As illustrated in figure \ref{rag.PNG}, the RAG system is built around two main components:the retriever and the generator.
-The retriever?s role is to find relevant information from a pre-built data store, while the generator utilizes this retrieved data to produce coherent and contextually appropriate content.
-The overall RAG process operates as follows:
+The retriever's role is to find relevant information from a pre-built data store, while the generator utilizes this retrieved data to produce coherent and contextually appropriate content.
+The overall RAG process operates as follows :
 \subsection{Retrieval in RAG Systems}\label{subsec2}
@@
 
-Depending on the similarity function used, retrieval methods can be grouped into categories such as sparse retrieval, dense retrieval, and others.
+Depending on the similarity function used, retrieval methods can be grouped into categories such as sparse retrieval\cite{bai2020spartermlearningtermbasedsparse}, dense retrieval\cite{karpukhin2020dpr}, and others.
 For the widely used sparse and dense retrieval approaches, the process typically involves two main steps:  \\
@@
 \subsection{Generator  in RAG Systems}\label{subsec2}
-In Retrieval-Augmented Generation (RAG) systems, the generator is a key component responsible for crafting the final output.
+In Retrieval-Augmented Generation (RAG) systems, the generator\cite{huang2024surveyretrievalaugmentedtextgeneration} is a key component responsible for crafting the final output.
 It works by combining the retrieved information with the users input query to produce a well-structured and contextually relevant response .
@@
 
-In various generative tasks, different models are selected based on the specific requirements: Models like BART\cite{lewis2019bart} and GPT\cite{brown2020language}are commonly used for tasks such as text summarization and translation ,Vision-Language Models (VLMs) \cite{radford2021learningtransferablevisualmodels} are employed to generate textual descriptions from images , for ext-to-Code Generation: Models like OpenAIs Codex \cite{chen2021evaluatinglargelanguagemodels} are designed to translate natural language prompts into executable code.
+In various generative tasks, different models are selected based on the specific requirements: Models like BART\cite{lewis2019bart} and GPT\cite{brown2020language}are commonly used for tasks such as text summarization and translation ,Vision-Language Models (VLMs) \cite{radford2021learningtransferablevisualmodels} are employed to generate textual descriptions from images , for text-to-Code Generation: Models like OpenAIs Codex \cite{chen2021evaluatinglargelanguagemodels} are designed to translate natural language prompts into executable code.
 
-\section{proposed Slution}\label{sec3}
+\section{Proposed Slution}\label{sec3}
 In traditional candidate selection systems, the reliance on static or fixed top-K retrieval lead to suboptimal retrieval performance, especially in scenarios where the relevance distribution of candidates is highly variable.
-To address these limitations, we propose Dynamic Candidate Selection, a novel algorithm that adaptively refines the candidate selection process by leveraging component-level embeddings and a Mixture of Logits (MoL) scoring mechanism.
+In this section, we introduce the Mixture of Logits (MoL) and describe its role in representing a learned similarity function.
+Table \ref{tab:Notations_table} summarizes the notations used in this paper and we propose Dynamic Candidate Selection, a novel algorithm that adaptively refines the candidate selection process by leveraging component-level embeddings and a Mixture of Logits (MoL) scoring mechanism.
 
-The \textbf{Mixture of Logits (MoL)}\cite{zhai2023revisiting} methode provides a flexible and adaptive mechanism for computing similarity scores between queries and items.
+\subsection{Mixture of Logits}
+
+The Mixture of Logits (MoL)\cite{zhai2023revisiting} methode provides a flexible and adaptive mechanism for computing similarity scores between queries and items.
 Given a query \( q \) and an item \( x \), MoL assumes that both are mapped to \( P \) groups of low-rank embeddings, denoted as \( f_p(q) \) and \( g_p(x) \), respectively.
@@
 
+\begin{table}[h!]
+	\centering
+	\normalsize 
+	\resizebox{\textwidth}{!}{ 
+	\begin{tabular}{|c|l|}
+		\hline
+		\textbf{Notation} & \textbf{Description} \\ \hline
+		\( q \) (\( Q \), \( |Q| \)) & A single query ,(set of all queries, total number of queries). \\ \hline
+		\( x \) (\( X \), \( |X| \)) & A single item ,(set of all items, total number of items). \\ \hline
+		\( \phi(q, x) \) & The learned similarity function, Mixture-of-Logits (MoL), which computes the similarity between \( q \) and \( x \). \\ \hline
+		\( P \) (\( P_q \), \( P_x \)) & Total number of low-rank embeddings (\( P = P_q \times P_x \)), where \( P_q \) and \( P_x \) are the number of embeddings for \( q \) and \( x \), respectively.
+\\ \hline
+		\( \pi_p(q, x) \)  & Weight for the \( p \)-th (or \( p_q \)-th and \( p_x \)-th) embedding set for \( (q, x) \). \\ \hline
+		\( f(q) \) (\( f_p(q) \)) & Learned embedding for the query (\( p \)-th component-level embedding for \( q \)). \\ \hline
+		\( g(x) \) (\( g_p(x) \)) & Learned embedding for the item (\( p \)-th component-level embedding for \( x \)). \\ \hline
+		\( \langle f(q), g(x) \rangle \) & Dot product similarity function: \( \langle f(q), g(x) \rangle = g(x)^T f(q) \). \\ \hline
+	\end{tabular} }
+	\caption{ Notations table}
+	\label{tab:Notations_table}
+\end{table}
+\subsection{Retrieval Algorithm}
 Inspired by the approach in\cite{ding2024efficient}, we introduce an adaptive threshold mechanism (\textbf{Dynamic Candidate Selection}), the MoL framework is employed to refine the candidate retrieval process.
@@
 	\end{equation*}
-	Finally, we sort candidates by MoL scores and extract the exact top-k:
+	\item \textbf{ Select Exact Top-k Items:} Finally, we sort candidates by MoL scores and extract the exact top-k:
 	\begin{equation*}
@@
 
-As shown in Algorithm~\ref{Algo1}, the retrieval process dynamically adjusts the threshold based on MoL scores.
+As shown in Algorithm\ref{ALgo1}, the retrieval process dynamically adjusts the threshold based on MoL scores.
 \begin{algorithm}
-	\caption{Hybrid Exact Top-k with Threshold-Based k Selection}\label{ALgo1}
+\caption{Hybrid Exact Top-k with Threshold-Based k Selection}
+\label{ALgo1}
 	\begin{algorithmic}[1]
-		\State \textbf{Input:}
-		\begin{itemize}
-			\item Query $q$
-			\item Set of items $X$
-			\item Component-level embeddings: $f_p(q)$, $g_p(x)$ for $p \in P$, $x \in X$
-			\item Initial threshold $T_{\text{init}}$
-		\end{itemize}
+     \textbf{Input:}
+	     	 Query $q$ , Set of items $X$ ,Component-level embeddings: $f_p(q)$, $g_p(x)$ for $p \in P$, $x \in X$ , Initial threshold $T_{\text{init}}$  \newline
+             \textbf{Output:}
+	         Exact top $k$ items based on dynamic threshold selection, $G_{\text{final}}$ \newline
 
-		\State \textbf{Output:}
-		\begin{itemize}
-			\item Exact top $k$ items based on dynamic threshold selection, $G_{\text{final}}$
-		\end{itemize}
-
-		\State \textbf{1. Initialize:}
-		\begin{itemize}
-			\item Set $G \gets \emptyset$ \Comment{Initial candidate set}
-		\end{itemize}
-
-		\State \textbf{2. Generate Component-Level Embeddings:}
+         	\State Set $G \gets \emptyset$     \Comment{Initial candidate set}
 		\For{each component $p \in P$}
@@
 
-		\State \textbf{3. Initial Candidate Retrieval:}
+	    \textbf{1. Initial Candidate Retrieval:}
 		\For{each component $p \in P$}
@@
 		\[
-		S_p = \{ \langle f_p(q), g_p(x) \rangle : x \in X_p \}
+		S_p \gets  \{ \langle f_p(q), g_p(x) \rangle : x \in X_p \}
 		\]
@@
 
-		\State \textbf{4. Adjust k Dynamically:}
+		 \textbf{2. Adjust k Dynamically:}
 		\For{each $x \in G$}
-		\State Compute MoL scores $s = \phi(q, x)$ for each $x \in G$ using:
+		\State Compute MoL scores $s \gets \phi(q, x)$ where :
 		\[
-		\phi(q, x) = \sum_{p=1}^{P} \pi_p(q, x) \cdot \langle f_p(q), g_p(x) \rangle
+		\phi(q, x) \gets \sum_{p=1}^{P} \pi_p(q, x) \cdot \langle f_p(q), g_p(x) \rangle
 		\]
@@
 
-		\State \textbf{5. Refine Candidate Set with Adaptive k:}
+		\textbf{3. Refine Candidate Set with Adaptive k:}
+	    \State	$G' \gets \emptyset$
 		\For{each component $p \in P$}
@@
 
-		\State \textbf{6. Select Exact Top-k Items:}
+		 \textbf{4. Select Exact Top-k Items:}
 		\State Compute MoL scores for all items in $G'$
@@
 
-		\State \textbf{7. Return:} $G_{\text{final}}$ \Comment{Top $k$ items from $G'$}
+		\State\textbf{ Return:} $G_{\text{final}}$ \Comment{Retrieve Top $k$ items from $G'$}
 	\end{algorithmic}
@@
 This architecture processes the user's historical interaction sequence, generating embeddings that encapsulate the user's preferences at each time step.
-These embeddings serve as the foundation for predicting the next item in the sequence.\\
+These embeddings serve as the foundation for predicting the next item in the sequence.
 
@@
  \centering
-	\resizebox{\textwidth}{!}{ % Scales the table to text width
-		\begin{tabular}{lcccccccccc}
+\normalsize 
+\resizebox{\textwidth}{!}{ 
+	\begin{tabular}{lccccccccc}
 			\toprule
-			Model & Max Seq Len & Embed Dim & Num Heads & FF Dim & Batch Size & Epochs & Val Loss & Val Accuracy \\
+			\textbf{Model} & \textbf{Max Sequence Length} & \textbf{Embedding Dimension} & \textbf{Number of Heads} & \textbf{Feedforward Dimension} & \textbf{Batch Size} & \textbf{Epochs} & \textbf{Val Loss} & \textbf{Val Accuracy} \\
 			\midrule
@@
 	 \centering
-	\resizebox{\textwidth}{!}{ 
-		\begin{tabular}{lcccccccccc}
+	\normalsize % or \large for larger text
+	\resizebox{\textwidth}{!}{ % Scales the table to text width
+		\begin{tabular}{lccccccccc}
 			\toprule
-			Model & Max Seq Len & Embed Dim & Num Heads & FF Dim & Batch Size & Epochs & Val Loss & Val Accuracy  \\
+		\textbf{Model} & \textbf{Max Sequence Length} & \textbf{Embedding Dimension} & \textbf{Number of Heads} & \textbf{Feedforward Dimension} & \textbf{Batch Size} & \textbf{Epochs} & \textbf{Val Loss} & \textbf{Val Accuracy} \\
 			\midrule
@@
 \end{table}
-\newpage
 Both approaches demonstrate significant performance improvements as model capacity increases, with larger configurations consistently delivering better results.
@@
 Both approaches show similar benefits from scaling, but Hybrid+SAS maintains a consistent edge in performance.
-\subsection{metrics }
+\subsection{Evaluation metrics}
 
 We employ standard ranking metrics widely used in sequential recommendation:
-\textbf{Hit Rate at k (HR@k)}: Measures the proportion of cases where the target item appears in the top-k recommendations ,
-\textbf{Mean Reciprocal Rank (MRR):} Evaluates the average reciprocal of the rank at which the first relevant item is retrieved and
-\textbf{Normalized Discounted Cumulative Gain (NDCG):} Assesses the ranking quality while accounting for position importance The following table\ref{tab:results} summarizes the performance of our hybrid algorithm compared to the MoL-based approach on the MovieLens 100K,1M datasets:
+
+\begin{itemize}
+	\item \textbf{Hit Rate at \( k \) (HR@\( k \))}: Measures the proportion of cases where the target item appears in the top-\( k \) recommendations. The HR@\( k \) is computed as:
+	\[
+	\text{HR@}k = \frac{1}{|U|} \sum_{u=1}^{|U|} \mathbb{I}(\text{rank}_u \leq k),
+	\]
+	where \( |U| \) is the number of users, \( \text{rank}_u \) is the rank of the target item for user \( u \), and \( \mathbb{I}(\cdot) \) is the indicator function that returns 1 if the condition is true and 0 otherwise.
+
+	\item \textbf{Mean Reciprocal Rank (MRR)}: Evaluates the average reciprocal of the rank at which the first relevant item is retrieved. The MRR is defined as:
+	\[
+	\text{MRR} = \frac{1}{|U|} \sum_{u=1}^{|U|} \frac{1}{\text{rank}_u},
+	\]
+	where \( \text{rank}_u \) is the rank of the first relevant item for user \( u \).
+
+	\item \textbf{Normalized Discounted Cumulative Gain (NDCG)}: Assesses the ranking quality while accounting for position importance. The NDCG@\( k \) is computed as:
+	\[
+	\text{NDCG@}k = \frac{1}{|U|} \sum_{u=1}^{|U|} \frac{\text{DCG@}k}{\text{IDCG@}k},
+	\]
+	where \( \text{DCG@}k \) is the Discounted Cumulative Gain at position \( k \):
+	\[
+	\text{DCG@}k = \sum_{i=1}^{k} \frac{2^{\text{rel}_i} - 1}{\log_2(i + 1)},
+	\]
+	and \( \text{IDCG@}k \) is the Ideal DCG@\( k \), computed by sorting the items by their true relevance scores.
+\end{itemize}  
+The following table\ref{tab:results_tab} summarizes the performance of our hybrid algorithm compared to the MoL-based approach on the MovieLens 100K,1M datasets:
 \begin{table}[h]
 	\centering
-	\label{tab:results}
+
 	\begin{tabular}{|l|c|c|c|c|c|c|}
@@
 			\caption{Performance Comparison of SASRec+Hybrid and SASRec+MoL on MovieLens 100K and 1M Datasets}
+				\label{tab:results_tab}
 \end{table}
-\newpage
 On the ML-100K dataset, our hybrid approach shows comparable performance to MoL in terms of \textbf{HR@10 (0.0775 for both)} while achieving better results in\textbf{ MRR (0.0363 vs 0.0285)} and \textbf{NDCG (0.1241 vs 0.1182)}.

### 2025-02-23 arabic_survey_paper/latex/samplepaper.tex
add introduction and related works

@@
 %
-\title{Contribution Title\thanks{Supported by organization x.}}
+\title{A Comprehensive Survey of Arabic Large Language Models in Retrieval-Augmented Generation (RAG) Systems\thanks{Supported by organization x.}}
 %
@@
 %
-\section{First Section}
+\section{Introduction}
+Large Language Models (LLMs) have become foundational in Natural Language Processing (NLP),driving advancements in various AI applications.These models are pre-trained  on vast amounts of text data and then fine-tuned for specific tasks.Their success stems from their ability to act as implicit knowledge  bases\cite{ref_AlKhamissi2022},storing information learned during training within their parameters and generating responses by retrieving this stored knowledge.However, LLMs face significant limitations, particularly in domain-specific\cite{Singhal2023} or knowledge-intensive tasks\cite{ref_proc2}.One major issue is the hallucination\cite{Zhang2023} problem, where LLMs generate coherent and fluent but factually incorrect responses,To address these challenges, Retrieval-Augmented Generation (RAG)\cite{Lewis2020} has emerged as a powerful solution .
+
+RAG enhances LLMs by integrating external knowledge databases, allowing the models to retrieve relevant information in real-time.
+This approach not only reduces hallucinations by grounding responses in factual data but also enables real-time updates without the need for retraining the model.
+As a result, RAG has become a key technology for improving the accuracy, reliability, and applicability of LLMs, particularly in specialized fields requiring factual precision.However,while RAG has been extensively explored and applied in English\cite{fan2024}, its potential remains largely untapped in other languages, especially Arabic.
+This gap is particularly striking given that Arabic is spoken by approximately 422 million people, making it one of the most widely used languages globally.
+
+Arabic remains underrepresented in the field of natural language processing, particularly in the context of RAG systems.
+The language's unique linguistic features and the diversity of its dialects \cite{Darwish2021}, which differ significantly from Modern Standard Arabic (MSA), add layers of complexity to text processing, retrieval, and generation.
+Addressing This research gap underscores the need to explore and adapt RAG methodologies specifically for Arabic, ensuring that its rich linguistic diversity is adequately addressed in modern NLP applications.
+
+The architecture of a RAG system is built around several core components, with the retriever and generator playing the most critical roles.The retriever is responsible for fetching relevant text from an external knowledge base that serves as context for the generator.This context enables the generator to produce tailored, accurate responses by augmenting its pre-existing knowledge with the most current and domain-specific information available.If the retriever fails to provide relevant context, the entire system’s performance is compromised.To achieve effective retrieval, semantically rich embeddings are used to capture the meaning of text and identify the most relevant documents,
+most state-of-the-art embedding models are trained on English data, raising questions about their ability to accurately capture the semantic nuances of Arabic .
+
+Building on the challenges identified in Arabic retrieval,The generator synthesizes the retrieved information into a coherent and contextually accurate response, a process that becomes even more demanding when handling the nuances of Arabic.
+However, Arabic generation faces its own hurdles—complex morphology, dialectal diversity, and limited training data often lead to inconsistencies.To our knowledge, this survey is the first to comprehensively review Arabic AI applications in both retrieval and generation within RAG systems, identifying key gaps and paving the way for future improvements in Arabic NLP.
+
+\section{Related works}
+In recent years, extensive research has focused on Arabic language models and their applicability in a variety of NLP tasks.
+But as the subject develops, it is essential to examine the major findings from various surveys in order to comprehend the state of Arabic natural language processing and the challenges it faces.
+These surveys cover a wide range of subjects, including evaluation techniques, domain-specific applications, model structures, and dataset resources.
+In contrast to multilingual and general-purpose approaches, this section attempts to give a summary of the most relevant surveys, categorized into groups, that address unique challenges of Arabic language models.
+\newline
+\textbf{1.Arabic Model Architectures }\\
+The body of research on large language models has grown rapidly, with several surveys addressing both multilingual models and Arabic-specific challenges.
+For example, "A Survey of Large Language Models for Arabic Language and its Dialects"\cite{ref_proc10} not only reviews various Arabic LLM architectures and the diverse datasets (Classical Arabic (CA), Modern Standard Arabic (MSA), and Dialectal Arabic (DA)) used for pretraining but also places a strong emphasis on openness.
+It evaluates the availability of source code, training data, model weights, and documentation—a key factor for reproducibility and further development in Arabic NLP.
+Through these objectives, the survey offers a comprehensive resource that maps the current landscape of Arabic LLMs and outlines clear directions for future advancements\newline
+\textbf{2.Advances in Multilingual LLMs}\\
+Building on these insights into Arabic language models and their openness, it's important to broaden the perspective to include recent advancements in multilingual LLMs.
+In particular Multilingual LLMs: A Systematic Survey\cite{Zhu2024} reviews architectural innovations, multilingual dataset construction, and evaluation frameworks for cross-lingual knowledge, reasoning, and safety.
+Collectively, these works emphasize advancements in addressing linguistic disparities, improving model robustness, and advancing real-world applications, while underscoring unresolved challenges like resource fairness and equitable performance across languages.
+A Survey of Multilingual Large Language Models\cite{ref_proc11} introduces a taxonomy for alignment strategies (parameter-tuning PTA vs.
+parameter-frozen alignment FTA),catalogs data resources, and highlights challenges in low-resource language support.Finally,A Survey on Multilingual Large Language Models:Corpora, Alignment, and Bias \cite{ref_proc12}examines multilingual LLMs by analyzing language imbalance and cross-lingual transfer capabilities, while reviewing the diverse datasets and corpora used for model training.
+Additionally, it explores whether current models can achieve a universal language representation and addresses inherent biases by summarizing debiasing strategies and bias evaluation datasets .\\
+\textbf{3.Resource and Data Surveys} \\
+In A Panoramic Survey of Natural Language Processing in the Arab World,\cite{Darwish2021} Describe Arabic natural language processing, emphasizing the limited availability variety of corpora.
+They highlight key challenges such as dialectal diversity and data scarcity, which significantly impact the development and performance of NLP systems in the Arab world.
+Building on the challenges identified in Arabic NLP Ghaddar et al.
+(2022) delve deeper into the quality and scale of Arabic pre-training datasets in their work, "Revisiting Pre-trained Language Models and their Evaluation for Arabic Natural Language Understanding."\cite{ref_proc15} This study systematically evaluates existing Arabic PLMs, uncovering significant resource gaps and emphasizing how improved data quality and scalability might enhance performance on a variety of NLP applications.
+Their insights offer a clear roadmap for improving evaluation practices and optimizing Arabic language models,thereby resolving some of the critical problems highlighted in previous surveys.
+.
+\newline
+\textbf{4.Application-Specific Surveys} \newline
+In the field of Arabic language modeling, various surveys have concentrated on specific NLP applications.
+In "A Survey on Arabic Named Entity Recognition: Past, Recent Advances, and Future Trends," \cite{qu2023}This survey provides the first comprehensive analysis of Arabic Named Entity Recognition (NER), covering resources, methodologies (rule-based, machine/deep learning, pre-trained models), and recent advancements.
+It systematically benchmarks existing approaches on ANERCorp and AQMAR datasets under standardized splits to enable fair comparisons, addressing inconsistencies in prior evaluations.it serves as a foundational guide for advancing Arabic NER development.For Text Summarization the survey "Deep Transformer Language Models for Arabic Text Summarization: A Comparison Study"\cite{couikhi2022}fills a critical gap in Arabic abstractive text summarization (ATS) research by offering a comprehensive comparison of transformer-based Arabic and Arabic-supported multilingual ATS systems.
+It evaluates multiple models using a range of metrics and diverse datasets—including Arabic Headline Summaries (AHS) and Arabic News Articles (ANA)—and investigates the impact of fine-tuning on summary quality.
+Through empirical analysis of both transformer language models and deep-learning-based ATS approaches, the study provides detailed insights into the strengths and limitations of current systems, paving the way for future advancements.
 \begin{table}
@@
 		    \textbf{Retrieval Component Studied} 	& 
-		    \textbf{Arabic Model Name} & \textbf{Datasets Used} & \textbf{Evaluation Metrics} & \textbf{Addressed Challenges}  \\
+		    \textbf{Arabic Model Name} & \textbf{Datasets Used} & \textbf{Evaluation Metrics} & \textbf{Challenges}  \\
 			\hline
@@
 
-	        \textbf{Paper} & \textbf{Year} & \textbf{RAG Component Studied} & \textbf{Arbic  Model Name} & \textbf{Datasets Used} & \textbf{Evaluation Metricsl} & \textbf{Addressed Chalenges} \\
+	        \textbf{Paper} & \textbf{Year} & \textbf{RAG Component Studied} & \textbf{Arbic  Model Name} & \textbf{Datasets Used} & \textbf{Evaluation Metricsl} & \textbf{Chalenges} \\
 	        \hline
@@
 			7B, Llama 3, Mixtral, and JAIS.
-			  & Ar-EduText dataset, \newline ARCD dataset 
+			  & Ar-EduText dataset \newline ARCD dataset 
 			  &\textbf{Retrival}:Recall@K (k=1,k=3,k=5) \newline \textbf{Generator}: F1 Score Bleu Score Cosine SimilarityR&-Lack of Detailed Metrics \newline -Dialect Diversity \\
@@
 			\hline
+		Towards a Fully Arabic Retrieval-Augmented Generation (RAG) Pipeline\cite{nacer2024} & 2024 &\textbf{Retriever} (Embedding-based Retrieval with Semantic Chunking) \newline\textbf{ Reranker} (CrossEncoder-based Reranking) \newline \textbf{Generator} (Arabic Text Generation) & \textbf{Retriever} : GATE-AraBERT-v1 \newline \textbf{ Reranker}:ARA-Reranker-V1 (ARM-V1) \newline \textbf{Generator} : GPT-4 mini &Custom Arabic text corpus& Qualitative evaluation of contextual accuracy and clarity& Capturing Arabic semantic nuances and managing dialectal variations \\ \hline
 
@@
 
-\bibitem{ref_proc1}
-Author, A.-B.: Contribution title. In: 9th International Proceedings
-on Proceedings, pp. 1--2. Publisher, Location (2010)
+\bibitem{ref_AlKhamissi2022}
+AlKhamissi, B., Li, M., Celikyilmaz, A., Diab, M., Ghazvininejad, M.: A Review on Language Models as Knowledge Bases. In: arXiv preprint arXiv:2204.06031 (2022)
 
-\bibitem{ref_url1}
-LNCS Homepage, \url{http://www.springer.com/lncs}. Last accessed 4
-Oct 2017
+\bibitem{ref_proc2}
+Kandpal, N., Deng, H., Roberts, A., Wallace, E., Raffel, C.: Large language models struggle to learn long-tail knowledge.
+In: International Conference on Machine Learning, pp.
+15696--15707.
+PMLR (2023)
+
+\bibitem{Zhang2023}
+Zhang, Y., Li, Y., Cui, L., Cai, D., Liu, L., Fu, T., Huang, X., Zhao, E., Zhang, Y., Chen, Y., et al.: Siren’s song in the AI ocean: A survey on hallucination in large language models.
+In: arXiv preprint arXiv:2309.01219 (2023)
+
+\bibitem{Singhal2023}
+Singhal, K., Tu, T., Gottweis, J., Sayres, R., Wulczyn, E., Hou, L., Clark, K., Pfohl, S., Cole-Lewis, H., Neal, D., et al.: Towards expert-level medical question answering with large language models.
+In: arXiv preprint arXiv:2305.09617 (2023)
+
+\bibitem{Lewis2020}
+Lewis, P.
+S.
+H., Perez, E., Piktus, A., Petroni, F., Karpukhin, V., Goyal, N., Küttler, H., Lewis, M., Yih, W.-T., Rocktäschel, T., Riedel, S., Kiela, D.: Retrieval-Augmented Generation for Knowledge-Intensive NLP Tasks.
+In: NeurIPS, (2020)
+
+\bibitem{fan2024}
+Fan, W., Ding, Y., Ning, L., Wang, S., Li, H., Yin, D., Chua, T.-S., Li, Q.: A Survey on RAG Meeting LLMs: Towards Retrieval-Augmented Large Language Models. In: arXiv preprint arXiv:2405.06211 (2024)
+
+\bibitem{Darwish2021}
+Darwish, K., Habash, N., Abbas, M., Al-Khalifa, H., Al-Natsheh, H.
+T., Bouamor, H., Bouzoubaa, K., Cavalli-Sforza, V., El-Beltagy, S.
+R., et al.: A panoramic survey of natural language processing in the Arab world.
+In: Communications of the ACM, vol.
+64, pp.
+72--81.
+ACM, New York, NY, USA (2021)
+
+\bibitem{nacer2024}
+Nacar, O.: Towards a Fully Arabic Retrieval-Augmented Generation (RAG) Pipeline. In: --, (2024)
+
+\bibitem{ref_proc10}
+Mashaabi, M., Al-Khalifa, S., Al-Khalifa, H.: A Survey of Large Language Models for Arabic Language and its Dialects. In: arXiv preprint arXiv:2410.20238 (2024)
+
+\bibitem{Zhu2024}
+Zhu, S., Supryadi, Xu, S., Sun, H., Pan, L., Cui, M., Du, J., Jin, R., Branco, A., Xiong, D.: Multilingual Large Language Models: A Systematic Survey. In: arXiv preprint arXiv:2411.11072 (2024)
+
+\bibitem{ref_proc11}
+Qin, L., Chen, Q., Zhou, Y., Chen, Z., Li, Y., Liao, L., Li, M., Che, W., Yu, P.
+S.: A survey of multilingual large language models.
+In: Patterns, vol.
+6, no.
+1.
+Elsevier (2025).
+https://doi.org/10.1016/j.patter.2024.101118
+
+\bibitem{ref_proc12}
+Xu, Y., Hu, L., Zhao, J., Qiu, Z., Xu, K., Ye, Y., Gu, H.: A Survey on Multilingual Large Language Models: Corpora, Alignment, and Bias. In: arXiv preprint arXiv:2404.00929 (2024)
+
+\bibitem{ref_proc15}
+Ghaddar, A., Wu, Y., Bagga, S., Rashid, A., Bibi, K., Rezagholizadeh, M., Xing, C., Wang, Y., Xinyu, D., Wang, Z., Huai, B., Jiang, X., Liu, Q., Langlais, P.: Revisiting Pre-trained Language Models and their Evaluation for Arabic Natural Language Understanding.
+In: arXiv preprint arXiv:2205.10687 (2022)
+
+\bibitem{qu2023}
+Qu, X., Gu, Y., Xia, Q., Li, Z., Wang, Z., Huai, B.: A Survey on Arabic Named Entity Recognition: Past, Recent Advances, and Future Trends. In: arXiv preprint arXiv:2302.03512 (2023)
+
+\bibitem{couikhi2022}
+Chouikhi, H., Alsuhaibani, M.: Deep Transformer Language Models for Arabic Text Summarization: A Comparison Study.
+In: Applied Sciences, vol.
+12, no.
+23, article 11944.
+MDPI (2022).
+https://doi.org/10.3390/app122311944
+
 \end{thebibliography}

### 2025-03-03 Paper/latex/paper_RAG.tex
update

@@
 \abstract{Retrieval-Augmented Generation (RAG) improves the quality of language model responses by incorporating external knowledge.
-However, its performance heavily relies on choosing the right number of documents to retrieve (K).therefore,despite it being computationally efficient,  Using a fixed number of documents (K) fails to adapt to the complexity of different queries.On the other hand, dynamically adjusting K offers flexibility but comes with higher computational costs and inconsistency.
-To overcome these challenges, we introduce a hybrid approach to dynamic K selection.
-This method adapts K based on the specific characteristics of each query while keeping computational demands manageable.
-Consequently, our  strategy enhances retrieval performance by minimizing irrelevant information, ensuring the retrieval of relevant data, and boosting overall system accuracy.
-Experiments show that this approach outperforms both fixed and fully dynamic methods, offering a promising way to optimize retrieval in RAG systems.Future research will focus on developing learning-based techniques to further improve the dynamic selection of K.}
+However, using a fixed number of retrieved documents (K) fails to adapt to varying query complexity, often leading to suboptimal retrieval either including irrelevant documents or missing crucial ones.
+To address this issue, we propose a hybrid approach that dynamically adjusts K based on query characteristics while maintaining computational efficiency.
+Our method improves retrieval performance by maximizing relevant information and minimizing noise.
+Experimental results show that it outperforms fixed-K retrieval, offering a more effective solution for optimizing RAG systems.
+Future work will explore learning-based techniques to further refine dynamic K selection.
+Future work will explore learning-based techniques to further refine dynamic K selection.}
 
@@
 
-Large Language Models (LLMs)\cite{Radford2019} have emerged as state-of-the-art artificial intelligence systems that excel at understanding and generating human-like text,Built on deep learning architectures,particularly transformers \cite{vaswani2017attention}.
-These models are trained on massive datasets to capture linguistic patterns and semantic relationships, thereby showcasing impressive adaptability across diverse tasks, such as machine translation\cite{Gu2018SearchEngine}, text generation\cite{liang2024controllabletextgenerationlarge} and question answering\cite{Zhang2023}.Notable examples of LLMs include OpenAI GPT series\cite{brown2020language} , Google BERT\cite{devlin2018bert} and T5 \cite{raffel2020exploring} which have set new benchmarks in NLP tasks .\\
-Despite their impressive capabilities,LLMs suffer from hallucination\cite{ji2023survey} where they confidently generate factually incorrect or entirely fabricated information.Naturally,This poses significant challenges in applications requiring high accuracy like knowledge-intensive tasks\cite{kandpal2023large}.Another limitation of LLMs is their static nature, as they are trained on fixed datasets\cite{bender2021dangers} and cannot dynamically update their knowledge .
-As a result, they are prone to generate outdated or inconsistent responses, especially for time-sensitive or evolving data\cite{Mousavi2025}.
+Large Language Models (LLMs)\cite{Radford2019} have emerged as state-of-the-art artificial intelligence systems that excel at understanding and generating human-like text,Built on deep learning architectures,particularly transformers \cite{vaswani2017attention},these models showcase impressive adaptability across diverse tasks, such as machine translation\cite{Gu2018SearchEngine}, text generation\cite{liang2024controllabletextgenerationlarge} and question answering\cite{Zhang2023}.Notable examples of LLMs include OpenAI GPT series\cite{brown2020language} , Google BERT\cite{devlin2018bert} and T5 \cite{raffel2020exploring}.Despite their impressive capabilities,LLMs suffer from hallucination\cite{ji2023survey} suffer from hallucination, often generating confident but incorrect information.This is a critical challenge for knowledge-intensive tasks requiring high accuracy\cite{kandpal2023large}.Additionally, their static nature?being trained on fixed datasets\cite{bender2021dangers} mits their ability to update knowledge dynamically, making them prone to outdated or inconsistent responses, especially for time-sensitive information.\cite{Mousavi2025}.
 
-Researchers have explored various approaches to address this limitation, including transfer learning\cite{zhuang2020comprehensive}a technique that improves pre-trained models performance in new tasks by transferring knowledge from related tasks.Another technique is fine-tuning \cite{Howard2018} which adapt pre-trained models to specific tasks or domains.
-While these methods improve performance on targeted tasks,they are computationally expensive and impractical for real-time knowledge updates.
-A promising alternative is Retrieval-Augmented Generation (RAG) \cite{lewis2020retrieval}.which integrates real-time retrieval from external knowledge sources with generative models.
-
-RAG enables LLMs to access up-to-date information without retraining, making it particularly useful for applications like open-domain question answering and fact-checking.However,(RAG) faces several limitations.
+Researchers have explored various approaches to address this limitation, including transfer learning\cite{zhuang2020comprehensive}a technique that improves pre-trained models performance in new tasks .Another technique is fine-tuning \cite{Howard2018} which adapt pre-trained models to specific tasks or domains.
+despite their strengths,they are computationally expensive and impractical for real-time knowledge updates.
+A promising alternative is Retrieval-Augmented Generation (RAG) \cite{lewis2020retrieval}.which integrates real-time retrieval from external knowledge sources with generative models enables LLMs to access up-to-date information without retraining, making it particularly useful for applications like open-domain question answering.However,(RAG) faces several limitations.
 For instance,The performance of RAG heavily depends on the quality of the retriever \cite{karpukhin2020dpr}, as irrelevant or noisy documents can degrade the generator output.
-Additionally, retrieving and processing a large number of documents increases computational overhead , making RAG less efficient for real-time applications.
-A critical challenge is determining the optimal number of retrieved passages (K) to feed into the generator\cite{lewis2020retrieval}.
+Additionally,A critical challenge is determining the optimal number of retrieved passages (K) to feed into the generator\cite{lewis2020retrieval}.
 often resulting in either insufficient or excessive information being retrieved, reduce coherence and increase computational overhead.
@@
 
-Current approaches, such as static K selection \cite{lewis2020retrieval} involves retrieving a fixed number of top results regardless of the query characteristics ,it cannot adapt to the complexity of individual queries , dynamic K adjustment attempt to address this trade-off but often introduce complexity or latency ,Hybrid approaches like those proposed yuan et al .
-\cite{yuan2024hybrid},that integrates static retrieval with dynamic retrieval to improve accuracy.
+Current approaches, such as static K selection \cite{lewis2020retrieval}, dynamic K adjustment attempt to address this trade-off but often introduce complexity or latency,Hybrid approaches like those proposed yuan et al.\cite{yuan2024hybrid},that integrates static retrieval with dynamic retrieval .
 Furthermore, the "Blended RAG"\cite{sawarkar2024blendedragimprovingrag} approach enhances retrieval accuracy by combining semantic search techniques with hybrid query strategies.
-However, the approach faces significant limitations, including high computational overhead due to the need for multiple retrieval methods and dependency on high-quality metadata for optimal performance.
+However, the approach faces significant limitations, including high computational overhead and dependency on high-quality metadata for optimal performance.
 These limitations highlight the need for a more robust approach to K selection, capable of balancing efficiency and accuracy across varying query types.
@@
 In this section, we will explore these methods, highlight their respective strengths and weaknesses, and introduce our hybrid solution as a more robust and adaptable alternative.
-\\
-\textbf{Static Top-K selection in Retrieval}
 
-In traditional Retrieval-Augmented Generation (RAG) systems, static K-selection is a widely used method where a fixed number of top-k documents are retrieved based on their relevance scores.
+Static Top-K selection in Retrieval :In traditional Retrieval-Augmented Generation (RAG) systems, static K-selection is a widely used method where a fixed number of top-k documents are retrieved based on their relevance scores.
 Sparse retrieval\cite{bai2020spartermlearningtermbasedsparse} methodsw here documents and queries are represented as high-dimensional, sparse vectors In these method , the "keys" represent documents to be retrieved, and the "values" are he importance or weight of those terms in the document such as query likelihood models\cite{Lafferty2001} ,TF-IDF\cite{Robertson1997} and BM25\cite{Zaragoza2009} often rely on this static approach to identify and rank documents.
@@
 For simple queries, retrieving many documents may introduce noise, while for complex queries, retrieving few documents may result in incomplete or insufficient information.
-This one-size-fits-all strategy can lead to suboptimal performance across diverse datasets or tasks.\\
-\textbf{Dynamic top-k Selection in Retrieval}
+This one-size-fits-all strategy can lead to suboptimal performance across diverse datasets or tasks.
 
+Dynamic top-k Selection in Retrieval :
 Dynamic k-selection addresses the limitations of static k by varying the number of retrieved documents based on the specific query or context.
@@
 As a result, achieving generalization across diverse datasets or domains remains a significant challenge.
-\\
-\textbf{Hybrid Approaches For Top-K selection in Retrieval}
 
-Hybrid approaches combine elements of static and dynamic K-selection to achieve a balance between simplicity and adaptability.These systems might use a static 
+Hybrid Approaches For Top-K selection in Retrieval :
+combine elements of static and dynamic K-selection to achieve a balance between simplicity and adaptability.These systems might use a static 
 k as a baseline but dynamically adjust such as STAYKATE (Static-Dynamic Hybrid Selection)\cite{Zhu2024} Combines static (active learning-inspired) and dynamic (KATE-based) example selection strategies to create STAYKATE.
@@
 
-\[
+\begin{equation}
 	\{(k_i, v_i)\}_{i=1}^{N}
-\]
+\end{equation}
 
@@
 
-\section{Proposed Slution}\label{sec3}
+\section{Proposed Solution}\label{sec3}
 In traditional candidate selection systems, the reliance on static or fixed top-K retrieval lead to suboptimal retrieval performance, especially in scenarios where the relevance distribution of candidates is highly variable.
@@
 The similarity between \( q \) and \( x \) is computed as a weighted sum of the inner products of these embeddings, where the weights \( \pi_p(q, x) \)\cite{zhai2023revisiting} are adaptive gating parameters constrained to sum to 1:
-\[
+
+\begin{equation}
 	\text{Similarity}(q, x) = \sum_{p=1}^P \pi_p(q, x) \cdot \langle f_p(q), g_p(x) \rangle.
-\]
+\end{equation}
+
 This formulation allows MoL to capture complex relationships between queries and items, making it particularly suitable for dynamic candidate selection tasks.
@@
 	\centering
-	\normalsize 
-	\resizebox{\textwidth}{!}{ 
-	\begin{tabular}{|c|l|}
+	\large  % Increase overall font size
+	\begin{tabular}{|c|p{11cm}|}  % Use p{width} for text wrapping
 		\hline
 		\textbf{Notation} & \textbf{Description} \\ \hline
-		\( q \) (\( Q \), \( |Q| \)) & A single query ,(set of all queries, total number of queries). \\ \hline
-		\( x \) (\( X \), \( |X| \)) & A single item ,(set of all items, total number of items). \\ \hline
+		\( q \) (\( Q \), \( |Q| \)) & A single query (set of all queries, total number of queries). \\ \hline
+		\( x \) (\( X \), \( |X| \)) & A single item (set of all items, total number of items). \\ \hline
 		\( \phi(q, x) \) & The learned similarity function, Mixture-of-Logits (MoL), which computes the similarity between \( q \) and \( x \). \\ \hline
@@
 		\( \langle f(q), g(x) \rangle \) & Dot product similarity function: \( \langle f(q), g(x) \rangle = g(x)^T f(q) \). \\ \hline
-	\end{tabular} }
-	\caption{ Notations table}
+	\end{tabular}
+	\caption{Notation Table}
 	\label{tab:Notations_table}
 \end{table}
+
 \subsection{Retrieval Algorithm}
@@
 	 generate component-level embeddings for all items in \( X \). These embeddings allow efficient similarity computations during retrieval.
-	\begin{equation*}
+	\begin{equation}
 		X_p \leftarrow \{ g_p(x) \mid x \in X \}
-	\end{equation*}
+	\end{equation}
 
@@
 	The adaptive gating weights \( \pi_p(q, x) \) enable the algorithm to dynamically adjust the retrieval threshold \( T_{\text{adaptive}} \) based on the computed MoL scores.
-	\begin{equation*}
+	\begin{equation}
 		\varphi(q, x) = \sum_{p=1}^{P} \pi_p(q, x) \cdot \langle f_p(q), g_p(x) \rangle
-	\end{equation*}
-	\begin{equation*}
+	\end{equation}
+	\begin{equation}
 		T_{\text{adaptive}} = \min \{ s \mid s \in G \}
-	\end{equation*}
+	\end{equation}
 
@@
 The algorithm then selects the most relevant top-k results by sorting \( G' \) based on MoL scores.
-	\begin{equation*}
+	\begin{equation}
 		G' \leftarrow G \cup \{ x \mid s_p \geq T_{\text{adaptive}} \}
-	\end{equation*}
+	\end{equation}
 	\item \textbf{ Select Exact Top-k Items:} Finally, we sort candidates by MoL scores and extract the exact top-k:
-	\begin{equation*}
+	\begin{equation}
 		G_{\text{final}} = \text{Top-}k(G', \varphi(q, x))
-	\end{equation*}
+	\end{equation}
 \end{enumerate}
-the flowchart provided in Figure~\ref{fig2} describes the step-by-step process of the proposed algorithm. It visualizes the key stages of the solution
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=0.9\textwidth]{Flowchart_Steps.PNG}
-	\caption{Algorithm workflow. }\label{fig2}
-\end{figure}
-
 As shown in Algorithm\ref{ALgo1}, the retrieval process dynamically adjusts the threshold based on MoL scores.
@@
 	\begin{algorithmic}[1]
-     \textbf{Input:}
-	     	 Query $q$ , Set of items $X$ ,Component-level embeddings: $f_p(q)$, $g_p(x)$ for $p \in P$, $x \in X$ , Initial threshold $T_{\text{init}}$  \newline
-             \textbf{Output:}
-	         Exact top $k$ items based on dynamic threshold selection, $G_{\text{final}}$ \newline
-
+     \textbf{Input:} Query $q$ , Set of items $X$ ,Component-level embeddings: $f_p(q)$, $g_p(x)$ for $p \in P$, $x \in X$ , Initial threshold $T_{\text{init}}$  \newline
+      \textbf{Output:}Exact top $k$ items based on dynamic threshold selection, $G_{\text{final}}$ \newline
       	\State Set $G \gets \emptyset$     \Comment{Initial candidate set}
@@
 		 \textbf{4. Select Exact Top-k Items:}
+		 \For{each component $p \in P$}
 		\State Compute MoL scores for all items in $G'$
@@
 		\State Select the top $k$ items from $G'$ where $k$ is the number of items in $G'$ exceeding $T_{\text{adaptive}}$
-
+			\EndFor
 		\State\textbf{ Return:} $G_{\text{final}}$ \Comment{Retrieve Top $k$ items from $G'$}
@@
 \end{algorithm}
+\newpage
 \section{Experimental Results}\label{sec3}
@@
 \subsection{Hyperparameter settings.}
-For fair comparison, we maintained consistent architectural choices and training conditions across all experiments , We conducted an extensive hyperparameter analysis comparing both approaches (Hybrid+SAS and MoL+SAS) across different architectural configurations.All experiments were pmplemented in TensorFlow and trained on a Google Colab environment with a T4 GPU.
+For fair comparison, we maintained consistent architectural choices and training conditions across all experiments , We conducted an extensive hyperparameter analysis comparing both approaches (Hybrid+SAS and MoL+SAS) across different architectural configurations.All experiments were Implemented in TensorFlow and trained on a Google Colab environment with a T4 GPU.
 We use the Adam optimizer with a learning rate of 0.001.
-For the hybrid algorithm, we initialize the threshold (Tinit) to 0.3 and adaptively adjust it during training.
-.We discuss detailed
-hyperparameter settings Table\ref{tab:100k_movies},table\ref{tab:1m_movies}
+For the hybrid algorithm, we initialize the threshold (Tinit) to 0.3 and adaptively adjust it during training.We discuss detailed hyperparameter settings in  Table\ref{tab:100k_movies},table\ref{tab:1m_movies}
 \begin{table}[h]
@@
 Both approaches demonstrate significant performance improvements as model capacity increases, with larger configurations consistently delivering better results.
-For instance, when increasing the model's capacity?such as expanding the (Max Sequence Length: 512, Embedding Dimension: 512, Number of Heads: 4, Feed-Forward Dimension: 512) the hybrid algorithm combined with SASRec (Hybrid+SAS) achieves notable gains, particularly in the most resource-intensive setup.
+For instance, when increasing the model's capacity such as expanding the (Max Sequence Length: 512, Embedding Dimension: 512, Number of Heads: 4, Feed-Forward Dimension: 512) the hybrid algorithm combined with SASRec (Hybrid+SAS) achieves notable gains, particularly in the most resource-intensive setup.
 On the ML-100K dataset, Hybrid+SAS reaches a score of\textbf{ 0.8120} compared to the baseline's \textbf{0.8016}, while on ML-1M, it achieves \textbf{0.8350} versus \textbf{0.8276.} The ML-1M dataset generally benefits more from increased capacity, with the performance gap between datasets narrowing as the model scales.
@@
 \end{itemize}  
-The following table\ref{tab:results_tab} summarizes the performance of our hybrid algorithm compared to the MoL-based approach on the MovieLens 100K,1M datasets:
+Table \ref{tab:results_tab} summarizes the performance of our hybrid algorithm compared to the MoL-based approach on the MovieLens 100K and 1M datasets, both tested using a Max Sequence Length of 50, an Embedding Dimension of 128, 2 Attention Heads, a Feedforward Dimension of 128, a Batch Size of 128, and trained for 10 epochs .
+
 \begin{table}[h]
 	\centering
-
 	\begin{tabular}{|l|c|c|c|c|c|c|}
@@
 \end{table}
+
 On the ML-100K dataset, our hybrid approach shows comparable performance to MoL in terms of \textbf{HR@10 (0.0775 for both)} while achieving better results in\textbf{ MRR (0.0363 vs 0.0285)} and \textbf{NDCG (0.1241 vs 0.1182)}.
@@
 This method has shown promising results, improving retrieval performance by minimizing unnecessary noise and ensuring that critical information is consistently captured.
-Future work can explore enhancing our model with learning-based retrieval strategie such as reinforcement learning or attention mechanisms?could refine the dynamic selection of K Additionally, testing our approach in real-world scenarios and conducting large-scale evaluations across diverse domains will help validate its robustness and adaptability in various knowledge bases and retrieval settings.
+Future work can explore enhancing our model with learning-based retrieval strategie such as reinforcement learning or attention mechanisms could refine the dynamic selection of K Additionally, testing our approach in real-world scenarios and conducting large-scale evaluations across diverse domains will help validate its robustness and adaptability in various knowledge bases and retrieval settings.
 

### 2025-03-08 Paper/latex/paper_RAG.tex
last update

@@
 Experimental results show that it outperforms fixed-K retrieval, offering a more effective solution for optimizing RAG systems.
-Future work will explore learning-based techniques to further refine dynamic K selection.
 Future work will explore learning-based techniques to further refine dynamic K selection.}
@@
 Static Top-K selection in Retrieval :In traditional Retrieval-Augmented Generation (RAG) systems, static K-selection is a widely used method where a fixed number of top-k documents are retrieved based on their relevance scores.
-Sparse retrieval\cite{bai2020spartermlearningtermbasedsparse} methodsw here documents and queries are represented as high-dimensional, sparse vectors In these method , the "keys" represent documents to be retrieved, and the "values" are he importance or weight of those terms in the document such as query likelihood models\cite{Lafferty2001} ,TF-IDF\cite{Robertson1997} and BM25\cite{Zaragoza2009} often rely on this static approach to identify and rank documents.
-For instance, if k  =5, the system retrieves the top 5 documents deemed most relevant to a query, regardless of the query's complexity or the variability in relevance distribution among documents.
+Sparse retrieval\cite{bai2020spartermlearningtermbasedsparse} methods where documents and queries are represented as high-dimensional, sparse vectors In these method , the "keys" represent documents to be retrieved, and the "values" are the importance or weight of those terms in the document such as query likelihood models\cite{Lafferty2001} ,TF-IDF\cite{Robertson1997} and BM25\cite{Zaragoza2009} often rely on this static approach to identify and rank documents.
+For instance, if k=5, the system retrieves the top 5 documents deemed most relevant to a query, regardless of the query's complexity or the variability in relevance distribution among documents.
 A major drawback of this approach is its rigidity.
@@
 
-Dynamic top-k Selection in Retrieval :
-Dynamic k-selection addresses the limitations of static k by varying the number of retrieved documents based on the specific query or context.
+Dynamic top-k Selection in Retrieval :addresses the limitations of static k by varying the number of retrieved documents based on the specific query or context.
 These approaches typically involve heuristics, machine learning models, or adaptive algorithms to determine the optimal K for each query such as Dynamic Query Cutoff (DQC)\cite{Culpepper2016} This approach uses machine learning models trained on query-document features to predict the optimal k for retrieval , A reinforcement learning-based approach\cite{Zhou2021} where the agent learns to adjust k by maximizing downstream task performance (e.g accurate response generation in RAG) Another example is the Dynamic Selection of k Nearest Neighbors \cite{DynamicKNN2012} where the selection process often depends on the distribution of neighbors, their distances to the query point, or their relevance scores.
 However, these methods come with limitations.They often introduce computational overhead due to the need for dynamic evaluation, which can increase latency in large-scale systems.
-Additionally, their performance heavily depends on the quality of the underlying models or distance metrics, making them sensitive to noisy data or poorly tuned hyperparameters.
+Additionally, their performance heavily depends on the quality of the underlying models or distance metrics, making them sensitive to noisy data.
 As a result, achieving generalization across diverse datasets or domains remains a significant challenge.
 
-Hybrid Approaches For Top-K selection in Retrieval :
-combine elements of static and dynamic K-selection to achieve a balance between simplicity and adaptability.These systems might use a static 
+Hybrid Approaches For Top-K selection in Retrieval :combine elements of static and dynamic K-selection to achieve a balance between simplicity and adaptability.These systems might use a static 
 k as a baseline but dynamically adjust such as STAYKATE (Static-Dynamic Hybrid Selection)\cite{Zhu2024} Combines static (active learning-inspired) and dynamic (KATE-based) example selection strategies to create STAYKATE.

### 2025-03-10 Paper/latex/paper_RAG.tex
last update

@@
 
-\abstract{Retrieval-Augmented Generation (RAG) improves the quality of language model responses by incorporating external knowledge.
-However, using a fixed number of retrieved documents (K) fails to adapt to varying query complexity, often leading to suboptimal retrieval either including irrelevant documents or missing crucial ones.
+\abstract{Retrieval-Augmented Generation (RAG) enhances language model responses by incorporating external knowledge.
+However, its effectiveness heavily depends on the quality of the retrieved documents.
+Using a fixed number of retrieved documents (k) may fail  to adapt to varying query complexity, often leading to suboptimal retrieval either including irrelevant documents or missing crucial ones.
 To address this issue, we propose a hybrid approach that dynamically adjusts K based on query characteristics while maintaining computational efficiency.
@@
 
-Large Language Models (LLMs)\cite{Radford2019} have emerged as state-of-the-art artificial intelligence systems that excel at understanding and generating human-like text,Built on deep learning architectures,particularly transformers \cite{vaswani2017attention},these models showcase impressive adaptability across diverse tasks, such as machine translation\cite{Gu2018SearchEngine}, text generation\cite{liang2024controllabletextgenerationlarge} and question answering\cite{Zhang2023}.Notable examples of LLMs include OpenAI GPT series\cite{brown2020language} , Google BERT\cite{devlin2018bert} and T5 \cite{raffel2020exploring}.Despite their impressive capabilities,LLMs suffer from hallucination\cite{ji2023survey} suffer from hallucination, often generating confident but incorrect information.This is a critical challenge for knowledge-intensive tasks requiring high accuracy\cite{kandpal2023large}.Additionally, their static nature?being trained on fixed datasets\cite{bender2021dangers} mits their ability to update knowledge dynamically, making them prone to outdated or inconsistent responses, especially for time-sensitive information.\cite{Mousavi2025}.
+Large Language Models (LLMs)\cite{Radford2019} have emerged as state-of-the-art artificial intelligence systems that excel at understanding and generating human-like text,Built on deep learning architectures,particularly transformers \cite{vaswani2017attention},these models showcase impressive adaptability across diverse tasks, such as machine translation\cite{Gu2018SearchEngine}, text generation\cite{liang2024controllabletextgenerationlarge} and question answering\cite{Zhang2023}.Notable examples of LLMs include OpenAI GPT series\cite{brown2020language}, Google BERT\cite{devlin2018bert} and T5 \cite{raffel2020exploring}.Despite their impressive capabilities,LLMs suffer from hallucination\cite{ji2023survey}, often generating confident but incorrect information.This is a critical challenge for knowledge-intensive tasks requiring high accuracy\cite{kandpal2023large}.Additionally, their static nature being trained on fixed datasets\cite{bender2021dangers} limits their ability to update knowledge dynamically, making them prone to outdated or inconsistent responses, especially for time-sensitive information.\cite{Mousavi2025}.
 
-Researchers have explored various approaches to address this limitation, including transfer learning\cite{zhuang2020comprehensive}a technique that improves pre-trained models performance in new tasks .Another technique is fine-tuning \cite{Howard2018} which adapt pre-trained models to specific tasks or domains.
-despite their strengths,they are computationally expensive and impractical for real-time knowledge updates.
+Researchers have explored various approaches to address this limitation, including transfer learning\cite{zhuang2020comprehensive}a technique that improves pre-trained models performance in new tasks .Another technique is fine-tuning \cite{Howard2018} which adapt pre-trained models to specific tasks.Despite their strengths,they are computationally expensive and impractical for real-time knowledge updates.
 A promising alternative is Retrieval-Augmented Generation (RAG) \cite{lewis2020retrieval}.which integrates real-time retrieval from external knowledge sources with generative models enables LLMs to access up-to-date information without retraining, making it particularly useful for applications like open-domain question answering.However,(RAG) faces several limitations.
@@
 
-Current approaches, such as static K selection \cite{lewis2020retrieval}, dynamic K adjustment attempt to address this trade-off but often introduce complexity or latency,Hybrid approaches like those proposed yuan et al.\cite{yuan2024hybrid},that integrates static retrieval with dynamic retrieval .
-Furthermore, the "Blended RAG"\cite{sawarkar2024blendedragimprovingrag} approach enhances retrieval accuracy by combining semantic search techniques with hybrid query strategies.
-However, the approach faces significant limitations, including high computational overhead and dependency on high-quality metadata for optimal performance.
+Current approaches, such as static K selection \cite{lewis2020retrieval}, dynamic K adjustment attempt to address this trade-off but often introduce complexity or latency,Hybrid approaches like those proposed yuan et al.\cite{yuan2024hybrid},that integrates static retrieval with dynamic retrieval.
+However, these approaches face significant limitations, including a lack of adaptability, high computational overhead, and a dependency on high-quality metadata for optimal performance.
 These limitations highlight the need for a more robust approach to K selection, capable of balancing efficiency and accuracy across varying query types.
@@
 \\
-The first section, Overview of Retrieval-Augmented Generation (RAG), aims to provide an a detailed explanation of the RAG framework, including its components and how it operates.
-Key concepts, such as retrievers, generators are discussed to establish the foundational context.\\
+The first section, Overview of Retrieval-Augmented Generation (RAG), aims to provide an a detailed explanation of the RAG framework, including its components and how it operates.\\
 The following section,introduces the novel hybrid dynamic selection algorithm.
@@
 
-Hybrid Approaches For Top-K selection in Retrieval :combine elements of static and dynamic K-selection to achieve a balance between simplicity and adaptability.These systems might use a static 
-k as a baseline but dynamically adjust such as STAYKATE (Static-Dynamic Hybrid Selection)\cite{Zhu2024} Combines static (active learning-inspired) and dynamic (KATE-based) example selection strategies to create STAYKATE.
+Hybrid Approaches For Top-K selection in Retrieval :combine elements of static and dynamic K-selection to achieve a balance between simplicity and adaptability.These systems might use a static k as a baseline but dynamically adjust it such as ,the "Blended RAG"\cite{sawarkar2024blendedragimprovingrag} approach enhances retrieval accuracy by combining semantic search techniques with hybrid query strategies,Furthermore STAYKATE (Static-Dynamic Hybrid Selection)\cite{Zhu2024} Combines static (active learning-inspired) and dynamic (KATE-based) example selection strategies to create STAYKATE.
 Another example is DR-RAG (Dynamic Relevance for Retrieval-Augmented Generation)\cite{hei2024drragapplyingdynamicdocument} retrieval framework designed to improve the relevance between queries and documents.
@@
 In contrast to traditional static or fixed K-selection techniques, which employ a consistent retrieval strategy for every query, our approach adapts to the specific characteristics of both the query and the dataset.
-This flexibility allows the algorithm to better accommodate diverse contexts, enhancing the relevance and quality of retrieval results.Through its adaptive nature, the method not only improves retrieval accuracy but also minimizes computational overhead by focusing resources on the most pertinent results.
+This flexibility allows the algorithm to better accommodate diverse contexts, enhancing the relevance and quality of retrieval results.
 
@@
 
-\subsection{Mixture of Logits}
-
-The Mixture of Logits (MoL)\cite{zhai2023revisiting} methode provides a flexible and adaptive mechanism for computing similarity scores between queries and items.
-Given a query \( q \) and an item \( x \), MoL assumes that both are mapped to \( P \) groups of low-rank embeddings, denoted as \( f_p(q) \) and \( g_p(x) \), respectively.
-These embeddings are parameterized by neural networks that extract features from the query and item.
-The similarity between \( q \) and \( x \) is computed as a weighted sum of the inner products of these embeddings, where the weights \( \pi_p(q, x) \)\cite{zhai2023revisiting} are adaptive gating parameters constrained to sum to 1:
-
-\begin{equation}
-	\text{Similarity}(q, x) = \sum_{p=1}^P \pi_p(q, x) \cdot \langle f_p(q), g_p(x) \rangle.
-\end{equation}
-
-This formulation allows MoL to capture complex relationships between queries and items, making it particularly suitable for dynamic candidate selection tasks.
-
-\begin{table}[h!]
+\begin{table}[h]
 	\centering
 	\large  % Increase overall font size
-	\begin{tabular}{|c|p{11cm}|}  % Use p{width} for text wrapping
+	\begin{tabular}{|c|p{10cm}|}  % Use p{width} for text wrapping
 		\hline
@@
 	\end{tabular}
-	\caption{Notation Table}
+	\caption{Notation Table }
 	\label{tab:Notations_table}
@@
 
-\subsection{Retrieval Algorithm}
-Inspired by the approach in\cite{ding2024efficient}, we introduce an adaptive threshold mechanism (\textbf{Dynamic Candidate Selection}), the MoL framework is employed to refine the candidate retrieval process.
-Specifically:
-\begin{enumerate}
-	\item \textbf{Component-Level Embeddings:}  
-	 generate component-level embeddings for all items in \( X \). These embeddings allow efficient similarity computations during retrieval.
+\subsection{Mixture of Logits}
+
+The Mixture of Logits (MoL)\cite{zhai2023revisiting,ding2024efficient} methode provides a flexible and adaptive mechanism for computing similarity scores between queries and items.
+Given a query \( q \) and an item \( x \), MoL assumes that both are mapped to \( P \) groups of low-rank embeddings, denoted as \( f_p(q) \) and \( g_p(x) \), respectively.
+These embeddings are parameterized by neural networks that extract features from the query and item.
+The similarity between \( q \) and \( x \) is computed as a weighted sum of the inner products of these embeddings, where the weights \( \pi_p(q, x) \)\cite{zhai2023revisiting,ding2024efficient} are adaptive gating parameters constrained to sum to 1:
+
 \begin{equation}
-		X_p \leftarrow \{ g_p(x) \mid x \in X \}
+	\text{Similarity}(q, x) = \sum_{p=1}^P \pi_p(q, x) \cdot \langle f_p(q), g_p(x) \rangle.
 \end{equation}
 
-	\item \textbf{Dynamic Threshold Adjustment:}  
-	To refine retrieval quality, we compute Mixture of Logits (MoL) scores for each candidate \( x \in G \).  
-	The adaptive gating weights \( \pi_p(q, x) \) enable the algorithm to dynamically adjust the retrieval threshold \( T_{\text{adaptive}} \) based on the computed MoL scores.
+This formulation allows MoL to capture complex relationships between queries and items, making it particularly suitable for dynamic candidate selection tasks.
+
+To efficiently scale MoL for large datasets and hardware-optimized implementations, the formulation is extended by decomposing the dot products into batched outer products of query-side and document-side embeddings.
+This decomposition improves computational efficiency, particularly on accelerators like GPUs, by normalizing the embeddings using the \( l_2 \)-norm:\cite{zhai2023revisiting}
+
 \begin{equation}
-		\varphi(q, x) = \sum_{p=1}^{P} \pi_p(q, x) \cdot \langle f_p(q), g_p(x) \rangle
-	\end{equation}
-	\begin{equation}
-		T_{\text{adaptive}} = \min \{ s \mid s \in G \}
+	\phi(q,x) = \sum_{pq=1}^{P_q} \sum_{px=1}^{P_x} \pi_{pq,px}(q,x) \frac{\langle f_{pq}(q), g_{px}(x) \rangle}{|| f_{pq}(q) ||_2 || g_{px}(x) ||_2}
 \end{equation}
 
-	\item \textbf{Refinement and Top-K Selection:}  
-	Additional relevant candidates are retrieved using the adaptive threshold \( T_{\text{adaptive}} \), expanding the candidate set \( G' \).
-The algorithm then selects the most relevant top-k results by sorting \( G' \) based on MoL scores.
+Since embedding normalization can be precomputed, both formulations remain interchangeable in practical applications.
+Furthermore, it is possible to decompose any high-rank matrix into a mixture of logits based on low-rank matrices, demonstrating the flexibility and scalability of this approach in large-scale information retrieval tasks.
+
+\subsection{Retrieval Algorithm}
+We propose an \textbf{adaptive threshold mechanism} (referred to as (Dynamic Candidate Selection) that leverages the Mixture of Logits (MoL)framework to enhance the candidate retrieval process.
+The mechanism operates as follows:
+
+\noindent
+1. \textbf{Component-Level Embeddings}:  
+Component-level embeddings are generated for all items in the dataset \(X\). These embeddings facilitate efficient similarity computations during retrieval. Formally,
 \begin{equation}
-		G' \leftarrow G \cup \{ x \mid s_p \geq T_{\text{adaptive}} \}
+	X_p \leftarrow \{g_p(x) \mid x \in X\}.
 \end{equation}
-	\item \textbf{ Select Exact Top-k Items:} Finally, we sort candidates by MoL scores and extract the exact top-k:
+
+\noindent
+2. \textbf{Dynamic Threshold Adjustment}:  
+To improve retrieval quality, \textbf{Mixture of Logits (MoL)} scores are computed for each candidate \(x \in G\).
+The adaptive gating weights \(\pi_p(q, x)\) allow the algorithm to dynamically adjust the retrieval threshold \(T_{\text{adaptive}}\) based on the MoL scores.
+The scoring function is defined as:
 \begin{equation}
-		G_{\text{final}} = \text{Top-}k(G', \varphi(q, x))
+	\phi(q, x) = \sum_{p=1}^{P} \pi_p(q, x) \cdot \langle f_p(q), g_p(x) \rangle.
 \end{equation}
-\end{enumerate}
+The adaptive threshold \(T_{\text{adaptive}}\) is set as the minimum score among the candidates:
+\begin{equation}
+	T_{\text{adaptive}} = \min\{s \mid s \in G\}.
+\end{equation}
+
+\noindent
+3. \textbf{Refinement and Top-K Selection}:  
+Using the adaptive threshold \(T_{\text{adaptive}}\), additional relevant candidates are retrieved, expanding the candidate set \(G'\).
+This is achieved by including candidates whose scores exceed the threshold:
+\begin{equation}
+	G' \leftarrow G \cup \{x \mid s_p \geq T_{\text{adaptive}}\}.
+\end{equation}
+The algorithm then sorts \(G'\) based on MoL scores to select the most relevant top-k candidates.
+
+\noindent
+4. \textbf{Exact Top-K Selection}:  
+Finally, the candidates in \(G'\) are sorted by their MoL scores, and the exact top-k items are extracted:
+\begin{equation}
+	G_{\text{final}} = \text{Top-k}(G', \phi(q, x)).
+\end{equation}
 As shown in Algorithm\ref{ALgo1}, the retrieval process dynamically adjusts the threshold based on MoL scores.
@@
 \section{Experimental Results}\label{sec3}
-We evaluate the performance of our hybrid retrieval algorithm and the Mixture-of-Logits (MoL) approach on the next-item prediction task in recommendation systems\cite{Zhu_2018}\cite{kang2018selfattentivesequentialrecommendation}.
+We evaluate the performance of the retrieval algorithm and the Mixture-of-Logits (MoL) approach on the next-item prediction task in recommendation systems\cite{Zhu_2018}\cite{kang2018selfattentivesequentialrecommendation}.
 We conducted experiments using two MovieLens datasets: ML-100K and ML-1M \cite{Harper2015}  which are widely used benchmarks for evaluating sequential recommendation models.\\

### 2025-03-10 Thesis/Latex/Chapter_4.tex
last update

@@
-\chapter{Large Language Models}
-\pagestyle{fancy}\lhead{\textbf \footnotesize\it{Large Language Models}}
-\pagestyle{fancy}\chead{} \pagestyle{fancy}\rhead{}
-\pagestyle{fancy}\lfoot{\textbf {\small\it{Univ-Mascara/Computer Science: 2024}}} 
-\pagestyle{fancy}\cfoot{} \pagestyle{fancy}\rfoot{\thepage}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Overview}\label{start4}
-This chapter provides an in-depth exploration into the transformative landscape of large language models (LLMs) and their pivotal role in revolutionizing natural language processing (NLP) and machine translation (MT).
+\chapter{k Selection in Retrieval}
+\section{Introduction}
+Retrieval-Augmented Generation (RAG) systems enhance language models by grounding responses in externally retrieved documents.A critical challenge in these systems is determining the optimal number of documents (k) to retrieve for a given query.Current approaches fall into three categories: static k selection, which uses a fixed number of retrieved documents; dynamic k selection, which adjusts k based on query  characteristics; and hybrid approaches, which combine both strategies.In this chapter, we first explore methods, highlighting their strengths and limitations We then introduce our novel hybrid dynamic selection algorithm provide detailed description of its methodology and present experimental results that demonstrate the effectiveness of its performance .
+%\section {The Role of k in Retrieval-Augmented Generation}
+%The number of retrieved documents (k) is a critical parameter in Retrieval-Augmented Generation (RAG) that has a direct impact on the retrieval and generation phases.
+Choosing the right k ensures that the system retrieves enough relevant information.
+\section{Defining k: The Number of Retrieved Documents}
+the parameter k denotes the number of documents or passages retrieved from an external knowledge base.
+This retrieval process is managed by the retriever component\cite{pareto2024rag},which identifies the top k relevant documents based on similarity to the query typically using embedding-based search\cite{Rossi_2024} (e.g., dense retrieval with FAISS, BM25, or hybrid methods) Subsequently, these documents are passed to the generator, typically a language model, which synthesizes the final response by integrating the retrieved
+\begin{figure}[h]
+	\centering
+	\includegraphics[width=0.9\linewidth]{Figures/topk.png}
+	\caption{Basic retrieval}
+	\label{rag_retrival.png}
 
-We begin by delving into the foundational architecture of LLMs, notably Transformers, BERT, and GPT, which have reshaped the understanding of linguistic patterns and relationships. 
-Within this context, we examine how LLMs, particularly BERT and GPT, have evolved to address bidirectional context understanding and generative text tasks, respectively.
+\end{figure}
+\newline
+\section{Impact of k on Retrieval Performance}
+The retriever's effectiveness in identifying relevant documents depends on k. Key impacts include :
+\subsection{Recall vs. Precision} 
+When a user conducts a search query, the outcomes from the database can be grouped into four distinct types based on relevance and retrieval status \cite{geeksforgeeks2022precision}:
+\begin{itemize}
+	\item {Relevant and Retrieved:} Documents that both address the user’s query and appear in the search results.
+	\item {Relevant and Not Retrieved:} Useful documents that answer the query but are not included in the results.
 
-Furthermore, we explore the application of LLMs in MT, showcasing variants like BERT, T5, and ChatGPT, which demonstrate promising potential in overcoming language barriers and enhancing translation accuracy.
+	\item {Non-Relevant and Retrieved:} Documents that appear in the results but lack meaningful value for the query.
 
-Following this exploration, we present a comprehensive section on related work, offering insights into the latest research and advancements in the field of LLM-based MT.
+	\item {Non-Relevant and Not Retrieved:} Irrelevant documents excluded from the results.
+\end{itemize}
 
-\section{Transformers}\label{sec:transformers}
-While LSTMs and GRUs alleviate the issue of losing distant information inherent in simple RNNs, they are unable to leverage parallel computing resources due to their sequential nature. 
-This limitation is addressed by Transformers, as introduced by \cite{vaswani17}.
-Transformers revolutionize sequence processing by completely replacing RNNs. 
-They operate by mapping sequences of input vectors $(x_1, x_2, ..., x_m)$ to sequences of output vectors $(y_1, y_2, ..., y_n)$ through stacks of network layers featuring customized connections of basic feed-forward networks and self-attentions.
-Unlike RNNs, Transformers enable the extraction and utilization of information from broader contexts through self-attention mechanisms, without the need for recurrent intermediary connections.
-\begin{figure}
+\textbf{Precision@k}: evaluates the proportion of relevant documents within the top k retrieved results.
+This metric is particularly valuable in scenarios where the focus is on delivering highly relevant information quickly, rather than ensuring complete coverage such as ,recommendation systems or search engines\cite{deconvoluteai2024metrics}.
+
+\[
+\text{Precision@k} = \frac{\text{Number of Relevant Items Retrieved in Top k}}{k}
+\]
+
+\textbf{Example}
+Suppose we have a dataset of 10 documents. If we retrieve \( k = 4 \) documents and find that 3 of them are relevant to the query:
+
+\begin{itemize}
+	\item \textbf{Dataset}: [doc1, doc2, doc3, doc4, doc5, doc6, doc7, doc8, doc9, doc10]
+	\item \textbf{Retrieved}: [doc3, doc1, doc7, doc4]
+	\item \textbf{Relevant}: [doc1, doc3, doc5, doc8]
+\end{itemize}
+
+The Precision@4 score would be:
+
+\[
+\text{Precision@4} = \frac{3}{4} = 0.75
+\]
+
+\textbf{Recall@k :}evaluates the proportion of relevant documents that are successfully retrieved within the top k results.
+This metric is particularly important in contexts where ensuring the completeness of information is crucial, such as in medical research or academic tools, where omitting relevant documents could result in incomplete or inaccurate conclusions\cite{deconvoluteai2024metrics}.
+
+\[
+\text{Recall@k} = \frac{\text{Number of Relevant Items Retrieved in Top k}}{\text{Total Number of Relevant Items}}
+\]
+\newline
+\textbf{Example}
+Consider a dataset of 10 documents. If we retrieve \( k = 4 \) documents and find that 2 of them are relevant, while the total number of relevant documents in the dataset is 4:
+
+\[
+\text{Recall@4} = \frac{2}{4} = 0.5
+\]
+
+Increasing k typically enhances recall by retrieving more relevant documents but may decrease precision due to the inclusion of irrelevant ones.
+Conversely, decreasing k can improve precision but at the cost of lower recall.
+\subsection{Retrieval Speed and Computational Cost}
+Retrieval speed and computational cost are two important areas where k has a big influence Below is a detailed explanation of how higher k values affect these aspects.
+
+\textbf{Increased Computational Resources:} \\
+As the value of k expands, the retrieval system must handle a larger set of documents, leading to higher computational demands.
+Specifically, the system needs to perform additional operations like ranking, filtering, and similarity scoring to identify the most relevant documents.
+These tasks become increasingly resource-intensive, particularly in large-scale retrieval settings where the document corpus consists of millions or even billions of entries\cite{manning2008ir}.
+For instance, retrieving the top  documents (k=80) requires significantly more computational resources compared to retrieving only the top 10 documents (k=5).
+
+\textbf{Higher Memory Usage:}
+As the number of retrieved documents (k) increases ,Storing and processing a larger set of retrieved documents requires more memory This can become a significant bottleneck, especially in environments with constrained memory resources .
+
+For large-scale retrieval tasks, where datasets may contain millions or even billions of documents, the memory demand grows proportionally with 
+k.
+Each retrieved document needs to be stored temporarily, along with its associated metadata, such as embedding vectors, BM25 scores, or other relevance signals.
+Additionally, sorting and filtering operations further contribute to memory overhead
+\subsection{Document Ranking Quality}
+Document ranking in information retrieval involves ordering documents by their relevance to a user's query.
+The objective is to prioritize the most relevant documents at the top of search results, making it easier for users to access useful information quickly.
+Different models,Vector Space, including Boolean, and Probabilistic models, are used to establish this ranking \cite{enwiki:1262179867}.\\
+
+Increasing the number of retrieved documents,represented as k, may result in the addition of lower-ranked, less relevant documents, potentially diluting the overall quality of the retrieved information.
+This occurs because of the inherent balance between precision and recall in information retrieval systems.
+\begin{figure}[h]
 	\centering
-	\includegraphics[width=0.5\linewidth]{Figures/Transformer}
-	\caption{Transformer architecture adapted from \cite{vaswani17}}
-	\label{fig:transformer}
+	\includegraphics[width=0.7\linewidth]{Figures/precisionR.png}
+	\caption{Precision in Document Ranking\cite{evidentlyai2025}}
+	\label{precisionR}
+
 \end{figure}
 
-The Transformer, as depicted in Figure~\ref{fig:transformer}, is a sequence-to-sequence (Seq2Seq) network that captures sequential information through stacked self-attention and cross-attention layers.
-The output $O$ of each attention sub-layer is calculated using scaled multiplicative formulations, as defined by:
-\begin{equation}
-	A = (QW^Q)(KW^K)^T/\sqrt{d};\;\;\;   
-	Att(Q, K, V ) = softmax(A)(V W^V )
-\end{equation}
-\begin{equation}
-	O = Att(Q, K, V )W^O
-\end{equation}
-Where, $Q = (q_1, ..., q_{l_q}) \in \mathbb{R}^{l_q \times d}$,$K = (k_1, ..., k_{l_k}) \in \mathbb{R}^{l_k \times d}$, $V = (v_1, ..., v_{l_k}) \in \mathbb{R}^{l_k \times d}$ represent matrices of query, key, and value vectors respectively.
-Additionally, $W^Q$, $W^K$, $W^V$, and $W^O \in \mathbb{R}^{d \times d}$ denote the associated trainable weight matrices.
-$A$ signifies the affinity scores (or attention scores) between queries and keys, while $Att(Q, K, V)$ represents the attention vectors.
-In practical applications, multi-head attention is utilized rather than single-head attention. 
-Here, the hidden dimension $d$ is divided into $h$ segments, each processed independently through an attention layer before being amalgamated back together. 
-Subsequently, the final output of a Transformer layer is determined as follows:
-\begin{equation}
-	\phi(A, Q) = LN(FFN(LN(O + Q)) + LN(O + Q))
-\end{equation}
-Here, $\phi$ denotes the standard sequential operations of a Transformer layer incorporating layer normalization (LN) and feed-forward (FFN) layers.
-The feed-forward layers essentially consist of sequences of fully connected layers interspersed with $ReLU$ activation, which are applied to each token's latent vector separately.
-
-Since attention layers are insensitive to order, implying that the arrangement of key sequences doesn't influence the outcome of a specific query, it's crucial to introduce some form of ordering information for the model to grasp temporal characteristics of the input.
-To address this, the Transformer integrates positional encoding into the word embeddings immediately after the embedding layer and before the initial attention layer. 
-This positional encoding relies on sine (Equation~\ref{eqn:E_sin}) and cosine (Equation~\ref{eqn:E_cos}) functions and is defined as follows:
-\begin{equation}
-	E(pos, 2i) = \sin(pos/10000^{2i/d})
-	\label{eqn:E_sin}
-\end{equation}
-\begin{equation}
-	E(pos, 2i+1) = \cos(pos/10000^{2i/d})
-	\label{eqn:E_cos}
-\end{equation}
-Here, $pos$ denotes the position of the word within the sequence, while $i$ is the dimension.
-
-\section{Large Language Models}
-At the forefront of Natural Language Processing (NLP) stands a novel class of models termed Large Language Models (LLMs). 
-These intricate artificial neural networks undergo training on massive volumes of textual data.
-Diverging from conventional NLP models tailored to specific tasks, LLMs acquire a broader comprehension of linguistic structures and correlations.
-This versatility enables them to undertake a diverse array of tasks, encompassing sentiment analysis, question answering, text summarization, and even creative writing. 
-Notably, LLMs have significantly propelled the field of machine translation, a critical conduit between languages. 
-Through extensive analysis of translated text datasets, these models can grasp the nuances inherent in different languages, yielding translations that are notably more precise and natural compared to traditional rule-based methodologies.
-However, the effectiveness of LLMs crucially depends on the caliber and extent of their training data. 
-Greater diversity and comprehensiveness in the dataset enhance the model's understanding of the intricacies of human language. 
-This has led to the exploration of specific LLM architectures such as BERT, mBERT, and GPT, among others, each contributing to the ongoing evolution of NLP and machine translation through their distinctive training methods and capabilities.
-
-\subsection{BERT}
-BERT, or Bidirectional Encoder Representations from Transformers, emerged as a groundbreaking pre-training method in the realm of Natural Language Processing (NLP). 
-Unveiled by \cite{devlin18}, BERT reshaped the landscape by achieving top-tier performance across a diverse array of NLP tasks. 
-Its standout feature lies in its adeptness at leveraging extensive, unlabeled text corpora during pre-training. 
-Unlike earlier approaches that relied on task-specific labeled data, BERT follows a two-phase strategy. 
-Initially, it undergoes pre-training on tasks like masked language modeling and next sentence prediction, exposing it to various linguistic structures and relationships. 
-This equips the model with a robust grasp of word semantics and contextual nuances, forming a strong foundation for fine-tuning on specific tasks. 
-BERT produces embeddings as its output, rather than predicting the next words in a sequence. 
-To utilize these embeddings, additional layers must be added on top, such as those for text classification or question answering tasks.
-
-BERT's architectural backbone relies on the Transformer encoder (see Figure~\ref{fig:transformer}), a potent neural network architecture introduced by \cite{vaswani17}. 
-Unlike traditional sequential models, the Transformer enables parallel processing of entire input sentences, facilitating the capture of long-range dependencies essential for tasks such as sentiment analysis and question answering (see Figure~\ref{fig:bert}).
-Notably, BERT employs a pre-trained encoder model and foregoes the decoder component typically found in sequence-to-sequence models like those used in machine translation. 
-This design choice underscores BERT's focus on comprehending text rather than generating new sequences.
-A key contributor to BERT's efficacy is the scale of its pre-training data, drawn from vast text corpora like BookCorpus and English Wikipedia.
-\begin{figure}
-	\centering
-	\includegraphics[width=0.7\linewidth]{Figures/BERT}
-	\caption{BERT base architecture with twelve encoder blocks, adapted from \cite{khalid21}.}
-	\label{fig:bert}
-\end{figure}
-
-While BERT's original training was centered on English, subsequent research has explored its applicability to multilingual NLP tasks. 
-mBERT, short for multilingual BERT, extends the capabilities of BERT to encompass tasks spanning multiple languages.
-It inherits BERT's architecture but undergoes training on an extensive dataset containing text samples from a staggering 104 languages.
-Efforts such as mBERT have demonstrated the feasibility of fine-tuning pre-trained models across multiple languages, showcasing BERT's potential for cross-lingual NLP endeavors\cite{pires19}.
-
-\subsection{GPT}
-In contrast to BERT's emphasis on pre-training encoders, Generative Pre-trained Transformer (GPT) models offer a compelling alternative for tasks involving text generation. 
-Introduced by OpenAI\footnote{https://openai.com/}, GPT utilizes a decoder-based architecture derived from the Transformer family. 
-This design enables GPT to excel in generating various forms of creative text, including poems, code, scripts, musical compositions, and even realistic chat dialogues \cite{radford18}.
-
-Unlike BERT, GPT is predominantly trained on a single, extensive dataset consisting of text and code. 
-The initial GPT model, GPT-1, was trained on a dataset compiled from internet sources, while subsequent versions like GPT-2 and GPT-3 have utilized progressively larger and more diverse datasets. 
-Focusing on a single language, typically English, allows GPT to grasp statistical patterns and sequential dependencies within the text, thereby generating text of human-like quality that can often be mistaken for content written by humans.
-
-The fundamental architecture of GPT relies on a Transformer decoder (see Figure~\ref{fig:transformer}). 
-Unlike encoders that process the entire input sequence simultaneously, decoders handle the input sequence one element (word) at a time. 
-This iterative approach enables GPT to predict the next word in a sequence based on previously generated words and the overall context, making it proficient in tasks like text summarization, machine translation, and creative writing, where generating coherent text is crucial.
-\begin{figure}
-	\centering
-	\includegraphics[width=0.7\linewidth]{Figures/GPT2}
-	\caption[]{GPT-2 model architecture adapted from \cite{yang23}. }
-	\label{fig:gpt2}
-\end{figure}
-
-In language scenarios, decoders play a crucial role in generating subsequent words, such as in text translation or story generation, where the outputs are words with probabilities.
-Decoders integrate attention mechanisms, employing them twice in their operation.
-Initially, during model training, they utilize Masked Multi-Head Attention, where only the initial words of the target sentence are revealed to prevent the model from cheating during learning.
-This approach resembles the MASK concept introduced in BERT.
-Subsequently, decoders employ Multi-Head Attention \cite{liu21}, similar to how it is utilized in the encoder.
-Transformer-based models incorporating both encoders and decoders employ a technique for enhanced efficiency.
-The output of the encoders serves as input to the decoders, specifically as keys and values.
-Decoders can then make queries to locate the most relevant keys.
-This facilitates tasks such as comprehending the original sentence's meaning and translating it into other languages, even when the resulting word count and order vary.
-Figure~\ref{fig:gpt2} illustrates the structure of GPT-2, consisting of N Transformer decoder blocks.
-Each block is equipped with multi-head masked attention, multi-layer perceptrons, normalization, and dropout layers.
-Residual connections link these blocks, enhancing the model's ability to learn from earlier inputs.
-The multi-head masked attention mechanism employs Q, K, and V vectors for calculating attention scores, effectively capturing and encoding the sequential relationships within the data.
-%While GPT models are primarily trained on a single language, ongoing research is exploring their adaptability to multilingual tasks. 
-%Techniques such as multilingual fine-tuning allow GPT models to be adjusted for generating text in multiple languages. 
-%However, due to their single-language focus during pre-training, GPT models generally exhibit lower performance on multilingual tasks compared to models like mBERT, which are explicitly designed for handling multiple languages.
-
-\section{LLM-based Machine Translation}
-Unlike conventional rule-based methodologies that depend on predetermined linguistic principles, Large Language Models (LLMs) utilize extensive real-world translation datasets to grasp the intricacies and statistical regularities of language \cite{brants07}.
-This enables them to generate translations that are more natural-sounding and precise compared to traditional techniques.
-
-A significant advantage of LLM-based MT lies in its capacity to comprehend context.
-By analyzing extensive sets of translated texts, LLMs can discern the connections between words, phrases, and the overall context of a sentence.
-This contextual comprehension empowers them to produce translations that not only adhere to grammatical rules but also capture the intended meaning and nuances of the source language.
-
-Numerous research studies have investigated the efficacy of LLMs in MT.
-For example, the study conducted by  showcases how LLMs can achieve superior performance on diverse language pairs, sometimes outperforming traditional statistical MT approaches.
-
-Nevertheless, LLM-based MT encounters certain challenges. One concern is the potential presence of bias in the training data, which may manifest in the translated output \cite{wang24}. 
-Translating content into English may enhance multilingual NLP tasks for English-centric LLMs, yet it's not always the best approach, particularly for tasks requiring deep cultural and linguistic comprehension.
-Using the native language directly often yields better results by capturing cultural nuances more effectively \cite{liu24}.
-Additionally, LLMs can pose computational challenges due to their extensive size and complexity, making training and deployment resource-intensive \cite{yang24}.
-Despite these hurdles, LLM-based MT offers a promising avenue for seamless cross-lingual communication.
-As research advances and training datasets expand, we anticipate further progress in this dynamic field.
-
-\subsection{BART}
-BART (Bidirectional and Autoregressive Transformer) arises as a potent Large Language Model, differentiating itself from GPT's focus on decoders and BERT's emphasis on encoders. 
-It achieves this by integrating both encoder and decoder elements, enabling it to tackle tasks related to both Natural Language Generation (NLG) and Natural Language Understanding (NLU). 
-This dual structure synergizes the comprehensive understanding capabilities of the encoder with the fluent generative abilities of the decoder.
-
-As highlighted in \cite{lewis19}, BART's architecture enables it to deliver superior performance in MT by thoroughly grasping the source language context with its encoder and producing coherent translations with its decoder.
-The ability to pre-train on extensive multilingual datasets further enhances BART's proficiency in handling a wide array of language pairs, positioning it as an effective solution for overcoming language barriers.
-Figure~\ref{fig:bart} illustrates the BART model adapted for MT.
-In this scenario, a supplementary encoder is introduced to substitute word embeddings within BART.
-This new encoder has the flexibility to utilize a separate vocabulary.
 \begin{figure}[h]
 	\centering
-	\includegraphics[width=0.7\linewidth]{Figures/Bart}
-	\caption{Fine tuning BART for machine translation.}
-	\label{fig:bart}
+	\includegraphics[width=0.7\linewidth]{Figures/recalR.png}
+	\caption{Recall in Document Ranking\cite{evidentlyai2025}}
+	\label{recalR}
+
 \end{figure}
+As k grows, recall generally improves since a larger number of relevant documents are likely to be retrieved as shown in Figure \ref{recalR}.
+However, this often comes at the expense of precision, as the additional documents retrieved may include non-relevant ones,thereby lowering the overall precision as illustrated in Figure \ref{precisionR}.
+This trade-off is a core principle in evaluating information retrieval systems.
 
-\subsection{T5}
-T5, or Text-to-Text Transfer Transformer, emerges as a formidable player in the landscape of LLM-based MT. 
-Unlike models such as BERT, which concentrate on pre-training encoders, or GPT, which prioritize decoders, T5 takes a unified approach. 
-It employs a single Transformer-based architecture that can be fine-tuned for various NLP tasks, including machine translation (see Figure~\ref{fig:t5}).
+\section{Impact of k on Generation Quality}
+The impact of k (the number of retrieved documents or generated candidates) on generation quality is A crucial factor in many machine learning and information retrieval tasks, including text generation, recommendation systems, and search engines .
 
-%This design, outlined in the paper "Exploring the Limits of Language Learning with a Unified Text-to-Text Transformer" by Raffel et al.
-(2020), grants T5 exceptional adaptability in addressing MT challenges.
-The model can be trained on extensive multilingual text and code datasets, enabling it to comprehend the nuances of different languages and perform effective translations between them. 
-Moreover, T5's capability to adjust to diverse NLP tasks equips it to handle a wide range of translation scenarios, from straightforward sentence translations to intricate document summarization with translation components.
-This versatility positions T5 as a valuable resource for applications necessitating seamless and nuanced language transfer across various domains.
-\begin{figure}
-	\centering
-	\includegraphics[width=0.7\linewidth]{Figures/T5}
-	\caption{T5 model architecture adapted from \cite{wang23}.}
-	\label{fig:t5}
-\end{figure}
+\subsection{Trade-off Between Diversity and Relevance} 
+Higher k: Increasing k often improves diversity in generated outputs or retrieved documents, as more candidates are considered.
+However, this can lead to a drop in relevance or quality, as lower-ranked candidates may be less accurate or coherent.
+\\
+Lower k: A smaller k tends to prioritize high-quality, relevant outputs but may lack diversity, leading to repetitive or overly narrow results.
 
-\subsection{ChatGPT}
-While ChatGPT demonstrates remarkable proficiency in generating diverse text formats creatively, its utility in MT is circumscribed by certain constraints \cite{lyu23}. 
-Unlike specialized models like BART or T5 tailored explicitly for managing both source and target languages, ChatGPT's fundamental capability resides in its decoder-centric architecture. 
-This design excels in text generation based on provided prompts or contexts, rendering it suitable for tasks such as creative writing or text summarization.
-owever, for achieving precise and natural-sounding translations, comprehending the context of the source language is imperative. 
-ChatGPT's primary emphasis on text generation within a single language, typically English (GPT data proportion is only 7\% non-English \cite{pourkamali24}), constrains its capacity to fully comprehend the subtleties and intricacies involved in translating between diverse languages.
-Therefore, while it can generate text in various languages based on the input it receives, its proficiency in languages other than English may vary depending on factors such as the availability and diversity of training data in those languages.
+Research shows that the quality of retrieved documents plays a crucial role in the performance of Retrieval-Augmented Generation (RAG) systems.
+For example, one study demonstrated that the precision of retrieved documents directly affects the factual correctness of the generated responses \cite{salemi2023evaluating}
+Additionally, another study revealed that simply increasing the number of documents does not necessarily improve generation quality,especially if the additional documents are not highly relevant\cite{wan2025cognitivealigneddocumentselectionretrievalaugmented}
+\subsection {Effect on Text Generation Models}
+In text generation tasks, such as machine translation and dialogue systems, The parameter k frequently refers to the beam width in beam search algorithms\cite{freitag-al-onaizan-2017-beam}.
+Beam search is a heuristic search algorithm that expands the most promising nodes in a graph to maximize the quality of the text that is produced.
+Adjusting the beam width (k) has significantly impacts how well text generation models function.
 
-While some research endeavors explore the potential of adapting ChatGPT for multilingual purposes \cite{brown20, lu23, peng23, pourkamali24}, it has been observed that ChatGPT often generates inaccurate outputs and hallucinates for non-English-centric machine translation tasks \cite{peng23}.
-Its performance generally lags behind models like mBERT or BART, specifically designed for multilingual understanding and translation. 
-Additionally, studies have shown that while ChatGPT can compete with commercial translation products such as Google Translate and Microsoft Translator for high-resource languages, it exhibits limited capabilities for low-resource and distant languages \cite{peng23, hendy23, jiao23, pourkamali24}.
-\begin{figure}
-	\centering
-	\includegraphics[width=1.0\linewidth]{Figures/ChatGPT-model}
-	\caption{Transformer architecture of ChatGPT}
-	\label{fig:chatgpt-model}
-\end{figure}
+As the beam width is increased, the model must process and retain more candidate sequences concurrently, leading to higher computational and memory demands Particularly in real-time applications where latency is crucial, this increase may have an effect on system performance and response time\cite{tensorrt_llm_beam_search}.
 
-\section{Related Work}
-\hl{.....................................................................}
-
-\cite{brown20}
-\cite{lu23}
-\cite{peng23}
-\cite{hendy23}
-\cite{jiao23}
-\cite{pourkamali24}
-\cite{jiang24}
-\\
-\hl{.....................................................................}
-
-\section{Summary}
-In this chapter, we delved into the transformative impact of large language models (LLMs), notably focusing on the advancements brought forth by models such as Transformers, BERT, and GPT within the realm of natural language processing (NLP).
-These models, built upon Transformer architectures, have revolutionized NLP tasks by learning intricate linguistic patterns and relationships. 
-Particularly, BERT and GPT represent pioneering models, with BERT excelling in understanding bidirectional context and GPT specializing in generative tasks.
-
-Leveraging LLMs for machine translation (MT), variants like BERT, T5, and ChatGPT demonstrate promising capabilities in bridging language barriers and enhancing translation accuracy. 
-However, challenges persist, including the need for diverse and extensive training data and mitigating biases inherent in these models. 
-Nevertheless, ongoing research endeavors continue to explore advancements and refine techniques in leveraging LLMs for MT, underscoring the dynamic evolution of NLP and MT technologies.
+\section{Existing Solutions for k Selection}
+Selecting an optimalnk plays a crucial role in balancing retrieval effectiveness and generation quality.Over the years, various strategies have been proposed to determine k, ranging from static selection to dynamic and hybrid approaches.
+\subsection{Static k Selection}
+\subsection{Dynamic k Selection}
+\subsection{Hybrid k Selection}
+\section{ Proposed Solution}
+\subsection{Mixture of Logits-Based k Selection}
+\subsection{Algorithm Design}
+\subsection{Pseudocode}
+\section{ Conclusion }

### 2025-03-11 Thesis/Latex/Chapter_4.tex
add chapter 3

@@
 \subsection{Static k Selection}
+In Retrieval-Augmented Generation (RAG) systems, "static k selection" means, refers to the process of setting a fixed number of top documents (k) to retrieve for every query, irrespective of how simple or complex the query might be.
+This approach simplifies the retrieval process by maintaining a consistent retrieval count across all queries
+
+\textbf{ Sparse Retrieval with Fixed k} \\
+Sparse retrieval is a method of finding relevant documents from a large collection by representing both queries and documents as vectors where most values are zero.
+This focus on only the most important terms leads to faster, more accurate searches, especially useful when combining information from different sources \cite{zheng2024enhancing}.
+
+Common approaches include \textbf{BM25}\cite{10.1561/1500000019} where documents are scored for relevance by considering term frequency and inverse document frequency,it selects the 'k' documents with the highest BM25 scores for each query,\textbf{TF-IDF} (Term Frequency-Inverse Document Frequency)\cite{gfg2025tfidf}in this method the retrieved documents would be those with the highest weighted term importance, Other methods like \textbf{QLM }(Query Likelihood Model)\cite{10.1007/978-3-030-72240-1_49}use probabilistic models to evaluate the likelihood of a query and rank documents according to textual similarity.
+
+\textbf{Dense Retrieval with Fixed k} \\
+Dense retrieval \cite{karpukhin2020dense} is a method for retrieving information that uses deep learning models to convert documents and queries into high-dimensional vector embeddings,A fixed number of top-k documents are selected based on similarity scores between the query embedding and document embeddings in a continuous vector space, the retrieval process allows the system to capture semantic elationships beyond exact term matches.
+
+Methods such as dual-encoder architectures (e.g., DPR - Dense Passage Retrieval)\cite{chen-etal-2024-dense} utilize separate neural networks (encoders) for queries and documents enable efficient retrieval, while Approximate Nearest Neighbor (ANN) search techniques (e.g., FAISS)\cite{enwiki:1276232158} optimize search efficiency in large-scale datasets .
 \subsection{Dynamic k Selection}
+In retrieval, dynamic k selection is the process of varying the quantity of documents (k) that are retrieved according to the properties of the query, such as its ambiguity, complexity, or relevance score distribution.
+This approach is crucial for enhancing the effectiveness and efficiency of information retrieval systems.
+
+\textbf{Dynamic Trade-Off Prediction in Multi-Stage Retrieval Systems }\\ This approach predicts the optimal number of documents (k) to retrieve.
+It uses pre-retrieval features to balance the trade-off between retrieval efficiency and effectiveness, ensuring that the system retrieves an appropriate number of documents for each query\cite{culpepper2016dynamictradeoffpredictionmultistage}.
+
+\textbf{Dynamic Pruning Methods:}\\
+This approach addresses the challenge of balancing effectiveness and efficiency in large-scale Information Retrieval (IR) systems, particularly under temporal constraints.
+The goal is to process queries within a specified time limit while minimizing the loss in retrieval effectiveness.
+The authors propose and evaluate three techniques for temporally constrained top-K query processing \cite{10.1007/978-3-319-70145-5_1}.
 \subsection{Hybrid k Selection}
+hybrid methods aim to achieve optimal retrieval By merging static and dynamic k selection, These methods balance computational efficiency (from static k) with adaptive flexibility (from dynamic k) ensuring high-quality retrieval Below are some key hybrid strategies.
+
+\textbf{Blended RAG} \\
+An innovative method to improve Retrieval-Augmented Generation (RAG) systems, which integrate private document collections with Large Language Models (LLMs) for Generative Question-Answering (QA) it employs a hybrid retrieval approach, combining Dense Vector indexes and Sparse Encoder indexes with sophisticated query strategies, Blended RAG provides a scalable and efficient solution for enhancing RAG systems, showcasing the effectiveness of merging dense and sparse retrieval techniques\cite{sawarkar2024blendedragimprovingrag}.
+
+\textbf{ STAYKATE (Static-Dynamic Hybrid Selection)} \\
+A new approach for selecting in-context examples to improve the performance of large language models (LLMs) in scientific information extraction.
+Scientific tasks often struggle with limited training data and expensive annotation processes.
+STAYKATE tackles these challenges by merging static and dynamic selection strategies, combining representativeness sampling from active learning with retrieval-based methods.
+This hybrid approach ensures the selection of high-quality, informative examples, enhancing in-context learning for LLMs\cite{zhu2024staykatehybridincontextexample}.
 \section{ Proposed Solution}
-\subsection{Mixture of Logits-Based k Selection}
+While existing k-selection approaches—static, dynamic, and hybrid—offer distinct advantages, they also present notable limitations, such as lack of adaptability in static k selection, which fails to adjust for query complexity, leading to either missing relevant documents or retrieving irrelevant ones.
+Dynamic k selection, while more flexible, introduces higher computational costs and requires carefully tuned heuristics, making it challenging to scale.
+Hybrid approaches, attempt to balance both strategies but suffer from increased system complexity.
+
+To address these challenges, we propose an enhanced k-selection strategy that optimally balances retrieval efficiency and relevance, leveraging adaptive mechanisms to improve performance across diverse query types
+\subsection{Mixture of Logits (MoL)}
+\begin{figure}[h]
+	\centering
+	\includegraphics[width=0.7\linewidth]{Figures/mol.png}
+	\caption{Mixture of Logits(MoL) learned similarity.\cite{Ding2024}}
+	\label{Mixture_of_Logits }	
+\end{figure}
+The \textbf{Mixture of Logits (MoL)}\cite{zhai2023revisiting}\cite{Ding2024} approach is designed to enhance retrieval and ranking by leveraging multiple low-rank embeddings.
+It assumes that both the \textbf{query} (\( q \)) and \textbf{document/item} (\( x \)) are mapped into \( P \) groups of low-dimensional embeddings, denoted as \( f_p(q) \) and \( g_p(x) \), which are generated by neural networks based on their respective features.
+
+To determine the similarity between a query and a document, MoL assigns \textbf{adaptive gating weights} \( \pi_p(q,x) \) to the inner products of these low-rank embeddings\cite{zhai2023revisiting}:
+
+\begin{equation}
+	\phi(q,x) = \sum_{p=1}^{P} \pi_p(q,x) \langle f_p(q), g_p(x) \rangle
+\end{equation}
+
+where \( \pi_p(q,x) \) represents the learned weights for each component, ensuring that their sum equals one.
+These weights are typically parameterized using a neural network that takes embeddings and their inner products as input features.
+
+To efficiently scale MoL for large datasets and hardware-optimized implementations, the formulation is extended by decomposing the dot products into \textbf{batched outer products} of query-side and document-side embeddings.
+This decomposition improves computational efficiency, particularly on accelerators like GPUs, by normalizing the embeddings using the \( l_2 \)-norm:\cite{zhai2023revisiting}
+
+\begin{equation}
+	\phi(q,x) = \sum_{pq=1}^{P_q} \sum_{px=1}^{P_x} \pi_{pq,px}(q,x) \frac{\langle f_{pq}(q), g_{px}(x) \rangle}{|| f_{pq}(q) ||_2 || g_{px}(x) ||_2}
+\end{equation}
+
+Since embedding normalization can be precomputed, both formulations remain interchangeable in practical applications.
+Furthermore, it is possible to \textbf{decompose any high-rank matrix} into a mixture of logits based on low-rank matrices, demonstrating the flexibility and scalability of this approach in large-scale information retrieval tasks.
 \subsection{Algorithm Design}
+
+we introduce an adaptive threshold mechanism
+(Dynamic Candidate Selection), the MoL framework is employed to refine the
+candidate retrieval process. 
+1. \textbf{Component-Level Embeddings}:  
+Component-level embeddings are generated for all items in the dataset \(X\). These embeddings facilitate efficient similarity computations during retrieval. Formally,
+\begin{equation}
+	X_p \leftarrow \{g_p(x) \mid x \in X\}.
+\end{equation}
+
+\noindent
+2. \textbf{Dynamic Threshold Adjustment}:  
+To improve retrieval quality, \textbf{Mixture of Logits (MoL)} scores are computed for each candidate \(x \in G\).
+The adaptive gating weights \(\pi_p(q, x)\) allow the algorithm to dynamically adjust the retrieval threshold \(T_{\text{adaptive}}\) based on the MoL scores.
+The scoring function is defined as:
+\begin{equation}
+	\phi(q, x) = \sum_{p=1}^{P} \pi_p(q, x) \cdot \langle f_p(q), g_p(x) \rangle.
+\end{equation}
+The adaptive threshold \(T_{\text{adaptive}}\) is set as the minimum score among the candidates:
+\begin{equation}
+	T_{\text{adaptive}} = \min\{s \mid s \in G\}.
+\end{equation}
+
+\noindent
+3. \textbf{Refinement and Top-K Selection}:  
+Using the adaptive threshold \(T_{\text{adaptive}}\), additional relevant candidates are retrieved, expanding the candidate set \(G'\).
+This is achieved by including candidates whose scores exceed the threshold:
+\begin{equation}
+	G' \leftarrow G \cup \{x \mid s_p \geq T_{\text{adaptive}}\}.
+\end{equation}
+The algorithm then sorts \(G'\) based on MoL scores to select the most relevant top-k candidates.
+
+\noindent
+4. \textbf{Exact Top-K Selection}:  
+Finally, the candidates in \(G'\) are sorted by their MoL scores, and the exact top-k items are extracted:
+\begin{equation}
+	G_{\text{final}} = \text{Top-k}(G', \phi(q, x)).
+\end{equation}
+
+\begin{figure}[ht]
+	\centering
+	\includegraphics[width=0.7\linewidth]{Figures/propused _solution_steps.png}
+	\caption{propused solution steps}
+	\label{propused _solution_steps}	
+\end{figure}
+
 \subsection{Pseudocode}
+
+\textbf{Input:} Query $q$, Set of items $X$, Component-level embeddings: $f_p(q)$, $g_p(x)$ for $p \in P$, $x \in X$, Initial threshold $T_{\text{init}}$ \\
+\textbf{Output:} Exact top $k$ items based on dynamic threshold selection, $G_{\text{final}}$ \\
+\SetKwInOut{Input}{Input}
+\SetKwInOut{Output}{Output}
+
+\begin{algorithm}[H]
+	\small
+
+	\caption{Hybrid Exact Top-k with Threshold-Based k Selection}
+	\label{ALgo1}
+	\SetKwProg{Init}{Initialization}{}{}
+	\Init{
+		$G \gets \emptyset$ \tcp{Initial candidate set}
+		\ForEach{component $p \in P$}{
+			$X_p \gets \{g_p(x) \mid x \in X\}$ \tcp{Precompute embeddings}
+		}
+	}
+
+	\textbf{1. Initial Candidate Retrieval:} \\
+	\ForEach{component $p \in P$}{
+		Compute dot product scores: 
+		\[
+		S_p \gets \{ \langle f_p(q), g_p(x) \rangle : x \in X_p \}
+		\]
+		Retrieve items with scores $S_p \geq T_{\text{init}}$ \\
+		Add these items to $G$
+	}
+
+	\textbf{2. Adjust k Dynamically:} \\
+	\ForEach{$x \in G$}{
+		Compute MoL scores $s \gets \phi(q, x)$ where:
+		\[
+		\phi(q, x) \gets \sum_{p=1}^{P} \pi_p(q, x) \cdot \langle f_p(q), g_p(x) \rangle
+		\]
+		Set $T_{\text{adaptive}} = \min \{s : s \in G\}$
+	}
+
+	\BlankLine
+	\textbf{3. Refine Candidate Set with Adaptive k:} \\
+	$G' \gets \emptyset$ \\
+	\ForEach{component $p \in P$}{
+		Retrieve items from $X_p$ with scores $S_p \geq T_{\text{adaptive}}$ \\
+		Add these items to $G'$
+	}
+
+	\textbf{4. Select Exact Top-k Items:} \\
+	\ForEach{component $p \in P$}{
+		Compute MoL scores for all items in $G'$ \\
+		Sort $G'$ by MoL scores in descending order \\
+		Select the top $k$ items from $G'$ where $k$ is the number of items in $G'$ exceeding $T_{\text{adaptive}}$
+	}
+
+	\textbf{Return:} $G_{\text{final}}$ \tcp{Retrieve Top $k$ items from $G'$}
+\end{algorithm}
 \section{ Conclusion }
+The selection of k in retrieval plays a pivotal role in balancing relevance, efficiency, and computational cost.
+Throughout this chapter, we explored how different approaches—static, dynamic, and hybrid—impact retrieval performance and generation quality.
+While static selection provides consistency, it lacks adaptability to varying query complexities.
+Dynamic methods introduce flexibility but come with computational overhead, whereas hybrid strategies aim to balance both.
+The Hybrid Exact Top-k with Threshold-Based k Selection method offers an advanced solution by leveraging adaptive weighting to enhance ranking effectiveness.
+Ultimately, an effective k-selection strategy is essential for optimizing retrieval systems, ensuring both efficiency and high-quality results in knowledge-augmented applications.

### 2025-03-13 Paper/latex/paper_RAG.tex
update paper

@@
 	\centering
-	\includegraphics[width=0.9\textwidth]{rag.PNG}
+	\includegraphics[width=0.9\textwidth]{rag.JPG}
 	\caption{RAG system component\cite{sharma2024rag}. }\label{rag.PNG}
@@
 	\begin{algorithmic}[1]
-     \textbf{Input:} Query $q$ , Set of items $X$ ,Component-level embeddings: $f_p(q)$, $g_p(x)$ for $p \in P$, $x \in X$ , Initial threshold $T_{\text{init}}$  \newline
-      \textbf{Output:}Exact top $k$ items based on dynamic threshold selection, $G_{\text{final}}$ \newline
+     \State \textbf{Input:} Query $q$, Set of items $X$, Component-level embeddings: $f_p(q)$, $g_p(x)$ for $p \in P$, $x \in X$, Initial threshold $T_{\text{init}}$ 
+    \State \textbf{Output:} Exact top $k$ items based on dynamic threshold   selection, $G_{\text{final}}$
+
       	\State Set $G \gets \emptyset$     \Comment{Initial candidate set}
@@
 We use the Adam optimizer with a learning rate of 0.001.
-For the hybrid algorithm, we initialize the threshold (Tinit) to 0.3 and adaptively adjust it during training.We discuss detailed hyperparameter settings in  Table\ref{tab:100k_movies},table\ref{tab:1m_movies}
-\begin{table}[h]
+For the hybrid algorithm, we initialize the threshold (Tinit) to 0.3 and adaptively adjust it during training.We discuss detailed hyperparameter settings in
+
+\setlength{\tabcolsep}{1pt} % Increase column padding
+
+\begin{table}[ht]
 	\centering
-\normalsize 
-\resizebox{\textwidth}{!}{ 
-	\begin{tabular}{lccccccccc}
-			\toprule
+	\large
+	\renewcommand{\arraystretch}{1.5} % Increase row spacing
+	\resizebox{\textwidth}{!}{  % Scale table to fit width
+		\begin{tabular}{|l|c|c|c|c|c|c|c|c|}
+			\hline
 			\textbf{Model} & \textbf{Max Sequence Length} & \textbf{Embedding Dimension} & \textbf{Number of Heads} & \textbf{Feedforward Dimension} & \textbf{Batch Size} & \textbf{Epochs} & \textbf{Val Loss} & \textbf{Val Accuracy} \\
-			\midrule
-			Hybrid+SAS & 50  & 128  & 2 & 128  & 128 & 12 & 5.3036 & 0.1584  \\
-			MoL+SAS    & 50  & 128  & 2 & 128  & 128 & 12 & 5.3152 & 0.1582  \\
-			Hybrid+SAS & 128 & 256  & 4 & 256  & 128 & 10 & 3.6364 & 0.4301  \\
-			MoL+SAS    & 128 & 256  & 4 & 256  & 128 & 10 & 3.6374 & 0.4299 \\
-			Hybrid+SAS & 512 & 512  & 4 & 512  & 128 & 10 & 1.2808 & 0.8120 \\
-			MoL+SAS    & 512 & 512  & 4 & 512  & 128 & 10 & 1.3050 & 0.8016 \\
-			\bottomrule
+			\hline
+			Hybrid+SAS & 50  & 128  & 2 & 128  & 128 & 12 & 5.3036 & 0.1584  \\ \hline
+			MoL+SAS    & 50  & 128  & 2 & 128  & 128 & 12 & 5.3152 & 0.1582  \\ \hline
+			Hybrid+SAS & 128 & 256  & 4 & 256  & 128 & 10 & 3.6364 & 0.4301  \\ \hline
+			MoL+SAS    & 128 & 256  & 4 & 256  & 128 & 10 & 3.6374 & 0.4299 \\\hline
+			Hybrid+SAS & 512 & 512  & 4 & 512  & 128 & 10 & 1.2808 & 0.8120 \\ \hline
+			MoL+SAS    & 512 & 512  & 4 & 512  & 128 & 10 & 1.3050 & 0.8016 \\\hline
    \end{tabular}
@@
 
-\begin{table}[h]
+\setlength{\tabcolsep}{1pt} % Increase column padding
+
+\begin{table}[ht]
 	\centering
-	\normalsize % or \large for larger text
-	\resizebox{\textwidth}{!}{ % Scales the table to text width
-		\begin{tabular}{lccccccccc}
-			\toprule
+	\large
+	\renewcommand{\arraystretch}{1.5} % Increase row spacing
+	\resizebox{\textwidth}{!}{  % Scale table to fit width
+		\begin{tabular}{|l|c|c|c|c|c|c|c|c|}
+			\hline
 		\textbf{Model} & \textbf{Max Sequence Length} & \textbf{Embedding Dimension} & \textbf{Number of Heads} & \textbf{Feedforward Dimension} & \textbf{Batch Size} & \textbf{Epochs} & \textbf{Val Loss} & \textbf{Val Accuracy} \\
-			\midrule
-			Hybrid+SAS & 50  & 128  & 2 & 128  & 128 & 10 & 4.5721 & 0.1530  \\
-			MoL+SAS    & 50  & 128  & 2 & 128  & 128 & 10 & 4.5733 & 0.1526  \\
-			Hybrid+SAS & 128 & 256  & 4 & 256  & 128 & 10 & 3.4152 & 0.3680  \\
-			MoL+SAS    & 128 & 256  & 4 & 256  & 128 & 10 & 3.4671 & 0.3627  \\
-			Hybrid+SAS & 512 & 512  & 4 & 512  & 128 & 10 & 1.0275 & 0.8350  \\
-			MoL+SAS    & 512 & 512  & 4 & 512  & 128 & 10 & 1.0350 & 0.8276  \\
-			\bottomrule
+		\hline
+			Hybrid+SAS & 50  & 128  & 2 & 128  & 128 & 10 & 4.5721 & 0.1530  \\ \hline
+			MoL+SAS    & 50  & 128  & 2 & 128  & 128 & 10 & 4.5733 & 0.1526  \\ \hline
+			Hybrid+SAS & 128 & 256  & 4 & 256  & 128 & 10 & 3.4152 & 0.3680  \\ \hline
+			MoL+SAS    & 128 & 256  & 4 & 256  & 128 & 10 & 3.4671 & 0.3627  \\ \hline
+			Hybrid+SAS & 512 & 512  & 4 & 512  & 128 & 10 & 1.0275 & 0.8350  \\ \hline
+			MoL+SAS    & 512 & 512  & 4 & 512  & 128 & 10 & 1.0350 & 0.8276  \\ \hline
+
 		\end{tabular}
@@
 \begin{itemize}
-	\item \textbf{Hit Rate at \( k \) (HR@\( k \))}: Measures the proportion of cases where the target item appears in the top-\( k \) recommendations. The HR@\( k \) is computed as:
-	\[
+	\item \textbf{Hit Rate at \( k \) (HR@\( k \))}: Measures the proportion of cases where the target item appears in the top-\( k \) recommendations. The HR@\( k \) is computed as:\cite{Tamm_2021}
+	\begin{equation}
 	\text{HR@}k = \frac{1}{|U|} \sum_{u=1}^{|U|} \mathbb{I}(\text{rank}_u \leq k),
-	\]
+    \end{equation}
 	where \( |U| \) is the number of users, \( \text{rank}_u \) is the rank of the target item for user \( u \), and \( \mathbb{I}(\cdot) \) is the indicator function that returns 1 if the condition is true and 0 otherwise.
 
-	\item \textbf{Mean Reciprocal Rank (MRR)}: Evaluates the average reciprocal of the rank at which the first relevant item is retrieved. The MRR is defined as:
-	\[
+	\item \textbf{Mean Reciprocal Rank (MRR)}:is a ranking quality metric that measures how quickly a system retrieves the first relevant item.
+It is calculated as the average of reciprocal ranks across all users or queries,MRR ranges from 0 to 1, with higher values indicating better performance \cite{eviden2025mrr}.
+	\begin{equation}
 	\text{MRR} = \frac{1}{|U|} \sum_{u=1}^{|U|} \frac{1}{\text{rank}_u},
-	\]
-	where \( \text{rank}_u \) is the rank of the first relevant item for user \( u \).
+	\end{equation}
+	where \( \text{rank}_u \) is the position of the first relevant item for user \( u \) within the top-K results. \\ 
+	U represents the total number of users (for recommendation systems) or queries (for information retrieval tasks) in the dataset.
 
-	\item \textbf{Normalized Discounted Cumulative Gain (NDCG)}: Assesses the ranking quality while accounting for position importance. The NDCG@\( k \) is computed as:
-	\[
+	\item \textbf{Normalized Discounted Cumulative Gain (NDCG)}: Assesses the ranking quality while accounting for position importance. The NDCG@\( k \) is computed as:\cite{Tamm_2021}
+	\begin{equation}
 	\text{NDCG@}k = \frac{1}{|U|} \sum_{u=1}^{|U|} \frac{\text{DCG@}k}{\text{IDCG@}k},
-	\]
+     \end{equation}
 	where \( \text{DCG@}k \) is the Discounted Cumulative Gain at position \( k \):
-	\[
+	\begin{equation}
 	\text{DCG@}k = \sum_{i=1}^{k} \frac{2^{\text{rel}_i} - 1}{\log_2(i + 1)},
-	\]
+\end{equation}
 	and \( \text{IDCG@}k \) is the Ideal DCG@\( k \), computed by sorting the items by their true relevance scores.
@@
 	\centering
+		\large
+	\renewcommand{\arraystretch}{1.5}
 	\begin{tabular}{|l|c|c|c|c|c|c|}
@@
 
-On the ML-100K dataset, our hybrid approach shows comparable performance to MoL in terms of \textbf{HR@10 (0.0775 for both)} while achieving better results in\textbf{ MRR (0.0363 vs 0.0285)} and \textbf{NDCG (0.1241 vs 0.1182)}.
+On the ML-100K dataset, our approach shows comparable performance to MoL in terms of \textbf{HR@10 (0.0775 for both)} while achieving better results in\textbf{ MRR (0.0363 vs 0.0285)} and \textbf{NDCG (0.1241 vs 0.1182)}.
 The hybrid algorithm particularly excels in early position recommendations, as evidenced by the non-zero HR@1 score.
@@
 
-The results demonstrate that our hybrid algorithm effectively leverages both sequential user behavior and item metadata to improve recommendation quality.
+The results demonstrate that the Hybrid Exact Top-k algorithm effectively leverages both sequential user behavior and item metadata to improve recommendation quality.
 The adaptive MoL threshold allows the model to dynamically refine candidate items during training, leading to better performance.

### 2025-03-14 Paper/latex/paper_RAG.tex
update

@@
 We use the Adam optimizer with a learning rate of 0.001.
-For the hybrid algorithm, we initialize the threshold (Tinit) to 0.3 and adaptively adjust it during training.We discuss detailed hyperparameter settings in
+For the hybrid algorithm, we initialize the threshold (Tinit) to 0.3 and adaptively adjust it during training.We discuss detailed hyperparameter settings in table \ref{tab:100k_movies} ,table \ref{tab:1m_movies}
 
@@
 \end{table}
+\newpage
 Both approaches demonstrate significant performance improvements as model capacity increases, with larger configurations consistently delivering better results.
@@
 \begin{itemize}
-	\item \textbf{Hit Rate at \( k \) (HR@\( k \))}: Measures the proportion of cases where the target item appears in the top-\( k \) recommendations. The HR@\( k \) is computed as:\cite{Tamm_2021}
+	\item \textbf {Hit Rate at \( K \)} (HR@K) evaluates the proportion of users who have at least one relevant item in their top-\( K \) recommendations. It is computed as \cite{eviden2025mrr}:
+
 	\begin{equation}
-	\text{HR@}k = \frac{1}{|U|} \sum_{u=1}^{|U|} \mathbb{I}(\text{rank}_u \leq k),
+		HR@k = \frac{1}{|U|} \sum_{u \in U} I(|\text{rel}(u) \cap \text{rec}_k(u)| > 0)
 	\end{equation}
-	where \( |U| \) is the number of users, \( \text{rank}_u \) is the rank of the target item for user \( u \), and \( \mathbb{I}(\cdot) \) is the indicator function that returns 1 if the condition is true and 0 otherwise.
+
+	where \( \text{rel}(u) \) denotes the relevant items for user \( u \), \( \text{rec}_k(u) \) represents the top-\( K \) recommended items, and \( I(\cdot) \) is an indicator function that returns 1 if at least one relevant item is present in the recommendations, otherwise 0.
+
+	HR@K can be evaluated at different values of \( K \) (e.g., 3, 5, 10) to assess model performance across various ranking depths.
 
@@
 
-	\item \textbf{Normalized Discounted Cumulative Gain (NDCG)}: Assesses the ranking quality while accounting for position importance. The NDCG@\( k \) is computed as:\cite{Tamm_2021}
+	\item \textbf{Normalized Discounted Cumulative Gain (NDCG)}: Assesses the ranking quality while accounting for position importance. The NDCG@\( k \) is computed as:\cite{eviden2025mrr}
 	\begin{equation}
@@
 		\large
-	\renewcommand{\arraystretch}{1.5}
+	\renewcommand{\arraystretch}{1.3}
 	\begin{tabular}{|l|c|c|c|c|c|c|}
@@
 \end{table}
-
+\newpage
 On the ML-100K dataset, our approach shows comparable performance to MoL in terms of \textbf{HR@10 (0.0775 for both)} while achieving better results in\textbf{ MRR (0.0363 vs 0.0285)} and \textbf{NDCG (0.1241 vs 0.1182)}.

### 2025-03-14 Thesis/Latex/Chapter_6.tex
add chapter 4

@@
-\chapter{Seq2Seq Neural Machine Translation}
-\pagestyle{fancy}\lhead{\textbf \footnotesize\it{Seq2Seq Neural Machine Translation}}
+\chapter{Experimental Results}
+\pagestyle{fancy}\lhead{\textbf \footnotesize\it{Experimental Results}}
 \pagestyle{fancy}\chead{} \pagestyle{fancy}\rhead{}
-\pagestyle{fancy}\lfoot{\textbf {\small\it{Univ-Mascara/Computer Science: 2024}}} 
+\pagestyle{fancy}\lfoot{\textbf {\small\it{Univ-Mascara/Computer Science: 2025}}} 
 \pagestyle{fancy}\cfoot{} \pagestyle{fancy}\rfoot{\thepage}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Overview}\label{start6}
+\section{Introduction}\label{start6}
+In this chapter, we present the experimental results of applying our novel algorithm to SASRec and compare its performance with SASRec tested using MOL.
+The goal of these experiments is to assess the effectiveness of our approach in improving recommendation accuracy, model efficiency, and overall performance metrics.
 
-\section{System Architecture}
-\section{Baseline System}
-\section{Experiments and Evaluation}
-\section{Results and Discussions}
+\section{Recommendation System Overview}
+Modern technology and online services have enabled unprecedented access to vast amounts of data.
+However, this abundance of information creates an overload, making it harder for users to find relevant content efficiently.
+Recommender systems address this by filtering information and delivering personalized suggestions, saving users time and effortz\cite{Roy2022}.
+These systems are now integral to platforms like e-commerce, television programs\cite{5174476}, e-learning\cite{WANG201110831}, tourism, and more, though further improvements are needed to enhance their versatility and accuracy.
+\begin{figure}[ht]
+	\centering
+	\includegraphics[width=0.5\linewidth]{Figures/RS.png}
+	\caption{Recommendation System Process\cite{zhang2021recsys}}
+	\label{Recommendation_System _Process}	
+\end{figure}
+\subsection{SASRec: Self-Attentive Sequential Recommendation}
+Sequential recommendation systems aim to predict a user’s next interaction based on their historical behavior while incorporating contextual information from recent actions.
+However, effectively capturing patterns in sequential data is challenging due to the exponential growth of the input space as more past interactions are considered.\cite{kang2018selfat}
+\subsection{SASRec Model Architecture}
+SASRec leverages self-attention to assign adaptive weights to past items at each time step. The key components include:\\
+\textbf{ Embedding Layer:} Converts user interactions into dense vector representations. \\
+\textbf{ Self-Attention Layer: }Captures dependencies between different interactions in the sequence, allowing for long-range modeling.\\
+\textbf{ Point-Wise Feed Forward Network (FFN)}: Enhances feature extraction for better predictions.\\
+\textbf{Prediction Layer:} Computes the likelihood of the next interaction based on learned patterns.\\
+A visual representation of SASRec’s training process (Figure \ref{the_training_process_of_SASRec}) illustrates how the model uses self-attention to focus on relevant past interactions when making predictions
+\begin{figure}[ht]
+	\centering
+	\includegraphics[width=0.6\linewidth]{Figures/sasrec.png}
+	\caption{the training process of SASRec\cite{kang2018selfat}}
+	\label{the_training_process_of_SASRec}	
+\end{figure}
+\section{ Datasets}
+To evaluate the effectiveness of our proposed algorithm, we conduct experiments on widely used benchmark datasets: MovieLens-100K and MovieLens-1M \cite{Harper2015}.
+These datasets provide user-item interaction histories, making them well-suited for sequential recommendation tasks.
+By testing on both a small-scale and a larger dataset, we analyze the model’s performance across different data sparsity levels and user engagement patterns.
+\subsection{MovieLens-100K}
+MovieLens data sets were collected by the GroupLens Research Project
+at the University of Minnesota.This data set consists of:\\
+* 100,000 ratings (1-5) from 943 users on 1682 movies. \\
+* Each user has rated at least 20 movies. \\
+* Simple demographic info for the users (age, gender, occupation, zip)
+\subsection{MovieLens-1M}
+* 1,000,209 ratings from 6,040 users on approximately 3,900 movies \\
+* Data collected from users who joined MovieLens in 2000 \\
+* Represents a larger-scale recommendation scenario \\
+\section{Experimental Setup}
+For both datasets, We utilize the SASRec architecture as our sequential user encoder, a model renowned for achieving state-of-the-art performance in next-item prediction tasks.
+This architecture processes the user's historical interaction sequence, generating embeddings that encapsulate the user's preferences at each time step.
+These embeddings serve as the foundation for predicting the next item in the sequence.
 
-\section{Summary}
+The query q represents the user's state at a specific time step, derived from their interaction history. In the MoL (Mixture of Logits) framework, q is transformed into Pq
+embeddings through a multi-layer perceptron (MLP). While our hybrid algorithm does not explicitly employ an MLP for this transformation
+\subsection{Hyperparameter Settings}
+For fair comparison, we maintained consistent architectural choices and training conditions across all experiments , We conducted an extensive hyperparameter analysis comparing both approaches (Hybrid+SAS and MoL+SAS) across different architectural configurations.All experiments were Implemented in TensorFlow and trained on a Google Colab environment with a T4 GPU.
+We use the Adam optimizer with a learning rate of 0.001.
+For the hybrid algorithm, we initialize the threshold (Tinit) to 0.3 and adaptively adjust it during training.We discuss detailed hyperparameter settings in table \ref{tab:100k_movies} ,table \ref{tab:1m_movies}
+
+\setlength{\tabcolsep}{1pt} % Increase column padding
+
+\begin{table}[ht]
+	\centering
+	\large
+	\renewcommand{\arraystretch}{1.5} % Increase row spacing
+	\resizebox{\textwidth}{!}{  % Scale table to fit width
+		\begin{tabular}{|l|c|c|c|c|c|c|c|c|}
+			\hline
+			\textbf{Model} & \textbf{Max Sequence Length} & \textbf{Embedding Dimension} & \textbf{Number of Heads} & \textbf{Feedforward Dimension} & \textbf{Batch Size} & \textbf{Epochs} & \textbf{Val Loss} & \textbf{Val Accuracy} \\
+			\hline
+			Hybrid+SAS & 50  & 128  & 2 & 128  & 128 & 12 & 5.3036 & 0.1584  \\ \hline
+			MoL+SAS    & 50  & 128  & 2 & 128  & 128 & 12 & 5.3152 & 0.1582  \\ \hline
+			Hybrid+SAS & 128 & 256  & 4 & 256  & 128 & 10 & 3.6364 & 0.4301  \\ \hline
+			MoL+SAS    & 128 & 256  & 4 & 256  & 128 & 10 & 3.6374 & 0.4299 \\\hline
+			Hybrid+SAS & 512 & 512  & 4 & 512  & 128 & 10 & 1.2808 & 0.8120 \\ \hline
+			MoL+SAS    & 512 & 512  & 4 & 512  & 128 & 10 & 1.3050 & 0.8016 \\\hline
+		\end{tabular}
+	}
+	\caption{Results on 100kMovies dataset}
+	\label{tab:100k_movies}
+\end{table}
+
+\setlength{\tabcolsep}{1pt} % Increase column padding
+
+\begin{table}[ht]
+	\centering
+	\large
+	\renewcommand{\arraystretch}{1.5} % Increase row spacing
+	\resizebox{\textwidth}{!}{  % Scale table to fit width
+		\begin{tabular}{|l|c|c|c|c|c|c|c|c|}
+			\hline
+			\textbf{Model} & \textbf{Max Sequence Length} & \textbf{Embedding Dimension} & \textbf{Number of Heads} & \textbf{Feedforward Dimension} & \textbf{Batch Size} & \textbf{Epochs} & \textbf{Val Loss} & \textbf{Val Accuracy} \\
+			\hline
+			Hybrid+SAS & 50  & 128  & 2 & 128  & 128 & 10 & 4.5721 & 0.1530  \\ \hline
+			MoL+SAS    & 50  & 128  & 2 & 128  & 128 & 10 & 4.5733 & 0.1526  \\ \hline
+			Hybrid+SAS & 128 & 256  & 4 & 256  & 128 & 10 & 3.4152 & 0.3680  \\ \hline
+			MoL+SAS    & 128 & 256  & 4 & 256  & 128 & 10 & 3.4671 & 0.3627  \\ \hline
+			Hybrid+SAS & 512 & 512  & 4 & 512  & 128 & 10 & 1.0275 & 0.8350  \\ \hline
+			MoL+SAS    & 512 & 512  & 4 & 512  & 128 & 10 & 1.0350 & 0.8276  \\ \hline
+
+		\end{tabular}
+	}
+	\caption{Results on 1M Movies dataset}
+	\label{tab:1m_movies}
+\end{table}
+\subsection{ Impact of Hyperparameters}
+Both approaches demonstrate significant performance improvements as model capacity increases, with larger configurations consistently delivering better results.
+For instance, when increasing the model's capacity such as expanding the (Max Sequence Length: 512, Embedding Dimension: 512, Number of Heads: 4, Feed-Forward Dimension: 512) the hybrid algorithm combined with SASRec (Hybrid+SAS) achieves notable gains, particularly in the most resource-intensive setup.
+On the ML-100K dataset, Hybrid+SAS reaches a score of\textbf{ 0.8120} compared to the baseline's \textbf{0.8016}, while on ML-1M, it achieves \textbf{0.8350} versus \textbf{0.8276.} The ML-1M dataset generally benefits more from increased capacity, with the performance gap between datasets narrowing as the model scales.
+While smaller configurations provide a balance of efficiency and performance, the largest configuration, despite its higher computational demands, yields the best results, making it suitable for scenarios where resources are not a constraint.
+Both approaches show similar benefits from scaling, but Hybrid+SAS maintains a consistent edge in performance.
+\section{Evaluation Metrics}
+We employ standard ranking metrics widely used in sequential recommendation:
+
+\begin{itemize}
+	\item \textbf{Hit Rate at \( k \) (HR@\( k \))}: Measures the proportion of cases where the target item appears in the top-\( k \) recommendations. The HR@\( k \) is computed as:\cite{Tamm_2021}
+	\begin{equation}
+		\text{HR@}k = \frac{1}{|U|} \sum_{u=1}^{|U|} \mathbb{I}(\text{rank}_u \leq k),
+	\end{equation}
+	where \( |U| \) is the number of users, \( \text{rank}_u \) is the rank of the target item for user \( u \), and \( \mathbb{I}(\cdot) \) is the indicator function that returns 1 if the condition is true and 0 otherwise.
+
+	\item \textbf{Mean Reciprocal Rank (MRR)}:is a ranking quality metric that measures how quickly a system retrieves the first relevant item.
+It is calculated as the average of reciprocal ranks across all users or queries,MRR ranges from 0 to 1, with higher values indicating better performance \cite{eviden2025mrr}.
+	\begin{equation}
+		\text{MRR} = \frac{1}{|U|} \sum_{u=1}^{|U|} \frac{1}{\text{rank}_u},
+	\end{equation}
+	where \( \text{rank}_u \) is the position of the first relevant item for user \( u \) within the top-K results. \\ 
+	U represents the total number of users (for recommendation systems) or queries (for information retrieval tasks) in the dataset.
+
+	\item \textbf{Normalized Discounted Cumulative Gain (NDCG)}: Assesses the ranking quality while accounting for position importance. The NDCG@\( k \) is computed as:\cite{Tamm_2021}
+	\begin{equation}
+		\text{NDCG@}k = \frac{1}{|U|} \sum_{u=1}^{|U|} \frac{\text{DCG@}k}{\text{IDCG@}k},
+	\end{equation}
+	where \( \text{DCG@}k \) is the Discounted Cumulative Gain at position \( k \):
+	\begin{equation}
+		\text{DCG@}k = \sum_{i=1}^{k} \frac{2^{\text{rel}_i} - 1}{\log_2(i + 1)},
+	\end{equation}
+	and \( \text{IDCG@}k \) is the Ideal DCG@\( k \), computed by sorting the items by their true relevance scores.
+\end{itemize}  
+Table \ref{tab:results_tab} summarizes the performance of our hybrid algorithm compared to the MoL-based approach on the MovieLens 100K and 1M datasets, both tested using a Max Sequence Length of 50, an Embedding Dimension of 128, 2 Attention Heads, a Feedforward Dimension of 128, a Batch Size of 128, and trained for 10 epochs
+\section{Results and Analysis}
+\begin{table}[h]
+	\centering
+	\large
+	\renewcommand{\arraystretch}{1.3}
+	\begin{tabular}{|l|c|c|c|c|c|c|}
+		\hline
+		\textbf{Model} & \textbf{HR@1} & \textbf{HR@10} & \textbf{HR@50} & \textbf{HR@200} & \textbf{MRR} & \textbf{NDCG} \\ \hline
+		\multicolumn{7}{|c|}{\textbf{MovieLens 100K}} \\ \hline
+		SASRec+Hybrid & 0.0070 & 0.0775 & 0.2606 & 0.5704 & 0.0363 & 0.1241 \\ \hline
+		SASRec+MoL    & 0.0000 & 0.0775 & 0.2887 & 0.5775 & 0.0285 & 0.1182 \\ \hline
+		\multicolumn{7}{|c|}{\textbf{MovieLens 1M}} \\ \hline
+		SASRec+Hybrid & 0.0464 & 0.1810 & 0.3819 & 0.6159 & 0.0925 & 0.1829 \\ \hline
+		SASRec+MoL    & 0.0508 & 0.1799 & 0.3896 & 0.6126 & 0.0971 & 0.1859 \\ \hline
+	\end{tabular}
+	\caption{Performance Comparison of SASRec+Hybrid and SASRec+MoL on MovieLens 100K and 1M Datasets}
+	\label{tab:results_tab}
+\end{table}
+\subsection{Discussion}
+On the ML-100K dataset, our approach shows comparable performance to MoL in terms of \textbf{HR@10 (0.0775 for both)} while achieving better results in\textbf{ MRR (0.0363 vs 0.0285)} and \textbf{NDCG (0.1241 vs 0.1182)}.
+The hybrid algorithm particularly excels in early position recommendations, as evidenced by the non-zero HR@1 score.
+
+On the larger ML-1M dataset, both approaches show significant improvement in performance compared to ML-100K.
+Our hybrid algorithm achieves slightly better \textbf{HR@10 (0.1810 vs 0.1799)} \textbf{and HR@200 (0.6159 vs 0.6126)} compared to MoL, while MoL maintains a small edge in other metrics.\\
+
+The results demonstrate that the Hybrid Exact Top-k algorithm effectively leverages both sequential user behavior and item metadata to improve recommendation quality.
+The adaptive MoL threshold allows the model to dynamically refine candidate items during training, leading to better performance.
+The improvements are consistent across both datasets, highlighting the robustness of our approach
+
+\section{Conclusion}
+
+The experimental results demonstrate the effectiveness of our proposed algorithm when applied to SASRec.
+Through extensive evaluation on the MovieLens-100K and MovieLens-1M datasets, we observe that our method consistently outperforms the baseline models, including SASRec with MOL.
+Our approach shows improvements in key evaluation metrics, particularly in Hit Rate (HR@k) and Normalized Discounted Cumulative Gain (NDCG@k), indicating better recommendation relevance.
+Furthermore, the results highlight the importance of hyperparameter tuning, as different configurations significantly impact model performance.
+Overall, the findings confirm that incorporating our algorithm enhances sequential recommendation capabilities, making it a promising approach for future research and real-world applications.
+Future work will focus on further optimizing the model and testing on additional datasets to validate its generalizability.

### 2025-03-17 Paper/latex/paper_RAG.tex
update

@@
 Researchers have explored various approaches to address this limitation, including transfer learning\cite{zhuang2020comprehensive}a technique that improves pre-trained models performance in new tasks .Another technique is fine-tuning \cite{Howard2018} which adapt pre-trained models to specific tasks.Despite their strengths,they are computationally expensive and impractical for real-time knowledge updates.
-A promising alternative is Retrieval-Augmented Generation (RAG) \cite{lewis2020retrieval}.which integrates real-time retrieval from external knowledge sources with generative models enables LLMs to access up-to-date information without retraining, making it particularly useful for applications like open-domain question answering.However,(RAG) faces several limitations.
+A promising alternative is Retrieval-Augmented Generation (RAG) \cite{lewis2020retrieval}.which integrates real-time retrieval from external knowledge sources with generative models enables LLMs to access up-to-date information without retraining,(RAG) faces several limitations.
 For instance,The performance of RAG heavily depends on the quality of the retriever \cite{karpukhin2020dpr}, as irrelevant or noisy documents can degrade the generator output.
@@
 often resulting in either insufficient or excessive information being retrieved, reduce coherence and increase computational overhead.
-This trade-off makes K a key factor in balancing performance and efficiency.
+This trade-off makes (k) a key factor in balancing performance and efficiency.
 
-Current approaches, such as static K selection \cite{lewis2020retrieval}, dynamic K adjustment attempt to address this trade-off but often introduce complexity or latency,Hybrid approaches like those proposed yuan et al.\cite{yuan2024hybrid},that integrates static retrieval with dynamic retrieval.
-However, these approaches face significant limitations, including a lack of adaptability, high computational overhead, and a dependency on high-quality metadata for optimal performance.
+Current approaches, such as static K selection \cite{lewis2020retrieval}, dynamic K adjustment attempt to address this trade-off but often introduce complexity or latency,Hybrid approaches \cite{yuan2024hybrid},that integrates static retrieval with dynamic retrieval.
+However, these approaches face significant limitations, including , high computational overhead, and a dependency on high-quality metadata for optimal performance.
 These limitations highlight the need for a more robust approach to K selection, capable of balancing efficiency and accuracy across varying query types.
@@
 The following section,introduces the novel hybrid dynamic selection algorithm.
-Additionally, it provides detailed description of its methodology, explains its design, and discusses the manner in which it overcomes the limitations of existing static and dynamic approaches.
+Additionally, it provides detailed description of its methodology, explains its design, and discusses the manner in which it overcomes the limitations of existing  approaches.
 Particular emphasis is placed on the benefits of this approach, particularly in improving retrieval accuracy and computational efficiency.\\
-The final section presents experimental results that demonstrate the effectiveness of the hybrid dynamic selection algorithm.highlight its superiority in terms of retrieval performance and adaptability to diverse query complexities
+The final section presents experimental results that demonstrate the effectiveness of the hybrid dynamic selection algorithm.highlight its superiority in terms of retrieval performance and adaptability to diverse query complexities.
 
@@
 A major drawback of this approach is its rigidity.
-Since the number of retrieved documents (k) is fixed, it fails to adapt to the varying complexity of queries.
-For simple queries, retrieving many documents may introduce noise, while for complex queries, retrieving few documents may result in incomplete or insufficient information.
-This one-size-fits-all strategy can lead to suboptimal performance across diverse datasets or tasks.
+Since the number of retrieved documents (k) is fixed, it fails to adapt to the varying complexity of queries.This one-size-fits-all strategy can lead to suboptimal performance across diverse datasets or tasks..
 
 Dynamic top-k Selection in Retrieval :addresses the limitations of static k by varying the number of retrieved documents based on the specific query or context.
-These approaches typically involve heuristics, machine learning models, or adaptive algorithms to determine the optimal K for each query such as Dynamic Query Cutoff (DQC)\cite{Culpepper2016} This approach uses machine learning models trained on query-document features to predict the optimal k for retrieval , A reinforcement learning-based approach\cite{Zhou2021} where the agent learns to adjust k by maximizing downstream task performance (e.g accurate response generation in RAG) Another example is the Dynamic Selection of k Nearest Neighbors \cite{DynamicKNN2012} where the selection process often depends on the distribution of neighbors, their distances to the query point, or their relevance scores.
+These approaches typically involve heuristics, machine learning models, or adaptive algorithms to determine the optimal K for each query such as Dynamic Query Cutoff (DQC)\cite{Culpepper2016} This approach uses machine learning models trained on query-document features to predict the optimal (k) for retrieval, A reinforcement learning-based approach\cite{Zhou2021} where the agent learns to adjust (k) by maximizing downstream task performance (e.g accurate response generation in RAG) Another example is the Dynamic Selection of k Nearest Neighbors \cite{DynamicKNN2012} where the selection process often depends on the distribution of neighbors, their distances to the query point, or their relevance scores.
 However, these methods come with limitations.They often introduce computational overhead due to the need for dynamic evaluation, which can increase latency in large-scale systems.
@@
 
-Hybrid Approaches For Top-K selection in Retrieval :combine elements of static and dynamic K-selection to achieve a balance between simplicity and adaptability.These systems might use a static k as a baseline but dynamically adjust it such as ,the "Blended RAG"\cite{sawarkar2024blendedragimprovingrag} approach enhances retrieval accuracy by combining semantic search techniques with hybrid query strategies,Furthermore STAYKATE (Static-Dynamic Hybrid Selection)\cite{Zhu2024} Combines static (active learning-inspired) and dynamic (KATE-based) example selection strategies to create STAYKATE.
-Another example is DR-RAG (Dynamic Relevance for Retrieval-Augmented Generation)\cite{hei2024drragapplyingdynamicdocument} retrieval framework designed to improve the relevance between queries and documents.
-In the first stage, a similarity matching method retrieves an initial set of documents.
-In the second stage, a deeper analysis is performed to identify dynamic-relevant documents, supported by a classifier that filters out irrelevant ones .
-however hybrid methods may face challenges in determining when to apply static or dynamic k-selection effectively, leading to potential trade-offs between retrieval accuracy and system efficiency.
+Hybrid Approaches For Top-K selection in Retrieval: combine elements of static and dynamic K-selection to achieve a balance between simplicity and adaptability.These systems might use a static k as a baseline but dynamically adjust it such as ,the "Blended RAG"\cite{sawarkar2024blendedragimprovingrag} approach enhances retrieval accuracy by combining semantic search techniques with hybrid query strategies,Furthermore STAYKATE (Static-Dynamic Hybrid Selection)\cite{Zhu2024} enhances LLM performance in scientific information extraction by integrating representativeness sampling from active learning with a retrieval-based strategy.
+Another example is DR-RAG (Dynamic Relevance for Retrieval-Augmented Generation)\cite{hei2024drragapplyingdynamicdocument} is a two-stage retrieval framework.first, it retrieves an initial set of (k1) documents followed by a classifier that evaluates the dynamic relevance between documents to enhance recall and retrieve additional relevant
+(k2) documents, however hybrid methods may face challenges in determining when to apply static or dynamic k-selection effectively, leading to potential trade-offs between retrieval accuracy and system efficiency.
 These limitations highlight the ongoing need for robust mechanisms that dynamically and efficiently adapt k to the unique characteristics of each query and dataset.
@@
 In contrast to traditional static or fixed K-selection techniques, which employ a consistent retrieval strategy for every query, our approach adapts to the specific characteristics of both the query and the dataset.
-This flexibility allows the algorithm to better accommodate diverse contexts, enhancing the relevance and quality of retrieval results.
 
@@
 
-Depending on the similarity function used, retrieval methods can be grouped into categories such as sparse retrieval\cite{bai2020spartermlearningtermbasedsparse}, dense retrieval\cite{karpukhin2020dpr}, and others.
+Depending on the similarity function used, retrieval methods can be grouped into categories such as sparse retrieval\cite{bai2020spartermlearningtermbasedsparse}which relies on exact term matching techniques like BM25 and TF-IDF, dense retrieval\cite{karpukhin2020dpr}which Uses deep learning models to encode queries and documents into dense vector embeddings, retrieving relevant documents based on semantic similarity rather than exact word matching, and others.
 For the widely used sparse and dense retrieval approaches, the process typically involves two main steps:  \\
-1. Encoding each object into a specific representation.\\
-2. Building an index to organize the data for efficient searching .
+1. Encoding each object into a specific representation (e.g., term-based for sparse retrieval, vector embeddings for dense retrieval).\\
+2. Building an index to organize and efficiently retrieve relevant data during search .
 
@@
 
-Once the retriever fetches relevant data from external sources,the generator processes and integrates this information into a well-structured and meaningful answer.
+Once the retriever fetches relevant data from external sources,the generator processes and integrates this information into a well-structured and meaningful  response.
 At the core of this process is a Large Language Model (LLM), which ensures that the generated text is not only fluent and accurate but also remains relevant to the original query.
 
-In various generative tasks, different models are selected based on the specific requirements: Models like BART\cite{lewis2019bart} and GPT\cite{brown2020language}are commonly used for tasks such as text summarization and translation ,Vision-Language Models (VLMs) \cite{radford2021learningtransferablevisualmodels} are employed to generate textual descriptions from images , for text-to-Code Generation: Models like OpenAIs Codex \cite{chen2021evaluatinglargelanguagemodels} are designed to translate natural language prompts into executable code.
+In various generative tasks, different models are selected based on the specific requirements: Models like BART\cite{lewis2019bart}, GPT\cite{brown2020language}and and T5 \cite{raffel2020exploring} are commonly used for tasks such as text summarization and translation ,Vision-Language Models (VLMs) \cite{radford2021learningtransferablevisualmodels} are employed to generate textual descriptions from images , for text-to-Code Generation: Models like OpenAIs Codex \cite{chen2021evaluatinglargelanguagemodels} are designed to translate natural language prompts into executable code.
 
@@
 \subsection{Retrieval Algorithm}
-We propose an \textbf{adaptive threshold mechanism} (referred to as (Dynamic Candidate Selection) that leverages the Mixture of Logits (MoL)framework to enhance the candidate retrieval process.
+We propose an \textbf{adaptive threshold mechanism} (referred to as (Dynamic Candidate Selection) that leverages the Mixture of Logits (MoL) framework to enhance the candidate retrieval process.
 The mechanism operates as follows:
@@
 
-The query q represents the user's state at a specific time step, derived from their interaction history. In the MoL (Mixture of Logits) framework, q is transformed into Pq
-embeddings through a multi-layer perceptron (MLP). While our hybrid algorithm does not explicitly employ an MLP for this transformation
+The query \( q \) represents the user's state at a specific time step, derived from their interaction history. In the MoL (Mixture of Logits) framework,\( q \)is transformed into \( Pq \)
+embeddings through a multi-layer perceptron (MLP).
 \subsection{Hyperparameter settings.}
@@
 \end{table}
-\newpage
+
 On the ML-100K dataset, our approach shows comparable performance to MoL in terms of \textbf{HR@10 (0.0775 for both)} while achieving better results in\textbf{ MRR (0.0363 vs 0.0285)} and \textbf{NDCG (0.1241 vs 0.1182)}.
@@
 While using a fixed K is straightforward, it often fails to adapt to the varying complexities of different queries.
-This can result in either missing important information or retrieving too much irrelevant data, which adds noise to the results.
+This can result in either missing important information or retrieving irrelevant data, which adds noise to the results.
 On the other hand, dynamically adjusting K offers greater flexibility but comes with its own set of challenges, such as increased computational costs and inconsistencies in determining the optimal retrieval size.

### 2025-03-22 Paper/latex/paper_RAG.tex
update

@@
 \usepackage{listings}%
+\usepackage{subcaption}
 
@@
 However, its effectiveness heavily depends on the quality of the retrieved documents.
-Using a fixed number of retrieved documents (k) may fail  to adapt to varying query complexity, often leading to suboptimal retrieval either including irrelevant documents or missing crucial ones.
-To address this issue, we propose a hybrid approach that dynamically adjusts K based on query characteristics while maintaining computational efficiency.
+Using a fixed number of retrieved documents K may fail  to adapt to varying query complexity, often leading to suboptimal retrieval either including irrelevant documents or missing crucial ones.
+To address this issue, we propose a hybrid method that dynamically adjusts K based on query characteristics while maintaining computational efficiency.
 Our method improves retrieval performance by maximizing relevant information and minimizing noise.
 Experimental results show that it outperforms fixed-K retrieval, offering a more effective solution for optimizing RAG systems.
-Future work will explore learning-based techniques to further refine dynamic K selection.}
+Compared to traditional methods, it dynamically adjusts based on the query and dataset characteristics, offering greater adaptability across various contexts.}
 
@@
 
-Large Language Models (LLMs)\cite{Radford2019} have emerged as state-of-the-art artificial intelligence systems that excel at understanding and generating human-like text,Built on deep learning architectures,particularly transformers \cite{vaswani2017attention},these models showcase impressive adaptability across diverse tasks, such as machine translation\cite{Gu2018SearchEngine}, text generation\cite{liang2024controllabletextgenerationlarge} and question answering\cite{Zhang2023}.Notable examples of LLMs include OpenAI GPT series\cite{brown2020language}, Google BERT\cite{devlin2018bert} and T5 \cite{raffel2020exploring}.Despite their impressive capabilities,LLMs suffer from hallucination\cite{ji2023survey}, often generating confident but incorrect information.This is a critical challenge for knowledge-intensive tasks requiring high accuracy\cite{kandpal2023large}.Additionally, their static nature being trained on fixed datasets\cite{bender2021dangers} limits their ability to update knowledge dynamically, making them prone to outdated or inconsistent responses, especially for time-sensitive information.\cite{Mousavi2025}.
+Large Language Models (LLMs) have emerged as state-of-the-art artificial intelligence systems that excel at understanding and generating human-like text\cite{Radford2019}, Built on deep learning architectures, particularly transformers \cite{vaswani2017attention},these models showcase impressive adaptability across diverse tasks, such as machine translation\cite{Gu2018SearchEngine}, text generation\cite{liang2024controllabletextgenerationlarge} and question answering\cite{Zhang2023}.
+Notable examples of LLMs include OpenAI GPT series\cite{brown2020language}, Google BERT\cite{devlin2018bert} and T5 \cite{raffel2020exploring}.
+Despite their impressive capabilities, LLMs suffer from hallucination\cite{ji2023survey}, often generating confident but incorrect information.
+This is a critical challenge for knowledge-intensive tasks requiring high accuracy\cite{kandpal2023large}.
+Additionally, their static nature being trained on fixed datasets\cite{bender2021dangers} limits their ability to update knowledge dynamically, making them prone to outdated or inconsistent responses, especially for time-sensitive information.\cite{Mousavi2025}.
 
-Researchers have explored various approaches to address this limitation, including transfer learning\cite{zhuang2020comprehensive}a technique that improves pre-trained models performance in new tasks .Another technique is fine-tuning \cite{Howard2018} which adapt pre-trained models to specific tasks.Despite their strengths,they are computationally expensive and impractical for real-time knowledge updates.
-A promising alternative is Retrieval-Augmented Generation (RAG) \cite{lewis2020retrieval}.which integrates real-time retrieval from external knowledge sources with generative models enables LLMs to access up-to-date information without retraining,(RAG) faces several limitations.
-For instance,The performance of RAG heavily depends on the quality of the retriever \cite{karpukhin2020dpr}, as irrelevant or noisy documents can degrade the generator output.
-Additionally,A critical challenge is determining the optimal number of retrieved passages (K) to feed into the generator\cite{lewis2020retrieval}.
-often resulting in either insufficient or excessive information being retrieved, reduce coherence and increase computational overhead.
-This trade-off makes (k) a key factor in balancing performance and efficiency.
+Researchers have explored various approaches to address this limitation, including transfer learning\cite{zhuang2020comprehensive}, fine-tuning \cite{Howard2018} which adapt pre-trained models to specific tasks.
+Despite their strengths, they are computationally expensive and impractical for real-time knowledge updates.
+A promising alternative is Retrieval-Augmented Generation (RAG) \cite{lewis2020retrieval}, which integrates real-time retrieval from external knowledge sources with generative models, enables LLMs to access up-to-date information without retraining,(RAG) faces several limitations.
+For instance, The performance of RAG heavily depends on the quality of the retriever \cite{karpukhin2020dpr}, as irrelevant or noisy documents can degrade the generator output.
+Additionally, a critical challenge is determining the optimal number of retrieved passages K to feed into the generator\cite{lewis2020retrieval}.
+Often resulting in either insufficient or excessive information being retrieved, reduce coherence and increase computational overhead.
+This trade-off makes K a key factor in balancing performance and efficiency.
 
-Current approaches, such as static K selection \cite{lewis2020retrieval}, dynamic K adjustment attempt to address this trade-off but often introduce complexity or latency,Hybrid approaches \cite{yuan2024hybrid},that integrates static retrieval with dynamic retrieval.
-However, these approaches face significant limitations, including , high computational overhead, and a dependency on high-quality metadata for optimal performance.
-These limitations highlight the need for a more robust approach to K selection, capable of balancing efficiency and accuracy across varying query types.
+Current approaches, such as static K selection \cite{lewis2020retrieval}, dynamic K adjustment attempt to address this trade-off but often introduce complexity or latency, Hybrid approaches \cite{yuan2024hybrid},integrates static retrieval with dynamic retrieval.
+However, these approaches face significant limitations, including, high computational overhead, and a dependency on high-quality metadata for optimal performance.
+These limitations highlight the need for a more K selection robust approach, capable of balancing efficiency and accuracy across varying query types.
 
 In this paper, we investigate the problem of selecting K in RAG and propose a novel approach to dynamically adjust K based on  thresholds and a Mixture of Logits (MoL)\cite{ding2024efficient} scoring mechanism, ensuring efficient and context-sensitive candidate selection while reducing computational costs.
-the use of  adaptive thresholds algorithm improves the relevance of retrieved documents, refining the set to include only the most contextually appropriate matches.
+The use of  adaptive thresholds algorithm improves the relevance of retrieved documents, refining the set to include only the most contextually appropriate matches.
 Unlike traditional approaches, it adapts to the query and dataset characteristics, making it more flexible for diverse contexts.
@@
 
-This paper contributes to the existing literature by reviewing the various approaches to Retrieval-Augmented Generation (RAG), with a specific focus on static, dynamic, and Hybrid selection strategies.
-Additionally, it provides a critical analysis of the strengths and limitations of these methods, as it identifies key gaps in their design and performance.
-On account of this, there arises a pressing need for an improved approach capable of addressing the challenges of both static and dynamic in retrieval-augmented systems.
-\\
-The first section, Overview of Retrieval-Augmented Generation (RAG), aims to provide an a detailed explanation of the RAG framework, including its components and how it operates.\\
-The following section,introduces the novel hybrid dynamic selection algorithm.
+The remainder of this paper is as follows :We start by presenting an overview of the research problem,  We summarize existing solutions, and highlight the gaps that our work addresses through a more dynamic approach to k selection.\\
+Section 2: summarizes existing solutions, additionally, it provides a critical analysis of the strengths and limitations of these methods, as it identifies key gaps in their design and performance.
+As a result, there is a critical need for a more effective approach.\\
+Section 3: aims to provide a detailed explanation of the RAG framework, including its components and how it operates.\\
+Section 4: introduces the novel hybrid dynamic selection algorithm.
 Additionally, it provides detailed description of its methodology, explains its design, and discusses the manner in which it overcomes the limitations of existing  approaches.
 Particular emphasis is placed on the benefits of this approach, particularly in improving retrieval accuracy and computational efficiency.\\
-The final section presents experimental results that demonstrate the effectiveness of the hybrid dynamic selection algorithm.highlight its superiority in terms of retrieval performance and adaptability to diverse query complexities.
+Section 5: presents experimental results that demonstrate the effectiveness of the hybrid dynamic selection algorithm.
+Highlight its superiority in terms of retrieval performance and adaptability to diverse query complexities.
 
 \section{Related works}\label{sec2}
-The selection of an optimal number of retrieved items (K) in Retrieval-Augmented Generation (RAG) systems plays a crucial role in ensuring accurate and efficient responses.Several methods, including static, dynamic, and hybrid approaches have been proposed to address this challenge While these methods show promise in improving performance,they often face trade-offs between adaptability, efficiency when applied to complex or diverse query patterns.
+The selection of an optimal number of retrieved items K in Retrieval-Augmented Generation (RAG) systems plays a crucial role in ensuring accurate and efficient responses.
+Several methods, including static, dynamic, and hybrid approaches have been proposed to address this challenge While these methods show promise in improving performance, they often face trade-offs between adaptability, efficiency when applied to complex or diverse query patterns.
 In this section, we will explore these methods, highlight their respective strengths and weaknesses, and introduce our hybrid solution as a more robust and adaptable alternative.
@@
 Static Top-K selection in Retrieval :In traditional Retrieval-Augmented Generation (RAG) systems, static K-selection is a widely used method where a fixed number of top-k documents are retrieved based on their relevance scores.
-Sparse retrieval\cite{bai2020spartermlearningtermbasedsparse} methods where documents and queries are represented as high-dimensional, sparse vectors In these method , the "keys" represent documents to be retrieved, and the "values" are the importance or weight of those terms in the document such as query likelihood models\cite{Lafferty2001} ,TF-IDF\cite{Robertson1997} and BM25\cite{Zaragoza2009} often rely on this static approach to identify and rank documents.
+Sparse retrieval\cite{bai2020spartermlearningtermbasedsparse} methods where documents and queries are represented as high-dimensional, sparse vectors In these method, the "keys" represent documents to be retrieved, and the "values" are the importance or weight of those terms in the document such as query likelihood models\cite{Lafferty2001}, TF-IDF\cite{Robertson1997} and BM25\cite{Zaragoza2009} often rely on this static approach to identify and rank documents.
 For instance, if k=5, the system retrieves the top 5 documents deemed most relevant to a query, regardless of the query's complexity or the variability in relevance distribution among documents.
 A major drawback of this approach is its rigidity.
-Since the number of retrieved documents (k) is fixed, it fails to adapt to the varying complexity of queries.This one-size-fits-all strategy can lead to suboptimal performance across diverse datasets or tasks..
+Since the number of retrieved documents K is fixed, it fails to adapt to the varying complexity of queries.
+This one-size-fits-all strategy can lead to suboptimal performance across diverse datasets or tasks.
 
-Dynamic top-k Selection in Retrieval :addresses the limitations of static k by varying the number of retrieved documents based on the specific query or context.
-These approaches typically involve heuristics, machine learning models, or adaptive algorithms to determine the optimal K for each query such as Dynamic Query Cutoff (DQC)\cite{Culpepper2016} This approach uses machine learning models trained on query-document features to predict the optimal (k) for retrieval, A reinforcement learning-based approach\cite{Zhou2021} where the agent learns to adjust (k) by maximizing downstream task performance (e.g accurate response generation in RAG) Another example is the Dynamic Selection of k Nearest Neighbors \cite{DynamicKNN2012} where the selection process often depends on the distribution of neighbors, their distances to the query point, or their relevance scores.
-However, these methods come with limitations.They often introduce computational overhead due to the need for dynamic evaluation, which can increase latency in large-scale systems.
+Dynamic top-K Selection in Retrieval :addresses the limitations of static K by varying the number of retrieved documents based on the specific query or context.
+These approaches typically involve heuristics, machine learning models, or adaptive algorithms to determine the optimal K for each query such as Dynamic Query Cutoff (DQC)\cite{Culpepper2016} This approach uses machine learning models trained on query-document features to predict the optimal K for retrieval, a reinforcement learning-based approach\cite{Zhou2021} where the agent learns to adjust K by maximizing downstream task performance (e.g accurate response generation in RAG) Another example is the Dynamic Selection of K Nearest Neighbors \cite{DynamicKNN2012} where the selection process often depends on the distribution of neighbors, their distances to the query point, or their relevance scores.
+However, these methods come with limitations.
+They often introduce computational overhead due to the need for dynamic evaluation, which can increase latency in large-scale systems.
 Additionally, their performance heavily depends on the quality of the underlying models or distance metrics, making them sensitive to noisy data.
@@
 
-Hybrid Approaches For Top-K selection in Retrieval: combine elements of static and dynamic K-selection to achieve a balance between simplicity and adaptability.These systems might use a static k as a baseline but dynamically adjust it such as ,the "Blended RAG"\cite{sawarkar2024blendedragimprovingrag} approach enhances retrieval accuracy by combining semantic search techniques with hybrid query strategies,Furthermore STAYKATE (Static-Dynamic Hybrid Selection)\cite{Zhu2024} enhances LLM performance in scientific information extraction by integrating representativeness sampling from active learning with a retrieval-based strategy.
-Another example is DR-RAG (Dynamic Relevance for Retrieval-Augmented Generation)\cite{hei2024drragapplyingdynamicdocument} is a two-stage retrieval framework.first, it retrieves an initial set of (k1) documents followed by a classifier that evaluates the dynamic relevance between documents to enhance recall and retrieve additional relevant
-(k2) documents, however hybrid methods may face challenges in determining when to apply static or dynamic k-selection effectively, leading to potential trade-offs between retrieval accuracy and system efficiency.
-These limitations highlight the ongoing need for robust mechanisms that dynamically and efficiently adapt k to the unique characteristics of each query and dataset.
+Hybrid Approaches For Top-K selection in Retrieval: combine elements of static and dynamic K-selection to achieve a balance between simplicity and adaptability.
+These systems might use a static K as a baseline but dynamically adjust it such as, the "Blended RAG"\cite{sawarkar2024blendedragimprovingrag} approach enhances retrieval accuracy by combining semantic search techniques with hybrid query strategies, Furthermore STAYKATE (Static-Dynamic Hybrid Selection)\cite{Zhu2024} enhances LLM performance in scientific information extraction by integrating representativeness sampling from active learning with a retrieval-based strategy.
+Another example is DR-RAG (Dynamic Relevance for Retrieval-Augmented Generation)\cite{hei2024drragapplyingdynamicdocument} First, it retrieves an initial set of $k_1$ documents.
+Then, a classifier dynamically evaluates their relevance and retrieves additional relevant $K_2$ documents to enhance recall.
+However hybrid methods may face challenges in determining when to apply static or dynamic K-selection effectively, leading to potential trade-offs between retrieval accuracy and system efficiency.
+These limitations highlight the ongoing need for robust mechanisms that dynamically and efficiently adapt K to the unique characteristics of each query and dataset.
 
@@
 \end{figure}
-As illustrated in figure \ref{rag.PNG}, the RAG system is built around two main components:the retriever and the generator.
+As illustrated in figure \ref{rag.PNG}, the RAG system is built around two main components: the retriever and the generator.
 The retriever's role is to find relevant information from a pre-built data store, while the generator utilizes this retrieved data to produce coherent and contextually appropriate content.
@@
 In Retrieval-Augmented Generation (RAG) systems, the generator\cite{huang2024surveyretrievalaugmentedtextgeneration} is a key component responsible for crafting the final output.
-It works by combining the retrieved information with the users input query to produce a well-structured and contextually relevant response .
+It works by combining the retrieved information with the users input query to produce a well-structured and contextually relevant response.
 
@@
 
-In various generative tasks, different models are selected based on the specific requirements: Models like BART\cite{lewis2019bart}, GPT\cite{brown2020language}and and T5 \cite{raffel2020exploring} are commonly used for tasks such as text summarization and translation ,Vision-Language Models (VLMs) \cite{radford2021learningtransferablevisualmodels} are employed to generate textual descriptions from images , for text-to-Code Generation: Models like OpenAIs Codex \cite{chen2021evaluatinglargelanguagemodels} are designed to translate natural language prompts into executable code.
+In various generative tasks, different models are selected based on the specific requirements: Models like BART\cite{lewis2019bart}, GPT\cite{brown2020language}and and T5 \cite{raffel2020exploring} are commonly used for tasks such as text summarization and translation, Vision-Language Models (VLMs) \cite{radford2021learningtransferablevisualmodels} are employed to generate textual descriptions from images, for text-to-Code Generation: Models like OpenAIs Codex \cite{chen2021evaluatinglargelanguagemodels} are designed to translate natural language prompts into executable code.
 
@@
 \noindent
-2. \textbf{Dynamic Threshold Adjustment}:  
+2.\textbf{ Initial Candidate Retrieval:} This step involves computing similarity scores between a query representation and item representations for each feature component $p \in P$.
+
+\begin{equation}
+	S_p \gets  \{ \langle f_p(q), g_p(x) \rangle : x \in X_p \}
+	\label{eq:initial_retrieval}
+\end{equation}
+Here, $f_p(q)$ and $g_p(x)$ represent the feature embeddings of the query $q$ and item $x$ for the $p$-th component, respectively.
+The dot product $\langle f_p(q), g_p(x) \rangle$ measures the relevance of each item $x \in X_p$ with respect to $q$, producing a set of scores $S_p$ .
+
+\noindent
+3. \textbf{Dynamic Threshold Adjustment}:  
 To improve retrieval quality, \textbf{Mixture of Logits (MoL)} scores are computed for each candidate \(x \in G\).
@@
 	\phi(q, x) = \sum_{p=1}^{P} \pi_p(q, x) \cdot \langle f_p(q), g_p(x) \rangle.
+	\label{similarity_function}
 \end{equation}
@@
 \noindent
-3. \textbf{Refinement and Top-K Selection}:  
+4. \textbf{Refinement and Top-K Selection}:  
 Using the adaptive threshold \(T_{\text{adaptive}}\), additional relevant candidates are retrieved, expanding the candidate set \(G'\).
@@
 \label{ALgo1}
+ \textbf{Input:} Query $q$, Set of items $X$, Component-level embeddings: $f_p(q)$, $g_p(x)$ for $p \in P$, $x \in X$, Initial threshold $T_{\text{init}}$ \\
+ \textbf{Output:} Exact top $k$ items based on dynamic threshold   selection, $G_{\text{final}}$
+
 	\begin{algorithmic}[1]
-     \State \textbf{Input:} Query $q$, Set of items $X$, Component-level embeddings: $f_p(q)$, $g_p(x)$ for $p \in P$, $x \in X$, Initial threshold $T_{\text{init}}$ 
-    \State \textbf{Output:} Exact top $k$ items based on dynamic threshold   selection, $G_{\text{final}}$
 
@@
 		\For{each component $p \in P$}
-		\State Compute dot product scores: 
-		\[
-		S_p \gets  \{ \langle f_p(q), g_p(x) \rangle : x \in X_p \}
-		\]
+		\State Compute dot product scores:  with (eq\eqref{eq:initial_retrieval})
+
 		\State Retrieve items with scores $S_p \geq T_{\text{init}}$
@@
 		\For{each $x \in G$}
-		\State Compute MoL scores $s \gets \phi(q, x)$ where :
-		\[
-		\phi(q, x) \gets \sum_{p=1}^{P} \pi_p(q, x) \cdot \langle f_p(q), g_p(x) \rangle
-		\]
+		\State Compute MoL scores $s \gets \phi(q, x)$ using : eq\eqref{similarity_function}
 		\State Set $T_{\text{adaptive}} = \min \{s : s \in G\}$
@@
 \newpage
-\section{Experimental Results}\label{sec3}
-We evaluate the performance of the retrieval algorithm and the Mixture-of-Logits (MoL) approach on the next-item prediction task in recommendation systems\cite{Zhu_2018}\cite{kang2018selfattentivesequentialrecommendation}.
-We conducted experiments using two MovieLens datasets: ML-100K and ML-1M \cite{Harper2015}  which are widely used benchmarks for evaluating sequential recommendation models.\\
-\textbf{MovieLens-100K} \\
-100,000 ratings (1-5 scale) from 943 users on 1,682 movies\\
-Each user has rated at least 20 movies \\
-Includes user demographic information (age, gender, occupation, zip code)\\
-\textbf{MovieLens-1M} \\
-1,000,209 ratings from 6,040 users on approximately 3,900 movies \\
-Data collected from users who joined MovieLens in 2000 \\
-Represents a larger-scale recommendation scenario \\
-\subsection{ Setup }
-For both datasets, We utilize the SASRec architecture as our sequential user encoder, a model renowned for achieving state-of-the-art performance in next-item prediction tasks.
-This architecture processes the user's historical interaction sequence, generating embeddings that encapsulate the user's preferences at each time step.
-These embeddings serve as the foundation for predicting the next item in the sequence.
+\subsection{Evaluation metrics in recommendation system}
 
-The query \( q \) represents the user's state at a specific time step, derived from their interaction history. In the MoL (Mixture of Logits) framework,\( q \)is transformed into \( Pq \)
-embeddings through a multi-layer perceptron (MLP).
-\subsection{Hyperparameter settings.}
-For fair comparison, we maintained consistent architectural choices and training conditions across all experiments , We conducted an extensive hyperparameter analysis comparing both approaches (Hybrid+SAS and MoL+SAS) across different architectural configurations.All experiments were Implemented in TensorFlow and trained on a Google Colab environment with a T4 GPU.
-We use the Adam optimizer with a learning rate of 0.001.
-For the hybrid algorithm, we initialize the threshold (Tinit) to 0.3 and adaptively adjust it during training.We discuss detailed hyperparameter settings in table \ref{tab:100k_movies} ,table \ref{tab:1m_movies}
-
-\setlength{\tabcolsep}{1pt} % Increase column padding
-
-\begin{table}[ht]
-	\centering
-	\large
-	\renewcommand{\arraystretch}{1.5} % Increase row spacing
-	\resizebox{\textwidth}{!}{  % Scale table to fit width
-		\begin{tabular}{|l|c|c|c|c|c|c|c|c|}
-			\hline
-			\textbf{Model} & \textbf{Max Sequence Length} & \textbf{Embedding Dimension} & \textbf{Number of Heads} & \textbf{Feedforward Dimension} & \textbf{Batch Size} & \textbf{Epochs} & \textbf{Val Loss} & \textbf{Val Accuracy} \\
-			\hline
-			Hybrid+SAS & 50  & 128  & 2 & 128  & 128 & 12 & 5.3036 & 0.1584  \\ \hline
-			MoL+SAS    & 50  & 128  & 2 & 128  & 128 & 12 & 5.3152 & 0.1582  \\ \hline
-			Hybrid+SAS & 128 & 256  & 4 & 256  & 128 & 10 & 3.6364 & 0.4301  \\ \hline
-			MoL+SAS    & 128 & 256  & 4 & 256  & 128 & 10 & 3.6374 & 0.4299 \\\hline
-			Hybrid+SAS & 512 & 512  & 4 & 512  & 128 & 10 & 1.2808 & 0.8120 \\ \hline
-			MoL+SAS    & 512 & 512  & 4 & 512  & 128 & 10 & 1.3050 & 0.8016 \\\hline
-   \end{tabular}
-}
-\caption{Results on 100kMovies dataset}
-\label{tab:100k_movies}
-\end{table}
-
-\setlength{\tabcolsep}{1pt} % Increase column padding
-
-\begin{table}[ht]
-	\centering
-	\large
-	\renewcommand{\arraystretch}{1.5} % Increase row spacing
-	\resizebox{\textwidth}{!}{  % Scale table to fit width
-		\begin{tabular}{|l|c|c|c|c|c|c|c|c|}
-			\hline
-		\textbf{Model} & \textbf{Max Sequence Length} & \textbf{Embedding Dimension} & \textbf{Number of Heads} & \textbf{Feedforward Dimension} & \textbf{Batch Size} & \textbf{Epochs} & \textbf{Val Loss} & \textbf{Val Accuracy} \\
-		\hline
-			Hybrid+SAS & 50  & 128  & 2 & 128  & 128 & 10 & 4.5721 & 0.1530  \\ \hline
-			MoL+SAS    & 50  & 128  & 2 & 128  & 128 & 10 & 4.5733 & 0.1526  \\ \hline
-			Hybrid+SAS & 128 & 256  & 4 & 256  & 128 & 10 & 3.4152 & 0.3680  \\ \hline
-			MoL+SAS    & 128 & 256  & 4 & 256  & 128 & 10 & 3.4671 & 0.3627  \\ \hline
-			Hybrid+SAS & 512 & 512  & 4 & 512  & 128 & 10 & 1.0275 & 0.8350  \\ \hline
-			MoL+SAS    & 512 & 512  & 4 & 512  & 128 & 10 & 1.0350 & 0.8276  \\ \hline
-
-		\end{tabular}
-	}
-	\caption{Results on 1M Movies dataset}
-	\label{tab:1m_movies}
-\end{table}
-\newpage
-Both approaches demonstrate significant performance improvements as model capacity increases, with larger configurations consistently delivering better results.
-For instance, when increasing the model's capacity such as expanding the (Max Sequence Length: 512, Embedding Dimension: 512, Number of Heads: 4, Feed-Forward Dimension: 512) the hybrid algorithm combined with SASRec (Hybrid+SAS) achieves notable gains, particularly in the most resource-intensive setup.
-On the ML-100K dataset, Hybrid+SAS reaches a score of\textbf{ 0.8120} compared to the baseline's \textbf{0.8016}, while on ML-1M, it achieves \textbf{0.8350} versus \textbf{0.8276.} The ML-1M dataset generally benefits more from increased capacity, with the performance gap between datasets narrowing as the model scales.
-While smaller configurations provide a balance of efficiency and performance, the largest configuration, despite its higher computational demands, yields the best results, making it suitable for scenarios where resources are not a constraint.
-Both approaches show similar benefits from scaling, but Hybrid+SAS maintains a consistent edge in performance.
-\subsection{Evaluation metrics}
-
-We employ standard ranking metrics widely used in sequential recommendation:
+The retrieval algorithm is integrated into a recommendation system, and its effectiveness is assessed using standard ranking metrics.
+These metrics are widely used in sequential recommendation tasks to evaluate the model's ability to rank relevant items higher in a user's preference list:
 
@@
 \end{itemize}  
-Table \ref{tab:results_tab} summarizes the performance of our hybrid algorithm compared to the MoL-based approach on the MovieLens 100K and 1M datasets, both tested using a Max Sequence Length of 50, an Embedding Dimension of 128, 2 Attention Heads, a Feedforward Dimension of 128, a Batch Size of 128, and trained for 10 epochs .
-
+\section{Experimental Results}\label{sec3}
+In order to measure the effectiveness of our proposed method, we perform a comprehensive evaluation of both the retrieval algorithm and the Mixture-of-Logits (MoL) approach.
+This assessment focuses on the next-item prediction task, a fundamental challenge in recommendation systems\cite{Zhu_2018}\cite{kang2018selfattentivesequentialrecommendation}, where the goal is to predict the most relevant item a user will interact with next based on their past interactions.
+\subsection{Dataset}
+We conducted experiments using two MovieLens datasets : ML-100K and ML-1M \cite{Harper2015}which are widely used benchmarks for evaluating sequential recommendation models as detailed in Table\ref{tab:movielens}.\\
 \begin{table}[h]
 	\centering
-		\large
-	\renewcommand{\arraystretch}{1.3}
-	\begin{tabular}{|l|c|c|c|c|c|c|}
+	\renewcommand{\arraystretch}{1.2}
+	\begin{tabular}{|l|p{10cm}|}
 		\hline
-		\textbf{Model} & \textbf{HR@1} & \textbf{HR@10} & \textbf{HR@50} & \textbf{HR@200} & \textbf{MRR} & \textbf{NDCG} \\ \hline
-		\multicolumn{7}{|c|}{\textbf{MovieLens 100K}} \\ \hline
-		SASRec+Hybrid & 0.0070 & 0.0775 & 0.2606 & 0.5704 & 0.0363 & 0.1241 \\ \hline
-		SASRec+MoL    & 0.0000 & 0.0775 & 0.2887 & 0.5775 & 0.0285 & 0.1182 \\ \hline
-		\multicolumn{7}{|c|}{\textbf{MovieLens 1M}} \\ \hline
-		SASRec+Hybrid & 0.0464 & 0.1810 & 0.3819 & 0.6159 & 0.0925 & 0.1829 \\ \hline
-		SASRec+MoL    & 0.0508 & 0.1799 & 0.3896 & 0.6126 & 0.0971 & 0.1859 \\ \hline
+		\textbf{Dataset} & \textbf{Description} \\
+		\hline
+		\textbf{MovieLens-100K} & 100,000 ratings (1-5 scale) from 943 users on 1,682 movies. \\
+		& Each user has rated at least 20 movies. \\
+		& Includes user demographic information (age, gender, occupation, zip code). \\
+		\hline
+		\textbf{MovieLens-1M} & 1,000,209 ratings from 6,040 users on approximately 3,900 movies. \\
+		& Data collected from users who joined MovieLens in 2000. \\
+		& Represents a larger-scale recommendation scenario. \\
+		\hline
 	\end{tabular}
-			\caption{Performance Comparison of SASRec+Hybrid and SASRec+MoL on MovieLens 100K and 1M Datasets}
-				\label{tab:results_tab}
+	\caption{Summary of MovieLens datasets}
+	\label{tab:movielens}
 \end{table}
 
-On the ML-100K dataset, our approach shows comparable performance to MoL in terms of \textbf{HR@10 (0.0775 for both)} while achieving better results in\textbf{ MRR (0.0363 vs 0.0285)} and \textbf{NDCG (0.1241 vs 0.1182)}.
-The hybrid algorithm particularly excels in early position recommendations, as evidenced by the non-zero HR@1 score.
+\subsection{ Setup }
+For both datasets, We utilize the SASRec architecture as our sequential user encoder, a model renowned for achieving state-of-the-art performance in next-item prediction tasks.
+This architecture processes the user's historical interaction sequence, generating embeddings that encapsulate the user's preferences at each time step.
+These embeddings serve as the foundation for predicting the next item in the sequence.
 
-On the larger ML-1M dataset, both approaches show significant improvement in performance compared to ML-100K.
-Our hybrid algorithm achieves slightly better \textbf{HR@10 (0.1810 vs 0.1799)} \textbf{and HR@200 (0.6159 vs 0.6126)} compared to MoL, while MoL maintains a small edge in other metrics.\\
+The query \( q \) represents the user's state at a specific time step, derived from their interaction history. In the MoL (Mixture of Logits) framework,\( q \)is transformed into \( Pq \)
+embeddings through a multi-layer perceptron (MLP).
+\subsection{Impact of Hyperparameters}
+For fair comparison, we maintained consistent architectural choices and training conditions across all experiments, We conducted an extensive hyperparameter analysis comparing both approaches (Hybrid+SAS and MoL+SAS) across different architectural configurations.
+All experiments were Implemented in TensorFlow and trained on a Google Colab environment with a T4 GPU.
+We use the Adam optimizer with a learning rate of 0.001.
+For the hybrid algorithm, we initialize the threshold (Tinit) to 0.3 and adaptively adjust it during training.
+We discuss detailed hyperparameter settings in table \ref{tab:100k_movies}, table \ref{tab:1m_movies}
+
+\setlength{\tabcolsep}{1pt} % Increase column padding
+
+\begin{table}[ht]
+	\centering
+	\large
+	\renewcommand{\arraystretch}{1.7} % Increase row spacing
+	\resizebox{\textwidth}{!}{  % Scale table to fit width
+		\begin{tabular}{|l|c|c|c|c|c|c|c|c|}
+		      \hline
+		 \textbf{Model} & \textbf{\shortstack{Max \\ Sequence Length}} & 
+		 \textbf{\shortstack{Embedding \\ Dimension}} & 
+		 \textbf{\shortstack{Number \\ of Heads}} & 
+		 \textbf{\shortstack{Feedforward \\ Dimension}} & 
+		 \textbf{\shortstack{Batch \\ Size}} & 
+		 \textbf{Epochs} & 
+		 \textbf{\shortstack{Val \\ Loss}} & 
+		 \textbf{\shortstack{Val \\ Accuracy}} \\
+		 \hline
+			Hybrid+SAS & 50  & 128  & 2 & 128  & 128 & 12 & 5.3036 & 0.1584  \\ \hline
+			MoL+SAS    & 50  & 128  & 2 & 128  & 128 & 12 & 5.3152 & 0.1582  \\ \hline
+			Hybrid+SAS & 128 & 256  & 4 & 256  & 128 & 10 & 3.6364 & 0.4301  \\ \hline
+			MoL+SAS    & 128 & 256  & 4 & 256  & 128 & 10 & 3.6374 & 0.4299 \\\hline
+			Hybrid+SAS & 512 & 512  & 4 & 512  & 128 & 10 & 1.2808 & 0.8120 \\ \hline
+			MoL+SAS    & 512 & 512  & 4 & 512  & 128 & 10 & 1.3050 & 0.8016 \\\hline
+   \end{tabular}
+}
+\caption{Results on 100kMovies dataset}
+\label{tab:100k_movies}
+\end{table}
+
+\setlength{\tabcolsep}{1pt} % Increase column padding
+
+\begin{table}[ht]
+	\centering
+	\large
+	\renewcommand{\arraystretch}{1.7} % Increase row spacing
+	\resizebox{\textwidth}{!}{  % Scale table to fit width
+		\begin{tabular}{|l|c|c|c|c|c|c|c|c|}
+		    \hline
+		\textbf{Model} & \textbf{\shortstack{Max \\ Sequence Length}} & 
+		\textbf{\shortstack{Embedding \\ Dimension}} & 
+		\textbf{\shortstack{Number \\ of Heads}} & 
+		\textbf{\shortstack{Feedforward \\ Dimension}} & 
+		\textbf{\shortstack{Batch \\ Size}} & 
+		\textbf{Epochs} & 
+		\textbf{\shortstack{Val \\ Loss}} & 
+		\textbf{\shortstack{Val \\ Accuracy}} \\
+		\hline
+			Hybrid+SAS & 50  & 128  & 2 & 128  & 128 & 10 & 4.5721 & 0.1530  \\ \hline
+			MoL+SAS    & 50  & 128  & 2 & 128  & 128 & 10 & 4.5733 & 0.1526  \\ \hline
+			Hybrid+SAS & 128 & 256  & 4 & 256  & 128 & 10 & 3.4152 & 0.3680  \\ \hline
+			MoL+SAS    & 128 & 256  & 4 & 256  & 128 & 10 & 3.4671 & 0.3627  \\ \hline
+			Hybrid+SAS & 512 & 512  & 4 & 512  & 128 & 10 & 1.0275 & 0.8350  \\ \hline
+			MoL+SAS    & 512 & 512  & 4 & 512  & 128 & 10 & 1.0350 & 0.8276  \\ \hline
+
+		\end{tabular}
+	}
+	\caption{Results on 1M Movies dataset}
+	\label{tab:1m_movies}
+\end{table}
+\newpage
+Both approaches demonstrate significant performance improvements as model capacity increases, with larger configurations consistently delivering better results.
+For instance, when increasing the model's capacity such as expanding the (Max Sequence Length: 512, Embedding Dimension: 512, Number of Heads: 4, Feed-Forward Dimension: 512) the hybrid algorithm combined with SASRec (Hybrid+SAS) achieves notable gains, particularly in the most resource-intensive setup.
+On the ML-100K dataset, Hybrid+SAS reaches a score of\textbf{0.8120} compared to the baseline's \textbf{0.8016}, while on ML-1M, it achieves \textbf{0.8350} versus \textbf{0.8276} The ML-1M dataset generally benefits more from increased capacity, with the performance gap between datasets narrowing as the model scales.
+While smaller configurations provide a balance of efficiency and performance, the largest configuration, despite its higher computational demands, yields the best results, making it suitable for scenarios where resources are not a constraint.
+Both approaches show similar benefits from scaling, but Hybrid+SAS maintains a consistent edge in performance.
+
+As shown in Figures~\ref{fig:k_changes_100K} and \ref{fig:k_changes_1M}, the value of $k$ fluctuates across epochs for both MovieLens 100K and MovieLens 1M datasets.
+These variations indicate the adaptive nature of $k$ in response to the dataset characteristics and training progress.
+In the following section, we analyze how these changes influence model performance.
+
+\begin{figure}[htbp]
+	\centering
+	\begin{subfigure}{0.48\textwidth}
+		\centering
+		\includegraphics[width=\linewidth]{CHAGES_OF_K_FOR_100k.png}
+		\caption{K changes across epochs for MovieLens 100K}
+		\label{fig:k_changes_100K}
+	\end{subfigure}
+	\hfill
+	\begin{subfigure}{0.48\textwidth}
+		\centering
+		\includegraphics[width=\linewidth]{chnages_inK_for_1M.png}
+		\caption{K changes across epochs for MovieLens 1M}
+		\label{fig:k_changes_1M}
+	\end{subfigure}
+	\caption{Comparison of K changes across epochs for different datasets}
+	\label{fig:k_changes}
+\end{figure}
+the value of K changed dynamically during training, with different behaviors.
+For the MovieLens 100K dataset, K showed a steady increase across epochs, whereas for the MovieLens 1M dataset, K began at a lower value, experienced fluctuations, peaked, and then stabilized.
+This indicates that the larger dataset necessitated more adaptive adjustments in retrieval compared to the smaller dataset.\\
+Figure~\ref{histo_p} summarizes the performance of our hybrid algorithm compared to the MoL-based approach on the MovieLens 100K and 1M datasets, both tested using a Max Sequence Length of 50, an Embedding Dimension of 128, 2 Attention Heads, a Feedforward Dimension of 128, a Batch Size of 128, and trained for 10 epochs .
+\\
+\begin{figure}[ht]
+	\centering
+	\includegraphics[width=0.7\textwidth]{HISTOGRAM.png}
+	\caption{Comparison of Models Across Evaluation Metrics}\label{histo_p}
+\end{figure}
+The variations in K help explain the performance differences seen in the fig \ref{histo_p}.
+The Hybrid+SAS model applied to the MovieLens 1M dataset, which exhibited more dynamic changes in K achieved higher HR@50(0.3631) and HR@200(0.6170) scores, reflecting better long-range recommendation quality.
+The fluctuations in K in the 1M dataset likely allowed the model to balance exploration and precision, improving its overall ranking performance.
+
+On the other hand, the more gradual K increase in the 100K dataset resulted in relatively lower scores, suggesting that a static or overly conservative 
+K selection might limit retrieval effectiveness.
+
+Additionally, the MoL+SAS model achieved better performance than Hybrid+SAS in HR@10 for both datasets.
+This aligns with the observation that MoL+SAS tended to retrieve fewer but more relevant items at shorter ranking positions.
+This behavior can be attributed to how K evolved during training?smaller K values in earlier epochs likely enabled MoL+SAS to maintain higher precision at shorter ranks.
+These findings highlight the importance of dynamically tuning K based on dataset characteristics to optimize recommendation effectiveness.
 
@@
 The adaptive MoL threshold allows the model to dynamically refine candidate items during training, leading to better performance.
-The improvements are consistent across both datasets, highlighting the robustness of our approach.
+The improvements are consistent across both datasets, highlighting the robustness of our approach.\\
 %%=============================================%%

### 2025-04-06 Thesis/Latex/Chapter_2.tex
update

@@
 
-\subsection{Historical development of NLP}
+\subsection{Historical development of (NLP)}
 The figure below presents a timeline outlining the key developments in Natural Language Processing (NLP)  beginning with rule-based approaches in the 1950s and progressing through the rise of statistical methods and early neural networks in the late 1980s.
@@
 \subsection{Basics Concepts in Neural Networks}
-\textbf{Weights} \\
+\textbf{Weights}
+
 Variables on the edges between nodes that are multiplied with node outputs to form the input for the next layer.Weights are crucial for training and tuning a neural network and are often initialized within a range of -1 to 1.
-\\ \\
-\textbf{Biases} \\
+\\
+\textbf{Biases}
+
 Additional nodes in hidden and output layers that connect to every node within their respective layers but not to the previous layer.
 Biases add a constant value (typically 1 or -1) to the input of a layer, helping shift the activation function and aiding in effective learning.
-\textbf{Activation Functions} \\
+\textbf{Activation Functions}
+
 These functions introduce non-linearity to the network, enabling it to learn and model complex data patterns.
@@
 \subsection{ Recurrent Neural Networks (RNNs)}
-Recurrent neural networks (RNNs) are a type of neural network designed for processing sequential data, such as text or time series, by effectively handling variable-length inputs.
-An RNN consists of a hidden state h and an optical output y, which operate on a variable  sequence x=($x_1$....$x_t$)At each time step t, the hidden state $h^t$
-of the RNN is updated according to the function $h^{(t)} = f(h^{(t-1)}, x_t)$
-where f is a non-linear activation function, such as a logistic sigmoid or a more complex long short-term memory (LSTM) unit.
-This design allows RNNs to remember past inputs and incorporate them into future outputs, enabling them to recognize patterns that occur at multiple positions within a sequence.
-Additionally, RNNs utilize parameter sharing across time steps, which helps in generalizing across sequences of varying lengths and learning complex dependencies over time.
-Despite challenges with long-term dependencies, RNNs are particularly effective for tasks requiring context or sequential understanding, making them valuable for various applications, including time series analysis and natural language processing\cite{cho2014learning}
+Recurrent Neural Networks (RNNs) are a type of neural networks specifically designed for modeling and processing sequential data, such as text or time series.
+They are particularly well-suited for handling inputs of variable length by maintaining a dynamic internal state.
+An RNN comprises a hidden state $h$ and an optional output $y$, and operates over an input sequence $x = (x_1, x_2, \dots, x_t)$.
+
+At each time step $t$, the hidden state $h^{(t)}$ is updated based on the previous hidden state $h^{(t-1)}$ and the current input $x_t$ using a non-linear transition function $f$, as shown in Equation~\ref{eq:rnn_update}:
+
+\begin{equation}
+	h^{(t)} = f(h^{(t-1)}, x_t)
+	\label{eq:rnn_update}
+\end{equation}
+
+The function $f$ can be a simple non-linear activation such as the logistic sigmoid or hyperbolic tangent, or a more complex function such as a Long Short-Term Memory (LSTM) unit or a Gated Recurrent Unit (GRU).
 \begin{figure}[htbp]
@@
 \end{figure}
+
+This architecture enables RNNs to retain information from previous time steps, allowing the model to capture temporal dependencies and patterns that span across various positions in the sequence.
+Furthermore, the use of shared parameters across time steps facilitates generalization over sequences of different lengths and enhances the network's ability to learn long-range dependencies.
+
+Despite known challenges, such as difficulties in capturing long-term dependencies due to issues like vanishing gradients, RNNs remain highly effective for tasks that require sequential reasoning or contextual awareness.
+As a result, they are widely used in applications ranging from time series forecasting to natural language processing.
+\cite{cho2014learning}
+
 \subsection{ Long-Short Term Memory (LSTM)}
@@
 \end{figure}
+
 Forget Gate: Decides what information to discard from the cell state
-\[
+% Forget Gate
+\begin{equation}
 	f_t = \sigma \left( W_f \cdot [h_{t-1}, x_t] + b_f \right)
-\]
+\end{equation}
 
-Input Gate: Determines what new information to add to the cell state
-\[
+% Input Gate: Determines what new information to add to the cell state
+\begin{equation}
 	i_t = \sigma \left( W_i \cdot [h_{t-1}, x_t] + b_i \right)
-\]
-\[
+\end{equation}
+\begin{equation}
 	\tilde{C}_t = \tanh \left( W_c \cdot [h_{t-1}, x_t] + b_c \right)
-\]
+\end{equation}
 
-Output Gate: Controls what information to output from the cell state
-\[
+% Cell state update (commonly used but missing in your snippet)
+\begin{equation}
+	C_t = f_t \odot C_{t-1} + i_t \odot \tilde{C}_t
+\end{equation}
+
+% Output Gate: Controls what information to output from the cell state
+\begin{equation}
 	o_t = \sigma \left( W_o \cdot [h_{t-1}, x_t] + b_o \right)
-\]
-\[
+\end{equation}
+\begin{equation}
 	h_t = o_t \odot \tanh(C_t)
-\]
+\end{equation}
 
@@
 This approach enables the network to capture dependencies from both past and future contexts, improving its ability to resolve temporal dependencies.
+
 Bidirectional LSTMs are particularly effective at handling multidimensional problems, encapsulating spatially and temporally distributed information, and dealing with incomplete data through flexible connection mechanisms \cite{hochreiter1997long}.
@@
 \subsection{Encoder and Decoder Stacks}
-\textbf{Encoder:} Each layer in the encoder consists of two sub-layers: \\
+\textbf{Encoder}\\ Each layer in the encoder consists of two sub-layers: \\
 Multi-head self-attention mechanism: This allows the model to focus on different parts of the input sequence. \\
@@
 Output Dimension: All sub-layers and embedding layers output vectors of dimension d-model = 512.\\
-\textbf{Decoder:} Similar to the encoder, the decoder also has 6 identical layers, each containing the same two sub-layers:
+\textbf{Decoder}\\ Similar to the encoder, the decoder also has 6 identical layers, each containing the same two sub-layers:
 Multi-head self-attention mechanism ,
@@
 \end{figure}
-\subsection{Self-Attention Mechanism}
-Self-attention is a mechanism that allows the model to weigh the importance of different words in a sequence relative to each other.
-This is crucial for understanding context and relationships between words \cite{vaswani2017attention}.
-\subsection{Query, Key, and Value Vectors:}
-\textbf{Query Vectors (Q):} Generated from the input word's embedding and represent the word's role in finding relevant information in other words.\\
-\textbf{Key Vectors (K):} Derived similarly from embedding and serve as a comparison standard for each word.\\
-\textbf{Value Vectors (V):} Contain the actual information to be passed on and are weighted according to relevance determined by the Query and Key comparison \cite{vaswani2017attention}.. 
-\subsection{Self-Attention Calculation:}
-The score is computed by taking the dot product of the Query vector of a word with the Key vectors of all other words.
-This score determines how much focus should be placed on each word when encoding the current word.
-The scores are scaled and passed through a softmax function, producing a distribution of attention weights. \\
-Each Value vector is weighted by these attention scores and summed to produce the final output of the self-attention layer\cite{vaswani2017attention}.
-\subsection{Role of Multi-Head Attention}
-Multi-head attention enhances the model’s ability to focus on different aspects of the input by creating multiple sets of Query, Key, and Value vectors.
-Each set is processed independently and then combined.
-\\
-This approach allows the model to attend to information from different representation subspaces at different positions, improving its ability to capture complex relationships\cite{vaswani2017attention}.
+\subsection{Self-Attention Mechanism in Transformers}
+
+Self-attention is a fundamental component of the Transformer architecture, enabling the model to dynamically assess the relevance of each token in a sequence relative to others.
+This mechanism facilitates the capture of contextual dependencies and semantic relationships among words.
+
+The self-attention operation involves three key components derived from the input embeddings: the \textit{Query} ($Q$), \textit{Key} ($K$), and \textit{Value} ($V$) matrices.
+The attention scores are computed by taking the dot product between the query vector of a token and the key vectors of all tokens in the sequence:
+
+\begin{equation}
+	\text{Attention}(Q, K, V) = \text{softmax}\left(\frac{QK^\top}{\sqrt{d_k}}\right)V
+\end{equation}
+
+Here, $d_k$ denotes the dimension of the key vectors, which is used to scale the dot product to prevent excessively large values that could hinder gradient flow.
+The softmax function transforms the scaled scores into a probability distribution, which serves as the attention weights.
+These weights are then applied to the value vectors to compute a weighted sum, yielding the output representation for each token.
+
+This mechanism allows each token to selectively focus on other relevant tokens, thereby enhancing the model’s capacity to understand complex linguistic patterns~\cite{vaswani2017attention}.
 \subsection{Positional Encoding:}
@@
 each word during processing\cite{alammar2018illustratedtransformer}.
-\subsection{Transformers in NLP}
+\subsection{Role of Multi-Head Attention}
+Multi-head attention is a key enhancement of the self-attention mechanism that significantly improves the model’s representational capacity.
+Instead of relying on a single attention operation, the mechanism projects the input into multiple distinct sets of Query, Key, and Value vectors, referred to as attention heads.
+Each head independently performs self-attention, allowing the model to capture diverse contextual relationships from different subspaces of the input representation.
+
+The outputs of all attention heads are then concatenated and linearly transformed, producing a unified representation that integrates the complementary information gathered by each head.
+This mechanism enhances the model’s ability to capture complex patterns across different positions in the sequence, thereby improving overall performance in a variety of natural language understanding tasks\cite{vaswani2017attention}.
+
+\subsection{The Rise of Transformers in Natural Language Processing}
 Transformers have revolutionized the field of Natural Language Processing (NLP) by introducing a powerful and efficient way to handle textual data.
@@
 This stage allows the model to recognize and internalize a wide range of linguistic patterns, laying the groundwork for fine-tuning on specific tasks.\\
-Several pre-training objectives have been implemented to maximize the effectiveness of this learning process, each offering distinct benefits to the model's performance
+Several pre-training objectives have been implemented to maximize the effectiveness of this learning process, each offering distinct benefits to the model's performance .\\
 \textbf{Full Language Modeling:} Used since GPT-2, this approach trains decoder-only models to predict the next token in a sequence based on previous tokens.
@@
 \textbf{Masked Language Modeling: }Popularized by BERT, this method involves masking certain tokens in the input text and training the model to predict them, helping the model understand word context.
-An extension, span corruption, masks entire text spans for prediction, further improving contextual comprehension\cite{wang2023language}.
-\begin{figure}[htbp]
-
-	\includegraphics[width=1\linewidth]{Figures/pretraining.png}
+An extension, span corruption, masks entire text spans for prediction, further improving contextual comprehension\cite{wang2023language}.\\
+\textbf{Unified Language Modeling} refers to an approach that integrates multiple training objectives, including causal, non-causal, and masked language modeling.
+In this framework, masked language modeling differs from traditional bidirectional attention mechanisms by employing unidirectional attention—processing context either from left-to-right or right-to-left, rather than considering both directions simultaneously\cite{dong2019unified}.
+\begin{figure}[h]
+	\centering
+	\includegraphics[width=0.6\linewidth]{Figures/pretraining.png}
 	\caption{Training Tokenization in Full, Prefix, and Masked Language Modeling.}
@@
 \subsection{Fine-tuning }
-Fine-tuning has been the most common approach for adapting LLMs to specific tasks.
-After pre-training on large datasets, these models are further trained on smaller, supervised datasets tailored to the desired task.
-Fine-tuning updates the weights of the pre-trained model, allowing it to excel at specific tasks like sentiment analysis, machine translation, or question answering.
-The main advantage of fine-tuning is that it often results in strong performance on task-specific benchmarks.
-However, it comes with challenges, such as the need for a new large dataset for every task, the risk of overfitting to narrow distributions, and potential poor generalization to out-of-distribution examples\cite{touvron2023llama}.
+Fine-tuning is a key technique for adapting pre-trained large language models (LLMs) to specific downstream tasks. It encompasses several strategies, each serving different objectives.
+
+One common method is \textbf{transfer learning}, where a pre-trained model is further trained on task-specific data to enhance its performance on targeted applications.
+This allows models to leverage general language understanding while adapting to more specialized domains.
+
+Another approach is \textbf{instruction tuning}, which involves fine-tuning the model on datasets formatted as instruction-response pairs.
+These datasets typically include multiple tasks described in natural language, enabling the model to better understand prompts and generalize across various instructions.
+This method has been particularly effective in improving zero-shot and few-shot performance.
+
+To address ethical and safety concerns, \textbf{alignment tuning} is employed.
+This involves adjusting the model’s behavior based on human feedback to ensure outputs are helpful, honest, and harmless (the "HHH" criteria).
+A widely adopted framework for this is reinforcement learning with human feedback (RLHF).
+RLHF combines reward modeling—where human preferences guide the ranking of responses—with reinforcement learning techniques, such as proximal policy optimization (PPO), to iteratively align model behavior with human values.
+
+These fine-tuning approaches, while powerful, come with trade-offs in terms of data requirements, computational cost, and generalization capabilities.
+Nonetheless, they remain essential for tailoring LLMs to both functional and ethical requirements across diverse applications.\cite{touvron2023llama}.
 \subsection{Few-Shot, One-Shot, and Zero-Shot Learning}
@@
 \end{itemize}
-\section{Popular Models and Datasets}
+\section{Popular Models}
 Large Language Models (LLMs) like GPT-3, GPT-4, and BERT have revolutionized NLP by leveraging vast datasets such as Common Crawl and WebText.
@@
 \subsection{BERT}
-BERT (Bidirectional Encoder Representations from Transformers) is a groundbreaking language representation model that pretrains deep bidirectional representations from unlabeled text by jointly conditioning on both left and right contexts in all layers.
-This bidirectional approach enables BERT to be fine-tuned with just a single additional output layer for various tasks, such as question answering and language inference, without the need for extensive architectural modifications.
-BERT has achieved remarkable performance, setting new state-of-the-art results across eleven NLP tasks, including a GLUE score of 80.5\%, MultiNLI accuracy of 86.7\%, SQuAD v1.1 F1 score of 93.2\%, and SQuAD v2.0 F1 score of 83.1\%.
-Its combination of conceptual simplicity and empirical effectiveness makes BERT a powerful and versatile tool for a wide range of natural language processing applications\cite{devlin2019bert}.
+BERT (Bidirectional Encoder Representations from Transformers) is a   language representation model that pretrains deep bidirectional representations from unlabeled text by jointly conditioning on both left and right contexts at every layer.
+This bidirectional strategy allows BERT to be fine-tuned for diverse downstream tasks, such as question answering and natural language inference, by simply appending a single output layer, thereby eliminating the need for extensive architectural modifications.
+
+BERT has demonstrated exceptional performance, establishing new state-of-the-art results across eleven natural language processing benchmarks.
+Notably, it achieved a GLUE score of 80.5\%, a MultiNLI accuracy of 86.7\%, a SQuAD v1.1 F1 score of 93.2\%, and a SQuAD v2.0 F1 score of 83.1\%.
+The model's elegant design, coupled with its empirical success, underscores its versatility and efficacy as a powerful tool for a wide array of natural language processing applications.
+\cite{devlin2019bert}.
 \subsection{T5}
@@
 \subsection{LLAMA2}
-\subsection{WebText }
-WebText is a dataset developed by OpenAI to encapsulate a wide variety of high-quality web content.
-It comprises text sourced from outbound links on popular websites like Reddit, chosen based on their popularity and the quality of their content.
-Unlike Common Crawl, which indiscriminately includes a broad range of web pages, WebText is more selective, focusing on high-quality, engaging material.
-Although smaller in size compared to Common Crawl, WebText is curated to ensure the text reflects diverse, well-written content, making it particularly valuable for training language models to produce coherent and contextually appropriate text\cite{radford2019language}.
-\subsection{RuleTaker}
-RuleTaker is a collection of datasets designed to assess the deductive reasoning capabilities of language models.
-Each dataset includes a set of facts, rules, and a boolean question that the model must use logical reasoning to answer.
-The datasets feature synthetic subsets requiring various levels of reasoning complexity, with different numbers of deduction steps necessary to reach an answer.
-Among the datasets are the Bird dataset, which illustrates McCarthy’s abnormality problem , the Electricity dataset, which models appliance functions, and the ParaRules corpus, where sentences like “Bob is cold” are paraphrased to “In the snow sits Bob, crying from being cold.” The collection comprises 580,000 training examples, 84,000 validation examples, and 173,000 testing examples\cite{helwe2024}.
-\begin{figure}
-	\centering
-	\includegraphics[width=0.7\linewidth]{Figures/dataset-exmple.png}
-	\caption{exammple of RuleTaker dataset }
-\end{figure}
-\section{Challenges and Limitations}
-LLMs have greatly advanced NLP, but they also present challenges such as high computational costs, issues with adversarial robustness, and difficulties in interpretability.
-As these models scale, they encounter new challenges in scalability, privacy, and real-time processing.
-Ongoing research is exploring multi-modality, transfer learning, and continuous learning, which bring additional complexities and highlight the evolving impact of LLMs in practical applications\cite{Naveed2024}.
+Llama 2 introduces a new family of transformer-based language models available in both foundation and fine-tuned chat variants.
+These models incorporate architectural enhancements through optimized training procedures, refined pretraining datasets, and advanced fine-tuning techniques including reinforcement learning from human feedback.
+Empirical evaluations reveal that Llama 2 outperforms its predecessors and other competitive benchmarks across a range of natural language processing tasks, demonstrating notable improvements in accuracy, safety, and generalization.
+Its open release not only promotes transparency and reproducibility but also fosters collaborative advancements in the field\cite{touvron2023llama2openfoundation}.
+
+\subsection{Jais}
+Jais and Jais-chat are advanced Arabic-centric large language models built on a GPT-3 decoder-only architecture.
+Pretrained on a diverse dataset of Arabic and English texts including programming language source code—these 13-billion-parameter models excel in Arabic knowledge and reasoning, outperforming existing open Arabic and multilingual models.
+Notably, they also remain competitive in English tasks despite limited English data.
+We detail their training, fine-tuning, safety alignment, and evaluation, and release both the foundational Jais and the instruction-tuned Jais-chat variants to foster further research on Arabic LLMs\cite{sengupta2023jaisjaischatarabiccentricfoundation}.
+
+\section{Limitations of Large Language Models}
+LLMs have greatly advanced NLP, but they also present challenges As these models scale, they encounter new challenges in scalability, privacy, and real-time processing.
+Several studies have examined the inherent limitations of large language models (LLMs).
+Notably, \cite{Bender2021} and \cite{Naveed2024} highlight critical challenges that affect the performance and reliability of these models applications.
 \subsection{ Computational Cost}
@@
 Although enhancing computational resources can boost performance, the gains diminish over time when the size of the model and dataset stay constant, adhering to the power law of diminishing returns.
-\subsection{Overfitting}
-Despite their advanced learning abilities, they can overfit to noisy or unusual patterns in their large training datasets, which may result in generating nonsensical responses.
-The ongoing discussion about Memorization versus Generalization in LLMs revolves around finding an optimal balance.
-Memorization helps the model retain specific details from its training, allowing it to answer precise questions accurately.
-On the other hand, generalization enables the model to make predictions and generate responses for new, unseen inputs, which is crucial for handling diverse real-world tasks.
-The challenge is to find the right balance: excessive memorization can lead to overfitting, reducing the model's flexibility and ability to handle novel inputs.
-\subsection{ Interpretability and Explainability}
-The "black-box" nature of LLMs makes it difficult to understand how they make decisions, which is vital for gaining broader acceptance and trust, particularly in sensitive areas.
-Although these models have advanced capabilities, the lack of transparency into their workings limits their effectiveness and reliability.
-Efforts are underway to enhance the explainability of LLMs to build user trust and promote responsible use of AI.
-Understanding how LLMs generate their responses is crucial for ensuring they align with human values and legal standards.
+\subsection{Bias and Ethical Concerns}
+The training data for LLMs often encapsulates various social, cultural, and linguistic biases, which the models can inadvertently learn and propagate.
+This bias can manifest in outputs that reinforce stereotypes or display partiality, raising significant ethical concerns.
+Addressing these issues is crucial for ensuring fairness and mitigating unintended consequences in real-world applications.
 \subsection{ Hallucinations}
@@
 \end{itemize}
-\subsection{Privacy Concerns}
-As Large Language Models (LLMs) have become more complex and sizable, privacy concerns have intensified, particularly regarding data sharing and potential misuse.
-Risks include the creation of harmful content, evasion of filters, and issues related to data privacy, especially in areas like e-commerce where safeguarding customer information is vital.
-When LLMs are trained on private data and then made publicly available, additional privacy risks arise.
-Since LLMs can memorize phrases from their training data, there's a danger that these phrases could be exploited by malicious actors to retrieve sensitive information, thus threatening personal privacy.
-\subsection{Real-Time Processing}
-Real-time processing in Large Language Models (LLMs) is crucial for many applications, particularly as mobile AI solutions become more popular and concerns about information security and privacy grow.
-However, LLMs typically consist of hundreds of layers and millions of parameters, which create significant challenges for real-time processing due to their high computational demands and the limited storage capacity of hardware platforms, especially in edge computing environments.
-Although efforts such as MobileBERT attempt to lessen memory usage, they still encounter considerable execution overhead because of the extensive number of model layers, resulting in high inference latency.
+\subsection{Overfitting}
+Despite their advanced learning abilities, they can overfit to noisy or unusual patterns in their large training datasets, which may result in generating nonsensical responses.
+The ongoing discussion about Memorization versus Generalization in LLMs revolves around finding an optimal balance.
+Memorization helps the model retain specific details from its training, allowing it to answer precise questions accurately.
+On the other hand, generalization enables the model to make predictions and generate responses for new, unseen inputs, which is crucial for handling diverse real-world tasks.
+The challenge is to find the right balance: excessive memorization can lead to overfitting, reducing the model's flexibility and ability to handle novel inputs.
+
 \section{Domains of Application}
@@
 These models, which excel at understanding and generating human-like text, are finding valuable applications across diverse fields.
-\subsection{Law} 
-In the legal field, Large Language Models (LLMs) can support the analysis of legal documents by assisting with tasks such as generating initial coding for datasets, identifying key themes, and classifying information accordingly.
-This synergy between legal professionals and LLMs has been effective in examining legal texts, such as court opinions on theft, enhancing both the efficiency and quality of legal research.
-Additionally, LLMs have been tested for their capability to generate explanations of legal terms, with a focus on improving factual accuracy by integrating sentences from relevant case law.
-By incorporating pertinent case law, these enhanced models can produce more accurate and relevant explanations with fewer factual errors.
-Furthermore, LLMs can be specialized with domain-specific knowledge to tackle legal reasoning tasks and address legal queries effectively \cite{Naveed2024}.
+\subsection{Legal Information and Law}
+Large language models (LLMs) are increasingly influential in the legal sector, enhancing the analysis and processing of extensive legal texts.
+Their ability to manage and interpret large datasets supports tasks such as document classification, information retrieval, and even judicial outcome prediction.
+For instance, \cite{Chalk2020} presents Legal-BERT, adapts the BERT architecture specifically for legal applications, yielding notable improvements in handling legal documents.
+Similarly, \cite{Bommarito2018} investigated the use of LLMs in forecasting Supreme Court decisions, demonstrating their potential to contribute to legal analytics and reasoning.
+These advancements underscore the transformative impact of LLMs on legal research, offering new tools to improve the efficiency and accuracy of legal practice.
+
 \subsection{Cybersecurity}
@@
 Furthermore, LLMs simulate patient interactions, aiding students in honing their clinical skills \cite{Naveed2024}.
-
 \subsection{journalism}
@@
 By enhancing their reasoning capabilities, LLMs can uncover deeper insights into propaganda and misinformation, making them a powerful tool for modern journalism\cite{helwe2024}.
-
 \section{Conclusion}
@@
 We covered key concepts in neural networks and the transformative impact of the Transformer architecture.
-The exploration of LLMs highlighted their emergence, scaling, and the intricacies of pre-training and fine-tuning.
-We also discussed various challenges, including computational costs, interpretability, and privacy concerns.
-Finally, the chapter reviewed the applications of LLMs in fields like law, cybersecurity, medicine, and journalism, showcasing their potential and the ongoing need for further development.
+We also discussed various challenges.
+Finally, the chapter reviewed the applications of LLMs in fields like Legal Information and Law, cybersecurity, medicine, and journalism, showcasing their potential and the ongoing need for further development.
+
+Although LLMs have demonstrated promising applications in the legal sector, their influence is still constrained by practical limitations and contextual challenges.
+Recognizing these gaps, the next chapter explores Retrieval-Augmented Generation (RAG) as a novel approach to enhance legal information processing.
+This discussion will particularly focus on how RAG can be applied within the Algerian legal framework to address existing shortcomings and improve accessibility and accuracy in legal data management.

### 2025-04-06 Thesis/Latex/Chapter_3.tex
add chapter 4
update

@@
 \chapter{RAG (Retrieval-Augmented Generation)}
-\pagestyle{fancy}\lhead{\textbf \footnotesize\it{Enhancing Legal Information Access with Retrieval Augmented LLMs for Juridical Data}}
+\pagestyle{fancy}\lhead{\textbf \footnotesize\it{RAG (Retrieval-Augmented Generation}}
 \pagestyle{fancy}\chead{} \pagestyle{fancy}\rhead{}
@@
 Understanding the individual components of retrieval and generation is essential to appreciate how their synergy improves the overall performance of RAG systems.
-\subsection{Definition of RAG}
+\subsection{Definition of (RAG)}
 Retrieval-Augmented Generation (RAG) is a technique that leverages the strengths of pre-trained large language models (LLMs) and external data sources.
 By merging the generative abilities of LLMs like GPT-3 or GPT-4 with the accuracy of specialized data search mechanisms, RAG systems can generate more sophisticated and contextually relevant responses\cite{selvaraj2024}.
-\subsection{Historical Development of RAG}
+\subsection{Historical Development of (RAG)}
 The foundations of Retrieval-Augmented Generation (RAG) can be traced back to traditional information retrieval (IR) systems,  Early models primarily relied on keyword matching and simple ranking mechanisms to retrieve relevant documents from databases, establishing a basic framework for information access.
@@
 Researchers explored how LLMs could be augmented with retrieval capabilities, leading to the development of RAG systems that leverage the strengths of both retrieval and generative capabilities\cite{gao2024retrieval}.
-\subsection{ Differences Between RAG and Fine-Tuning}
-The enhancement of large language models (LLMs) has gained significant interest due to their increasing use.
-Among the various optimization strategies for LLMs, Retrieval-Augmented Generation (RAG) is often compared to fine-tuning (FT) and prompt engineering.
-Each method possesses unique attributes,
-\begin{enumerate}
-	\item Methodological Characteristics :\\
-	RAG is compared to providing a tailored textbook for information retrieval, making it suitable for precise information retrieval tasks.
-It excels in dynamic environments with real-time knowledge updates.
-\\
-	Fine-tuning is likened to a student internalizing knowledge, making it more static and suitable for replicating specific structures, styles, or formats.
-	\item External Knowledge and Adaptation: \\
-	RAG relies on external knowledge sources and allows for high interpretability but may involve higher latency and ethical considerations regarding data retrieval. \\
-	Fine-tuning requires retraining for updates and involves significant computational resources for dataset preparation and training.
-It enables deep customization of the model’s behavior but may struggle with unfamiliar data.
-	\item Performance: \\ RAG consistently outperforms unsupervised fine-tuning in knowledge-intensive tasks, particularly for both previously encountered and new knowledge.
-	LLMs often struggle to learn new factual information through unsupervised fine-tuning.
-	\item Use Cases and Combination: \\
-	The choice between RAG and fine-tuning depends on specific needs for data dynamics, customization, and computational capabilities.
-They are not mutually exclusive; their combined use may yield optimal performance and often requires multiple iterations to refine the results \cite{gao2024retrieval}.
+\subsection{Differences between (RAG), fine-tuning and transfer learning}
+Recent advances in natural language processing have led to the development of three prominent methodologies for enhancing language models: Retrieval-Augmented Generation (RAG), fine-tuning, and transfer learning.
 
-\end{enumerate}
-\section{Types of RAG Systems}
-The RAG research field is continuously evolving, with three main stages: Naive RAG, Advanced RAG, and Modular RAG.
-\subsection{Naive RAG}
+RAG, as described by Lewis et al.
+\cite{lewis2020retrieval}, integrates external retrieval mechanisms with generative models.
+This approach enables the model to dynamically access and incorporate relevant information from external knowledge bases during the generation process, thereby addressing limitations in its internal memory and providing more accurate responses for knowledge-intensive tasks.
+
+Fine-tuning, detailed by Howard and Ruder \cite{howard2018universal}, involves adapting a pre-trained model to a specific task by further training it on a targeted dataset.
+This process refines the model’s parameters to better capture the nuances of the task at hand, leading to improved performance even when the available domain-specific data is relatively limited.
+
+Transfer learning, as exemplified by Devlin et al.
+\cite{devlin2018bert}, refers to the broader strategy of leveraging a model that has been pre-trained on large, diverse datasets for use in various downstream tasks.
+This method capitalizes on the rich, generalized representations learned during the pre-training phase, and then applies task-specific adaptations to achieve state-of-the-art results across a range of applications.
+
+Collectively, these methodologies offer complementary strategies to enhance language model performance, each contributing unique advantages and addressing different challenges in the field of natural language processing.
+
+\section{Types of (RAG) Systems}
+As detailed in \cite{gao2024retrieval}, the evolution of Retrieval-Augmented Generation (RAG) can be categorized into three distinct phases: Naive RAG, Advanced RAG, and Modular RAG .
+\begin{figure}[h]
+	\centering
+	\includegraphics[width=0.9\linewidth]{Figures/rag_types.png}
+	\caption{Types of RAG Systems}
+	\label{rag_architectur.png}
+\end{figure}
+\subsection{Naive (RAG)}
 Naive RAG, a foundational approach in Retrieval-Augmented Generation, operates on a straightforward "Retrieve-Read-Generate" paradigm. This method involves three primary steps:
@@
 \end{itemize}
-To address these limitations, more sophisticated RAG techniques have emerged, which we will explore in the next section.
-\subsection{Advanced RAG}
-Taking aim at the shortcomings of Naive RAG, Advanced RAG introduces specific improvements to enhance retrieval quality. This approach utilizes pre-retrieval and post-retrieval strategies.
-\subsubsection{Pre-retrieval Strategies:}
+
+\subsection{Advanced (RAG)}
+Taking aim at the shortcomings of Naive RAG, Advanced RAG introduces specific improvements to enhance retrieval quality. This approach utilizes pre-retrieval and post-retrieval strategies.\\
+\textbf{Pre-retrieval Strategies:}
 \begin{itemize}
@@
 \end{itemize}
-\subsubsection{Post-Retrieval Strategies:}
+\textbf{Post-Retrieval Strategies:}
 \begin{itemize}
@@
 By addressing indexing issues and refining the query and retrieved information, Advanced RAG aims to improve the overall accuracy and relevance of the generated response.
-\subsection{Modular RAG} 
-Modular RAG represents the latest evolution in RAG, offering greater adaptability. it introduces specialized modules and innovative patterns to enhance retrieval and processing capabilities.
-\subsubsection{New Modules}
+\subsection{Modular (RAG)} 
+Modular RAG represents the latest evolution in RAG, offering greater adaptability. it introduces specialized modules and innovative patterns to enhance retrieval and processing capabilities.\\
+\textbf{New Modules}
 \begin{itemize}
@@
 \end{itemize}
-\subsubsection{New Patterns} 
+\textbf{New Patterns} 
 \begin{itemize}
@@
 \\
-Two models have been proposed in RAG:
-\begin{itemize}
-	\item \textbf{RAG-Sequence} uses the same document to generate the entire sequence of output. It marginalizes over the top K retrieved documents to produce a final answer.
-	\item \textbf{RAG-Token} allows different tokens in the output sequence to be generated based on different documents, making it more flexible when combining information from various
-\end{itemize}
+\begin{figure}[h]
+	\centering
+	\includegraphics[width=0.9\linewidth]{Figures/rag_architectur.png}
+	\caption{Rag architecture}
+	\label{rag_architectur.png}
+\end{figure}
 \subsection{ Augmentation Techniques}
@@
 \end{itemize}
-\begin{figure}
-	\centering
-	\includegraphics[width=0.9\linewidth]{Figures/rag_architectur.png}
-	\caption{Rag architecture}
-	\label{rag_architectur.png}
-\end{figure}
-\section{Training and Fine-Tuning RAG Models}
-Training and fine-tuning are essential steps in enhancing the performance of RAG models by optimizing both the retriever and generator components leading to more coherent, relevant, and accurate responses.
-\subsection{Training the Retriever}
-A powerful technique for improving retrieval relevance involves training the retriever model using \textbf{contrastive learning}.
-This approach aims to maximize the similarity between query and relevant document embeddings while minimizing the similarity between query and irrelevant document embeddings.
-By training the model on numerous positive (relevant) and negative (irrelevant) pairs, the retriever learns to effectively distinguish between relevant and irrelevant information\cite{sbert2024}.
+\section{RAG for Legal Information Retrieval}
+
+Retrieval-Augmented Generation (RAG) holds significant promise for transforming the way legal information is accessed, processed, and utilized.
+In legal contexts, particularly where precision and comprehensiveness are critical, RAG enables the dynamic integration of relevant external knowledge into language models during inference.
+This is especially valuable for retrieving and synthesizing content from case law, legislative texts, legal commentaries, and scholarly analyses\cite{lexemoRAG}.
+
+Unlike traditional keyword-based or Boolean legal search systems, which rely on exact matches and rigid query structures, RAG models operate by embedding both queries and documents into a shared vector space, allowing for more semantic and contextual retrieval.
+This enables users to pose natural language queries and obtain responses grounded in specific, retrieved legal documents.
+For instance, rather than retrieving a list of documents containing the keyword "contract termination," a RAG model can retrieve multiple relevant legal provisions, judgments, or commentaries and synthesize a coherent and accurate summary or answer.
+
+Moreover, RAG is well-suited to handle the diversity and complexity of legal texts.
+It can retrieve from multiple sources—statutes, judicial rulings, legal doctrines—and present consolidated outputs that reduce the time and effort needed for legal research.
+In contexts where laws are updated frequently or where legal reasoning requires multi-document understanding, such as in complex case analysis, RAG provides an efficient, scalable alternative to traditional systems.
+
+The potential benefits are particularly evident when applied to under-resourced legal systems or multilingual environments, such as in Algeria, where legal texts exist in both Arabic and French, and digital legal resources remain fragmented.
+Here, RAG could offer a pathway toward more accessible, interpretable, and comprehensive legal information services\cite{asteraRAG}.
+
 \section{Task and Evaluation}
@@
 \end{itemize}
-\section{Limitations}
-While RAG has gained significant traction across diverse applications, it still faces certain limitations in terms of effectiveness and efficiency\cite{zhao2024retrieval}.
+\section{ Challenges and Limitations in (RAG)}
+While RAG has gained significant traction across diverse applications, it still faces certain limitations in terms of effectiveness and efficiency\cite{zhao2024retrieval} \cite{gupta2024comprehensive}.
+\subsection{Scalability and Efficiency}
+A major challenge for RAG models lies in their scalability.
+Since retrieval components depend on external databases, efficiently handling large and constantly expanding datasets demands robust retrieval algorithms.
+Additionally, the high computational and memory requirements pose difficulties for deploying RAG models in real-time or resource-limited environments
 \subsection{Noisy Retrieval Results}
-\begin{itemize}
-	\item Retrieval Quality: The quality of retrieved information can be affected by factors like indexing techniques, query formulation, and the underlying dataset.
-	\item Hallucinations: Noisy or irrelevant information can lead to the generation of hallucinated or factually incorrect responses.
-	\item Contextual Understanding: The LLM may struggle to understand the context of the retrieved information, especially if it's poorly formatted or contains inconsistencies.
-\end{itemize}
+The quality of retrieval results can significantly impact the performance of Retrieval-Augmented Generation (RAG) models.
+Several factors contribute to noisy retrieval, including the methods used for indexing, the way queries are formulated, and the characteristics of the underlying datasets.
+When irrelevant or noisy information is retrieved, it may result in the generation of hallucinated or factually inaccurate responses, further compromising the reliability of the system.
+Additionally, the model’s ability to correctly interpret the retrieved information can be hindered, particularly when the information lacks proper formatting or contains inconsistencies.
+\subsection{Bias and Fairness} 
+Like other machine learning models, Retrieval-Augmented Generation (RAG) systems can inherit biases from the datasets they rely on for retrieval.
+These biases may be amplified during the generation process, potentially leading to biased or harmful outputs.
+Addressing this issue by developing effective techniques to mitigate bias in both the retrieval and generation phases remains an ongoing challenge.
 \subsection{Extra Overhead}
@@
 \end{itemize}
-\subsection{Interaction of Retrieval and Generation}
-\begin{itemize}
-	\item Aligning the goals of the retriever and generator is challenging.
-	\item Optimizing the interaction between the two components requires careful design and tuning.
-	\item The impact of various factors, such as metric selection and hyperparameter tuning, on RAG performance is still not fully understood.
-	\item
 
-\end{itemize}
 \subsection{Long Context Generation:}
@@
 \section{Conclusion}
-This chapter provided an overview of Retrieval-Augmented Generation (RAG), highlighting its significance in natural language processing.
-We discussed the key concepts of retrieval and generation, emphasizing the advantages of their combination in enhancing language model capabilities.
-The historical development of RAG was explored, along with its core components, including the retrieval mechanism, generation process, and augmentation techniques.
-Additionally, we examined various downstream tasks and evaluation targets, illustrating RAG's versatility in applications like open-domain question answering and fact verification.
-Overall, RAG represents a promising advancement in knowledge-intensive tasks, paving the way for further research and innovation in the field.
+
+In this chapter, we explored the core principles and mechanisms of Retrieval-Augmented Generation (RAG), emphasizing its potential to enhance language models through access to external knowledge sources during inference.
+We examined its general architecture, key stages, and its evolution from naive to more modular frameworks.
+
+A particular focus was placed on the application of RAG in the legal domain, where we highlighted its potential for improving legal information retrieval by enabling more contextualized and semantically rich access to case law, statutes, and scholarly texts.
+We also outlined the main challenges and limitations facing RAG models, including noisy retrieval, bias, scalability.
+
+Following this foundational understanding, the next chapter delves into a critical aspect of RAG performance: the problem of knowledge selection.
+Specifically, we explore how the effectiveness of retrieved documents directly influences the quality of generated responses, and we investigate strategies to improve relevance and ranking within RAG pipelines.

### 2025-04-06 Thesis/Latex/Chapter_4.tex
add chapter 4
update

@@
 \chapter{k Selection in Retrieval}
+\pagestyle{fancy}\lhead{\textbf \footnotesize\it{k Selection in Retrieval}}
+\pagestyle{fancy}\chead{} \pagestyle{fancy}\rhead{}
+\pagestyle{fancy}\lfoot{\textbf {\small\it{Univ-Mascara/Computer Science: 2025}}} 
+\pagestyle{fancy}\cfoot{} \pagestyle{fancy}\rfoot{\thepage}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Introduction}
@@
 \end{algorithm}
+\section{Recommendation System Overview}
+Modern technology and online services have enabled unprecedented access to vast amounts of data.
+However, this abundance of information creates an overload, making it harder for users to find relevant content efficiently.
+Recommender systems address this by filtering information and delivering personalized suggestions, saving users time and effortz\cite{Roy2022}.
+These systems are now integral to platforms like e-commerce, television programs\cite{5174476}, e-learning\cite{WANG201110831}, tourism, and more, though further improvements are needed to enhance their versatility and accuracy.
+\begin{figure}[ht]
+	\centering
+	\includegraphics[width=0.5\linewidth]{Figures/RS.png}
+	\caption{Recommendation System Process\cite{zhang2021recsys}}
+	\label{Recommendation_System _Process}	
+\end{figure}
+\subsection{SASRec: Self-Attentive Sequential Recommendation}
+Sequential recommendation systems aim to predict a user’s next interaction based on their historical behavior while incorporating contextual information from recent actions.
+However, effectively capturing patterns in sequential data is challenging due to the exponential growth of the input space as more past interactions are considered.\cite{kang2018selfat}
+\subsection{SASRec Model Architecture}
+SASRec leverages self-attention to assign adaptive weights to past items at each time step. The key components include:\\
+\textbf{ Embedding Layer:} Converts user interactions into dense vector representations. \\
+\textbf{ Self-Attention Layer: }Captures dependencies between different interactions in the sequence, allowing for long-range modeling.\\
+\textbf{ Point-Wise Feed Forward Network (FFN)}: Enhances feature extraction for better predictions.\\
+\textbf{Prediction Layer:} Computes the likelihood of the next interaction based on learned patterns.\\
+A visual representation of SASRec’s training process (Figure \ref{the_training_process_of_SASRec}) illustrates how the model uses self-attention to focus on relevant past interactions when making predictions
+\begin{figure}[ht]
+	\centering
+	\includegraphics[width=0.6\linewidth]{Figures/sasrec.png}
+	\caption{the training process of SASRec\cite{kang2018selfat}}
+	\label{the_training_process_of_SASRec}	
+\end{figure}
+\section{Evaluation Metrics}
+We employ standard ranking metrics widely used in sequential recommendation:
+
+\begin{itemize}
+	\item \textbf{Hit Rate at \( k \) (HR@\( k \))}: Measures the proportion of cases where the target item appears in the top-\( k \) recommendations. The HR@\( k \) is computed as:\cite{Tamm_2021}
+	\begin{equation}
+		\text{HR@}k = \frac{1}{|U|} \sum_{u=1}^{|U|} \mathbb{I}(\text{rank}_u \leq k),
+	\end{equation}
+	where \( |U| \) is the number of users, \( \text{rank}_u \) is the rank of the target item for user \( u \), and \( \mathbb{I}(\cdot) \) is the indicator function that returns 1 if the condition is true and 0 otherwise.
+
+	\item \textbf{Mean Reciprocal Rank (MRR)}:is a ranking quality metric that measures how quickly a system retrieves the first relevant item.
+It is calculated as the average of reciprocal ranks across all users or queries,MRR ranges from 0 to 1, with higher values indicating better performance \cite{eviden2025mrr}.
+	\begin{equation}
+		\text{MRR} = \frac{1}{|U|} \sum_{u=1}^{|U|} \frac{1}{\text{rank}_u},
+	\end{equation}
+	where \( \text{rank}_u \) is the position of the first relevant item for user \( u \) within the top-K results. \\ 
+	U represents the total number of users (for recommendation systems) or queries (for information retrieval tasks) in the dataset.
+
+	\item \textbf{Normalized Discounted Cumulative Gain (NDCG)}: Assesses the ranking quality while accounting for position importance. The NDCG@\( k \) is computed as:\cite{Tamm_2021}
+	\begin{equation}
+		\text{NDCG@}k = \frac{1}{|U|} \sum_{u=1}^{|U|} \frac{\text{DCG@}k}{\text{IDCG@}k},
+	\end{equation}
+	where \( \text{DCG@}k \) is the Discounted Cumulative Gain at position \( k \):
+	\begin{equation}
+		\text{DCG@}k = \sum_{i=1}^{k} \frac{2^{\text{rel}_i} - 1}{\log_2(i + 1)},
+	\end{equation}
+	and \( \text{IDCG@}k \) is the Ideal DCG@\( k \), computed by sorting the items by their true relevance scores.
+\end{itemize}  
+Table \ref{tab:results_tab} summarizes the performance of our hybrid algorithm compared to the MoL-based approach on the MovieLens 100K and 1M datasets, both tested using a Max Sequence Length of 50, an Embedding Dimension of 128, 2 Attention Heads, a Feedforward Dimension of 128, a Batch Size of 128, and trained for 10 epochs
 \section{ Conclusion }

### 2025-04-06 Thesis/Latex/Chapter_6.tex
update

@@
 
-\section{Recommendation System Overview}
-Modern technology and online services have enabled unprecedented access to vast amounts of data.
-However, this abundance of information creates an overload, making it harder for users to find relevant content efficiently.
-Recommender systems address this by filtering information and delivering personalized suggestions, saving users time and effortz\cite{Roy2022}.
-These systems are now integral to platforms like e-commerce, television programs\cite{5174476}, e-learning\cite{WANG201110831}, tourism, and more, though further improvements are needed to enhance their versatility and accuracy.
-\begin{figure}[ht]
-	\centering
-	\includegraphics[width=0.5\linewidth]{Figures/RS.png}
-	\caption{Recommendation System Process\cite{zhang2021recsys}}
-	\label{Recommendation_System _Process}	
-\end{figure}
-\subsection{SASRec: Self-Attentive Sequential Recommendation}
-Sequential recommendation systems aim to predict a user’s next interaction based on their historical behavior while incorporating contextual information from recent actions.
-However, effectively capturing patterns in sequential data is challenging due to the exponential growth of the input space as more past interactions are considered.\cite{kang2018selfat}
-\subsection{SASRec Model Architecture}
-SASRec leverages self-attention to assign adaptive weights to past items at each time step. The key components include:\\
-\textbf{ Embedding Layer:} Converts user interactions into dense vector representations. \\
-\textbf{ Self-Attention Layer: }Captures dependencies between different interactions in the sequence, allowing for long-range modeling.\\
-\textbf{ Point-Wise Feed Forward Network (FFN)}: Enhances feature extraction for better predictions.\\
-\textbf{Prediction Layer:} Computes the likelihood of the next interaction based on learned patterns.\\
-A visual representation of SASRec’s training process (Figure \ref{the_training_process_of_SASRec}) illustrates how the model uses self-attention to focus on relevant past interactions when making predictions
-\begin{figure}[ht]
-	\centering
-	\includegraphics[width=0.6\linewidth]{Figures/sasrec.png}
-	\caption{the training process of SASRec\cite{kang2018selfat}}
-	\label{the_training_process_of_SASRec}	
-\end{figure}
 \section{ Datasets}
@@
 Both approaches show similar benefits from scaling, but Hybrid+SAS maintains a consistent edge in performance.
-\section{Evaluation Metrics}
-We employ standard ranking metrics widely used in sequential recommendation:
 
-\begin{itemize}
-	\item \textbf{Hit Rate at \( k \) (HR@\( k \))}: Measures the proportion of cases where the target item appears in the top-\( k \) recommendations. The HR@\( k \) is computed as:\cite{Tamm_2021}
-	\begin{equation}
-		\text{HR@}k = \frac{1}{|U|} \sum_{u=1}^{|U|} \mathbb{I}(\text{rank}_u \leq k),
-	\end{equation}
-	where \( |U| \) is the number of users, \( \text{rank}_u \) is the rank of the target item for user \( u \), and \( \mathbb{I}(\cdot) \) is the indicator function that returns 1 if the condition is true and 0 otherwise.
-
-	\item \textbf{Mean Reciprocal Rank (MRR)}:is a ranking quality metric that measures how quickly a system retrieves the first relevant item.
-It is calculated as the average of reciprocal ranks across all users or queries,MRR ranges from 0 to 1, with higher values indicating better performance \cite{eviden2025mrr}.
-	\begin{equation}
-		\text{MRR} = \frac{1}{|U|} \sum_{u=1}^{|U|} \frac{1}{\text{rank}_u},
-	\end{equation}
-	where \( \text{rank}_u \) is the position of the first relevant item for user \( u \) within the top-K results. \\ 
-	U represents the total number of users (for recommendation systems) or queries (for information retrieval tasks) in the dataset.
-
-	\item \textbf{Normalized Discounted Cumulative Gain (NDCG)}: Assesses the ranking quality while accounting for position importance. The NDCG@\( k \) is computed as:\cite{Tamm_2021}
-	\begin{equation}
-		\text{NDCG@}k = \frac{1}{|U|} \sum_{u=1}^{|U|} \frac{\text{DCG@}k}{\text{IDCG@}k},
-	\end{equation}
-	where \( \text{DCG@}k \) is the Discounted Cumulative Gain at position \( k \):
-	\begin{equation}
-		\text{DCG@}k = \sum_{i=1}^{k} \frac{2^{\text{rel}_i} - 1}{\log_2(i + 1)},
-	\end{equation}
-	and \( \text{IDCG@}k \) is the Ideal DCG@\( k \), computed by sorting the items by their true relevance scores.
-\end{itemize}  
-Table \ref{tab:results_tab} summarizes the performance of our hybrid algorithm compared to the MoL-based approach on the MovieLens 100K and 1M datasets, both tested using a Max Sequence Length of 50, an Embedding Dimension of 128, 2 Attention Heads, a Feedforward Dimension of 128, a Batch Size of 128, and trained for 10 epochs
 \section{Results and Analysis}

### 2025-04-07 Thesis/Latex/Chapter_2.tex
update

--- 2025-01-22 Thesis/Latex/Chapter_2.tex
@@
-\chapter{Large Language Models (LLMs)}
-\pagestyle{fancy}\lhead{\textbf \footnotesize\it{Enhancing Legal Information Access with Retrieval Augmented LLMs for Juridical Data}}
+\chapter{Retrieval-Augmented Generation (RAG)}
+\pagestyle{fancy}\lhead{\textbf \footnotesize\it{ Retrieval-Augmented Generation (RAG)}}
 \pagestyle{fancy}\chead{} \pagestyle{fancy}\rhead{}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Introduction} \label{start1}
-Large Language Models (LLMs) represent a transformative leap in Natural Language Processing (NLP), allowing machines to perform complex language tasks with remarkable accuracy.
-From generating coherent text to answering nuanced questions, LLMs have pushed the boundaries of what AI can achieve.
-This chapter delves into the architecture and principles behind LLMs, their development through scaling and training techniques, and the models that have defined this field.
-It also addresses the challenges these models face and their expanding role in various real-world applications.
+\section{Introduction}\label{start 1}
 
-\section{Natural language processing (NLP)}
-Natural Language Processing (NLP) is a field within artificial intelligence (AI) that concentrates on the interaction between computers and human language.
-It involves the development of algorithms and models  that allow machines to comprehend, interpret, and generate human language in a meaningful context.NLP is crucial for enabling computers to process and respond to human language effectively, as demonstrated by features like Google's predictive text in keyboards and language translation systems that manage multiple languages efficiently \cite{murugan2024nlp}
+Large language models (LLMs) have seen significant advancements but face challenges, particularly in tasks that demand extensive knowledge or deal with queries beyond their training data.
+These limitations often result in inaccuracies or “hallucinations.” To address this, Retrieval-Augmented Generation (RAG) supplements LLMs by retrieving relevant document chunks from external knowledge sources based on semantic similarity.
+By doing so, RAG helps reduce factual errors, allowing LLMs to produce more accurate content.
+This has led to the widespread adoption of RAG, particularly in chatbots and other real-world applications, making it a crucial technology in advancing the capabilities of LLMs.
+\section{Fundamentals of Retrieval-Augmented Generation}
+RAG represents a significant advancement in the capabilities of large language models (LLMs) by integrating external knowledge retrieval with the generation of text.
+Understanding the individual components of retrieval and generation is essential to appreciate how their synergy improves the overall performance of RAG systems.
+\subsection{Definition of (RAG)}
+Retrieval-Augmented Generation (RAG) is a technique that leverages the strengths of pre-trained large language models (LLMs) and external data sources.
+By merging the generative abilities of LLMs like GPT-3 or GPT-4 with the accuracy of specialized data search mechanisms, RAG systems can generate more sophisticated and contextually relevant responses\cite{selvaraj2024}.
+\subsection{Historical Development of (RAG)}
+The foundations of Retrieval-Augmented Generation (RAG) can be traced back to traditional information retrieval (IR) systems,  Early models primarily relied on keyword matching and simple ranking mechanisms to retrieve relevant documents from databases, establishing a basic framework for information access.
+The landscape began to shift with the rise of neural networks in the 2010s.
+Notable innovations like Word2Vec and Transformer-based architectures enabled retrieval methods to incorporate deeper semantic understanding, paving the way for more sophisticated document retrieval processes.
+A significant advancement occurred with the introduction of dense passage retrieval (DPR) in 2020, which utilized bi-encoder architectures to map both queries and documents into dense vector spaces. 
+The integration of these advanced retrieval techniques with large language models (LLMs) became a focal point as models like GPT-3 gained prominence.
+Researchers explored how LLMs could be augmented with retrieval capabilities, leading to the development of RAG systems that leverage the strengths of both retrieval and generative capabilities\cite{gao2024retrieval}.
+\subsection{Differences between (RAG), fine-tuning and transfer learning}
+Recent advances in natural language processing have led to the development of three prominent methodologies for enhancing language models: Retrieval-Augmented Generation (RAG), fine-tuning, and transfer learning.
 
-\subsection{Historical development of NLP}
-The figure below presents a timeline outlining the key developments in Natural Language Processing (NLP)  beginning with rule-based approaches in the 1950s and progressing through the rise of statistical methods and early neural networks in the late 1980s.
-It then highlights the impact of deep learning from the 2000s onward, leading to the development of pre-trained models like BERT and GPT.
-The timeline concludes with the emergence of large language models (LLMs) from 2019 to the present, marking a significant shift in NLP research and applications.
-\begin{figure}[htbp]
-	\centerline{\includegraphics[scale=.7]{Figures/the_historyOf_nlp.png}}
-	\caption{The Vauquois triangle, illustrating the foundations of machine translation.}
-	\label{the_historyOf_nlp.png}
+RAG, as described by Lewis et al.
+\cite{lewis2020retrieval}, integrates external retrieval mechanisms with generative models.
+This approach enables the model to dynamically access and incorporate relevant information from external knowledge bases during the generation process, thereby addressing limitations in its internal memory and providing more accurate responses for knowledge-intensive tasks.
+
+Fine-tuning, detailed by Howard and Ruder \cite{howard2018universal}, involves adapting a pre-trained model to a specific task by further training it on a targeted dataset.
+This process refines the model’s parameters to better capture the nuances of the task at hand, leading to improved performance even when the available domain-specific data is relatively limited.
+
+Transfer learning, as exemplified by Devlin et al.
+\cite{devlin2018bert}, refers to the broader strategy of leveraging a model that has been pre-trained on large, diverse datasets for use in various downstream tasks.
+This method capitalizes on the rich, generalized representations learned during the pre-training phase, and then applies task-specific adaptations to achieve state-of-the-art results across a range of applications.
+
+Collectively, these methodologies offer complementary strategies to enhance language model performance, each contributing unique advantages and addressing different challenges in the field of natural language processing.
+
+\section{Types of (RAG) Systems}
+As detailed in \cite{gao2024retrieval}, the evolution of Retrieval-Augmented Generation (RAG) can be categorized into three distinct phases: Naive RAG, Advanced RAG, and Modular RAG .
+\begin{figure}[h]
+	\centering
+	\includegraphics[width=0.9\linewidth]{Figures/rag_types.png}
+	\caption{Types of RAG Systems}
+	\label{rag_architectur.png}
 \end{figure}
-
-\subsection{Importance of natural language processing}
-
-\begin{enumerate}
-	\item \textbf{Efficient Data Processing:} NLP enables businesses to process and analyze large volumes of unstructured, text-heavy data that would be challenging to handle otherwise..
-
-	\item \textbf{Understanding Human Language:}NLP can interpret complex language elements, such as acronyms, abbreviations, and context, improving the accuracy of machine learning models..
-
-	\item \textbf{Advancements in Technology:}Improvements in deep learning and machine learning have made NLP more effective, expanding the types of data that can be analyzed.
-
-	\item \textbf{Natural Interactions:} NLP allows users to interact naturally with AI chatbots and voice assistants, like Siri, without needing to use specific,predefined\cite{techtarget_nlp}.
-
-\end{enumerate}
-
-\section{Neural Networks in NLP}
-A neural network, or artificial neural network, is a machine learning algorithm inspired by the human brain.
-It is a key component of deep learning, a branch of machine learning effective in solving complex problems like image recognition and language processing.
-Unlike traditional computer programs that use a step-by-step algorithmic approach, neural networks learn from examples, mimicking the way neurons in the human brain operate.
-They consist of interconnected nodes (processing elements) that work together in parallel to solve specific problems.
-\\
-A neural network has three basic sections, or parts, each composed of "nodes." \\ The input layer is the first part, receiving raw data, with each node (or neuron) representing a feature of this data.
-The hidden layers, which form the intermediate section, perform various transformations and computations, enabling the network to learn complex patterns and relationships.
-Finally, the output layer, which is the last part, produces the network's output, with the number of nodes corresponding to the desired output classes or regression values \cite{murugan2024nlp}.
-\begin{figure}[htbp]
-
-	\centerline{\includegraphics[width=.7\linewidth]{Figures/ANN.png}}
-	\caption{Artificial Neural Network (ANN).}
-	\label{ANN.png}
-\end{figure}
-
-\subsection{Basics Concepts in Neural Networks}
-\textbf{Weights} \\
-Variables on the edges between nodes that are multiplied with node outputs to form the input for the next layer.Weights are crucial for training and tuning a neural network and are often initialized within a range of -1 to 1.
-\\ \\
-\textbf{Biases} \\
-Additional nodes in hidden and output layers that connect to every node within their respective layers but not to the previous layer.
-Biases add a constant value (typically 1 or -1) to the input of a layer, helping shift the activation function and aiding in effective learning.
-\textbf{Activation Functions} \\
-These functions introduce non-linearity to the network, enabling it to learn and model complex data patterns.
-Common activation functions include the sigmoid, hyperbolic tangent (tanh), and rectified linear unit (ReLU).
-\cite{taylor2017neural}.
-
-\subsection{ Recurrent Neural Networks (RNNs)}
-Recurrent neural networks (RNNs) are a type of neural network designed for processing sequential data, such as text or time series, by effectively handling variable-length inputs.
-An RNN consists of a hidden state h and an optical output y, which operate on a variable  sequence x=($x_1$....$x_t$)At each time step t, the hidden state $h^t$
-of the RNN is updated according to the function $h^{(t)} = f(h^{(t-1)}, x_t)$
-where f is a non-linear activation function, such as a logistic sigmoid or a more complex long short-term memory (LSTM) unit.
-This design allows RNNs to remember past inputs and incorporate them into future outputs, enabling them to recognize patterns that occur at multiple positions within a sequence.
-Additionally, RNNs utilize parameter sharing across time steps, which helps in generalizing across sequences of varying lengths and learning complex dependencies over time.
-Despite challenges with long-term dependencies, RNNs are particularly effective for tasks requiring context or sequential understanding, making them valuable for various applications, including time series analysis and natural language processing\cite{cho2014learning}
-\begin{figure}[htbp]
-
-	\centerline{\includegraphics[width=1\linewidth]{Figures/RNN.png}}
-	\caption{RNN Architecture.}
-	\label{RNN.png}
-\end{figure}
-\subsection{ Long-Short Term Memory (LSTM)}
-Long Short-Term Memory (LSTM) networks, designed by Hochreiter and Schmidhuber, are an improved version of recurrent neural networks (RNNs) designed to learn long-term dependencies in sequential data, making them suitable for tasks like time series forecasting and language translation
-They address the limitations of traditional RNNs by introducing a memory cell that maintains information over extended periods.
-This memory cell is regulated by three gates—input, forget, and output gates—which control the flow of information in and out of the cell as shown in the fig \cite{geeksforgeeks_lstm}
-\begin{figure}[htbp]
-	\centering\includegraphics[width=0.9\linewidth]{Figures/LSTM.png}
-	\caption{LSTM Architecture.}
-	\label{/LSTM.png}
-\end{figure}
-Forget Gate: Decides what information to discard from the cell state
-\[
-f_t = \sigma \left( W_f \cdot [h_{t-1}, x_t] + b_f \right)
-\]
-
-Input Gate: Determines what new information to add to the cell state
-\[
-i_t = \sigma \left( W_i \cdot [h_{t-1}, x_t] + b_i \right)
-\]
-\[
-\tilde{C}_t = \tanh \left( W_c \cdot [h_{t-1}, x_t] + b_c \right)
-\]
-
-Output Gate: Controls what information to output from the cell state
-\[
-o_t = \sigma \left( W_o \cdot [h_{t-1}, x_t] + b_o \right)
-\]
-\[
-h_t = o_t \odot \tanh(C_t)
-\]
-
-It is worth mentioning that Bidirectional LSTM networks extend the traditional LSTM architecture by processing data in both forward and backward directions.
-This approach enables the network to capture dependencies from both past and future contexts, improving its ability to resolve temporal dependencies.
-Bidirectional LSTMs are particularly effective at handling multidimensional problems, encapsulating spatially and temporally distributed information, and dealing with incomplete data through flexible connection mechanisms \cite{hochreiter1997long}.
-\section{ Transformer Architecture}
-The Transformer architecture is a deep learning model introduced in June 2017 by Vaswani et al.
-from Google Brain.
-Their paper, titled "Attention Is All You Need," presented a groundbreaking approach to processing sequential data through the use of a self-attention mechanism.
-This innovative method allows the model to assign different levels of importance to various parts of the input, enabling it to capture long-range dependencies much more effectively than earlier models like RNNs and LSTMs.The original Transformer model is structured as a stack of six layers, where the output of each layer i serves as the input to the subsequent layer i+1, continuing this process until the final prediction is reached.It features a six-layer encoder on the left and a corresponding six-layer decoder on the right,  both of which work together to transform input sequences into meaningful outputs.
-Each encoder and decoder consists of six identical layers that allow the model to process and generate language efficiently \cite{rothman2021transformers}.
-\begin{figure}[htbp]
-
-	\centerline{\includegraphics[width=.6\linewidth]{
-			Figures/trasformers.png}}
-	\caption{The Transformer - model architecture..}
-	\label{trasformers.png}
-
-\end{figure}
-\subsection{Encoder and Decoder Stacks}
-\textbf{Encoder:} Each layer in the encoder consists of two sub-layers: \\
-Multi-head self-attention mechanism: This allows the model to focus on different parts of the input sequence. \\
-Position-wise feed-forward network: A fully connected network applied to each position separately and identically. \\
-Residual Connections: Residual connections are applied around each sub-layer, followed by layer normalization. The output of each sub-layer is computed as Layer Norm(x + Sublayer(x)). \\
-Output Dimension: All sub-layers and embedding layers output vectors of dimension d-model = 512.\\
-\textbf{Decoder:} Similar to the encoder, the decoder also has 6 identical layers, each containing the same two sub-layers:
-Multi-head self-attention mechanism ,
-Position-wise feed-forward network. \\
-Additional Sub-layer: The decoder includes a third sub-layer that performs multi-head attention over the output of the encoder stack. \\
-Residual Connections and Normalization: Like the encoder, residual connections are applied around each sub-layer, followed by layer normalization. \\
-Masked Self-Attention: The self-attention mechanism in the decoder is modified to prevent positions from attending to subsequent positions, ensuring that predictions for position i depend only on the known outputs at positions before i \cite{vaswani2017attention}.
-\begin{figure}[htbp]
-
-	\centerline{\includegraphics[width=.8\linewidth]{
-			Figures/incoderDecoder.png}}
-	\caption{Encoder and Decoder Stacks}
-	\label{incoderDecoder.png}
-
-\end{figure}
-\subsection{Self-Attention Mechanism}
-Self-attention is a mechanism that allows the model to weigh the importance of different words in a sequence relative to each other.
-This is crucial for understanding context and relationships between words \cite{vaswani2017attention}.
-\subsection{Query, Key, and Value Vectors:}
-\textbf{Query Vectors (Q):} Generated from the input word's embedding and represent the word's role in finding relevant information in other words.\\
-\textbf{Key Vectors (K):} Derived similarly from embedding and serve as a comparison standard for each word.\\
-\textbf{Value Vectors (V):} Contain the actual information to be passed on and are weighted according to relevance determined by the Query and Key comparison \cite{vaswani2017attention}.. 
-\subsection{Self-Attention Calculation:}
-The score is computed by taking the dot product of the Query vector of a word with the Key vectors of all other words.
-This score determines how much focus should be placed on each word when encoding the current word.
-The scores are scaled and passed through a softmax function, producing a distribution of attention weights. \\
-Each Value vector is weighted by these attention scores and summed to produce the final output of the self-attention layer\cite{vaswani2017attention}.
-\subsection{Role of Multi-Head Attention}
-Multi-head attention enhances the model’s ability to focus on different aspects of the input by creating multiple sets of Query, Key, and Value vectors.
-Each set is processed independently and then combined.
-\\
-This approach allows the model to attend to information from different representation subspaces at different positions, improving its ability to capture complex relationships\cite{vaswani2017attention}.
-\subsection{Positional Encoding:}
-Since self-attention mechanisms do not inherently capture the order of words, positional encoding is added to the word embeddings to provide information about the relative position of each word in the sequence.
-These encodings are added directly to the input embeddings, allowing the model to consider both the position and the content of 
-each word during processing\cite{alammar2018illustratedtransformer}.
-\subsection{Transformers in NLP}
-Transformers have revolutionized the field of Natural Language Processing (NLP) by introducing a powerful and efficient way to handle textual data.
-This advancement has led to the creation of highly effective models like BERT and GPT, with its deep bidirectional context, excels at understanding and improving performance across various NLP tasks.
-GPT, on the other hand, is renowned for generating coherent and contextually relevant text, significantly advancing applications such as text generation and translation\cite{devlin2019bert}.\\
-\textbf{Scalability:} Transformers efficiently handle large datasets and long sequences, overcoming the limitations of RNNs.
-This scalability allows for training with billions of parameters, enhancing model capabilities.\\
-\textbf{Rich Contextual Understanding: }The self-attention mechanism in transformers captures relationships between words across the entire sequence, enabling deep contextual understanding and more accurate language processing.\\
-\textbf{Model Efficiency:} Transformers enable parallel processing, which speeds up training and makes them more efficient than sequential models like RNNs.
-This efficiency supports the rapid development and deployment of advanced language models\cite{vaswani2017attention}.
-\section{ Emergence of Large Language Models (LLMs)}
-
-Large Language Models (LLMs) are advanced artificial intelligence systems designed to process and generate human-like text.
-They are typically based on transformer architectures and are characterized by their enormous scale, with billions of parameters.
-LLMs are pre-trained on vast amounts of text data in a self-supervised manner, enabling them to develop a broad understanding of language.
-They are capable of performing a wide range of tasks with minimal task-specific fine-tuning, often achieving significant performance improvements through few-shot or zero-shot learning\cite{brown2020language}.
-\subsection{Scaling in Large Language Models (LLMs)}
-Scaling is crucial in the evolution of large language models.
-The history of scaling shows that increasing both model size and dataset size leads to significant improvements in performance across various NLP tasks.
-For instance, early work by Brants et al.
-(2007) demonstrated the benefits of using language models trained on vast datasets, such as 2 trillion tokens, which led to significant advancements in machine translation quality.
-This was followed by efforts like those of Heafield et al.
-(2013), who scaled traditional models to Web-scale data, and Jozefowicz et al.
-(2016), who scaled LSTMs to 1 billion parameters, achieving state-of-the-art results on large benchmarks.The advent of transformer-based models marked a significant shift.Models like BERT, GPT-2, and GPT-3, with their enormous parameter counts—up to 175 billion for GPT-3—demonstrated that scaling up not only the model but also the dataset size yields substantial gains in performance.
-Researchers like Kaplan et al.
-(2020) and Hoffmann et al.
-(2022) studied how scaling affects model performance, proposing power laws that show a predictable relationship between model size, dataset size, and performance.
-These studies emphasized the importance of scaling for the continued progress of LLMs\cite{touvron2023llama}.
-\subsection{Pre-training}
-Pre-training is a crucial stage in developing Large Language Models (LLMs), where the model learns from extensive unlabeled datasets through a process called self-supervision.
-This stage allows the model to recognize and internalize a wide range of linguistic patterns, laying the groundwork for fine-tuning on specific tasks.\\
-Several pre-training objectives have been implemented to maximize the effectiveness of this learning process, each offering distinct benefits to the model's performance
-\textbf{Full Language Modeling:} Used since GPT-2, this approach trains decoder-only models to predict the next token in a sequence based on previous tokens.
-This autoregressive method enables models like GPT-3 to generate coherent and contextually relevant text.\\
-\textbf{Prefix Language Modeling: }Employed in encoder-decoder and non-causal decoder-only models, this technique uses a non-causal (considering both past and future tokens) prefix for predicting subsequent tokens, offering more flexibility and enhancing the model's adaptability across various language tasks.\\
-\textbf{Masked Language Modeling: }Popularized by BERT, this method involves masking certain tokens in the input text and training the model to predict them, helping the model understand word context.
-An extension, span corruption, masks entire text spans for prediction, further improving contextual comprehension\cite{wang2023language}.
-\begin{figure}[htbp]
-
-	\includegraphics[width=1\linewidth]{Figures/pretraining.png}
-	\caption{Training Tokenization in Full, Prefix, and Masked Language Modeling.}
-\end{figure}
-\subsection{Fine-tuning }
-Fine-tuning has been the most common approach for adapting LLMs to specific tasks.
-After pre-training on large datasets, these models are further trained on smaller, supervised datasets tailored to the desired task.
-Fine-tuning updates the weights of the pre-trained model, allowing it to excel at specific tasks like sentiment analysis, machine translation, or question answering.
-The main advantage of fine-tuning is that it often results in strong performance on task-specific benchmarks.
-However, it comes with challenges, such as the need for a new large dataset for every task, the risk of overfitting to narrow distributions, and potential poor generalization to out-of-distribution examples\cite{touvron2023llama}.
-\subsection{Few-Shot, One-Shot, and Zero-Shot Learning}
-Few-shot, one-shot, and zero-shot learning represent different paradigms of using LLMs without extensive fine-tuning.
-These methods involve using pre-trained models to perform tasks with minimal task-specific data.\\
-\textbf{Few-Shot Learning} involves giving the model a few examples (typically 10-100) of the task during inference.
-This approach reduces the need for large, task-specific datasets and limits the risk of overfitting to narrow data distributions.
-However, few-shot results are generally not as strong as those from fully fine-tuned models.\\
-\textbf{One-Shot Learning} is similar to few-shot learning but uses only a single example alongside a natural language description of the task.
-This method mirrors how humans are often instructed for tasks, making it an interesting approach for tasks where providing multiple examples is impractical.\\
-\textbf{Zero-Shot Learning} requires the model to perform a task based solely on a natural language description, with no examples provided.
-While this method offers maximum flexibility and robustness, it is also the most challenging for models.
-Zero-shot performance is often weaker than few-shot or one-shot, but it represents a significant step towards task-agnostic AI, where models can generalize across a wide range of tasks with minimal human intervention\cite{brown2020language}.
-\subsection{Evaluation Datasets and Tasks}
-Evaluating Large Language Models (LLMs) is essential for determining their effectiveness and limitations in comprehending and generating human language.
-This evaluation typically falls into two primary categories:
-\begin{itemize}
-	\item Natural Language Understanding (NLU):This measures the model's proficiency in comprehending language, encompassing tasks such as sentiment analysis, text classification, natural language inference (NLI), question answering (QA), commonsense reasoning (CR), mathematical reasoning (MR), and reading comprehension (RC).
-	\item Natural Language Generation (NLG):This assesses the model's capability to produce text based on given context.
-It includes tasks like summarization, sentence completion, machine translation (MT), and dialogue generation.
-\end{itemize}
-\textbf{Benchmarks} play a critical role in evaluating LLMs, providing standardized tests to measure their performance across various tasks:
-\begin{itemize}
-	\item MMLU: Measures model knowledge from pretraining and evaluates performance in zero-shot and few-shot scenarios across 57 subjects, testing world knowledge and problem-solving abilities.
-	\item SuperGLUE: An advanced benchmark that builds on GLUE, assessing tasks like question answering and natural language inference.
-It is designed to test deeper aspects of language understanding and requires significant advancements in various learning methodologies.
-	\item BIG-bench: A large-scale benchmark for evaluating LLMs across diverse tasks including reasoning, creativity, ethics, and domain-specific knowledge.
-	\item GLUE: A foundational benchmark for evaluating and analyzing natural language understanding, offering a range of resources for model assessment\cite{Naveed2024}.
-\end{itemize}
-\section{Popular Models and Datasets}
-Large Language Models (LLMs) like GPT-3, GPT-4, and BERT have revolutionized NLP by leveraging vast datasets such as Common Crawl and WebText.
-These datasets provide a diverse linguistic foundation, enabling models to perform a wide range of tasks with remarkable accuracy and contextual understanding.
-\subsection{GPT-N Models}
-GPT models are advanced autoregressive language models that generate substantial and complex machine-produced text from minimal input.
-They leverage deep learning techniques to mimic human text generation by predicting the current value based on preceding values.
-NLP models initially struggled with tasks outside their training sets due to data restrictions. OpenAI addressed this with GPT-1,introduced in 2018.
-\begin{itemize}
-	\item GPT-1, trained on the BooksCorpus dataset, utilized a 12-layer transformer decoder with self-attention.
-Its pre-training allowed for zero-shot performance on various tasks, demonstrating the potential of generative language models.
-	\item In 2019, GPT-2 improved upon GPT-1 by using a larger dataset and 1.5 billion parameters (compared to GPT-1’s 117 million).
-It excelled in tasks like translation and summarization, enhancing accuracy in recognizing long-distance relationships.
-	\item GPT-3, released later, featured around 175 billion parameters and was trained on the Common Crawl dataset.
-It could generate human-like text, perform basic math, and write code.
-Despite its capabilities, its size and cost made it challenging to implement.
-	\item GPT-4, launched in March 2023, advanced further with multimodal capabilities and context windows of up to 32,768 tokens.
-It incorporates reinforcement learning for better alignment with human input and policy\cite{yenduri2023gpt}
-\end{itemize}
-\subsection{BERT}
-BERT (Bidirectional Encoder Representations from Transformers) is a groundbreaking language representation model that pretrains deep bidirectional representations from unlabeled text by jointly conditioning on both left and right contexts in all layers.
-This bidirectional approach enables BERT to be fine-tuned with just a single additional output layer for various tasks, such as question answering and language inference, without the need for extensive architectural modifications.
-BERT has achieved remarkable performance, setting new state-of-the-art results across eleven NLP tasks, including a GLUE score of 80.5\%, MultiNLI accuracy of 86.7\%, SQuAD v1.1 F1 score of 93.2\%, and SQuAD v2.0 F1 score of 83.1\%.
-Its combination of conceptual simplicity and empirical effectiveness makes BERT a powerful and versatile tool for a wide range of natural language processing applications\cite{devlin2019bert}.
-\subsection{T5}
-The T5 (Text-To-Text Transfer Transformer) model represents a unified framework for natural language processing (NLP), designed to address various text processing tasks by treating them as "text-to-text" problems.
-This approach involves converting all tasks into a format where both input and output are text, allowing the same model, objective, and training procedure to be applied across different tasks.
-T5 leverages extensive pre-training on a large, unlabeled dataset, enabling it to develop general-purpose knowledge that enhances its performance on a range of NLP tasks, such as question answering, document summarization, and sentiment classification\cite{raffel2023exploring}.
-\subsection{LLAMA2}
-\subsection{WebText }
-WebText is a dataset developed by OpenAI to encapsulate a wide variety of high-quality web content.
-It comprises text sourced from outbound links on popular websites like Reddit, chosen based on their popularity and the quality of their content.
-Unlike Common Crawl, which indiscriminately includes a broad range of web pages, WebText is more selective, focusing on high-quality, engaging material.
-Although smaller in size compared to Common Crawl, WebText is curated to ensure the text reflects diverse, well-written content, making it particularly valuable for training language models to produce coherent and contextually appropriate text\cite{radford2019language}.
-\subsection{RuleTaker}
-RuleTaker is a collection of datasets designed to assess the deductive reasoning capabilities of language models.
-Each dataset includes a set of facts, rules, and a boolean question that the model must use logical reasoning to answer.
-The datasets feature synthetic subsets requiring various levels of reasoning complexity, with different numbers of deduction steps necessary to reach an answer.
-Among the datasets are the Bird dataset, which illustrates McCarthy’s abnormality problem , the Electricity dataset, which models appliance functions, and the ParaRules corpus, where sentences like “Bob is cold” are paraphrased to “In the snow sits Bob, crying from being cold.” The collection comprises 580,000 training examples, 84,000 validation examples, and 173,000 testing examples\cite{helwe2024}.
-\begin{figure}
-	\centering
-	\includegraphics[width=0.7\linewidth]{Figures/dataset-exmple.png}
-	\caption{exammple of RuleTaker dataset }
-\end{figure}
-\section{Challenges and Limitations}
-LLMs have greatly advanced NLP, but they also present challenges such as high computational costs, issues with adversarial robustness, and difficulties in interpretability.
-As these models scale, they encounter new challenges in scalability, privacy, and real-time processing.
-Ongoing research is exploring multi-modality, transfer learning, and continuous learning, which bring additional complexities and highlight the evolving impact of LLMs in practical applications\cite{Naveed2024}.
-\subsection{ Computational Cost}
-Training large language models (LLMs) demands significant computational resources, leading to higher production costs and environmental concerns due to the substantial energy used in large-scale training.
-Although enhancing computational resources can boost performance, the gains diminish over time when the size of the model and dataset stay constant, adhering to the power law of diminishing returns.
-\subsection{Overfitting}
-Despite their advanced learning abilities, they can overfit to noisy or unusual patterns in their large training datasets, which may result in generating nonsensical responses.
-The ongoing discussion about Memorization versus Generalization in LLMs revolves around finding an optimal balance.
-Memorization helps the model retain specific details from its training, allowing it to answer precise questions accurately.
-On the other hand, generalization enables the model to make predictions and generate responses for new, unseen inputs, which is crucial for handling diverse real-world tasks.
-The challenge is to find the right balance: excessive memorization can lead to overfitting, reducing the model's flexibility and ability to handle novel inputs.
-\subsection{ Interpretability and Explainability}
-The "black-box" nature of LLMs makes it difficult to understand how they make decisions, which is vital for gaining broader acceptance and trust, particularly in sensitive areas.
-Although these models have advanced capabilities, the lack of transparency into their workings limits their effectiveness and reliability.
-Efforts are underway to enhance the explainability of LLMs to build user trust and promote responsible use of AI.
-Understanding how LLMs generate their responses is crucial for ensuring they align with human values and legal standards.
-\subsection{ Hallucinations}
-LLMs sometimes produce "hallucinations," or responses that, despite appearing plausible, are incorrect or do not match the provided information.
-These hallucinations can be classified into three types:
+\subsection{Naive (RAG)}
+Naive RAG, a foundational approach in Retrieval-Augmented Generation, operates on a straightforward "Retrieve-Read-Generate" paradigm. This method involves three primary steps:
 
 \begin{itemize}
-	\item Input-conflicting hallucination: When the model generates responses that do not align with the user's input.
-	\item Context-conflicting hallucination: When the model produces content that contradicts information it has previously generated.
-	\item Fact-conflicting hallucination: When the model creates responses that conflict with established knowledge.
+	\item Indexing: Raw data is cleaned, extracted, and converted into a uniform text format.
+It's then segmented into smaller chunks and encoded into vector representations using an embedding model.
+These vectors are stored in a vector database for efficient similarity searches.
+	\item Retrieval: When a user query is received, it's encoded into a vector and compared to the indexed chunks.
+The top K most similar chunks are retrieved and included in the prompt for the language model.
+	\item Generation: The query and retrieved chunks are combined into a prompt, which is fed to a large language model.
+The model generates a response based on the provided context and its internal knowledge.
 \end{itemize}
-\subsection{Privacy Concerns}
-As Large Language Models (LLMs) have become more complex and sizable, privacy concerns have intensified, particularly regarding data sharing and potential misuse.
-Risks include the creation of harmful content, evasion of filters, and issues related to data privacy, especially in areas like e-commerce where safeguarding customer information is vital.
-When LLMs are trained on private data and then made publicly available, additional privacy risks arise.
-Since LLMs can memorize phrases from their training data, there's a danger that these phrases could be exploited by malicious actors to retrieve sensitive information, thus threatening personal privacy.
-\subsection{Real-Time Processing}
-Real-time processing in Large Language Models (LLMs) is crucial for many applications, particularly as mobile AI solutions become more popular and concerns about information security and privacy grow.
-However, LLMs typically consist of hundreds of layers and millions of parameters, which create significant challenges for real-time processing due to their high computational demands and the limited storage capacity of hardware platforms, especially in edge computing environments.
-Although efforts such as MobileBERT attempt to lessen memory usage, they still encounter considerable execution overhead because of the extensive number of model layers, resulting in high inference latency.
-\section{Domains of Application}
-The use of Large Language Models (LLMs) in various downstream tasks has become increasingly prevalent in both AI research and industry, with new applications being identified and explored regularly.
-These models, which excel at understanding and generating human-like text, are finding valuable applications across diverse fields.
-\subsection{Law} 
-In the legal field, Large Language Models (LLMs) can support the analysis of legal documents by assisting with tasks such as generating initial coding for datasets, identifying key themes, and classifying information accordingly.
-This synergy between legal professionals and LLMs has been effective in examining legal texts, such as court opinions on theft, enhancing both the efficiency and quality of legal research.
-Additionally, LLMs have been tested for their capability to generate explanations of legal terms, with a focus on improving factual accuracy by integrating sentences from relevant case law.
-By incorporating pertinent case law, these enhanced models can produce more accurate and relevant explanations with fewer factual errors.
-Furthermore, LLMs can be specialized with domain-specific knowledge to tackle legal reasoning tasks and address legal queries effectively \cite{Naveed2024}.
-\subsection{Cybersecurity}
-large Language Models (LLMs) have garnered significant attention in the field of cybersecurity.
-Recent research has highlighted their potential in addressing software bugs created by human developers and identifying cybersecurity threats.
-For example, Arora et al.
-have proposed methods for utilizing LLMs to evaluate cyber threats on social media through sentiment analysis.
-LLMs are also employed to detect cybersecurity-related information in Open Source Intelligence (OSINT), aiding in the identification of potential cyber threats.
-Additionally, LLMs have shown promise in detecting scams, such as phishing.
-Initial tests with models like GPT-3.5 and GPT-4 have demonstrated their ability to recognize common phishing indicators in emails.
-While LLMs exhibit considerable potential in cybersecurity, improving their reasoning abilities could enhance their effectiveness further, such as in uncovering zero-day vulnerabilities in open-source software by analyzing logic and source code\cite{helwe2024}.
-\subsection{Medicine}
-The integration of Large Language Models (LLMs) into medicine is transforming both healthcare delivery and research.
-In clinical settings, LLMs are increasingly utilized in decision support systems to offer evidence-based treatment recommendations.
-By analyzing patient data and medical literature, these models can assist in diagnosing conditions, suggesting relevant tests, and proposing effective treatment options.
-Additionally, LLMs improve patient interactions through applications like chatbots that answer questions about symptoms and medications, schedule appointments, and provide health advice.\\
-In medical research, LLMs help sift through vast amounts of literature to extract, filter, and summarize relevant information, identify key studies, and predict future research directions.
-They also play a role in medical education by generating training materials, creating exam questions, explaining complex topics, and offering personalized feedback.
-Furthermore, LLMs simulate patient interactions, aiding students in honing their clinical skills \cite{Naveed2024}.
+While Naive RAG offers a basic framework, it faces several challenges in 
+\begin{itemize}
+	\item Retrieval Issues: The retrieval process can be imprecise, leading to the selection of irrelevant or missing information.
+	\item Generation Challenges: The language model may hallucinate, generating content not supported by the retrieved context. It may also produce irrelevant, toxic, or biased outputs.
+	\item Augmentation Difficulties: Integrating retrieved information into the generation process can be challenging, leading to disjointed or redundant responses.
+\end{itemize}
 
-\subsection{journalism}
-Large Language Models (LLMs) offer valuable support to journalists, especially in fact-checking and news verification.
-They can process and cross-reference large volumes of data with established knowledge bases.
-Research has demonstrated that LLMs, such as GPT-3.5, can be used to detect fake news by providing rationales that enhance other models like BERT, which can then be fine-tuned for this purpose .
-In addition to fact-checking, LLMs are useful for analyzing political debates, helping journalists to identify key themes, monitor how discussions evolve, and evaluate sentiments.
-They can also assist in detecting logical fallacies and underlying motives in political discourse and propaganda .
-By enhancing their reasoning capabilities, LLMs can uncover deeper insights into propaganda and misinformation, making them a powerful tool for modern journalism\cite{helwe2024}.
+\subsection{Advanced (RAG)}
+Taking aim at the shortcomings of Naive RAG, Advanced RAG introduces specific improvements to enhance retrieval quality. This approach utilizes pre-retrieval and post-retrieval strategies.\\
+\textbf{Pre-retrieval Strategies:}
+\begin{itemize}
+	\item Enhanced Indexing: Advanced RAG tackles indexing issues through a sliding window approach, finer segmentation of data, and inclusion of metadata.
+Additionally, it optimizes the retrieval process by employing various methods.
+	\item Query Optimization: This stage focuses on refining the user's initial query to make it clearer and more suitable for retrieval.
+Techniques like query rewriting, transformation, and expansion are commonly used.
+\end{itemize}
+\textbf{Post-Retrieval Strategies:}
+\begin{itemize}
+	\item Re-ranking Chunks: After relevant information is retrieved, Advanced RAG prioritizes the most relevant content by re-ranking the retrieved chunks and placing them strategically within the prompt.
+	\item Context Compression: To avoid overwhelming the LLM with too much information, post-retrieval efforts focus on selecting the most essential parts of the retrieved context, highlighting critical sections, and compressing the data to be processed.
+\end{itemize}
+By addressing indexing issues and refining the query and retrieved information, Advanced RAG aims to improve the overall accuracy and relevance of the generated response.
+\subsection{Modular (RAG)} 
+Modular RAG represents the latest evolution in RAG, offering greater adaptability. it introduces specialized modules and innovative patterns to enhance retrieval and processing capabilities.\\
+\textbf{New Modules}
+\begin{itemize}
+	\item Search Module: Adapts to specific scenarios by leveraging LLM-generated code and query languages to search across various data sources.
+	\item RAG Fusion: Employs a multi-query strategy to expand user queries, uncover both explicit and implicit knowledge, and improve retrieval results.
+	\item Memory Module: Leverages the LLM's memory to guide retrieval and create an unbounded memory pool, aligning the text more closely with data distribution.
+	\item Routing Module: Navigates through diverse data sources, selecting the optimal pathway for a query based on its specific needs.
+	\item Predict Module: Reduces redundancy and noise by generating relevant context directly through the LLM.
+	\item Task Adapter Module: Tailors RAG to various downstream tasks by automating prompt retrieval and creating task-specific retrievers.
+\end{itemize}
+\textbf{New Patterns} 
+\begin{itemize}
+	\item Flexible Module Arrangement: Modular RAG allows for the substitution and reconfiguration of modules to address specific challenges, surpassing the fixed structures of previous RAG paradigms.
+	\item Innovative Retrieval Strategies: Techniques like Rewrite-Retrieve-Read, Generate-Read, and Recite-Read leverage the LLM's capabilities to refine queries, generate content, and retrieve information from model weights.
+	\item Hybrid Retrieval: Combines keyword, semantic, and vector searches to cater to diverse queries, improving retrieval relevance.
+	\item Dynamic Module Interaction: Frameworks like Demonstrate-Search-Predict and ITERRETGEN demonstrate the dynamic use of module outputs to enhance each other's functionality.
+	\item Adaptive Retrieval: Techniques like FLARE and Self-RAG evaluate the necessity of retrieval based on different scenarios, allowing for a more flexible and efficient approach.
 
+\end{itemize}
+\section{Core Components of RAG}
+The core components of Retrieval-Augmented Generation (RAG) systems consist of a retrieval mechanism, a generation process, and augmentation techniques.
+These elements work together to enhance the model's ability to access relevant external information, generate coherent and contextually appropriate responses, and improve overall performance in knowledge-intensive tasks.
+\subsection{ Retrieval Mechanism}
+RAG systems combine parametric memory (a pre-trained language model) with non-parametric memory (a retrieval mechanism).
+The retrieval mechanism allows RAG models to access external information sources (e.g., Wikipedia), and this process is central to improving the model's ability to generate factual and accurate outputs \cite{lewis2020retrieval}.
+\begin{itemize}
+	\item Traditional Techniques: \\
+	Classical retrieval methods include \textbf{TF-IDF and BM25}, which rely on sparse vector representations based on term frequencies.
+These methods use exact keyword matching, making them limited in semantic understanding.
+	\item Modern Retrieval Approaches:\\
+	\textbf{ Dense Passage Retrieval (DPR): }This approach utilizes dense vector representations learned by neural models like BERT to encode both the query and document.
+It allows for a more semantic understanding of the content, making retrieval more effective.
+DPR, for example, computes the similarity between the query and documents using Maximum Inner Product Search (MIPS), which finds the closest passages based on the dense vector space.
+	\item Implementation in RAG:\\ In RAG, dense retrieval methods are often paired with techniques like \textbf{Maximum Inner Product Search} (MIPS), which efficiently matches query and document embeddings.
+This enables the system to return the most relevant documents for subsequent generation \cite{karpukhin2020dense}.
+
+\end{itemize}
+\subsection{Generation Process}
+The generation in RAG occurs through a combination of retrieved passages and the input query.\\
+Large Language Models (LLMs) like BART or T5 are utilized for this generation, leveraging their advanced capabilities in understanding and producing human-like text.
+The retrieval component supplies external context, which the generator conditions on to formulate a response.This context is crucial, as it enriches the LLM's understanding, enabling it to incorporate real-time, relevant information.
+Moreover, the generation process benefits from the LLM's ability to synthesize information, allowing it to create responses that are not only coherent but also informed by the latest data, thus improving accuracy and relevance in knowledge-intensive tasks \cite{lewis2020retrieval}.
+\\
+\begin{figure}[h]
+	\centering
+	\includegraphics[width=0.9\linewidth]{Figures/rag_architectur.png}
+	\caption{Rag architecture}
+	\label{rag_architectur.png}
+\end{figure}
+\subsection{ Augmentation Techniques}
+The augmentation of the retrieval process in Retrieval-Augmented Generation (RAG) systems focuses on improving how queries are refined and how relevant information is retrieved for downstream generation tasks.
+Key augmentation methods include.
+\begin{itemize}
+	\item \textbf{Query Augmentation:} In traditional retrieval pipelines, user queries are often under-specified or ambiguous, leading to poor retrieval performance.
+Query augmentation involves dynamically rewriting the user's query to better match the documents in the knowledge base.
+This can be done by leveraging large language models (LLMs) to generate tailored queries or synthetic questions and answers (QAs) that better align with the search objective​
+	\item \textbf{Synthetic QA Generation:} Instead of using raw document chunks, retrieval is enhanced by generating and embedding synthetic QA pairs from documents.
+This helps to capture the semantic essence of long texts more effectively, reducing noise and improving retrieval precision.
+These synthetic QAs can be used to rewrite the user query, making it more specific to the task at hand \cite{mombaerts2024meta}.
+\end{itemize}
+\section{RAG for Legal Information Retrieval}
+
+Retrieval-Augmented Generation (RAG) holds significant promise for transforming the way legal information is accessed, processed, and utilized.
+In legal contexts, particularly where precision and comprehensiveness are critical, RAG enables the dynamic integration of relevant external knowledge into language models during inference.
+This is especially valuable for retrieving and synthesizing content from case law, legislative texts, legal commentaries, and scholarly analyses\cite{lexemoRAG}.
+
+Unlike traditional keyword-based or Boolean legal search systems, which rely on exact matches and rigid query structures, RAG models operate by embedding both queries and documents into a shared vector space, allowing for more semantic and contextual retrieval.
+This enables users to pose natural language queries and obtain responses grounded in specific, retrieved legal documents.
+For instance, rather than retrieving a list of documents containing the keyword "contract termination," a RAG model can retrieve multiple relevant legal provisions, judgments, or commentaries and synthesize a coherent and accurate summary or answer.
+
+Moreover, RAG is well-suited to handle the diversity and complexity of legal texts.
+It can retrieve from multiple sources—statutes, judicial rulings, legal doctrines—and present consolidated outputs that reduce the time and effort needed for legal research.
+In contexts where laws are updated frequently or where legal reasoning requires multi-document understanding, such as in complex case analysis, RAG provides an efficient, scalable alternative to traditional systems.
+
+The potential benefits are particularly evident when applied to under-resourced legal systems or multilingual environments, such as in Algeria, where legal texts exist in both Arabic and French, and digital legal resources remain fragmented.
+Here, RAG could offer a pathway toward more accessible, interpretable, and comprehensive legal information services\cite{asteraRAG}.
+
+\section{Task and Evaluation}
+The rapid development and increasing use of Retrieval-Augmented Generation (RAG) models in NLP have made their evaluation a critical area of research within the LLM community.
+This section covers the primary downstream tasks associated with RAG, the datasets used, and the methods for evaluating RAG systems \cite{zhou2020trustworthiness}.
+\subsection{Factuality}
+\begin{itemize}
+	\item Fact-Checking Benchmarks: Using datasets like FEVER and SQuAD to evaluate the model's ability to identify and correct factual errors.
+	\item Adversarial Testing: Creating adversarial examples to test the model's robustness against misleading information.
+	\item Contextual Understanding: Assessing the model's ability to understand the context of a query and provide accurate answers.
+\end{itemize}
+\subsection{ Robustness}
+\begin{itemize}
+	\item Noise Injection: Introducing noise into the retrieved documents to test the model's ability to handle imperfect information.
+	\item Adversarial Attacks: Evaluating the model's resilience to attacks that aim to manipulate its outputs.
+	\item Domain Adaptation: Testing the model's ability to adapt to new domains and data distributions.
+\end{itemize}
+\subsection{Fairness}
+\begin{itemize}
+
+	\item Bias Detection: Identifying and quantifying biases in the training data and model outputs.
+	\item Fairness Metrics: Using metrics like demographic parity and equalized odds to evaluate the model's fairness.
+	\item Mitigation Techniques: Implementing techniques like debiasing and fairness constraints to mitigate biases.
+\end{itemize}
+\subsection{Objective Metrics:}
+\begin{itemize}
+	\item Accuracy: Precision, recall, and F1-score for evaluating the correctness of the generated responses.
+	\item Consistency: Measuring the consistency of the model's outputs across different queries and contexts.
+	\item Coherence: Assessing the coherence and fluency of the generated text.
+\end{itemize}
+\subsection{Subjective Metrics}
+\begin{itemize}
+	\item Human Evaluation: Using human raters to evaluate the quality of the generated responses.
+	\item User Studies: Conducting user studies to gather feedback on the user experience
+\end{itemize}
+\section{ Challenges and Limitations in (RAG)}
+While RAG has gained significant traction across diverse applications, it still faces certain limitations in terms of effectiveness and efficiency\cite{zhao2024retrieval} \cite{gupta2024comprehensive}.
+\subsection{Scalability and Efficiency}
+A major challenge for RAG models lies in their scalability.
+Since retrieval components depend on external databases, efficiently handling large and constantly expanding datasets demands robust retrieval algorithms.
+Additionally, the high computational and memory requirements pose difficulties for deploying RAG models in real-time or resource-limited environments
+\subsection{Noisy Retrieval Results}
+The quality of retrieval results can significantly impact the performance of Retrieval-Augmented Generation (RAG) models.
+Several factors contribute to noisy retrieval, including the methods used for indexing, the way queries are formulated, and the characteristics of the underlying datasets.
+When irrelevant or noisy information is retrieved, it may result in the generation of hallucinated or factually inaccurate responses, further compromising the reliability of the system.
+Additionally, the model’s ability to correctly interpret the retrieved information can be hindered, particularly when the information lacks proper formatting or contains inconsistencies.
+\subsection{Bias and Fairness} 
+Like other machine learning models, Retrieval-Augmented Generation (RAG) systems can inherit biases from the datasets they rely on for retrieval.
+These biases may be amplified during the generation process, potentially leading to biased or harmful outputs.
+Addressing this issue by developing effective techniques to mitigate bias in both the retrieval and generation phases remains an ongoing challenge.
+\subsection{Extra Overhead}
+\begin{itemize}
+	\item Computational Cost: Retrieval and processing additional information can increase computational costs, especially for large-scale models and complex queries.
+	\item Latency: Retrieval and processing can introduce latency, impacting the real-time performance of RAG systems.
+	\item Complexity: Implementing and deploying RAG systems requires careful consideration of various factors, including data preparation, model selection, and system architecture.
+\end{itemize}
+
+\subsection{Long Context Generation:}
+
+\begin{itemize}
+	\item Context Length Limits: LLMs have limitations on the amount of context they can process at once.
+	\item Information Loss: Long documents may be truncated or summarized, leading to loss of important information.
+	\item Computational Cost: Processing long contexts can be computationally expensive.
+\end{itemize}
+\newpage
 \section{Conclusion}
-This chapter has outlined the evolution and significance of Natural Language Processing (NLP), from early neural networks to advanced Large Language Models (LLMs).
-We covered key concepts in neural networks and the transformative impact of the Transformer architecture.
-The exploration of LLMs highlighted their emergence, scaling, and the intricacies of pre-training and fine-tuning.
-We also discussed various challenges, including computational costs, interpretability, and privacy concerns.
-Finally, the chapter reviewed the applications of LLMs in fields like law, cybersecurity, medicine, and journalism, showcasing their potential and the ongoing need for further development.
+
+In this chapter, we explored the core principles and mechanisms of Retrieval-Augmented Generation (RAG), emphasizing its potential to enhance language models through access to external knowledge sources during inference.
+We examined its general architecture, key stages, and its evolution from naive to more modular frameworks.
+
+A particular focus was placed on the application of RAG in the legal domain, where we highlighted its potential for improving legal information retrieval by enabling more contextualized and semantically rich access to case law, statutes, and scholarly texts.
+We also outlined the main challenges and limitations facing RAG models, including noisy retrieval, bias, scalability.
+
+Following this foundational understanding, the next chapter delves into a critical aspect of RAG performance: the problem of knowledge selection.
+Specifically, we explore how the effectiveness of retrieved documents directly influences the quality of generated responses, and we investigate strategies to improve relevance and ranking within RAG pipelines.

### 2025-04-07 Thesis/Latex/Chapter_3.tex
update

--- 2025-01-22 Thesis/Latex/Chapter_3.tex
@@
-\chapter{RAG (Retrieval-Augmented Generation)}
-\pagestyle{fancy}\lhead{\textbf \footnotesize\it{Enhancing Legal Information Access with Retrieval Augmented LLMs for Juridical Data}}
+\chapter{k Selection in Retrieval}
+\pagestyle{fancy}\lhead{\textbf \footnotesize\it{k Selection in Retrieval}}
 \pagestyle{fancy}\chead{} \pagestyle{fancy}\rhead{}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{ Introduction}\label{start2}
+\section{Introduction}
+Retrieval-Augmented Generation (RAG) systems enhance language models by grounding responses in externally retrieved documents.A critical challenge in these systems is determining the optimal number of documents (k) to retrieve for a given query.Current approaches fall into three categories: static k selection, which uses a fixed number of retrieved documents; dynamic k selection, which adjusts k based on query  characteristics; and hybrid approaches, which combine both strategies.In this chapter, we first explore methods, highlighting their strengths and limitations We then introduce our novel hybrid dynamic selection algorithm provide detailed description of its methodology and present experimental results that demonstrate the effectiveness of its performance .
+%\section {The Role of k in Retrieval-Augmented Generation}
+%The number of retrieved documents (k) is a critical parameter in Retrieval-Augmented Generation (RAG) that has a direct impact on the retrieval and generation phases.
+Choosing the right k ensures that the system retrieves enough relevant information.
+\section{Defining k: The Number of Retrieved Documents}
+the parameter k denotes the number of documents or passages retrieved from an external knowledge base.
+This retrieval process is managed by the retriever component\cite{pareto2024rag},which identifies the top k relevant documents based on similarity to the query typically using embedding-based search\cite{Rossi_2024} (e.g., dense retrieval with FAISS, BM25, or hybrid methods) Subsequently, these documents are passed to the generator, typically a language model, which synthesizes the final response by integrating the retrieved
+\begin{figure}[h]
+	\centering
+	\includegraphics[width=0.9\linewidth]{Figures/topk.png}
+	\caption{Basic retrieval}
+	\label{rag_retrival.png}
 
-Large language models (LLMs) have seen significant advancements but face challenges, particularly in tasks that demand extensive knowledge or deal with queries beyond their training data.
-These limitations often result in inaccuracies or “hallucinations.” To address this, Retrieval-Augmented Generation (RAG) supplements LLMs by retrieving relevant document chunks from external knowledge sources based on semantic similarity.
-By doing so, RAG helps reduce factual errors, allowing LLMs to produce more accurate content.
-This has led to the widespread adoption of RAG, particularly in chatbots and other real-world applications, making it a crucial technology in advancing the capabilities of LLMs.
-\section{Fundamentals of Retrieval-Augmented Generation}
-RAG represents a significant advancement in the capabilities of large language models (LLMs) by integrating external knowledge retrieval with the generation of text.
-Understanding the individual components of retrieval and generation is essential to appreciate how their synergy improves the overall performance of RAG systems.
-\subsection{Definition of RAG}
-Retrieval-Augmented Generation (RAG) is a technique that leverages the strengths of pre-trained large language models (LLMs) and external data sources.
-By merging the generative abilities of LLMs like GPT-3 or GPT-4 with the accuracy of specialized data search mechanisms, RAG systems can generate more sophisticated and contextually relevant responses\cite{selvaraj2024}.
-\subsection{Historical Development of RAG}
-The foundations of Retrieval-Augmented Generation (RAG) can be traced back to traditional information retrieval (IR) systems,  Early models primarily relied on keyword matching and simple ranking mechanisms to retrieve relevant documents from databases, establishing a basic framework for information access.
-The landscape began to shift with the rise of neural networks in the 2010s.
-Notable innovations like Word2Vec and Transformer-based architectures enabled retrieval methods to incorporate deeper semantic understanding, paving the way for more sophisticated document retrieval processes.
-A significant advancement occurred with the introduction of dense passage retrieval (DPR) in 2020, which utilized bi-encoder architectures to map both queries and documents into dense vector spaces. 
-The integration of these advanced retrieval techniques with large language models (LLMs) became a focal point as models like GPT-3 gained prominence.
-Researchers explored how LLMs could be augmented with retrieval capabilities, leading to the development of RAG systems that leverage the strengths of both retrieval and generative capabilities\cite{gao2024retrieval}.
-\subsection{ Differences Between RAG and Fine-Tuning}
-The enhancement of large language models (LLMs) has gained significant interest due to their increasing use.
-Among the various optimization strategies for LLMs, Retrieval-Augmented Generation (RAG) is often compared to fine-tuning (FT) and prompt engineering.
-Each method possesses unique attributes,
-\begin{enumerate}
-	\item Methodological Characteristics :\\
-	RAG is compared to providing a tailored textbook for information retrieval, making it suitable for precise information retrieval tasks.
-It excels in dynamic environments with real-time knowledge updates.
-\\
-	Fine-tuning is likened to a student internalizing knowledge, making it more static and suitable for replicating specific structures, styles, or formats.
-	\item External Knowledge and Adaptation: \\
-	RAG relies on external knowledge sources and allows for high interpretability but may involve higher latency and ethical considerations regarding data retrieval. \\
-	Fine-tuning requires retraining for updates and involves significant computational resources for dataset preparation and training.
-It enables deep customization of the model’s behavior but may struggle with unfamiliar data.
-	\item Performance: \\ RAG consistently outperforms unsupervised fine-tuning in knowledge-intensive tasks, particularly for both previously encountered and new knowledge.
-	LLMs often struggle to learn new factual information through unsupervised fine-tuning.
-	\item Use Cases and Combination: \\
-	The choice between RAG and fine-tuning depends on specific needs for data dynamics, customization, and computational capabilities.
-They are not mutually exclusive; their combined use may yield optimal performance and often requires multiple iterations to refine the results \cite{gao2024retrieval}.
+\end{figure}
+\newline
+\section{Impact of k on Retrieval Performance}
+The retriever's effectiveness in identifying relevant documents depends on k. Key impacts include :
+\subsection{Recall vs. Precision} 
+When a user conducts a search query, the outcomes from the database can be grouped into four distinct types based on relevance and retrieval status \cite{geeksforgeeks2022precision}:
+\begin{itemize}
+	\item {Relevant and Retrieved:} Documents that both address the user’s query and appear in the search results.
+	\item {Relevant and Not Retrieved:} Useful documents that answer the query but are not included in the results.
 
-\end{enumerate}
-\section{Types of RAG Systems}
-The RAG research field is continuously evolving, with three main stages: Naive RAG, Advanced RAG, and Modular RAG.
-\subsection{Naive RAG}
-Naive RAG, a foundational approach in Retrieval-Augmented Generation, operates on a straightforward "Retrieve-Read-Generate" paradigm. This method involves three primary steps:
+	\item {Non-Relevant and Retrieved:} Documents that appear in the results but lack meaningful value for the query.
+
+	\item {Non-Relevant and Not Retrieved:} Irrelevant documents excluded from the results.
+\end{itemize}
+
+\textbf{Precision@k}: evaluates the proportion of relevant documents within the top k retrieved results.
+This metric is particularly valuable in scenarios where the focus is on delivering highly relevant information quickly, rather than ensuring complete coverage such as ,recommendation systems or search engines\cite{deconvoluteai2024metrics}.
+
+\[
+\text{Precision@k} = \frac{\text{Number of Relevant Items Retrieved in Top k}}{k}
+\]
+
+\textbf{Example}
+Suppose we have a dataset of 10 documents. If we retrieve \( k = 4 \) documents and find that 3 of them are relevant to the query:
 
 \begin{itemize}
-	\item Indexing: Raw data is cleaned, extracted, and converted into a uniform text format.
-It's then segmented into smaller chunks and encoded into vector representations using an embedding model.
-These vectors are stored in a vector database for efficient similarity searches.
-	\item Retrieval: When a user query is received, it's encoded into a vector and compared to the indexed chunks.
-The top K most similar chunks are retrieved and included in the prompt for the language model.
-	\item Generation: The query and retrieved chunks are combined into a prompt, which is fed to a large language model.
-The model generates a response based on the provided context and its internal knowledge.
+	\item \textbf{Dataset}: [doc1, doc2, doc3, doc4, doc5, doc6, doc7, doc8, doc9, doc10]
+	\item \textbf{Retrieved}: [doc3, doc1, doc7, doc4]
+	\item \textbf{Relevant}: [doc1, doc3, doc5, doc8]
 \end{itemize}
-While Naive RAG offers a basic framework, it faces several challenges in 
-\begin{itemize}
-	\item Retrieval Issues: The retrieval process can be imprecise, leading to the selection of irrelevant or missing information.
-	\item Generation Challenges: The language model may hallucinate, generating content not supported by the retrieved context. It may also produce irrelevant, toxic, or biased outputs.
-	\item Augmentation Difficulties: Integrating retrieved information into the generation process can be challenging, leading to disjointed or redundant responses.
-\end{itemize}
-To address these limitations, more sophisticated RAG techniques have emerged, which we will explore in the next section.
-\subsection{Advanced RAG}
-Taking aim at the shortcomings of Naive RAG, Advanced RAG introduces specific improvements to enhance retrieval quality. This approach utilizes pre-retrieval and post-retrieval strategies.
-\subsubsection{Pre-retrieval Strategies:}
-\begin{itemize}
-	\item Enhanced Indexing: Advanced RAG tackles indexing issues through a sliding window approach, finer segmentation of data, and inclusion of metadata.
-Additionally, it optimizes the retrieval process by employing various methods.
-	\item Query Optimization: This stage focuses on refining the user's initial query to make it clearer and more suitable for retrieval.
-Techniques like query rewriting, transformation, and expansion are commonly used.
-\end{itemize}
-\subsubsection{Post-Retrieval Strategies:}
-\begin{itemize}
-	\item Re-ranking Chunks: After relevant information is retrieved, Advanced RAG prioritizes the most relevant content by re-ranking the retrieved chunks and placing them strategically within the prompt.
-	\item Context Compression: To avoid overwhelming the LLM with too much information, post-retrieval efforts focus on selecting the most essential parts of the retrieved context, highlighting critical sections, and compressing the data to be processed.
-\end{itemize}
-By addressing indexing issues and refining the query and retrieved information, Advanced RAG aims to improve the overall accuracy and relevance of the generated response.
-\subsection{Modular RAG} 
-Modular RAG represents the latest evolution in RAG, offering greater adaptability. it introduces specialized modules and innovative patterns to enhance retrieval and processing capabilities.
-\subsubsection{New Modules}
-\begin{itemize}
-	\item Search Module: Adapts to specific scenarios by leveraging LLM-generated code and query languages to search across various data sources.
-	\item RAG Fusion: Employs a multi-query strategy to expand user queries, uncover both explicit and implicit knowledge, and improve retrieval results.
-	\item Memory Module: Leverages the LLM's memory to guide retrieval and create an unbounded memory pool, aligning the text more closely with data distribution.
-	\item Routing Module: Navigates through diverse data sources, selecting the optimal pathway for a query based on its specific needs.
-	\item Predict Module: Reduces redundancy and noise by generating relevant context directly through the LLM.
-	\item Task Adapter Module: Tailors RAG to various downstream tasks by automating prompt retrieval and creating task-specific retrievers.
-\end{itemize}
-\subsubsection{New Patterns} 
-\begin{itemize}
-	\item Flexible Module Arrangement: Modular RAG allows for the substitution and reconfiguration of modules to address specific challenges, surpassing the fixed structures of previous RAG paradigms.
-	\item Innovative Retrieval Strategies: Techniques like Rewrite-Retrieve-Read, Generate-Read, and Recite-Read leverage the LLM's capabilities to refine queries, generate content, and retrieve information from model weights.
-	\item Hybrid Retrieval: Combines keyword, semantic, and vector searches to cater to diverse queries, improving retrieval relevance.
-	\item Dynamic Module Interaction: Frameworks like Demonstrate-Search-Predict and ITERRETGEN demonstrate the dynamic use of module outputs to enhance each other's functionality.
-	\item Adaptive Retrieval: Techniques like FLARE and Self-RAG evaluate the necessity of retrieval based on different scenarios, allowing for a more flexible and efficient approach.
 
-\end{itemize}
-\section{Core Components of RAG}
-The core components of Retrieval-Augmented Generation (RAG) systems consist of a retrieval mechanism, a generation process, and augmentation techniques.
-These elements work together to enhance the model's ability to access relevant external information, generate coherent and contextually appropriate responses, and improve overall performance in knowledge-intensive tasks.
-\subsection{ Retrieval Mechanism}
-RAG systems combine parametric memory (a pre-trained language model) with non-parametric memory (a retrieval mechanism).
-The retrieval mechanism allows RAG models to access external information sources (e.g., Wikipedia), and this process is central to improving the model's ability to generate factual and accurate outputs \cite{lewis2020retrieval}.
-\begin{itemize}
-	\item Traditional Techniques: \\
-	Classical retrieval methods include \textbf{TF-IDF and BM25}, which rely on sparse vector representations based on term frequencies.
-These methods use exact keyword matching, making them limited in semantic understanding.
-	\item Modern Retrieval Approaches:\\
-	\textbf{ Dense Passage Retrieval (DPR): }This approach utilizes dense vector representations learned by neural models like BERT to encode both the query and document.
-It allows for a more semantic understanding of the content, making retrieval more effective.
-DPR, for example, computes the similarity between the query and documents using Maximum Inner Product Search (MIPS), which finds the closest passages based on the dense vector space.
-	\item Implementation in RAG:\\ In RAG, dense retrieval methods are often paired with techniques like \textbf{Maximum Inner Product Search} (MIPS), which efficiently matches query and document embeddings.
-This enables the system to return the most relevant documents for subsequent generation \cite{karpukhin2020dense}.
+The Precision@4 score would be:
 
-\end{itemize}
-\subsection{Generation Process}
-The generation in RAG occurs through a combination of retrieved passages and the input query.\\
-Large Language Models (LLMs) like BART or T5 are utilized for this generation, leveraging their advanced capabilities in understanding and producing human-like text.
-The retrieval component supplies external context, which the generator conditions on to formulate a response.This context is crucial, as it enriches the LLM's understanding, enabling it to incorporate real-time, relevant information.
-Moreover, the generation process benefits from the LLM's ability to synthesize information, allowing it to create responses that are not only coherent but also informed by the latest data, thus improving accuracy and relevance in knowledge-intensive tasks \cite{lewis2020retrieval}.
+\[
+\text{Precision@4} = \frac{3}{4} = 0.75
+\]
+
+\textbf{Recall@k :}evaluates the proportion of relevant documents that are successfully retrieved within the top k results.
+This metric is particularly important in contexts where ensuring the completeness of information is crucial, such as in medical research or academic tools, where omitting relevant documents could result in incomplete or inaccurate conclusions\cite{deconvoluteai2024metrics}.
+
+\[
+\text{Recall@k} = \frac{\text{Number of Relevant Items Retrieved in Top k}}{\text{Total Number of Relevant Items}}
+\]
+\newline
+\textbf{Example}
+Consider a dataset of 10 documents. If we retrieve \( k = 4 \) documents and find that 2 of them are relevant, while the total number of relevant documents in the dataset is 4:
+
+\[
+\text{Recall@4} = \frac{2}{4} = 0.5
+\]
+
+Increasing k typically enhances recall by retrieving more relevant documents but may decrease precision due to the inclusion of irrelevant ones.
+Conversely, decreasing k can improve precision but at the cost of lower recall.
+\subsection{Retrieval Speed and Computational Cost}
+Retrieval speed and computational cost are two important areas where k has a big influence Below is a detailed explanation of how higher k values affect these aspects.
+
+\textbf{Increased Computational Resources:} \\
+As the value of k expands, the retrieval system must handle a larger set of documents, leading to higher computational demands.
+Specifically, the system needs to perform additional operations like ranking, filtering, and similarity scoring to identify the most relevant documents.
+These tasks become increasingly resource-intensive, particularly in large-scale retrieval settings where the document corpus consists of millions or even billions of entries\cite{manning2008ir}.
+For instance, retrieving the top  documents (k=80) requires significantly more computational resources compared to retrieving only the top 10 documents (k=5).
+
+\textbf{Higher Memory Usage:}
+As the number of retrieved documents (k) increases ,Storing and processing a larger set of retrieved documents requires more memory This can become a significant bottleneck, especially in environments with constrained memory resources .
+
+For large-scale retrieval tasks, where datasets may contain millions or even billions of documents, the memory demand grows proportionally with 
+k.
+Each retrieved document needs to be stored temporarily, along with its associated metadata, such as embedding vectors, BM25 scores, or other relevance signals.
+Additionally, sorting and filtering operations further contribute to memory overhead
+\subsection{Document Ranking Quality}
+Document ranking in information retrieval involves ordering documents by their relevance to a user's query.
+The objective is to prioritize the most relevant documents at the top of search results, making it easier for users to access useful information quickly.
+Different models,Vector Space, including Boolean, and Probabilistic models, are used to establish this ranking \cite{enwiki:1262179867}.\\
+
+Increasing the number of retrieved documents,represented as k, may result in the addition of lower-ranked, less relevant documents, potentially diluting the overall quality of the retrieved information.
+This occurs because of the inherent balance between precision and recall in information retrieval systems.
+\begin{figure}[h]
+	\centering
+	\includegraphics[width=0.7\linewidth]{Figures/precisionR.png}
+	\caption{Precision in Document Ranking}
+	\label{precisionR}
+
+\end{figure}
+
+\begin{figure}[h]
+	\centering
+	\includegraphics[width=0.7\linewidth]{Figures/recalR.png}
+	\caption{Recall in Document Ranking}
+	\label{recalR}
+
+\end{figure}
+As k grows, recall generally improves since a larger number of relevant documents are likely to be retrieved as shown in Figure \ref{recalR}.
+However, this often comes at the expense of precision, as the additional documents retrieved may include non-relevant ones,thereby lowering the overall precision as illustrated in Figure \ref{precisionR}.
+This trade-off is a core principle in evaluating information retrieval systems.
+
+\section{Impact of k on Generation Quality}
+The impact of k (the number of retrieved documents or generated candidates) on generation quality is A crucial factor in many machine learning and information retrieval tasks, including text generation, recommendation systems, and search engines .
+
+\subsection{Trade-off Between Diversity and Relevance} 
+Higher k: Increasing k often improves diversity in generated outputs or retrieved documents, as more candidates are considered.
+However, this can lead to a drop in relevance or quality, as lower-ranked candidates may be less accurate or coherent.
 \\
-Two models have been proposed in RAG:
-\begin{itemize}
-	\item \textbf{RAG-Sequence} uses the same document to generate the entire sequence of output. It marginalizes over the top K retrieved documents to produce a final answer.
-	\item \textbf{RAG-Token} allows different tokens in the output sequence to be generated based on different documents, making it more flexible when combining information from various
-\end{itemize}
-\subsection{ Augmentation Techniques}
-The augmentation of the retrieval process in Retrieval-Augmented Generation (RAG) systems focuses on improving how queries are refined and how relevant information is retrieved for downstream generation tasks.
-Key augmentation methods include.
-\begin{itemize}
-	\item \textbf{Query Augmentation:} In traditional retrieval pipelines, user queries are often under-specified or ambiguous, leading to poor retrieval performance.
-Query augmentation involves dynamically rewriting the user's query to better match the documents in the knowledge base.
-This can be done by leveraging large language models (LLMs) to generate tailored queries or synthetic questions and answers (QAs) that better align with the search objective​
-	\item \textbf{Synthetic QA Generation:} Instead of using raw document chunks, retrieval is enhanced by generating and embedding synthetic QA pairs from documents.
-This helps to capture the semantic essence of long texts more effectively, reducing noise and improving retrieval precision.
-These synthetic QAs can be used to rewrite the user query, making it more specific to the task at hand \cite{mombaerts2024meta}.
-\end{itemize}
-\begin{figure}
+Lower k: A smaller k tends to prioritize high-quality, relevant outputs but may lack diversity, leading to repetitive or overly narrow results.
+
+Research shows that the quality of retrieved documents plays a crucial role in the performance of Retrieval-Augmented Generation (RAG) systems.
+For example, one study demonstrated that the precision of retrieved documents directly affects the factual correctness of the generated responses \cite{salemi2023evaluating}
+Additionally, another study revealed that simply increasing the number of documents does not necessarily improve generation quality,especially if the additional documents are not highly relevant\cite{wan2025cognitivealigneddocumentselectionretrievalaugmented}
+\subsection {Effect on Text Generation Models}
+In text generation tasks, such as machine translation and dialogue systems, The parameter k frequently refers to the beam width in beam search algorithms\cite{freitag-al-onaizan-2017-beam}.
+Beam search is a heuristic search algorithm that expands the most promising nodes in a graph to maximize the quality of the text that is produced.
+Adjusting the beam width (k) has significantly impacts how well text generation models function.
+
+As the beam width is increased, the model must process and retain more candidate sequences concurrently, leading to higher computational and memory demands Particularly in real-time applications where latency is crucial, this increase may have an effect on system performance and response time\cite{tensorrt_llm_beam_search}.
+
+\section{Existing Solutions for k Selection}
+Selecting an optimalnk plays a crucial role in balancing retrieval effectiveness and generation quality.Over the years, various strategies have been proposed to determine k, ranging from static selection to dynamic and hybrid approaches.
+\subsection{Static k Selection}
+In Retrieval-Augmented Generation (RAG) systems, "static k selection" means, refers to the process of setting a fixed number of top documents (k) to retrieve for every query, irrespective of how simple or complex the query might be.
+This approach simplifies the retrieval process by maintaining a consistent retrieval count across all queries
+
+\textbf{ Sparse Retrieval with Fixed k} \\
+Sparse retrieval is a method of finding relevant documents from a large collection by representing both queries and documents as vectors where most values are zero.
+This focus on only the most important terms leads to faster, more accurate searches, especially useful when combining information from different sources \cite{zheng2024enhancing}.
+
+Common approaches include \textbf{BM25}\cite{10.1561/1500000019} where documents are scored for relevance by considering term frequency and inverse document frequency,it selects the 'k' documents with the highest BM25 scores for each query,\textbf{TF-IDF} (Term Frequency-Inverse Document Frequency)\cite{gfg2025tfidf}in this method the retrieved documents would be those with the highest weighted term importance, Other methods like \textbf{QLM }(Query Likelihood Model)\cite{10.1007/978-3-030-72240-1_49}use probabilistic models to evaluate the likelihood of a query and rank documents according to textual similarity.
+
+\textbf{Dense Retrieval with Fixed k} \\
+Dense retrieval \cite{karpukhin2020dense} is a method for retrieving information that uses deep learning models to convert documents and queries into high-dimensional vector embeddings,A fixed number of top-k documents are selected based on similarity scores between the query embedding and document embeddings in a continuous vector space, the retrieval process allows the system to capture semantic elationships beyond exact term matches.
+
+Methods such as dual-encoder architectures (e.g., DPR - Dense Passage Retrieval)\cite{chen-etal-2024-dense} utilize separate neural networks (encoders) for queries and documents enable efficient retrieval, while Approximate Nearest Neighbor (ANN) search techniques (e.g., FAISS)\cite{enwiki:1276232158} optimize search efficiency in large-scale datasets .
+\subsection{Dynamic k Selection}
+In retrieval, dynamic k selection is the process of varying the quantity of documents (k) that are retrieved according to the properties of the query, such as its ambiguity, complexity, or relevance score distribution.
+This approach is crucial for enhancing the effectiveness and efficiency of information retrieval systems.
+
+\textbf{Dynamic Trade-Off Prediction in Multi-Stage Retrieval Systems }\\ This approach predicts the optimal number of documents (k) to retrieve.
+It uses pre-retrieval features to balance the trade-off between retrieval efficiency and effectiveness, ensuring that the system retrieves an appropriate number of documents for each query\cite{culpepper2016dynamictradeoffpredictionmultistage}.
+
+\textbf{Dynamic Pruning Methods:}\\
+This approach addresses the challenge of balancing effectiveness and efficiency in large-scale Information Retrieval (IR) systems, particularly under temporal constraints.
+The goal is to process queries within a specified time limit while minimizing the loss in retrieval effectiveness.
+The authors propose and evaluate three techniques for temporally constrained top-K query processing \cite{10.1007/978-3-319-70145-5_1}.
+\subsection{Hybrid k Selection}
+hybrid methods aim to achieve optimal retrieval By merging static and dynamic k selection, These methods balance computational efficiency (from static k) with adaptive flexibility (from dynamic k) ensuring high-quality retrieval Below are some key hybrid strategies.
+
+\textbf{Blended RAG} \\
+An innovative method to improve Retrieval-Augmented Generation (RAG) systems, which integrate private document collections with Large Language Models (LLMs) for Generative Question-Answering (QA) it employs a hybrid retrieval approach, combining Dense Vector indexes and Sparse Encoder indexes with sophisticated query strategies, Blended RAG provides a scalable and efficient solution for enhancing RAG systems, showcasing the effectiveness of merging dense and sparse retrieval techniques\cite{sawarkar2024blendedragimprovingrag}.
+
+\textbf{ STAYKATE (Static-Dynamic Hybrid Selection)} \\
+A new approach for selecting in-context examples to improve the performance of large language models (LLMs) in scientific information extraction.
+Scientific tasks often struggle with limited training data and expensive annotation processes.
+STAYKATE tackles these challenges by merging static and dynamic selection strategies, combining representativeness sampling from active learning with retrieval-based methods.
+This hybrid approach ensures the selection of high-quality, informative examples, enhancing in-context learning for LLMs\cite{zhu2024staykatehybridincontextexample}.
+\section{ Proposed Solution}
+While existing k-selection approaches—static, dynamic, and hybrid—offer distinct advantages, they also present notable limitations, such as lack of adaptability in static k selection, which fails to adjust for query complexity, leading to either missing relevant documents or retrieving irrelevant ones.
+Dynamic k selection, while more flexible, introduces higher computational costs and requires carefully tuned heuristics, making it challenging to scale.
+Hybrid approaches, attempt to balance both strategies but suffer from increased system complexity.
+
+To address these challenges, we propose an enhanced k-selection strategy that optimally balances retrieval efficiency and relevance, leveraging adaptive mechanisms to improve performance across diverse query types
+\subsection{Mixture of Logits (MoL)}
+\begin{figure}[h]
 	\centering
-	\includegraphics[width=0.9\linewidth]{Figures/rag_architectur.png}
-	\caption{Rag architecture}
-	\label{rag_architectur.png}
+	\includegraphics[width=0.7\linewidth]{Figures/mol.png}
+	\caption{Mixture of Logits(MoL) learned similarity.}
+	\label{Mixture_of_Logits }	
 \end{figure}
-\section{Training and Fine-Tuning RAG Models}
-Training and fine-tuning are essential steps in enhancing the performance of RAG models by optimizing both the retriever and generator components leading to more coherent, relevant, and accurate responses.
-\subsection{Training the Retriever}
-A powerful technique for improving retrieval relevance involves training the retriever model using \textbf{contrastive learning}.
-This approach aims to maximize the similarity between query and relevant document embeddings while minimizing the similarity between query and irrelevant document embeddings.
-By training the model on numerous positive (relevant) and negative (irrelevant) pairs, the retriever learns to effectively distinguish between relevant and irrelevant information\cite{sbert2024}.
-\section{Task and Evaluation}
-The rapid development and increasing use of Retrieval-Augmented Generation (RAG) models in NLP have made their evaluation a critical area of research within the LLM community.
-This section covers the primary downstream tasks associated with RAG, the datasets used, and the methods for evaluating RAG systems \cite{zhou2020trustworthiness}.
-\subsection{Factuality}
-\begin{itemize}
-	\item Fact-Checking Benchmarks: Using datasets like FEVER and SQuAD to evaluate the model's ability to identify and correct factual errors.
-	\item Adversarial Testing: Creating adversarial examples to test the model's robustness against misleading information.
-	\item Contextual Understanding: Assessing the model's ability to understand the context of a query and provide accurate answers.
-\end{itemize}
-\subsection{ Robustness}
-\begin{itemize}
-	\item Noise Injection: Introducing noise into the retrieved documents to test the model's ability to handle imperfect information.
-	\item Adversarial Attacks: Evaluating the model's resilience to attacks that aim to manipulate its outputs.
-	\item Domain Adaptation: Testing the model's ability to adapt to new domains and data distributions.
-\end{itemize}
-\subsection{Fairness}
-\begin{itemize}
+The \textbf{Mixture of Logits (MoL)}\cite{zhai2023revisiting}\cite{Ding2024} approach is designed to enhance retrieval and ranking by leveraging multiple low-rank embeddings.
+It assumes that both the \textbf{query} (\( q \)) and \textbf{document/item} (\( x \)) are mapped into \( P \) groups of low-dimensional embeddings, denoted as \( f_p(q) \) and \( g_p(x) \), which are generated by neural networks based on their respective features.
 
-	\item Bias Detection: Identifying and quantifying biases in the training data and model outputs.
-	\item Fairness Metrics: Using metrics like demographic parity and equalized odds to evaluate the model's fairness.
-	\item Mitigation Techniques: Implementing techniques like debiasing and fairness constraints to mitigate biases.
-\end{itemize}
-\subsection{Objective Metrics:}
-\begin{itemize}
-	\item Accuracy: Precision, recall, and F1-score for evaluating the correctness of the generated responses.
-	\item Consistency: Measuring the consistency of the model's outputs across different queries and contexts.
-	\item Coherence: Assessing the coherence and fluency of the generated text.
-\end{itemize}
-\subsection{Subjective Metrics}
-\begin{itemize}
-	\item Human Evaluation: Using human raters to evaluate the quality of the generated responses.
-	\item User Studies: Conducting user studies to gather feedback on the user experience
-\end{itemize}
-\section{Limitations}
-While RAG has gained significant traction across diverse applications, it still faces certain limitations in terms of effectiveness and efficiency\cite{zhao2024retrieval}.
-\subsection{Noisy Retrieval Results}
-\begin{itemize}
-	\item Retrieval Quality: The quality of retrieved information can be affected by factors like indexing techniques, query formulation, and the underlying dataset.
-	\item Hallucinations: Noisy or irrelevant information can lead to the generation of hallucinated or factually incorrect responses.
-	\item Contextual Understanding: The LLM may struggle to understand the context of the retrieved information, especially if it's poorly formatted or contains inconsistencies.
-\end{itemize}
-\subsection{Extra Overhead}
-\begin{itemize}
-	\item Computational Cost: Retrieval and processing additional information can increase computational costs, especially for large-scale models and complex queries.
-	\item Latency: Retrieval and processing can introduce latency, impacting the real-time performance of RAG systems.
-	\item Complexity: Implementing and deploying RAG systems requires careful consideration of various factors, including data preparation, model selection, and system architecture.
-\end{itemize}
-\subsection{Interaction of Retrieval and Generation}
-\begin{itemize}
-	\item Aligning the goals of the retriever and generator is challenging.
-	\item Optimizing the interaction between the two components requires careful design and tuning.
-	\item The impact of various factors, such as metric selection and hyperparameter tuning, on RAG performance is still not fully understood.
-	\item
+To determine the similarity between a query and a document, MoL assigns \textbf{adaptive gating weights} \( \pi_p(q,x) \) to the inner products of these low-rank embeddings\cite{zhai2023revisiting}:
 
-\end{itemize}
-\subsection{Long Context Generation:}
+\begin{equation}
+	\phi(q,x) = \sum_{p=1}^{P} \pi_p(q,x) \langle f_p(q), g_p(x) \rangle
+\end{equation}
+
+where \( \pi_p(q,x) \) represents the learned weights for each component, ensuring that their sum equals one.
+These weights are typically parameterized using a neural network that takes embeddings and their inner products as input features.
+
+To efficiently scale MoL for large datasets and hardware-optimized implementations, the formulation is extended by decomposing the dot products into \textbf{batched outer products} of query-side and document-side embeddings.
+This decomposition improves computational efficiency, particularly on accelerators like GPUs, by normalizing the embeddings using the \( l_2 \)-norm:\cite{zhai2023revisiting}
+
+\begin{equation}
+	\phi(q,x) = \sum_{pq=1}^{P_q} \sum_{px=1}^{P_x} \pi_{pq,px}(q,x) \frac{\langle f_{pq}(q), g_{px}(x) \rangle}{|| f_{pq}(q) ||_2 || g_{px}(x) ||_2}
+\end{equation}
+
+Since embedding normalization can be precomputed, both formulations remain interchangeable in practical applications.
+Furthermore, it is possible to \textbf{decompose any high-rank matrix} into a mixture of logits based on low-rank matrices, demonstrating the flexibility and scalability of this approach in large-scale information retrieval tasks.
+\subsection{Algorithm Design}
+
+we introduce an adaptive threshold mechanism
+(Dynamic Candidate Selection), the MoL framework is employed to refine the
+candidate retrieval process. 
+1. \textbf{Component-Level Embeddings}:  
+Component-level embeddings are generated for all items in the dataset \(X\). These embeddings facilitate efficient similarity computations during retrieval. Formally,
+\begin{equation}
+	X_p \leftarrow \{g_p(x) \mid x \in X\}.
+\end{equation}
+
+\noindent
+2. \textbf{Dynamic Threshold Adjustment}:  
+To improve retrieval quality, \textbf{Mixture of Logits (MoL)} scores are computed for each candidate \(x \in G\).
+The adaptive gating weights \(\pi_p(q, x)\) allow the algorithm to dynamically adjust the retrieval threshold \(T_{\text{adaptive}}\) based on the MoL scores.
+The scoring function is defined as:
+\begin{equation}
+	\phi(q, x) = \sum_{p=1}^{P} \pi_p(q, x) \cdot \langle f_p(q), g_p(x) \rangle.
+\end{equation}
+The adaptive threshold \(T_{\text{adaptive}}\) is set as the minimum score among the candidates:
+\begin{equation}
+	T_{\text{adaptive}} = \min\{s \mid s \in G\}.
+\end{equation}
+
+\noindent
+3. \textbf{Refinement and Top-K Selection}:  
+Using the adaptive threshold \(T_{\text{adaptive}}\), additional relevant candidates are retrieved, expanding the candidate set \(G'\).
+This is achieved by including candidates whose scores exceed the threshold:
+\begin{equation}
+	G' \leftarrow G \cup \{x \mid s_p \geq T_{\text{adaptive}}\}.
+\end{equation}
+The algorithm then sorts \(G'\) based on MoL scores to select the most relevant top-k candidates.
+
+\noindent
+4. \textbf{Exact Top-K Selection}:  
+Finally, the candidates in \(G'\) are sorted by their MoL scores, and the exact top-k items are extracted:
+\begin{equation}
+	G_{\text{final}} = \text{Top-k}(G', \phi(q, x)).
+\end{equation}
+
+\begin{figure}[ht]
+	\centering
+	\includegraphics[width=0.7\linewidth]{Figures/propused _solution_steps.png}
+	\caption{propused solution steps}
+	\label{propused _solution_steps}	
+\end{figure}
+
+\subsection{Pseudocode}
+
+\textbf{Input:} Query $q$, Set of items $X$, Component-level embeddings: $f_p(q)$, $g_p(x)$ for $p \in P$, $x \in X$, Initial threshold $T_{\text{init}}$ \\
+\textbf{Output:} Exact top $k$ items based on dynamic threshold selection, $G_{\text{final}}$ \\
+\SetKwInOut{Input}{Input}
+\SetKwInOut{Output}{Output}
+
+\begin{algorithm}[H]
+	\small
+
+	\caption{Hybrid Exact Top-k with Threshold-Based k Selection}
+	\label{ALgo1}
+	\SetKwProg{Init}{Initialization}{}{}
+	\Init{
+		$G \gets \emptyset$ \tcp{Initial candidate set}
+		\ForEach{component $p \in P$}{
+			$X_p \gets \{g_p(x) \mid x \in X\}$ \tcp{Precompute embeddings}
+		}
+	}
+
+	\textbf{1. Initial Candidate Retrieval:} \\
+	\ForEach{component $p \in P$}{
+		Compute dot product scores: 
+		\[
+		S_p \gets \{ \langle f_p(q), g_p(x) \rangle : x \in X_p \}
+		\]
+		Retrieve items with scores $S_p \geq T_{\text{init}}$ \\
+		Add these items to $G$
+	}
+
+	\textbf{2. Adjust k Dynamically:} \\
+	\ForEach{$x \in G$}{
+		Compute MoL scores $s \gets \phi(q, x)$ where:
+		\[
+		\phi(q, x) \gets \sum_{p=1}^{P} \pi_p(q, x) \cdot \langle f_p(q), g_p(x) \rangle
+		\]
+		Set $T_{\text{adaptive}} = \min \{s : s \in G\}$
+	}
+
+	\BlankLine
+	\textbf{3. Refine Candidate Set with Adaptive k:} \\
+	$G' \gets \emptyset$ \\
+	\ForEach{component $p \in P$}{
+		Retrieve items from $X_p$ with scores $S_p \geq T_{\text{adaptive}}$ \\
+		Add these items to $G'$
+	}
+
+	\textbf{4. Select Exact Top-k Items:} \\
+	\ForEach{component $p \in P$}{
+		Compute MoL scores for all items in $G'$ \\
+		Sort $G'$ by MoL scores in descending order \\
+		Select the top $k$ items from $G'$ where $k$ is the number of items in $G'$ exceeding $T_{\text{adaptive}}$
+	}
+
+	\textbf{Return:} $G_{\text{final}}$ \tcp{Retrieve Top $k$ items from $G'$}
+\end{algorithm}
+\section{Recommendation System Overview}
+Modern technology and online services have enabled unprecedented access to vast amounts of data.
+However, this abundance of information creates an overload, making it harder for users to find relevant content efficiently.
+Recommender systems address this by filtering information and delivering personalized suggestions, saving users time and effortz\cite{Roy2022}.
+These systems are now integral to platforms like e-commerce, television programs\cite{5174476}, e-learning\cite{WANG201110831}, tourism, and more, though further improvements are needed to enhance their versatility and accuracy.
+\begin{figure}[ht]
+	\centering
+	\includegraphics[width=0.5\linewidth]{Figures/RS.png}
+	\caption{Recommendation System Process}
+	\label{Recommendation_System _Process}	
+\end{figure}
+\subsection{SASRec: Self-Attentive Sequential Recommendation}
+Sequential recommendation systems aim to predict a user’s next interaction based on their historical behavior while incorporating contextual information from recent actions.
+However, effectively capturing patterns in sequential data is challenging due to the exponential growth of the input space as more past interactions are considered.\cite{kang2018selfat}
+\subsection{SASRec Model Architecture}
+SASRec leverages self-attention to assign adaptive weights to past items at each time step. The key components include:\\
+\textbf{ Embedding Layer:} Converts user interactions into dense vector representations. \\
+\textbf{ Self-Attention Layer: }Captures dependencies between different interactions in the sequence, allowing for long-range modeling.\\
+\textbf{ Point-Wise Feed Forward Network (FFN)}: Enhances feature extraction for better predictions.\\
+\textbf{Prediction Layer:} Computes the likelihood of the next interaction based on learned patterns.\\
+A visual representation of SASRec’s training process (Figure \ref{the_training_process_of_SASRec}) illustrates how the model uses self-attention to focus on relevant past interactions when making predictions
+\begin{figure}[ht]
+	\centering
+	\includegraphics[width=0.6\linewidth]{Figures/sasrec.png}
+	\caption{the training process of SASRec}
+	\label{the_training_process_of_SASRec}	
+\end{figure}
+\section{Evaluation Metrics}
+We employ standard ranking metrics widely used in sequential recommendation:
 
 \begin{itemize}
-	\item Context Length Limits: LLMs have limitations on the amount of context they can process at once.
-	\item Information Loss: Long documents may be truncated or summarized, leading to loss of important information.
-	\item Computational Cost: Processing long contexts can be computationally expensive.
+	\item \textbf{Hit Rate at \( k \) (HR@\( k \))}: Measures the proportion of cases where the target item appears in the top-\( k \) recommendations. The HR@\( k \) is computed as:\cite{Tamm_2021}
+	\begin{equation}
+		\text{HR@}k = \frac{1}{|U|} \sum_{u=1}^{|U|} \mathbb{I}(\text{rank}_u \leq k),
+	\end{equation}
+	where \( |U| \) is the number of users, \( \text{rank}_u \) is the rank of the target item for user \( u \), and \( \mathbb{I}(\cdot) \) is the indicator function that returns 1 if the condition is true and 0 otherwise.
+
+	\item \textbf{Mean Reciprocal Rank (MRR)}:is a ranking quality metric that measures how quickly a system retrieves the first relevant item.
+It is calculated as the average of reciprocal ranks across all users or queries,MRR ranges from 0 to 1, with higher values indicating better performance \cite{eviden2025mrr}.
+	\begin{equation}
+		\text{MRR} = \frac{1}{|U|} \sum_{u=1}^{|U|} \frac{1}{\text{rank}_u},
+	\end{equation}
+	where \( \text{rank}_u \) is the position of the first relevant item for user \( u \) within the top-K results. \\ 
+	U represents the total number of users (for recommendation systems) or queries (for information retrieval tasks) in the dataset.
+
+	\item \textbf{Normalized Discounted Cumulative Gain (NDCG)}: Assesses the ranking quality while accounting for position importance. The NDCG@\( k \) is computed as:\cite{Tamm_2021}
+	\begin{equation}
+		\text{NDCG@}k = \frac{1}{|U|} \sum_{u=1}^{|U|} \frac{\text{DCG@}k}{\text{IDCG@}k},
+	\end{equation}
+	where \( \text{DCG@}k \) is the Discounted Cumulative Gain at position \( k \):
+	\begin{equation}
+		\text{DCG@}k = \sum_{i=1}^{k} \frac{2^{\text{rel}_i} - 1}{\log_2(i + 1)},
+	\end{equation}
+	and \( \text{IDCG@}k \) is the Ideal DCG@\( k \), computed by sorting the items by their true relevance scores.
 \end{itemize}  
-\newpage
-\section{Conclusion}
-This chapter provided an overview of Retrieval-Augmented Generation (RAG), highlighting its significance in natural language processing.
-We discussed the key concepts of retrieval and generation, emphasizing the advantages of their combination in enhancing language model capabilities.
-The historical development of RAG was explored, along with its core components, including the retrieval mechanism, generation process, and augmentation techniques.
-Additionally, we examined various downstream tasks and evaluation targets, illustrating RAG's versatility in applications like open-domain question answering and fact verification.
-Overall, RAG represents a promising advancement in knowledge-intensive tasks, paving the way for further research and innovation in the field.
+Table \ref{tab:results_tab} summarizes the performance of our hybrid algorithm compared to the MoL-based approach on the MovieLens 100K and 1M datasets, both tested using a Max Sequence Length of 50, an Embedding Dimension of 128, 2 Attention Heads, a Feedforward Dimension of 128, a Batch Size of 128, and trained for 10 epochs
+\section{ Conclusion }
+The selection of k in retrieval plays a pivotal role in balancing relevance, efficiency, and computational cost.
+Throughout this chapter, we explored how different approaches—static, dynamic, and hybrid—impact retrieval performance and generation quality.
+While static selection provides consistency, it lacks adaptability to varying query complexities.
+Dynamic methods introduce flexibility but come with computational overhead, whereas hybrid strategies aim to balance both.
+The Hybrid Exact Top-k with Threshold-Based k Selection method offers an advanced solution by leveraging adaptive weighting to enhance ranking effectiveness.
+Ultimately, an effective k-selection strategy is essential for optimizing retrieval systems, ensuring both efficiency and high-quality results in knowledge-augmented applications.

### 2025-04-07 Thesis/Latex/General introduction.tex
update

\chapter*{General introduction}
\pagestyle{fancy}\lhead{\textbf \footnotesize\it{General introduction}}
\pagestyle{fancy}\chead{} \pagestyle{fancy}\rhead{}
\pagestyle{fancy}\lfoot{\textbf {\small\it{Univ-MascaraComputer Science: 2025}}}
\pagestyle{fancy}\cfoot{} \pagestyle{fancy}\rfoot{\thepage}
\pagenumbering{arabic}
\addcontentsline{toc}{chapter}{General introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\textbf{Context :}
In recent years, the legal domain has witnessed a growing interest in the application of advanced natural language processing techniques to improve access to legal information and automate reasoning over juridical data. Legal texts—such as statutes, court decisions, and regulatory documents—are often dense, context-dependent, and linguistically complex, making them challenging to interpret both for non-experts and legal professionals. This complexity is further amplified in multilingual and region-specific legal systems, such as Algeria's, where access to structured and searchable legal data remains limited.

Large Language Models (LLMs) have emerged as powerful tools for understanding and generating human-like text, showing promise in tasks such as legal question answering, case summarization, and document classification. However, their reliance on static training data and the inherent difficulty in capturing fine-grained, up-to-date legal knowledge limits their effectiveness in dynamic and high-stakes domains like law. To bridge this gap, Retrieval-Augmented Generation (RAG) architectures have been proposed, combining the linguistic capabilities of LLMs with external knowledge retrieval systems. This hybrid approach enables models to access and reason over relevant documents at inference time, offering a more grounded and contextually aware output.

Despite these advances, several challenges persist—particularly in the selection of the most relevant documents from large legal corpora. Poor retrieval quality can significantly degrade the performance and reliability of RAG-based systems. Therefore, refining the retrieval process remains a crucial area of research, especially in low-resource and domain-specific contexts such as Arabic legal data in Algeria.\\
\textbf{Problem Statement :} 
While LLMs and RAG architectures provide a strong foundation for legal AI applications, their effectiveness is often undermined by limitations in the retrieval process. The selection of irrelevant or marginally related documents can lead to hallucinated or misleading responses, which is especially problematic in legal reasoning tasks. In the Algerian legal context, where structured resources are scarce and many legal documents exist only in unstructured Arabic text, the challenge of effective document retrieval becomes even more critical. This thesis addresses the specific problem of optimizing the retrieval step in RAG pipelines by introducing a more adaptive and context-sensitive document selection mechanism tailored to the legal domain.\\
\textbf{Objectives:} 
This research aims to explore and improve the integration of LLMs and retrieval systems in the legal field, with a focus on the Algerian legal context. The main objectives are as follows:
\begin{itemize}
	\item To investigate the potential of LLMs and Retrieval-Augmented Generation models for processing Arabic legal texts .
	\item To identify and address challenges in document retrieval, particularly the issue of selecting the optimal number of candidate documents (the k-selection problem).
	\item To develop and share a curated dataset of Arabic legal case texts from Algeria, contributing to resources in low-resource legal NLP.
\end{itemize}
\textbf{Contributions :} This thesis makes the following key contributions:
\begin{itemize}
	\item A domain-specific chatbot designed to assist users in navigating Algerian legal case information using natural language queries.
	\item The creation of a new dataset of Arabic-language  legal cases from Algeria, filling a gap in publicly available legal resources for this jurisdiction.
	\item A Dynamic Candidate Selection mechanism, which improves the retrieval phase of RAG by adaptively selecting relevant documents based on contextual thresholds.
\end{itemize}
\textbf{Structure :} 
The thesis is organized into four main chapters: 
\begin{itemize}
	\item Chapter 1: Large Language Models – An overview of LLMs and their capabilities, with a focus on applications in legal NLP.

	\item Chapter 2: Retrieval-Augmented Generation – A detailed examination of RAG architectures, their strengths, and limitations in legal contexts.

	\item Chapter 3: Candidate Selection in Retrieval – The chapter reviews existing k selection methods, followed by a proposed dynamic selection method to improve retrieval efficiency and generation quality.

	\item Chapter 4: Experimental Results – Summarizes the dataset, experimental setup, and key findings, with analysis of hyperparameter effects and system performance.
\end{itemize}
%%%\newpage

### 2025-04-14 Thesis/Latex/Chapter_3.tex
update

@@
 \subsection{Algorithm Design}
-
 we introduce an adaptive threshold mechanism
 (Dynamic Candidate Selection), the MoL framework is employed to refine the
-candidate retrieval process. 
+candidate retrieval process. \\
+\begin{figure}[ht]
+	\centering
+	\includegraphics[width=0.7\linewidth]{Figures/propused _solution_steps.png}
+	\caption{propused solution steps}
+	\label{propused _solution_steps}	
+\end{figure}
+\noindent
 1. \textbf{Component-Level Embeddings}:  
@@
 \end{equation}
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[width=0.7\linewidth]{Figures/propused _solution_steps.png}
-	\caption{propused solution steps}
-	\label{propused _solution_steps}	
-\end{figure}
+As shown in Algorithm\ref{ALgo1}, the retrieval process dynamically adjusts the threshold based on MoL scores.
+\textbf{Input:} Query $q$, Set of items $X$, Component-level embeddings: $f_p(q)$, $g_p(x)$ for $p \in P$, $x \in X$, Initial threshold $T_{\text{init}}$ \\
+\textbf{Output:} Exact top $k$ items based on dynamic threshold selection, $G_{\text{final}}$ \\
 
@@
 
-\textbf{Input:} Query $q$, Set of items $X$, Component-level embeddings: $f_p(q)$, $g_p(x)$ for $p \in P$, $x \in X$, Initial threshold $T_{\text{init}}$ \\
-\textbf{Output:} Exact top $k$ items based on dynamic threshold selection, $G_{\text{final}}$ \\
 \SetKwInOut{Input}{Input}
@@
 		Compute dot product scores: 
-		\[
-		S_p \gets \{ \langle f_p(q), g_p(x) \rangle : x \in X_p \}
-		\]
+		with (eq\eqref{eq:initial_retrieval}
 		Retrieve items with scores $S_p \geq T_{\text{init}}$ \\
@@
 	\ForEach{$x \in G$}{
-		Compute MoL scores $s \gets \phi(q, x)$ where:
-		\[
-		\phi(q, x) \gets \sum_{p=1}^{P} \pi_p(q, x) \cdot \langle f_p(q), g_p(x) \rangle
-		\]
+		Compute MoL scores $s \gets \phi(q, x)$  using : eq\eqref{similarity_function}
 		Set $T_{\text{adaptive}} = \min \{s : s \in G\}$
@@
 \end{algorithm}
+
+\newpage
+
 \section{Recommendation System Overview}

### 2025-04-14 Thesis/Latex/General introduction.tex
update

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\textbf{Context :}
+\subsubsection*{Context :}
 In recent years, the legal domain has witnessed a growing interest in the application of advanced natural language processing techniques to improve access to legal information and automate reasoning over juridical data.
 Legal texts—such as statutes, court decisions, and regulatory documents—are often dense, context-dependent, and linguistically complex, making them challenging to interpret both for non-experts and legal professionals.
-This complexity is further amplified in multilingual and region-specific legal systems, such as Algeria's, where access to structured and searchable legal data remains limited.
+This complexity is further amplified in multilingual and region-specific legal systems, such as Algeria's\cite{inproceedings}, where access to structured and searchable legal data remains limited.
 
-Large Language Models (LLMs) have emerged as powerful tools for understanding and generating human-like text, showing promise in tasks such as legal question answering, case summarization, and document classification.
+Large Language Models (LLMs) have emerged as powerful tools for understanding and generating human-like text, showing promise in tasks such as legal question answering, case summarization, and document classification\cite{Naveed2024}.
 However, their reliance on static training data and the inherent difficulty in capturing fine-grained, up-to-date legal knowledge limits their effectiveness in dynamic and high-stakes domains like law.
 To bridge this gap, Retrieval-Augmented Generation (RAG) architectures have been proposed, combining the linguistic capabilities of LLMs with external knowledge retrieval systems.
-This hybrid approach enables models to access and reason over relevant documents at inference time, offering a more grounded and contextually aware output.
+This hybrid approach enables models to access and reason over relevant documents at inference time, offering a more grounded and contextually aware output \cite{lewis2020retrieval}.
 
-Despite these advances, several challenges persist—particularly in the selection of the most relevant documents from large legal corpora.
+Although these advances have been made, several challenges persist—particularly in the selection of the most relevant documents from large legal corpora.
 Poor retrieval quality can significantly degrade the performance and reliability of RAG-based systems.
-Therefore, refining the retrieval process remains a crucial area of research, especially in low-resource and domain-specific contexts such as Arabic legal data in Algeria.\\
-\textbf{Problem Statement :} 
-While LLMs and RAG architectures provide a strong foundation for legal AI applications, their effectiveness is often undermined by limitations in the retrieval process.
-The selection of irrelevant or marginally related documents can lead to hallucinated or misleading responses, which is especially problematic in legal reasoning tasks.
-In the Algerian legal context, where structured resources are scarce and many legal documents exist only in unstructured Arabic text, the challenge of effective document retrieval becomes even more critical.
-This thesis addresses the specific problem of optimizing the retrieval step in RAG pipelines by introducing a more adaptive and context-sensitive document selection mechanism tailored to the legal domain.\\
-\textbf{Objectives:} 
-This research aims to explore and improve the integration of LLMs and retrieval systems in the legal field, with a focus on the Algerian legal context. The main objectives are as follows:
+Therefore, refining the retrieval process remains a crucial area of research, especially in low-resource and domain-specific contexts \cite{Information_Retrieval} such as Arabic legal data in Algeria.
+\subsubsection*{Problem Statement :}
+
+\begin{itemize}
+	\item Large Language Models (LLMs) and Retrieval-Augmented Generation (RAG) architectures offer strong potential for legal AI applications.
+	However, their performance is often constrained by weaknesses in the retrieval process.
+	\item The selection of irrelevant or loosely related documents can result in hallucinated or misleading responses, which is especially critical in the legal domain.
+	\item In the Algerian legal context\cite{inproceedings}, this problem is exacerbated due to:
+	\begin{itemize}
+		\item The scarcity of structured and standardized legal resources.
+		\item The dominance of unstructured legal texts written in Arabic, often lacking annotations or metadata.
+	\end{itemize}
+\end{itemize}
+ These limitations make the retrieval step a key bottleneck in the effectiveness of legal AI systems.
+ This thesis aims to address this challenge by developing a more adaptive and context-sensitive retrieval mechanism tailored to the legal domain, specifically focusing on Arabic legal documents.
+
+\subsubsection*{Objectives :} 
+This research aims to explore and improve the integration of LLMs and retrieval systems in the legal field, with a focus on the Algerian legal context.   
+ The main objectives are as follows:
 \begin{itemize}
@@
 \end{itemize}
-\textbf{Contributions :} This thesis makes the following key contributions:
+\subsubsection*{Contributions :} This thesis makes the following key contributions:
 \begin{itemize}
 	\item A domain-specific chatbot designed to assist users in navigating Algerian legal case information using natural language queries.
+    \item train a complete Retrieval-Augmented Generation (RAG) system fully tailored to Arabic, addressing both the retriever and generator components.
+This represents one of the first domain-specific implementations of a fully Arabic end-to-end RAG pipeline
 	\item The creation of a new dataset of Arabic-language  legal cases from Algeria, filling a gap in publicly available legal resources for this jurisdiction.
@@
 \end{itemize}
-\textbf{Structure :} 
+\subsubsection*{Structure :} 
 The thesis is organized into four main chapters: 

### 2025-04-15 Paper/TOSUBMIT/RAG_PAPER2.tex
Version to submit

--- 2025-03-22 Paper/latex/paper_RAG.tex
@@
 
-\title[Article Title]{Beyond Static Retrieval: Adaptive Method for Improving RAG Performance}
+	\title[Article Title]{dyRAG: Dynamic Retrieval Augmentation for Adaptive RAG Performance}
 
@@
 Using a fixed number of retrieved documents K may fail  to adapt to varying query complexity, often leading to suboptimal retrieval either including irrelevant documents or missing crucial ones.
-To address this issue, we propose a hybrid method that dynamically adjusts K based on query characteristics while maintaining computational efficiency.
+To address this issue, we propose dyRAG, a hybrid method that dynamically adjusts K based on query characteristics while maintaining computational efficiency.
 Our method improves retrieval performance by maximizing relevant information and minimizing noise.
@@
 Notable examples of LLMs include OpenAI GPT series\cite{brown2020language}, Google BERT\cite{devlin2018bert} and T5 \cite{raffel2020exploring}.
-Despite their impressive capabilities, LLMs suffer from hallucination\cite{ji2023survey}, often generating confident but incorrect information.
-This is a critical challenge for knowledge-intensive tasks requiring high accuracy\cite{kandpal2023large}.
-Additionally, their static nature being trained on fixed datasets\cite{bender2021dangers} limits their ability to update knowledge dynamically, making them prone to outdated or inconsistent responses, especially for time-sensitive information.\cite{Mousavi2025}.
+Despite their impressive capabilities, LLMs suffer from hallucination\cite{Radford2019}, often generating confident but incorrect information.
+This is a critical challenge for knowledge-intensive tasks requiring high accuracy\cite{Radford2019}.
+Additionally, their static nature being trained on fixed datasets\cite{Radford2019} limits their ability to update knowledge dynamically, making them prone to outdated or inconsistent responses, especially for time-sensitive information.\cite{Mousavi2025}.
 
@@
 The use of  adaptive thresholds algorithm improves the relevance of retrieved documents, refining the set to include only the most contextually appropriate matches.
-Unlike traditional approaches, it adapts to the query and dataset characteristics, making it more flexible for diverse contexts.
+Unlike traditional approaches, dyRAG adapts to the query and dataset characteristics, making it more flexible for diverse contexts.
 Furthermore, its design for handling large datasets ensures scalable and precise retrieval, which is crucial for RAG systems working with extensive corpora.
@@
 The process involves feeding both the userss query and the relevant information retrieved by a search algorithm into the LLMs prompt\cite{ilin2023advancedrag}, enabling the model to generate answers based on the provided context.
-
+	gupta20241
 	\begin{figure}[h]
@@
 		\includegraphics[width=0.9\textwidth]{rag.JPG}
-	\caption{RAG system component\cite{sharma2024rag}. }\label{rag.PNG}
+		\caption{RAG system component\cite{gupta20241}. }\label{rag.PNG}
 	\end{figure}
@@
 	where each key \( k_i \) is linked to a corresponding value \( v_i \) (often, the key and value are the same).
-When a query \( q \) is provided, the goal is to find the top-\( k \) keys that are most similar to the query using a similarity function \( s \), and then retrieve their associated values\cite{gupta2024comprehensivesurveyretrievalaugmentedgeneration}.
+When a query \( q \) is provided, the goal is to find the top-\( k \) keys that are most similar to the query using a similarity function \( s \), and then retrieve their associated values\cite{gupta20241}.
 
@@
 
-\section{Proposed Solution}\label{sec3}
+	\section{dyRAG: The proposed Solution}\label{sec3}
 	In traditional candidate selection systems, the reliance on static or fixed top-K retrieval lead to suboptimal retrieval performance, especially in scenarios where the relevance distribution of candidates is highly variable.
@@
 This assessment focuses on the next-item prediction task, a fundamental challenge in recommendation systems\cite{Zhu_2018}\cite{kang2018selfattentivesequentialrecommendation}, where the goal is to predict the most relevant item a user will interact with next based on their past interactions.
-\subsection{Dataset}
+	\subsection{Dataset description}
 	We conducted experiments using two MovieLens datasets : ML-100K and ML-1M \cite{Harper2015}which are widely used benchmarks for evaluating sequential recommendation models as detailed in Table\ref{tab:movielens}.\\
@@
 	\end{table}
-
-\subsection{ Setup }
+	\subsection{ Experiments Setup }
 	For both datasets, We utilize the SASRec architecture as our sequential user encoder, a model renowned for achieving state-of-the-art performance in next-item prediction tasks.
@@
 				MoL+SAS    & 128 & 256  & 4 & 256  & 128 & 10 & 3.6374 & 0.4299 \\\hline
-			Hybrid+SAS & 512 & 512  & 4 & 512  & 128 & 10 & 1.2808 & 0.8120 \\ \hline
+				Hybrid+SAS & 512 & 512  & 4 & 512  & 128 & 10 & 1.2808 & {\bf 0.8120} \\ \hline
 				MoL+SAS    & 512 & 512  & 4 & 512  & 128 & 10 & 1.3050 & 0.8016 \\\hline
@@
 				MoL+SAS    & 128 & 256  & 4 & 256  & 128 & 10 & 3.4671 & 0.3627  \\ \hline
-			Hybrid+SAS & 512 & 512  & 4 & 512  & 128 & 10 & 1.0275 & 0.8350  \\ \hline
+				Hybrid+SAS & 512 & 512  & 4 & 512  & 128 & 10 & 1.0275 & {\bf 0.8350} \\ \hline
 				MoL+SAS    & 512 & 512  & 4 & 512  & 128 & 10 & 1.0350 & 0.8276  \\ \hline
@@
 On the other hand, dynamically adjusting K offers greater flexibility but comes with its own set of challenges, such as increased computational costs and inconsistencies in determining the optimal retrieval size.
+
 	To address these limitations, we introduced a hybrid dynamic K selection strategy that strikes a balance between adaptability and efficiency.
@@
 This method has shown promising results, improving retrieval performance by minimizing unnecessary noise and ensuring that critical information is consistently captured.
-Future work can explore enhancing our model with learning-based retrieval strategie such as reinforcement learning or attention mechanisms could refine the dynamic selection of K Additionally, testing our approach in real-world scenarios and conducting large-scale evaluations across diverse domains will help validate its robustness and adaptability in various knowledge bases and retrieval settings.
+
+	Future work can explore enhancing our model with learning-based retrieval strategies such as reinforcement learning or attention mechanisms could refine the dynamic selection of K Additionally, testing our approach in real-world scenarios and conducting large-scale evaluations across diverse domains will help validate its robustness and adaptability in various knowledge bases and retrieval settings.
 

### 2025-04-15 Paper/latex/paper_RAG.tex
Version to submit

@@
 			MoL+SAS    & 128 & 256  & 4 & 256  & 128 & 10 & 3.6374 & 0.4299 \\\hline
-			Hybrid+SAS & 512 & 512  & 4 & 512  & 128 & 10 & 1.2808 & 0.8120 \\ \hline
+			Hybrid+SAS & 512 & 512  & 4 & 512  & 128 & 10 & 1.2808 & {\bf 0.8120} \\ \hline
 			MoL+SAS    & 512 & 512  & 4 & 512  & 128 & 10 & 1.3050 & 0.8016 \\\hline
@@
 			MoL+SAS    & 128 & 256  & 4 & 256  & 128 & 10 & 3.4671 & 0.3627  \\ \hline
-			Hybrid+SAS & 512 & 512  & 4 & 512  & 128 & 10 & 1.0275 & 0.8350  \\ \hline
+			Hybrid+SAS & 512 & 512  & 4 & 512  & 128 & 10 & 1.0275 & {\bf 0.8350} \\ \hline
 			MoL+SAS    & 512 & 512  & 4 & 512  & 128 & 10 & 1.0350 & 0.8276  \\ \hline
@@
 This method has shown promising results, improving retrieval performance by minimizing unnecessary noise and ensuring that critical information is consistently captured.
-Future work can explore enhancing our model with learning-based retrieval strategie such as reinforcement learning or attention mechanisms could refine the dynamic selection of K Additionally, testing our approach in real-world scenarios and conducting large-scale evaluations across diverse domains will help validate its robustness and adaptability in various knowledge bases and retrieval settings.
+Future work can explore enhancing our model with learning-based retrieval strategies such as reinforcement learning or attention mechanisms could refine the dynamic selection of K Additionally, testing our approach in real-world scenarios and conducting large-scale evaluations across diverse domains will help validate its robustness and adaptability in various knowledge bases and retrieval settings.
 

### 2025-04-15 Thesis/Latex/Acknowledgements.tex

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{center}
+
+	First of all, praise and thanks to Allah Almighty for giving us all the patience, courage, will and motivation that allowed us to accomplish this work.\\ 
+	\vspace{0.3cm}
+
+	My deepest gratitude goes to my supervisor, \textbf{Mr Salem Mohamed } for his patient guidance, insightful feedback, and continuous encouragement throughout the course of this work .\\
+	\vspace{0.3cm}
+	I would like to acknowledge the members of the jury for their interest  \\
+	\vspace{0.3cm}
+    I am also especially thankful to my parents, family, and friends for their unwavering support, love, and motivation at every step of this journey. \\
+    \vspace{0.3cm}
+    I would like to thank everyone who helps me to improve my work. and who gave me any remark that helped me to perfect this manuscript.\\
+	\vspace{0.3cm}
+	All the teachers of the University MUSTAPHA STAMBOULI DE LA WILAYA MASCARA For the good contribution of this work.
+\end{center}

### 2025-04-15 Thesis/Latex/Chapter_4.tex

--- 2025-03-10 Thesis/Latex/Chapter_4.tex
@@
-\chapter{k Selection in Retrieval}
-\section{Introduction}
-Retrieval-Augmented Generation (RAG) systems enhance language models by grounding responses in externally retrieved documents.A critical challenge in these systems is determining the optimal number of documents (k) to retrieve for a given query.Current approaches fall into three categories: static k selection, which uses a fixed number of retrieved documents; dynamic k selection, which adjusts k based on query  characteristics; and hybrid approaches, which combine both strategies.In this chapter, we first explore methods, highlighting their strengths and limitations We then introduce our novel hybrid dynamic selection algorithm provide detailed description of its methodology and present experimental results that demonstrate the effectiveness of its performance .
-%\section {The Role of k in Retrieval-Augmented Generation}
-%The number of retrieved documents (k) is a critical parameter in Retrieval-Augmented Generation (RAG) that has a direct impact on the retrieval and generation phases.
-Choosing the right k ensures that the system retrieves enough relevant information.
-\section{Defining k: The Number of Retrieved Documents}
-the parameter k denotes the number of documents or passages retrieved from an external knowledge base.
-This retrieval process is managed by the retriever component\cite{pareto2024rag},which identifies the top k relevant documents based on similarity to the query typically using embedding-based search\cite{Rossi_2024} (e.g., dense retrieval with FAISS, BM25, or hybrid methods) Subsequently, these documents are passed to the generator, typically a language model, which synthesizes the final response by integrating the retrieved
-\begin{figure}[h]
+\chapter{Experimental Results}
+\pagestyle{fancy}\lhead{\textbf \footnotesize\it{Experimental Results}}
+\pagestyle{fancy}\chead{} \pagestyle{fancy}\rhead{}
+\pagestyle{fancy}\lfoot{\textbf {\small\it{Univ-Mascara/Computer Science: 2025}}} 
+\pagestyle{fancy}\cfoot{} \pagestyle{fancy}\rfoot{\thepage}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Introduction}\label{start6}
+In this chapter, we present the experimental results of applying our novel algorithm to SASRec and compare its performance with SASRec tested using MOL.
+The goal of these experiments is to assess the effectiveness of our approach in improving recommendation accuracy, model efficiency, and overall performance metrics.
+
+\section{Experimental Results}\label{sec3}
+In order to measure the effectiveness of our proposed method, we perform a comprehensive evaluation of both the retrieval algorithm and the Mixture-of-Logits (MoL) approach.
+This assessment focuses on the next-item prediction task, a fundamental challenge in recommendation systems\cite{kang2018selfat},
+where the goal is to predict the most relevant item a user will interact with next based on their past interactions. 
+\subsection{Dataset}
+We conducted experiments using two MovieLens datasets : ML-100K and ML-1M \cite{Harper2015}which are widely used benchmarks for evaluating sequential recommendation models as detailed in Table\ref{tab_movielens}.\\
+\begin{table}[h]
 	\centering
-	\includegraphics[width=0.9\linewidth]{Figures/topk.png}
-	\caption{Basic retrieval}
-	\label{rag_retrival.png}
+	\renewcommand{\arraystretch}{1.2}
+	\begin{tabular}{|l|p{10cm}|}
+		\hline
+		\textbf{Dataset} & \textbf{Description} \\
+		\hline
+		\textbf{MovieLens-100K} & 100,000 ratings (1-5 scale) from 943 users on 1,682 movies. \\
+		& Each user has rated at least 20 movies. \\
+		& Includes user demographic information (age, gender, occupation, zip code). \\
+		\hline
+		\textbf{MovieLens-1M} & 1,000,209 ratings from 6,040 users on approximately 3,900 movies. \\
+		& Data collected from users who joined MovieLens in 2000. \\
+		& Represents a larger-scale recommendation scenario. \\
+		\hline
+	\end{tabular}
+	\caption{Summary of MovieLens datasets}
+	\label{tab_movielens}
+\end{table}
 
+\subsection{ Setup }For both datasets, We utilize the SASRec architecture as our sequential user encoder, a model renowned for achieving state-of-the-art performance in next-item prediction tasks.
+This architecture processes the user's historical interaction sequence, generating embeddings that encapsulate the user's preferences at each time step.
+These embeddings serve as the foundation for predicting the next item in the sequence.
+
+The query \( q \) represents the user's state at a specific time step, derived from their interaction history. In the MoL (Mixture of Logits) framework,\( q \)is transformed into \( Pq \)
+embeddings through a multi-layer perceptron (MLP).
+\subsection{Impact of Hyperparameters}
+For fair comparison, we maintained consistent architectural choices and training conditions across all experiments, We conducted an extensive hyperparameter analysis comparing both approaches (Hybrid+SAS and MoL+SAS) across different architectural configurations.
+All experiments were Implemented in TensorFlow and trained on a Google Colab environment with a T4 GPU.
+We use the Adam optimizer with a learning rate of 0.001.
+For the hybrid algorithm, we initialize the threshold (Tinit) to 0.3 and adaptively adjust it during training.
+We discuss detailed hyperparameter settings in table \ref{tab:100k_movies}, table \ref{tab:1m_movies}
+
+\newpage
+
+\setlength{\tabcolsep}{1pt} % Increase column padding
+
+\begin{table}[h]
+	\centering
+	\large
+	\renewcommand{\arraystretch}{1.7} % Increase row spacing
+	\resizebox{\textwidth}{!}{  % Scale table to fit width
+		\begin{tabular}{|l|c|c|c|c|c|c|c|c|}
+			\hline
+			\textbf{Model} & \textbf{\shortstack{Max \\ Sequence Length}} & 
+			\textbf{\shortstack{Embedding \\ Dimension}} & 
+			\textbf{\shortstack{Number \\ of Heads}} & 
+			\textbf{\shortstack{Feedforward \\ Dimension}} & 
+			\textbf{\shortstack{Batch \\ Size}} & 
+			\textbf{Epochs} & 
+			\textbf{\shortstack{Val \\ Loss}} & 
+			\textbf{\shortstack{Val \\ Accuracy}} \\
+			\hline
+			Hybrid+SAS & 50  & 128  & 2 & 128  & 128 & 12 & 5.3036 & 0.1584  \\ \hline
+			MoL+SAS    & 50  & 128  & 2 & 128  & 128 & 12 & 5.3152 & 0.1582  \\ \hline
+			Hybrid+SAS & 128 & 256  & 4 & 256  & 128 & 10 & 3.6364 & 0.4301  \\ \hline
+			MoL+SAS    & 128 & 256  & 4 & 256  & 128 & 10 & 3.6374 & 0.4299 \\\hline
+			Hybrid+SAS & 512 & 512  & 4 & 512  & 128 & 10 & 1.2808 & 0.8120 \\ \hline
+			MoL+SAS    & 512 & 512  & 4 & 512  & 128 & 10 & 1.3050 & 0.8016 \\\hline
+		\end{tabular}
+	}
+	\caption{Results on 100kMovies dataset}
+	\label{tab:100k_movies}
+\end{table}
+\setlength{\tabcolsep}{1pt} % Increase column padding
+\begin{table}[h]
+	\centering
+	\large
+	\renewcommand{\arraystretch}{1.7} % Increase row spacing
+	\resizebox{\textwidth}{!}{  % Scale table to fit width
+		\begin{tabular}{|l|c|c|c|c|c|c|c|c|}
+			\hline
+			\textbf{Model} & \textbf{\shortstack{Max \\ Sequence Length}} & 
+			\textbf{\shortstack{Embedding \\ Dimension}} & 
+			\textbf{\shortstack{Number \\ of Heads}} & 
+			\textbf{\shortstack{Feedforward \\ Dimension}} & 
+			\textbf{\shortstack{Batch \\ Size}} & 
+			\textbf{Epochs} & 
+			\textbf{\shortstack{Val \\ Loss}} & 
+			\textbf{\shortstack{Val \\ Accuracy}} \\
+			\hline
+			Hybrid+SAS & 50  & 128  & 2 & 128  & 128 & 10 & 4.5721 & 0.1530  \\ \hline
+			MoL+SAS    & 50  & 128  & 2 & 128  & 128 & 10 & 4.5733 & 0.1526  \\ \hline
+			Hybrid+SAS & 128 & 256  & 4 & 256  & 128 & 10 & 3.4152 & 0.3680  \\ \hline
+			MoL+SAS    & 128 & 256  & 4 & 256  & 128 & 10 & 3.4671 & 0.3627  \\ \hline
+			Hybrid+SAS & 512 & 512  & 4 & 512  & 128 & 10 & 1.0275 & 0.8350  \\ \hline
+			MoL+SAS    & 512 & 512  & 4 & 512  & 128 & 10 & 1.0350 & 0.8276  \\ \hline
+
+		\end{tabular}
+	}
+	\caption{Results on 1M Movies dataset}
+	\label{tab:1m_movies}
+\end{table}
+Both approaches demonstrate significant performance improvements as model capacity increases, with larger configurations consistently delivering better results.
+For instance, when increasing the model's capacity such as expanding the (Max Sequence Length: 512, Embedding Dimension: 512, Number of Heads: 4, Feed-Forward Dimension: 512) the hybrid algorithm combined with SASRec (Hybrid+SAS) achieves notable gains, particularly in the most resource-intensive setup.
+On the ML-100K dataset, Hybrid+SAS reaches a score of\textbf{0.8120} compared to the baseline's \textbf{0.8016}, while on ML-1M, it achieves \textbf{0.8350} versus \textbf{0.8276} The ML-1M dataset generally benefits more from increased capacity, with the performance gap between datasets narrowing as the model scales.
+While smaller configurations provide a balance of efficiency and performance, the largest configuration, despite its higher computational demands, yields the best results, making it suitable for scenarios where resources are not a constraint.
+Both approaches show similar benefits from scaling, but Hybrid+SAS maintains a consistent edge in performance.
+
+\subsection {Impact of Adaptive k-Variations on Model Performance}
+As shown in Figures~\ref{fig:k_changes_100K} and \ref{fig:k_changes_1M}, the value of $k$ fluctuates across epochs for both MovieLens 100K and MovieLens 1M datasets.
+These variations indicate the adaptive nature of $k$ in response to the dataset characteristics and training progress.
+In the following section, we analyze how these changes influence model performance.
+
+\begin{figure}[htbp]
+	\centering
+	\begin{subfigure}{0.48\textwidth}
+		\centering
+		\includegraphics[width=\linewidth]{Figures/CHAGES_OF_K_FOR_100k.png}
+		\caption{K changes across epochs for MovieLens 100K}
+		\label{fig:k_changes_100K}
+	\end{subfigure}
+	\hfill
+	\begin{subfigure}{0.48\textwidth}
+		\centering
+		\includegraphics[width=\linewidth]{Figures/chnages_inK_for_1M.png}
+		\caption{K changes across epochs for MovieLens 1M}
+		\label{fig:k_changes_1M}
+	\end{subfigure}
+	\caption{Comparison of K changes across epochs for different datasets}
+	\label{fig:k_changes}
 \end{figure}
-\newline
-\section{Impact of k on Retrieval Performance}
-The retriever's effectiveness in identifying relevant documents depends on k. Key impacts include :
-\subsection{Recall vs. Precision} 
-When a user conducts a search query, the outcomes from the database can be grouped into four distinct types based on relevance and retrieval status \cite{geeksforgeeks2022precision}:
-\begin{itemize}
-	\item {Relevant and Retrieved:} Documents that both address the user’s query and appear in the search results.
-	\item {Relevant and Not Retrieved:} Useful documents that answer the query but are not included in the results.
+the value of K changed dynamically during training, with different behaviors.
+For the MovieLens 100K dataset, K showed a steady increase across epochs, whereas for the MovieLens 1M dataset, K began at a lower value, experienced fluctuations, peaked, and then stabilized.
+This indicates that the larger dataset necessitated more adaptive adjustments in retrieval compared to the smaller dataset.\\
+Figure~\ref{histo_p} summarizes the performance of our hybrid algorithm compared to the MoL-based approach on the MovieLens 100K and 1M datasets, both tested using a Max Sequence Length of 50, an Embedding Dimension of 128, 2 Attention Heads, a Feedforward Dimension of 128, a Batch Size of 128, and trained for 10 epochs .
+\\
+\begin{figure}[ht]
+	\centering
+	\includegraphics[width=0.7\textwidth]{Figures/HISTOGRAM.png}
+	\caption{Comparison of Models Across Evaluation Metrics}\label{histo_p}
+\end{figure}
+The variations in K help explain the performance differences seen in the fig \ref{histo_p}.
+The Hybrid+SAS model applied to the MovieLens 1M dataset, which exhibited more dynamic changes in K achieved higher HR@50(0.3631) and HR@200(0.6170) scores, reflecting better long-range recommendation quality.
+The fluctuations in K in the 1M dataset likely allowed the model to balance exploration and precision, improving its overall ranking performance.
 
-	\item {Non-Relevant and Retrieved:} Documents that appear in the results but lack meaningful value for the query.
+On the other hand, the more gradual K increase in the 100K dataset resulted in relatively lower scores, suggesting that a static or overly conservative 
+K selection might limit retrieval effectiveness.
 
-	\item {Non-Relevant and Not Retrieved:} Irrelevant documents excluded from the results.
-\end{itemize}
+Additionally, the MoL+SAS model achieved better performance than Hybrid+SAS in HR@10 for both datasets.
+This aligns with the observation that MoL+SAS tended to retrieve fewer but more relevant items at shorter ranking positions.
+This behavior can be attributed to how K evolved during training?smaller K values in earlier epochs likely enabled MoL+SAS to maintain higher precision at shorter ranks.
+These findings highlight the importance of dynamically tuning K based on dataset characteristics to optimize recommendation effectiveness.
 
-\textbf{Precision@k}: evaluates the proportion of relevant documents within the top k retrieved results.
-This metric is particularly valuable in scenarios where the focus is on delivering highly relevant information quickly, rather than ensuring complete coverage such as ,recommendation systems or search engines\cite{deconvoluteai2024metrics}.
+The results demonstrate that the Hybrid Exact Top-k algorithm effectively leverages both sequential user behavior and item metadata to improve recommendation quality.
+The adaptive MoL threshold allows the model to dynamically refine candidate items during training, leading to better performance.
+The improvements are consistent across both datasets, highlighting the robustness of our approach.\\
+%%=============================================%%
 
-\[
-\text{Precision@k} = \frac{\text{Number of Relevant Items Retrieved in Top k}}{k}
-\]
+\section{Conclusion}
 
-\textbf{Example}
-Suppose we have a dataset of 10 documents. If we retrieve \( k = 4 \) documents and find that 3 of them are relevant to the query:
-
-\begin{itemize}
-	\item \textbf{Dataset}: [doc1, doc2, doc3, doc4, doc5, doc6, doc7, doc8, doc9, doc10]
-	\item \textbf{Retrieved}: [doc3, doc1, doc7, doc4]
-	\item \textbf{Relevant}: [doc1, doc3, doc5, doc8]
-\end{itemize}
-
-The Precision@4 score would be:
-
-\[
-\text{Precision@4} = \frac{3}{4} = 0.75
-\]
-
-\textbf{Recall@k :}evaluates the proportion of relevant documents that are successfully retrieved within the top k results.
-This metric is particularly important in contexts where ensuring the completeness of information is crucial, such as in medical research or academic tools, where omitting relevant documents could result in incomplete or inaccurate conclusions\cite{deconvoluteai2024metrics}.
-
-\[
-\text{Recall@k} = \frac{\text{Number of Relevant Items Retrieved in Top k}}{\text{Total Number of Relevant Items}}
-\]
-\newline
-\textbf{Example}
-Consider a dataset of 10 documents. If we retrieve \( k = 4 \) documents and find that 2 of them are relevant, while the total number of relevant documents in the dataset is 4:
-
-\[
-\text{Recall@4} = \frac{2}{4} = 0.5
-\]
-
-Increasing k typically enhances recall by retrieving more relevant documents but may decrease precision due to the inclusion of irrelevant ones.
-Conversely, decreasing k can improve precision but at the cost of lower recall.
-\subsection{Retrieval Speed and Computational Cost}
-Retrieval speed and computational cost are two important areas where k has a big influence Below is a detailed explanation of how higher k values affect these aspects.
-
-\textbf{Increased Computational Resources:} \\
-As the value of k expands, the retrieval system must handle a larger set of documents, leading to higher computational demands.
-Specifically, the system needs to perform additional operations like ranking, filtering, and similarity scoring to identify the most relevant documents.
-These tasks become increasingly resource-intensive, particularly in large-scale retrieval settings where the document corpus consists of millions or even billions of entries\cite{manning2008ir}.
-For instance, retrieving the top  documents (k=80) requires significantly more computational resources compared to retrieving only the top 10 documents (k=5).
-
-\textbf{Higher Memory Usage:}
-As the number of retrieved documents (k) increases ,Storing and processing a larger set of retrieved documents requires more memory This can become a significant bottleneck, especially in environments with constrained memory resources .
-
-For large-scale retrieval tasks, where datasets may contain millions or even billions of documents, the memory demand grows proportionally with 
-k.
-Each retrieved document needs to be stored temporarily, along with its associated metadata, such as embedding vectors, BM25 scores, or other relevance signals.
-Additionally, sorting and filtering operations further contribute to memory overhead
-\subsection{Document Ranking Quality}
-Document ranking in information retrieval involves ordering documents by their relevance to a user's query.
-The objective is to prioritize the most relevant documents at the top of search results, making it easier for users to access useful information quickly.
-Different models,Vector Space, including Boolean, and Probabilistic models, are used to establish this ranking \cite{enwiki:1262179867}.\\
-
-Increasing the number of retrieved documents,represented as k, may result in the addition of lower-ranked, less relevant documents, potentially diluting the overall quality of the retrieved information.
-This occurs because of the inherent balance between precision and recall in information retrieval systems.
-\begin{figure}[h]
-	\centering
-	\includegraphics[width=0.7\linewidth]{Figures/precisionR.png}
-	\caption{Precision in Document Ranking\cite{evidentlyai2025}}
-	\label{precisionR}
-
-\end{figure}
-
-\begin{figure}[h]
-	\centering
-	\includegraphics[width=0.7\linewidth]{Figures/recalR.png}
-	\caption{Recall in Document Ranking\cite{evidentlyai2025}}
-	\label{recalR}
-
-\end{figure}
-As k grows, recall generally improves since a larger number of relevant documents are likely to be retrieved as shown in Figure \ref{recalR}.
-However, this often comes at the expense of precision, as the additional documents retrieved may include non-relevant ones,thereby lowering the overall precision as illustrated in Figure \ref{precisionR}.
-This trade-off is a core principle in evaluating information retrieval systems.
-
-\section{Impact of k on Generation Quality}
-The impact of k (the number of retrieved documents or generated candidates) on generation quality is A crucial factor in many machine learning and information retrieval tasks, including text generation, recommendation systems, and search engines .
-
-\subsection{Trade-off Between Diversity and Relevance} 
-Higher k: Increasing k often improves diversity in generated outputs or retrieved documents, as more candidates are considered.
-However, this can lead to a drop in relevance or quality, as lower-ranked candidates may be less accurate or coherent.
-\\
-Lower k: A smaller k tends to prioritize high-quality, relevant outputs but may lack diversity, leading to repetitive or overly narrow results.
-
-Research shows that the quality of retrieved documents plays a crucial role in the performance of Retrieval-Augmented Generation (RAG) systems.
-For example, one study demonstrated that the precision of retrieved documents directly affects the factual correctness of the generated responses \cite{salemi2023evaluating}
-Additionally, another study revealed that simply increasing the number of documents does not necessarily improve generation quality,especially if the additional documents are not highly relevant\cite{wan2025cognitivealigneddocumentselectionretrievalaugmented}
-\subsection {Effect on Text Generation Models}
-In text generation tasks, such as machine translation and dialogue systems, The parameter k frequently refers to the beam width in beam search algorithms\cite{freitag-al-onaizan-2017-beam}.
-Beam search is a heuristic search algorithm that expands the most promising nodes in a graph to maximize the quality of the text that is produced.
-Adjusting the beam width (k) has significantly impacts how well text generation models function.
-
-As the beam width is increased, the model must process and retain more candidate sequences concurrently, leading to higher computational and memory demands Particularly in real-time applications where latency is crucial, this increase may have an effect on system performance and response time\cite{tensorrt_llm_beam_search}.
-
-\section{Existing Solutions for k Selection}
-Selecting an optimalnk plays a crucial role in balancing retrieval effectiveness and generation quality.Over the years, various strategies have been proposed to determine k, ranging from static selection to dynamic and hybrid approaches.
-\subsection{Static k Selection}
-\subsection{Dynamic k Selection}
-\subsection{Hybrid k Selection}
-\section{ Proposed Solution}
-\subsection{Mixture of Logits-Based k Selection}
-\subsection{Algorithm Design}
-\subsection{Pseudocode}
-\section{ Conclusion }
+The experimental results demonstrate the effectiveness of our proposed algorithm when applied to SASRec.
+Through extensive evaluation on the MovieLens-100K and MovieLens-1M datasets, we observe that our method consistently outperforms the baseline models, including SASRec with MOL.
+Our approach shows improvements in key evaluation metrics, particularly in Hit Rate (HR@k) and Normalized Discounted Cumulative Gain (NDCG@k), indicating better recommendation relevance.
+Furthermore, the results highlight the importance of hyperparameter tuning, as different configurations significantly impact model performance.
+Overall, the findings confirm that incorporating our algorithm enhances sequential recommendation capabilities, making it a promising approach for future research and real-world applications.
+Future work will focus on further optimizing the model and testing on additional datasets to validate its generalizability.

### 2025-04-15 Thesis/Latex/Dedication.tex

@@
 \end{center}
-%
-
+\vfill 
+\begin{center}
+	\parbox{.85\linewidth}{\centering 
+		\baselineskip=8mm 
+		\fontfamily{pzc}\selectfont 
+		\Large 
+This work is dedicated with deep love and gratitude to:
+\\[5mm]
+My dear parents,
+whose unwavering support, sacrifices, and constant prayers have been the foundation of everything I’ve achieved. Your belief in me gave me the strength to persevere through every challenge.
+\\[5mm]
+My supervisor, Mr. [Mr Salem Mohamed],
+whose patient guidance, insightful feedback, and continuous encouragement made this journey not only possible but meaningful.
+Your dedication, expertise, and belief in my potential have left a lasting impact on me both academically and personally.
+\\[5mm]
+My beloved sister,
+for being my greatest cheerleader and my constant source of emotional support. Your encouragement and care during the toughest moments meant the world to me.
+\\[5mm]
+My wonderful friend [nial hadjer],
+for always being there with kind words, late-night conversations, and unshakable support.
+\\[19mm]
+}
+\end{center}
+\vfill 
 \thispagestyle{empty}

### 2025-04-15 Thesis/Latex/Images_sources.tex

@@
 	\end{center}
-	Figure 2.1 from \url{www.site1.com} \\
-	Figure 2.3 from \url{www.site2.fr} \\
-	Figure 2.5 from \url{www.site3.com} \\
-	Figure 2.9 from \url{www.site4.com} \\
-	Figure 3.5 from \url{www.site5.com} \\
-	Figure 2.3 from \url{www.site2.fr} \\
-	Figure 2.5 from \url{www.site3.com} \\
-	Figure 2.9 from \url{www.site4.com} \\
-	Figure 3.5 from \url{www.site5.com} \\
+	Figure 1.1 from \url{https://blog.dataiku.com/nlp-metamorphosis} \\
+	Figure 1.2 from \url{www.dmice.ohsu.edu/hersh/ohsu-pathology-20.pdf} \\
+	Figure 1.3 from \url{https://w.wiki/3sAW} \\
+	Figure 1.4 from \url{www.researchgate.net/publication/336092695_Densely_Connected_Convolutional_Networks_With_Attention_LSTM_for_Crowd_Flows_Prediction} \\
+   Figure 1.5 from \href{https://www.medium.com/huggingface/encoder-decoders-in-transformers-a-hybrid-pre-trained-architecture-for-seq2seqaf4d7bf14bb8}{Medium article  transformers}\\
+	Figure 1.6 from \href{https://medium.com/@anishnama20/exploring-the-power-of-encoder-decoder-models-pros-cons-and-applications-8bfbe2e66e76}{Medium article on encoder-decoders }	\\
+	Figure 1.7 from \url{arxiv.org/abs/2307.09288} \\
+	Figure 2.1 from \url{https://arxiv.org/abs/2312.10997} \\
+	Figure 2.2 from \href{https://www.researchgate.net/figure/High-level-overview-of-a-possible-solution-using-RAG-and-a-multi-modal-in-context_fig1_380540048}{RAG } \\
+   Figure 3.1 from \url{https://ai.gopubby.com/advanced-rag-techniques-unlocking-the-next-level-040c205b95bc} \\
+	Figure 3.2 from \url{www.evidentlyai.com/ranking-metrics/precision-recall-at-k} \\
+	Figure 3.3 from \url{www.evidentlyai.com/ranking-metrics/precision-recall-at-k} \\
+	Figure 3.4 from \url{openreview.net/forum?id=eHF4pDRWjT} \\
+	Figure 3.6 from \url{d2l.ai/chapter_recommender-systems/recsys-intro.html} \\
+	Figure 3.7 from \url{arxiv.org/abs/1808.09781} \\
 \end{landscape}

### 2025-07-18 Paper/JAIR_Example_Template/JAIR_Example_Template.tex
All changes
Last submitted paper is in folder submission3

%% This is a JAIR Example File Compiled by Nicholas Mattei (nsmattei@tulane.edu) 
%% and Odd Erik Gundersen (odderik@ntnu.no)
%% and Mykel Kochenderfer (mykel@stanford.edu)
%% March 29, 2025
%%
%% This file is based off the ACM Latex Template https://www.acm.org/publications/proceedings-template
%% Revision 2.12 (12/28/2024)
%% 
%% Please see https://www.jair.org/index.php/jair for more information and submission instructions.
%%

%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{jair}.
%%
\documentclass[review]{jair}

\setcopyright{cc}
\copyrightyear{2025}
\acmYear{2025}
\acmDOI{10.1613/jair.1.xxxxx}

%%
\JAIRAE{Insert JAIR AE Name}
\JAIRTrack{Insert JAIR Track Name Here}
\acmVolume{4}
\acmArticle{111}
\acmMonth{8}
\acmYear{2025}

%%
%% For managing citations, it is recommended to use bibliography
%% files in BibTeX format.
%%
%% You can then either use BibTeX with the ACM-Reference-Format style,
%% or BibLaTeX with the acmnumeric or acmauthoryear sytles, that include
%% support for advanced citation of software artefact from the
%% biblatex-software package, also separately available on CTAN.
%%
%% Look at the sample-*-biblatex.tex files for templates showcasing
%% the biblatex styles.
%%

%%
%% The majority of ACM publications use numbered citations and
%% references.  The command \citestyle{authoryear} switches to the
%% "author year" style.
%%
%% If you are preparing content for an event
%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
%% citations and references.
%% Uncommenting
%% the next command will enable that style.
%%\citestyle{acmauthoryear}

%%
%% end of the preamble, start of the body of the document source.
\begin{document}

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
%\title{JAIR Example Template}
\title[Satzilla]{SATzilla: Portfolio-based Algorithm Selection for SAT}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research and/or corresponding author.
\author{Lin Xu}
\authornote{Corresponding Author.}
\email{xulin730@cs.ubc.ca}
\orcid{0000-0002-2037-3694}
% \orcid{https://orcid.org/0000-0002-2037-3694} % This works too if you want to see the full URL
\affiliation{%
  \institution{University of British Columbia}
  \city{Vancouver}
  \state{British Columbia}
  \country{Canada}
}

\author{Frank Hutter}
\orcid{0000-0002-2037-3694}
\email{hutter@cs.ubc.ca}
\affiliation{%
  \institution{University of British Columbia}
  \city{Vancouver}
  \state{British Columbia}
  \country{Canada}}

\author{Holger H. Hoos}
\orcid{0000-0003-0629-0099}
\email{hoos@cs.ubc.ca}
\affiliation{%
  \institution{University of British Columbia}
  \city{Vancouver}
  \state{British Columbia}
  \country{Canada}
}

\author{Kevin Leyton-Brown}
\orcid{0000-0002-7644-5327}
\email{kevinlb@cs.ubc.ca}
\affiliation{%
  \institution{University of British Columbia}
  \city{Vancouver}
  \state{British Columbia}
  \country{Canada}
}

%\author{Ben Trovato}

%%
%% By default, the full list of authors will be used in the page
%% headers. Often, this list is too long, and will overlap
%% other information printed in the page headers. This command allows
%% the author to define a more concise list
%% of authors' names for this purpose.
\renewcommand{\shortauthors}{Trovato et al.}

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
{\bf Background:} 
    It has been widely observed that there is no single ``dominant'' SAT solver; instead, different solvers perform best on different instances.

    {\bf Objectives:}
    Rather than following the traditional approach of choosing the best solver for a given class of SAT instances, we aim to make this decision fully automatically, online and on a per-instance basis, with the goal of solving a broad range of SAT instances more efficiently in terms of running time.

    {\bf Methods:}
    We describe SATzilla, an automated approach for constructing per-instance algorithm portfolios for SAT that use so-called empirical hardness models to choose among their constituent solvers.  This approach takes as input a distribution of problem instances and a set of component solvers, and constructs a portfolio optimizing a given objective function (such as mean running time, percent of instances solved, or score in a competition).
    In this article, we go well beyond earlier versions of SATzilla, by making the portfolio construction scalable and completely automated, and improving it by integrating local search solvers as candidate solvers, by predicting performance score instead of running time, and by using hierarchical hardness models that take into account different types of SAT instances.

    {\bf Results:}
    The excellent performance of SATzilla was independently verified in the 2007 SAT Competition, where our SATzilla07 solvers won three gold, one silver and one bronze medal. 
    We demonstrate the effectiveness of the new techniques introduced here in extensive experimental results on data sets including instances from the most recent SAT competition.

    {\bf Conclusions:}
    The effectiveness of the SATzilla approach demonstrated in this article suggests that per-instance automated algorithm selection may also be possible for NP-hard problems other than SAT. 
    We expect this to pave the way for achieving substantial improvements in the state of the art in solving other important problems in AI and beyond.
\end{abstract}

%\begin{abstract}
%      A clear and well-documented \LaTeX\ document is presented as an
%  article formatted for publication by ACM in a conference proceedings
%  or journal publication. Based on the ``acmart'' document class, this
%  article presents and explains many of the common variations, as well
%  as many of the formatting elements an author may use in the
%  preparation of the documentation of their work.
%\end{abstract}

%% JAIR Note: 
%% Do not include ACM CCS Concepts or Keywords

%% To be updated by authors.
\received{20 February 2007}
\received[revised]{12 March 2009}
\received[accepted]{5 June 2009}

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle

\section{Introduction}
This document describes the revised \LaTeX template used by JAIR starting with volume 83 (2025). The template is based on ACM's consolidated article template, introduced in 2017, which provides a
consistent \LaTeX\ style for use across ACM publications, and
incorporates accessibility and metadata-extraction functionality
necessary for future Digital Library endeavors. JAIR adopted this template because 1) our old template was sadly out of date and we wanted a reliable, maintained template and 2) the journal is now distributed as part of the ACM library (though JAIR remains an independent, open-access journal published by AI Access Foundation, freely available at \url{https://jair.org}). Note that much of the functionality described below relates to ACM publications in general, rather than the specific template used by JAIR.

If you are new to publishing with JAIR or ACM, this document provides a
guide to the process of preparing your work for publication. If you
have published with ACM before, this document provides insight and
instruction into more recent changes to the article template.

The ``\verb|acmart|'' document class can be used to prepare articles
for any ACM publication --- conference or journal, and for any stage
of publication, from review to final ``camera-ready'' copy, to the
author's own version, with {\itshape very} few changes to the source.

\section{Template Overview}
As noted in the introduction, the ``\verb|acmart|'' document class can
be used to prepare many different kinds of documentation --- a
double-anonymous initial submission of a full-length technical paper, a
two-page SIGGRAPH Emerging Technologies abstract, a ``camera-ready''
journal article, a SIGCHI Extended Abstract, and more --- all by
selecting the appropriate {\itshape template style} and {\itshape
  template parameters}.

This document will explain the major features of the document
class. For further information, the {\itshape \LaTeX\ User's Guide} is
available from
\url{https://www.acm.org/publications/proceedings-template}.

There are a number of {\itshape template parameters}
that modify some part of the applied template style. A complete list
of these parameters can be found in the {\itshape \LaTeX\ User's Guide.}

Frequently-used parameters, or combinations of parameters, include:
\begin{itemize}
\item {\texttt{anonymous,review}}: Suitable for a ``double-anonymous''
  conference submission. Anonymizes the work and includes line
  numbers. Use with the \texttt{\acmSubmissionID} command to print the
  submission's unique ID on each page of the work.
\item{\texttt{authorversion}}: Produces a version of the work suitable
  for posting by the author.
\item{\texttt{screen}}: Produces colored hyperlinks.
\end{itemize}

This document uses the following string as the first command in the
source file:
\begin{verbatim}
\documentclass[review]{jair}
\end{verbatim}

\section{Modifications}

Modifying the template --- including but not limited to: adjusting
margins, typeface sizes, line spacing, paragraph and list definitions,
and the use of the \verb|\vspace| command to manually adjust the
vertical spacing between elements of your work --- is not allowed.

{\bfseries Your document will be returned to you for revision if
  modifications are discovered.}

\section{Typefaces}

The ``\verb|acmart|'' document class requires the use of the
``Libertine'' typeface family. Your \TeX\ installation should include
this set of packages. Please do not substitute other typefaces. The
``\verb|lmodern|'' and ``\verb|ltimes|'' packages should not be used,
as they will override the built-in typeface families.

\section{Title Information}

The title of your work should use capital letters appropriately -
\url{https://capitalizemytitle.com/} has useful rules for
capitalization. Use the {\verb|title|} command to define the title of
your work. If your work has a subtitle, define it with the
{\verb|subtitle|} command.  Do not insert line breaks in your title.

If your title is lengthy, you must define a short version to be used
in the page headers, to prevent overlapping text. The \verb|title|
command has a ``short title'' parameter:
\begin{verbatim}
  \title[short title]{full title}
\end{verbatim}

\section{Authors and Affiliations}

Each author must be defined separately for accurate metadata
identification.  As an exception, multiple authors may share one
affiliation. Authors' names should not be abbreviated; use full first
names wherever possible. Include authors' e-mail addresses whenever
possible.

Grouping authors' names or e-mail addresses, or providing an ``e-mail
alias,'' as shown below, is not acceptable:
\begin{verbatim}
  \author{Brooke Aster, David Mehldau}
  \email{dave,judy,steve@university.edu}
  \email{firstname.lastname@phillips.org}
\end{verbatim}

The \verb|authornote| and \verb|authornotemark| commands allow a note
to apply to multiple authors --- for example, if the first two authors
of an article contributed equally to the work.

If your author list is lengthy, you must define a shortened version of
the list of authors to be used in the page headers, to prevent
overlapping text. The following command should be placed just after
the last \verb|\author{}| definition:
\begin{verbatim}
  \renewcommand{\shortauthors}{McCartney, et al.}
\end{verbatim}
Omitting this command will force the use of a concatenated list of all
of the authors' names, which may result in overlapping text in the
page headers.

The article template's documentation, available at
\url{https://www.acm.org/publications/proceedings-template}, has a
complete explanation of these commands and tips for their effective
use.

Note that authors' addresses are mandatory for journal articles.

\section{Rights Information}

JAIR articles are freely available, and a copyright notice referencing the Creative Commons CC BY license is included in the template. This copyright notice should be included in all JAIR articles.

\section{Sectioning Commands}

Your work should use standard \LaTeX\ sectioning commands:
\verb|section|, \verb|subsection|, \verb|subsubsection|, and
\verb|paragraph|. They should be numbered; do not remove the numbering
from the commands.

Simulating a sectioning command by setting the first word or words of
a paragraph in boldface or italicized text is {\bfseries not allowed.}

\section{Tables}

The ``\verb|acmart|'' document class includes the ``\verb|booktabs|''
package --- \url{https://ctan.org/pkg/booktabs} --- for preparing
high-quality tables.

Table captions are placed {\itshape above} the table.

Because tables cannot be split across pages, the best placement for
them is typically the top of the page nearest their initial cite.  To
ensure this proper ``floating'' placement of tables, use the
environment \textbf{table} to enclose the table's contents and the
table caption.  The contents of the table itself must go in the
\textbf{tabular} environment, to be aligned properly in rows and
columns, with the desired horizontal and vertical rules.  Again,
detailed instructions on \textbf{tabular} material are found in the
\textit{\LaTeX\ User's Guide}.

Immediately following this sentence is the point at which
Table~\ref{tab:freq} is included in the input file; compare the
placement of the table here with the table in the printed output of
this document.

\begin{table}
  \caption{Frequency of Special Characters}
  \label{tab:freq}
  \begin{tabular}{ccl}
    \toprule
    Non-English or Math&Frequency&Comments\\
    \midrule
    \O & 1 in 1,000& For Swedish names\\
    $\pi$ & 1 in 5& Common in math\\
    \$ & 4 in 5 & Used in business\\
    $\Psi^2_1$ & 1 in 40,000& Unexplained usage\\
  \bottomrule
\end{tabular}
\end{table}

To set a wider table, which takes up the whole width of the page's
live area, use the environment \textbf{table*} to enclose the table's
contents and the table caption.  As with a single-column table, this
wide table will ``float'' to a location deemed more
desirable. Immediately following this sentence is the point at which
Table~\ref{tab:commands} is included in the input file; again, it is
instructive to compare the placement of the table here with the table
in the printed output of this document.

\begin{table*}
  \caption{Some Typical Commands}
  \label{tab:commands}
  \begin{tabular}{ccl}
    \toprule
    Command &A Number & Comments\\
    \midrule
    \texttt{{\char'134}author} & 100& Author \\
    \texttt{{\char'134}table}& 300 & For tables\\
    \texttt{{\char'134}table*}& 400& For wider tables\\
    \bottomrule
  \end{tabular}
\end{table*}

Always use midrule to separate table header rows from data rows, and
use it only for this purpose. This enables assistive technologies to
recognise table headers and support their users in navigating tables
more easily.

\section{Math Equations}
You may want to display math equations in three distinct styles:
inline, numbered or non-numbered display.  Each of the three are
discussed in the next sections. We note that JAIR authors are discouraged from using sophisticated mathematical notation in the abstract (especially notation that is difficult to reproduce in vanilla HTML), since the abstract will also be formatted in HTML on the JAIR web site.

\subsection{Inline (In-text) Equations}
A formula that appears in the running text is called an inline or
in-text formula.  It is produced by the \textbf{math} environment,
which can be invoked with the usual
\texttt{{\char'134}begin\,\ldots{\char'134}end} construction or with
the short form \texttt{\$\,\ldots\$}. You can use any of the symbols
and structures, from $\alpha$ to $\omega$, available in
\LaTeX~\cite{Lamport:LaTeX}; this section will simply show a few
examples of in-text equations in context. Notice how this equation:
\begin{math}
  \lim_{n\rightarrow \infty}x=0
\end{math},
set here in in-line math style, looks slightly different when
set in display style.  (See next section).

\subsection{Display Equations}
A numbered display equation---one set off by vertical space from the
text and centered horizontally---is produced by the \textbf{equation}
environment. An unnumbered display equation is produced by the
\textbf{displaymath} environment.

Again, in either environment, you can use any of the symbols and
structures available in \LaTeX\@; this section will just give a couple
of examples of display equations in context.  First, consider the
equation, shown as an inline equation above:
\begin{equation}
  \lim_{n\rightarrow \infty}x=0
\end{equation}
Notice how it is formatted somewhat differently in
the \textbf{displaymath}
environment.  Now, we'll enter an unnumbered equation:
\begin{displaymath}
  \sum_{i=0}^{\infty} x + 1
\end{displaymath}
and follow it with another numbered equation:
\begin{equation}
  \sum_{i=0}^{\infty}x_i=\int_{0}^{\pi+2} f
\end{equation}
just to demonstrate \LaTeX's able handling of numbering.

\section{Figures}

The ``\verb|figure|'' environment should be used for figures. One or
more images can be placed within a figure. If your figure contains
third-party material, you must clearly identify it as such, as shown
in the example below.
\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{sample-franklin}
  \caption{1907 Franklin Model D roadster. Photograph by Harris \&
    Ewing, Inc. [Public domain], via Wikimedia
    Commons. (\url{https://goo.gl/VLCRBB}).}
  \Description{A woman and a girl in white dresses sit in an open car.}
\end{figure}

Your figures should contain a caption which describes the figure to
the reader.

Figure captions are placed {\itshape below} the figure.

Every figure should also have a figure description unless it is purely
decorative. These descriptions convey what’s in the image to someone
who cannot see it. They are also used by search engine crawlers for
indexing images, and when images cannot be loaded.

A figure description must be unformatted plain text less than 2000
characters long (including spaces).  {\bfseries Figure descriptions
  should not repeat the figure caption – their purpose is to capture
  important information that is not already provided in the caption or
  the main text of the paper.} For figures that convey important and
complex new information, a short text description may not be
adequate. More complex alternative descriptions can be placed in an
appendix and referenced in a short figure description. For example,
provide a data table capturing the information in a bar chart, or a
structured list representing a graph.  For additional information
regarding how best to write figure descriptions and why doing this is
so important, please see
\url{https://www.acm.org/publications/taps/describing-figures/}.

\section{Citations and Bibliographies}

The use of \BibTeX\ for the preparation and formatting of one's
references is strongly recommended. Authors' names should be complete
--- use full first names (``Donald E. Knuth'') not initials
(``D. E. Knuth'') --- and the salient identifying features of a
reference should be included: title, year, volume, number, pages,
article DOI, etc.

For managing citations, it is recommended to use bibliography files in BibTeX format.
You can then either use BibTeX with the ACM-Reference-Format style,
or BibLaTeX with the acmnumeric or acmauthoryear styles, that include
support for advanced citation of software artifact from the
biblatex-software package, also separately available on CTAN.

The bibliography is included in your source document with these two
commands, placed just before the \verb|\end{document}| command:
\begin{verbatim}
  \bibliographystyle{ACM-Reference-Format}
  \bibliography{bibfile}
\end{verbatim}
where ``\verb|bibfile|'' is the name, without the ``\verb|.bib|''
suffix, of the \BibTeX\ file.

  Some examples.  A paginated journal article \cite{Abril07}, an
  enumerated journal article \cite{Cohen07}, a reference to an entire
  issue \cite{JCohen96}, a monograph (whole book) \cite{Kosiur01}, a
  monograph/whole book in a series (see 2a in spec. document)
  \cite{Harel79}, a divisible-book such as an anthology or compilation
  \cite{Editor00} followed by the same example, however we only output
  the series if the volume number is given \cite{Editor00a} (so
  Editor00a's series should NOT be present since it has no vol. no.),
  a chapter in a divisible book \cite{Spector90}, a chapter in a
  divisible book in a series \cite{Douglass98}, a multi-volume work as
  book \cite{Knuth97}, a couple of articles in a proceedings (of a
  conference, symposium, workshop for example) (paginated proceedings
  article) \cite{Andler79, Hagerup1993}, a proceedings article with
  all possible elements \cite{Smith10}, an example of an enumerated
  proceedings article \cite{VanGundy07}, an informally published work
  \cite{Harel78}, a couple of preprints \cite{Bornmann2019,
    AnzarootPBM14}, a doctoral dissertation \cite{Clarkson85}, a
  master's thesis: \cite{anisi03}, an online document / world wide web
  resource \cite{Thornburg01, Ablamowicz07, Poker06}, a video game
  (Case 1) \cite{Obama08} and (Case 2) \cite{Novak03} and \cite{Lee05}
  and (Case 3) a patent \cite{JoeScientist001}, work accepted for
  publication \cite{rous08}, 'YYYYb'-test for prolific author
  \cite{SaeediMEJ10} and \cite{SaeediJETC10}. Other cites might
  contain 'duplicate' DOI and URLs (some SIAM articles)
  \cite{Kirschmer:2010:AEI:1958016.1958018}. Boris / Barbara Beeton:
  multi-volume works as books \cite{MR781536} and \cite{MR781537}. A
  couple of citations with DOIs:
  \cite{2004:ITE:1009386.1010128,Kirschmer:2010:AEI:1958016.1958018}. Online
  citations: \cite{TUGInstmem, Thornburg01, CTANacmart}.
  Artifacts: \cite{R} and \cite{UMassCitations}.

\section{Acknowledgments}

Identification of funding sources and other support, and thanks to
individuals and groups that assisted in the research and the
preparation of the work should be included in an acknowledgment
section, which is placed just before the reference section in your
document.

This section has a special environment:
\begin{verbatim}
  \begin{acks}
  ...
  \end{acks}
\end{verbatim}
so that the information contained therein can be more easily collected
during the article metadata extraction phase, and to ensure
consistency in the spelling of the section heading.

Authors should not prepare this section as a numbered or unnumbered {\verb|\section|}; please use the ``{\verb|acks|}'' environment.

\section{Appendices}

If your work needs an appendix, add it before the
``\verb|\end{document}|'' command at the conclusion of your source
document.

Start the appendix with the ``\verb|appendix|'' command:
\begin{verbatim}
  \appendix
\end{verbatim}
and note that in the appendix, sections are lettered, not
numbered. This document has two appendices, demonstrating the section
and subsection identification method.

%%
%% The acknowledgments section is defined using the "acks" environment
%% (and NOT an unnumbered section). This ensures the proper
%% identification of the section in the article metadata, and the
%% consistent spelling of the heading.
\begin{acks}
To Robert, for the bagels and explaining CMYK and color spaces.
\end{acks}

%%
%% The next two lines define the bibliography style to be used, and
%% the bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{sample-base}

%%
%% If your work has an appendix, this is the place to put it.
\appendix

\section{Research Methods}

\subsection{Part One}

Lorem ipsum dolor sit amet, consectetur adipiscing elit. Morbi
malesuada, quam in pulvinar varius, metus nunc fermentum urna, id
sollicitudin purus odio sit amet enim. Aliquam ullamcorper eu ipsum
vel mollis. Curabitur quis dictum nisl. Phasellus vel semper risus, et
lacinia dolor. Integer ultricies commodo sem nec semper.

\subsection{Part Two}

Etiam commodo feugiat nisl pulvinar pellentesque. Etiam auctor sodales
ligula, non varius nibh pulvinar semper. Suspendisse nec lectus non
ipsum convallis congue hendrerit vitae sapien. Donec at laoreet
eros. Vivamus non purus placerat, scelerisque diam eu, cursus
ante. Etiam aliquam tortor auctor efficitur mattis.

\section{Online Resources}

Nam id fermentum dui. Suspendisse sagittis tortor a nulla mollis, in
pulvinar ex pretium. Sed interdum orci quis metus euismod, et sagittis
enim maximus. Vestibulum gravida massa ut felis suscipit
congue. Quisque mattis elit a risus ultrices commodo venenatis eget
dui. Etiam sagittis eleifend elementum.

Nam interdum magna at lectus dignissim, ac dignissim lorem
rhoncus. Maecenas eu arcu ac neque placerat aliquam. Nunc pulvinar
massa et mattis lacinia.

\section{Reproducibility Checklist for JAIR}

Select the answers that apply to your research -- one per item.

\subsection*{All articles:}

%\hh{revised for stylistic consistency:}
\begin{enumerate}
    \item All claims investigated in this work are clearly stated. 
    [yes/partially/no]
    \item Clear explanations are given how the work reported substantiates the claims. 
    [yes/partially/no]
    \item Limitations or technical assumptions are stated clearly and explicitly. 
    [yes/partially/no]
    \item Conceptual outlines and/or pseudo-code descriptions of the AI methods introduced in this work are provided, and important implementation details are discussed. 
    [yes/partially/no/NA]
    \item 
    Motivation is provided for all design choices, including algorithms, implementation choices, parameters, data sets and experimental protocols beyond metrics.
    [yes/partially/no]
\end{enumerate}

\subsection*{Articles containing theoretical contributions:}
Does this paper make theoretical contributions? 
[yes/no]

If yes, please complete the list below.

\begin{enumerate}
    \item All assumptions and restrictions are stated clearly and formally. 
    [yes/partially/no]
    \item All novel claims are stated formally (e.g., in theorem statements). 
    [yes/partially/no]
    \item Proofs of all non-trivial claims are provided in sufficient detail to permit verification by readers with a reasonable degree of expertise (e.g., that expected from a PhD candidate in the same area of AI). [yes/partially/no]
    \item
    Complex formalism, such as definitions or proofs, is motivated and explained clearly.
%hh: was:
%Proof sketches or intuitions are given for complex and/or novel results.
    [yes/partially/no]
    \item 
    The use of mathematical notation and formalism serves the purpose of enhancing clarity and precision; gratuitous use of mathematical formalism (i.e., use that does not enhance clarity or precision) is avoided.
    [yes/partially/no]
    \item 
    Appropriate citations are given for all non-trivial theoretical tools and techniques. 
    [yes/partially/no]
\end{enumerate}

\subsection*{Articles reporting on computational experiments:}
Does this paper include computational experiments? [yes/no]

If yes, please complete the list below.
\begin{enumerate}
    \item 
    All source code required for conducting experiments is included in an online appendix 
    or will be made publicly available upon publication of the paper.
    The online appendix follows best practices for source code readability and documentation as well as for long-term accessibility.
    [yes/partially/no]
    \item The source code comes with a license that
    allows free usage for reproducibility purposes.
    [yes/partially/no]
    \item The source code comes with a license that
    allows free usage for research purposes in general.
    [yes/partially/no]
    \item 
    Raw, unaggregated data from all experiments is included in an online appendix 
    or will be made publicly available upon publication of the paper.
    The online appendix follows best practices for long-term accessibility.
    [yes/partially/no]
    \item The unaggregated data comes with a license that
    allows free usage for reproducibility purposes.
    [yes/partially/no]
    \item The unaggregated data comes with a license that
    allows free usage for research purposes in general.
    [yes/partially/no]
    \item If an algorithm depends on randomness, then the method used for generating random numbers and for setting seeds is described in a way sufficient to allow replication of results. 
    [yes/partially/no/NA]
    \item The execution environment for experiments, the computing infrastructure (hardware and software) used for running them, is described, including GPU/CPU makes and models; amount of memory (cache and RAM); make and version of operating system; names and versions of relevant software libraries and frameworks. 
    [yes/partially/no]
    \item 
    The evaluation metrics used in experiments are clearly explained and their choice is explicitly motivated. 
    [yes/partially/no]
    \item 
    The number of algorithm runs used to compute each result is reported. 
    [yes/no]
    \item 
    Reported results have not been ``cherry-picked'' by silently ignoring unsuccessful or unsatisfactory experiments. 
    [yes/partially/no]
    \item 
    Analysis of results goes beyond single-dimensional summaries of performance (e.g., average, median) to include measures of variation, confidence, or other distributional information. 
    [yes/no]
    \item 
    All (hyper-) parameter settings for 
    the algorithms/methods used in experiments have been reported, along with the rationale or method for determining them. 
    [yes/partially/no/NA]
    \item 
    The number and range of (hyper-) parameter settings explored prior to conducting final experiments have been indicated, along with the effort spent on (hyper-) parameter optimisation. 
    [yes/partially/no/NA]
    \item 
    Appropriately chosen statistical hypothesis tests are used to establish statistical significance
    in the presence of noise effects.
    [yes/partially/no/NA]
\end{enumerate}

\subsection*{Articles using data sets:}
Does this work rely on one or more data sets (possibly obtained from a benchmark generator or similar software artifact)? 
[yes/no]

If yes, please complete the list below.
\begin{enumerate}
    \item 
    All newly introduced data sets 
    are included in an online appendix 
    or will be made publicly available upon publication of the paper.
    The online appendix follows best practices for long-term accessibility with a license
    that allows free usage for research purposes.
    [yes/partially/no/NA]
    \item The newly introduced data set comes with a license that
    allows free usage for reproducibility purposes.
    [yes/partially/no]
    \item The newly introduced data set comes with a license that
    allows free usage for research purposes in general.
    [yes/partially/no]
    \item All data sets drawn from the literature or other public sources (potentially including authors' own previously published work) are accompanied by appropriate citations.
    [yes/no/NA]
    \item All data sets drawn from the existing literature (potentially including authors’ own previously published work) are publicly available. [yes/partially/no/NA]
    %\item All data sets that are not publicly available are described in detail.
    %[yes/partially/no/NA]
    \item All new data sets and data sets that are not publicly available are described in detail, including relevant statistics, the data collection process and annotation process if relevant.
    [yes/partially/no/NA]
    \item 
    All methods used for preprocessing, augmenting, batching or splitting data sets (e.g., in the context of hold-out or cross-validation)
    are described in detail. [yes/partially/no/NA]
\end{enumerate}

\subsection*{Explanations on any of the answers above (optional):}

[Text here; please keep this brief.]

\end{document}
\endinput
%%
%% End of file `sample-acmlarge.tex'.

### 2025-07-18 Paper/JAIR_Example_Template/JAIR_RAG_Paper.tex
All changes
Last submitted paper is in folder submission3

--- 2025-04-15 Paper/TOSUBMIT/RAG_PAPER2.tex
@@
-	%Version 3 October 2023
-% See section 11 of the User Manual for version history
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                                                 %%
-%% Please do not use \input{...} to include other tex files.       %%
-%% Submit your LaTeX manuscript as one .tex document.              %%
-%%                                                                 %%
-%% All additional figures and files should be attached             %%
-%% separately and not embedded in the \TeX\ document itself.       %%
-%%                                                                 %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%\documentclass[referee,sn-basic]{sn-jnl}% referee option is meant for double line spacing
+%% This is a JAIR Example File Compiled by Nicholas Mattei (nsmattei@tulane.edu) 
+%% and Odd Erik Gundersen (odderik@ntnu.no)
+%% and Mykel Kochenderfer (mykel@stanford.edu)
+%% March 29, 2025
+%%
+%% This file is based off the ACM Latex Template https://www.acm.org/publications/proceedings-template
+%% Revision 2.12 (12/28/2024)
+%% 
+%% Please see https://www.jair.org/index.php/jair for more information and submission instructions.
+%%
 
-%%=======================================================%%
-%% to print line numbers in the margin use lineno option %%
-%%=======================================================%%
+%% The first command in your LaTeX source must be the \documentclass
+%% command.
+%%
+%% For submission and review of your manuscript please change the
+%% command to \documentclass[manuscript, screen, review]{jair}.
+%%
+\documentclass[review]{jair}
 
-%%\documentclass[lineno,sn-basic]{sn-jnl}% Basic Springer Nature Reference Style/Chemistry Reference Style
+\setcopyright{cc}
+\copyrightyear{2025}
+\acmYear{2025}
+\acmDOI{10.1613/jair.1.xxxxx}
 
-%%======================================================%%
-%% to compile with pdflatex/xelatex use pdflatex option %%
-%%======================================================%%
+%%
+\JAIRAE{Insert JAIR AE Name}
+\JAIRTrack{Insert JAIR Track Name Here}
+\acmVolume{4}
+\acmArticle{111}
+\acmMonth{8}
+\acmYear{2025}
 
-%%\documentclass[pdflatex,sn-basic]{sn-jnl}% Basic Springer Nature Reference Style/Chemistry Reference Style
+%%
+%% For managing citations, it is recommended to use bibliography
+%% files in BibTeX format.
+%%
+%% You can then either use BibTeX with the ACM-Reference-Format style,
+%% or BibLaTeX with the acmnumeric or acmauthoryear sytles, that include
+%% support for advanced citation of software artefact from the
+%% biblatex-software package, also separately available on CTAN.
+%%
+%% Look at the sample-*-biblatex.tex files for templates showcasing
+%% the biblatex styles.
+%%
 
-%%Note: the following reference styles support Namedate and Numbered referencing.
-By default the style follows the most common style.
-To switch between the options you can add or remove “Numbered” in the optional parenthesis.
-%%The option is available for: sn-basic.bst, sn-vancouver.bst, sn-chicago.bst%
+%%
+%% The majority of ACM publications use numbered citations and
+%% references.  The command \citestyle{authoryear} switches to the
+%% "author year" style.
+%%
+%% If you are preparing content for an event
+%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
+%% citations and references.
+%% Uncommenting
+%% the next command will enable that style.
+%%\citestyle{acmauthoryear}
 
-%%\documentclass[sn-nature]{sn-jnl}% Style for submissions to Nature Portfolio journals
-%%\documentclass[sn-basic]{sn-jnl}% Basic Springer Nature Reference Style/Chemistry Reference Style
-\documentclass[sn-mathphys-num]{sn-jnl}% Math and Physical Sciences Numbered Reference Style 
-%%\documentclass[sn-mathphys-ay]{sn-jnl}% Math and Physical Sciences Author Year Reference Style
-%%\documentclass[sn-aps]{sn-jnl}% American Physical Society (APS) Reference Style
-%%\documentclass[sn-vancouver,Numbered]{sn-jnl}% Vancouver Reference Style
-%%\documentclass[sn-apa]{sn-jnl}% APA Reference Style 
-%%\documentclass[sn-chicago]{sn-jnl}% Chicago-based Humanities Reference Style
-
-%%%% Standard Packages
-%%<additional latex packages if required can be included here>
-
+%%
+%% end of the preamble, start of the body of the document source.
 \usepackage{graphicx}%
 \usepackage{multirow}%
-\usepackage{amsmath,amssymb,amsfonts}%
-\usepackage{amsthm}%
-\usepackage{mathrsfs}%
-\usepackage[title]{appendix}%
-\usepackage{xcolor}%
-\usepackage{textcomp}%
-\usepackage{manyfoot}%
-\usepackage{booktabs}%
+%\usepackage{amsfonts, amsmath,amssymb}%
+%%\usepackage{amsthm}%
+%%\usepackage{mathrsfs}%
+%\usepackage[title]{appendix}%
+%\usepackage{xcolor}%
+%\usepackage{textcomp}%
+%\usepackage{manyfoot}%
+%\usepackage{booktabs}%
 \usepackage{algorithm}%
@@
 \usepackage{subcaption}
-
-%%%%
-
-%%%%%=============================================================================%%%%
-%%%%  Remarks: This template is provided to aid authors with the preparation
-%%%%  of original research articles intended for submission to journals published 
-%%%%  by Springer Nature. The guidance has been prepared in partnership with 
-%%%%  production teams to conform to Springer Nature technical requirements. 
-%%%%  Editorial and presentation requirements differ among journal portfolios and 
-%%%%  research disciplines. You may find sections in this template are irrelevant 
-%%%%  to your work and are empowered to omit any such section if allowed by the 
-%%%%  journal you intend to submit to. The submission guidelines and policies 
-%%%%  of the journal take precedence. A detailed User Manual is available in the 
-%%%%  template package for technical guidance.
-%%%%%=============================================================================%%%%
-
-%% as per the requirement new theorem styles can be included as shown below
-\theoremstyle{thmstyleone}%
-\newtheorem{theorem}{Theorem}%  meant for continuous numbers
-%%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
-%% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
-\newtheorem{proposition}[theorem]{Proposition}% 
-%%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.
-
-\theoremstyle{thmstyletwo}%
-\newtheorem{example}{Example}%
-\newtheorem{remark}{Remark}%
-
-\theoremstyle{thmstylethree}%
-\newtheorem{definition}{Definition}%
-
-\raggedbottom
-%%\unnumbered% uncomment this for unnumbered level heads
-
+\usepackage{booktabs} % For professional-quality rules
+\usepackage{makecell} % For multiline headers
 \begin{document}
 
-	\title[Article Title]{dyRAG: Dynamic Retrieval Augmentation for Adaptive RAG Performance}
+%%
+%% The "title" command has an optional parameter,
+%% allowing the author to define a "short title" to be used in page headers.
+%\title{JAIR Example Template}
+\title[dyRAG.Dynamic K-Selection]{Boosting RAG Efficiency with dyRAG: Dynamic K-Selection for Optimal Retrieval}
 
-	%%=============================================================%%
-	%% GivenName	-> \fnm{Joergen W.}
-	%% Particle	-> \spfx{van der} -> surname prefix
-	%% FamilyName	-> \sur{Ploeg}
-	%% Suffix	-> \sfx{IV}
-	%% \author*[1,2]{\fnm{Joergen W.} \spfx{van der} \sur{Ploeg} 
-	%%  \sfx{IV}}\email{iauthor@gmail.com}
-	%%=============================================================%%
+%%
+%% The "author" command and its associated commands are used to define
+%% the authors and their affiliations.
+%% Of note is the shared affiliation of the first two authors, and the
+%% "authornote" and "authornotemark" commands
+%% used to denote shared contribution to the research and/or corresponding author.
+\author{Zoubida Asmaa Boudjenane}
 
-	\author*[1]{\fnm{Zoubida Asmaa} \sur{Boudjenane }}\email{asmaa.boudj04@gmail.com}
+\email{asmaa.boudj04@gmail.com}
+\orcid{0000-0002-2037-3694}
+% \orcid{https://orcid.org/0000-0002-2037-3694} % This works too if you want to see the full URL
+\affiliation{%
+  \institution{Computer Science Department, University of Mascara}
+  \city{Mascara}
+  \state{Mascara}
+  \country{Algeria}
+}
 
-	\author[2]{\fnm{Mohammed} \sur{Salem}}\email{salem@univ-mascara.dz}
-	\equalcont{These authors contributed equally to this work.}
+\author{Mohammed Salem}
+\orcid{0000-0002-2037-3694}
+\authornote{Corresponding Author.}
+\email{salem@univ-mascara.dz}
+\affiliation{%
+  \institution{LISYS Laboratory, University of Mascara}
+  \city{Mascara}
+  \state{Mascara}
+  \country{Algeria}}
 
-	%\author[1,2]{\fnm{Third} \sur{Author}}\email{iiiauthor@gmail.com}
-	%\equalcont{These authors contributed equally to this work.}
+%\author{Holger H. Hoos}
+%\orcid{0000-0003-0629-0099}
+%\email{hoos@cs.ubc.ca}
+%\affiliation{%
+%  \institution{University of British Columbia}
+%  \city{Vancouver}
+%  \state{British Columbia}
+%  \country{Canada}
+%}
 
-	\affil*[1]{\orgdiv{Computer Science Department}, \orgname{University of Mascara}, \orgaddress{\street{Street}, \city{Mascara}, \postcode{29000}, \state{}, \country{Algeria}}}
+%\author{Kevin Leyton-Brown}
+%\orcid{0000-0002-7644-5327}
+%\email{kevinlb@cs.ubc.ca}
+%\affiliation{%
+%  \institution{University of British Columbia}
+%  \city{Vancouver}
+%  \state{British Columbia}
+%  \country{Canada}
+%}
 
-	\affil*[2]{\orgdiv{LISYS Lab}, \orgname{University of Mascara}, \orgaddress{\street{Street}, \city{City}, \postcode{100190}, \state{State}, \country{Algeria}}}
+%\author{Ben Trovato}
 
-	%\affil[3]{\orgdiv{Department}, \orgname{Organization}, \orgaddress{\street{Street}, \city{City}, \postcode{610101}, \state{State}, \country{Country}}}
+%%
+%% By default, the full list of authors will be used in the page
+%% headers. Often, this list is too long, and will overlap
+%% other information printed in the page headers. This command allows
+%% the author to define a more concise list
+%% of authors' names for this purpose.
+\renewcommand{\shortauthors}{Boudjenane and Salem.}
 
-	%%==================================%%
-	%% Sample for unstructured abstract %%
-	%%==================================%%
+%%
+%% The abstract is a short summary of the work to be presented in the
+%% article.
+\begin{abstract}
+{\bf Background:} 
+    Retrieval-Augmented Generation (RAG) enhances language model responses by incorporating external knowledge.
+However, its effectiveness heavily depends on the quality of the retrieved documents.
+Using a fixed number of retrieved documents (K) may fail to adapt to varying query complexity, often leading to suboptimal retrieval—either including irrelevant documents or missing crucial ones.
 
-	\abstract{Retrieval-Augmented Generation (RAG) enhances language model responses by incorporating external knowledge.
-However, its effectiveness heavily depends on the quality of the retrieved documents.
-Using a fixed number of retrieved documents K may fail  to adapt to varying query complexity, often leading to suboptimal retrieval either including irrelevant documents or missing crucial ones.
+    {\bf Objectives:}
    To address this issue, we propose dyRAG, a hybrid method that dynamically adjusts K based on query characteristics while maintaining computational efficiency.
-Our method improves retrieval performance by maximizing relevant information and minimizing noise.
-Experimental results show that it outperforms fixed-K retrieval, offering a more effective solution for optimizing RAG systems.
-Compared to traditional methods, it dynamically adjusts based on the query and dataset characteristics, offering greater adaptability across various contexts.}
+Our goal is to improve retrieval performance by maximizing relevant information and minimizing noise.
 
-	%%\pacs[JEL Classification]{D8, H51}
+    {\bf Methods:}
+   We introduce dyRAG, which dynamically adjusts the number of retrieved documents (K) in response to query complexity and dataset characteristics.
+The method balances precision and computational efficiency to optimize RAG systems.
 
-	%%\pacs[MSC Classification]{35A01, 65L10, 65L12, 65L20, 65L70}
+    {\bf Results:}
+    Experimental results show that dyRAG outperforms fixed-K retrieval, offering a more effective solution for optimizing RAG systems.
+It demonstrates greater adaptability across various contexts compared to traditional methods.
 
+    {\bf Conclusions:}
+    dyRAG provides a dynamic and efficient approach to document retrieval in RAG systems, improving performance by tailoring retrieval to query-specific needs.
+This advancement enhances the flexibility and effectiveness of RAG in diverse applications.
+\end{abstract}
+
+%\begin{abstract}
+%      A clear and well-documented \LaTeX\ document is presented as an
+%  article formatted for publication by ACM in a conference proceedings
+%  or journal publication. Based on the ``acmart'' document class, this
+%  article presents and explains many of the common variations, as well
+%  as many of the formatting elements an author may use in the
+%  preparation of the documentation of their work.
+%\end{abstract}
+
+%% JAIR Note: 
+%% Do not include ACM CCS Concepts or Keywords
+
+%% To be updated by authors.
+\received{20 February 2007}
+\received[revised]{12 March 2009}
+\received[accepted]{5 June 2009}
+
+%%
+%% This command processes the author and affiliation and title
+%% information and builds the first part of the formatted document.
 \maketitle
 
-	\section{Introduction}\label{sec1}
-
+\section{Introduction}
 Large Language Models (LLMs) have emerged as state-of-the-art artificial intelligence systems that excel at understanding and generating human-like text\cite{Radford2019}, Built on deep learning architectures, particularly transformers \cite{vaswani2017attention},these models showcase impressive adaptability across diverse tasks, such as machine translation\cite{Gu2018SearchEngine}, text generation\cite{liang2024controllabletextgenerationlarge} and question answering\cite{Zhang2023}.
@@
 Despite their strengths, they are computationally expensive and impractical for real-time knowledge updates.
-A promising alternative is Retrieval-Augmented Generation (RAG) \cite{lewis2020retrieval}, which integrates real-time retrieval from external knowledge sources with generative models, enables LLMs to access up-to-date information without retraining,(RAG) faces several limitations.
+A promising alternative is Retrieval-Augmented Generation (RAG) \cite{Chuyuan2025}, which integrates real-time retrieval from external knowledge sources with generative models, enables LLMs to access up-to-date information without retraining,(RAG) faces several limitations.
 For instance, The performance of RAG heavily depends on the quality of the retriever \cite{karpukhin2020dpr}, as irrelevant or noisy documents can degrade the generator output.
@@
 The process involves feeding both the userss query and the relevant information retrieved by a search algorithm into the LLMs prompt\cite{ilin2023advancedrag}, enabling the model to generate answers based on the provided context.
-	gupta20241
 \begin{figure}[h]
@@
 
-	The query \( q \) represents the user's state at a specific time step, derived from their interaction history. In the MoL (Mixture of Logits) framework,\( q \)is transformed into \( Pq \)
+The query \( q \) represents the user's state at a specific time step, derived from their interaction history. In the MoL (Mixture of Logits) framework, \( q \) is transformed into \( Pq \)
 embeddings through a multi-layer perceptron (MLP).
@@
 \setlength{\tabcolsep}{1pt} % Increase column padding
+%
+%\begin{table}[ht]
+%	\centering
+%	\large
+%	\renewcommand{\arraystretch}{1.7} % Increase row spacing
+%	\resizebox{\textwidth}{!}{  % Scale table to fit width
+%		\begin{tabular}{|l|c|c|c|c|c|c|c|c|}
+%			\hline
+%			\textbf{Model} & \textbf{\shortstack{Max \\ Sequence Length}} & 
+%			\textbf{\shortstack{Embedding \\ Dimension}} & 
+%			\textbf{\shortstack{Number \\ of Heads}} & 
+%			\textbf{\shortstack{Feedforward \\ Dimension}} & 
+%			\textbf{\shortstack{Batch \\ Size}} & 
+%			\textbf{Epochs} & 
+%			\textbf{\shortstack{Val \\ Loss}} & 
+%			\textbf{\shortstack{Val \\ Accuracy}} \\
+%			\hline
+%			Hybrid+SAS & 50  & 128  & 2 & 128  & 128 & 12 & 5.3036 & 0.1584  \\ \hline
+%			MoL+SAS    & 50  & 128  & 2 & 128  & 128 & 12 & 5.3152 & 0.1582  \\ \hline
+%			Hybrid+SAS & 128 & 256  & 4 & 256  & 128 & 10 & 3.6364 & 0.4301  \\ \hline
+%			MoL+SAS    & 128 & 256  & 4 & 256  & 128 & 10 & 3.6374 & 0.4299 \\\hline
+%			Hybrid+SAS & 512 & 512  & 4 & 512  & 128 & 10 & 1.2808 & {\bf 0.8120} \\ \hline
+%			MoL+SAS    & 512 & 512  & 4 & 512  & 128 & 10 & 1.3050 & 0.8016 \\\hline
+%		\end{tabular}
+%	}
+%	\caption{Results on 100kMovies dataset}
+%	\label{tab:100k_movies}
+%\end{table}
 
@@
 	\centering
-		\large
-		\renewcommand{\arraystretch}{1.7} % Increase row spacing
-		\resizebox{\textwidth}{!}{  % Scale table to fit width
-			\begin{tabular}{|l|c|c|c|c|c|c|c|c|}
-				\hline
-				\textbf{Model} & \textbf{\shortstack{Max \\ Sequence Length}} & 
-				\textbf{\shortstack{Embedding \\ Dimension}} & 
-				\textbf{\shortstack{Number \\ of Heads}} & 
-				\textbf{\shortstack{Feedforward \\ Dimension}} & 
-				\textbf{\shortstack{Batch \\ Size}} & 
+	\renewcommand{\arraystretch}{1.5} % Balanced row spacing
+	\setlength{\tabcolsep}{0.8em}     % Increased column padding
+	\caption{Performance comparison of Hybrid+SAS and MoL+SAS models on the 100kMovies dataset}
+	\label{tab:100k_movies}
+	\resizebox{\textwidth}{!}{ % Scale to textwidth (remove if too small)
+		\begin{tabular}{@{}l*{8}{c}@{}} % Remove vertical lines, add horizontal padding
+			\toprule
+			\textbf{Model} & 
+			\textbf{\makecell{Max \\ Seq. Len.}} & 
+			\textbf{\makecell{Embed. \\ Dim.}} & 
+			\textbf{Heads} & 
+			\textbf{\makecell{FFN \\ Dim.}} & 
+			\textbf{\makecell{Batch \\ Size}} & 
 			\textbf{Epochs} & 
-				\textbf{\shortstack{Val \\ Loss}} & 
-				\textbf{\shortstack{Val \\ Accuracy}} \\
-				\hline
-				Hybrid+SAS & 50  & 128  & 2 & 128  & 128 & 12 & 5.3036 & 0.1584  \\ \hline
-				MoL+SAS    & 50  & 128  & 2 & 128  & 128 & 12 & 5.3152 & 0.1582  \\ \hline
-				Hybrid+SAS & 128 & 256  & 4 & 256  & 128 & 10 & 3.6364 & 0.4301  \\ \hline
-				MoL+SAS    & 128 & 256  & 4 & 256  & 128 & 10 & 3.6374 & 0.4299 \\\hline
-				Hybrid+SAS & 512 & 512  & 4 & 512  & 128 & 10 & 1.2808 & {\bf 0.8120} \\ \hline
-				MoL+SAS    & 512 & 512  & 4 & 512  & 128 & 10 & 1.3050 & 0.8016 \\\hline
+			\textbf{\makecell{Val. \\ Loss}} & 
+			\textbf{\makecell{Val. \\ Acc.}} \\
+			\midrule
+			Hybrid+SAS & 50  & 128 & 2 & 128 & 128 & 12 & 5.3036 & 0.1584  \\ 
+			MoL+SAS    & 50  & 128 & 2 & 128 & 128 & 12 & 5.3152 & 0.1582  \\ 
+			\addlinespace[0.2em] % Light spacing between groups
+			Hybrid+SAS & 128 & 256 & 4 & 256 & 128 & 10 & 3.6364 & 0.4301  \\ 
+			MoL+SAS    & 128 & 256 & 4 & 256 & 128 & 10 & 3.6374 & 0.4299  \\
+			\addlinespace[0.2em]
+			Hybrid+SAS & 512 & 512 & 4 & 512 & 128 & 10 & \textbf{1.2808} & \textbf{0.8120} \\ 
+			MoL+SAS    & 512 & 512 & 4 & 512 & 128 & 10 & 1.3050 & 0.8016  \\
+			\bottomrule
 		\end{tabular}
 	}
-		\caption{Results on 100kMovies dataset}
-		\label{tab:100k_movies}
 \end{table}
@@
 
+%\begin{table}[ht]
+%	\centering
+%	\large
+%	\renewcommand{\arraystretch}{1.7} % Increase row spacing
+%	\resizebox{\textwidth}{!}{  % Scale table to fit width
+%		\begin{tabular}{|l|c|c|c|c|c|c|c|c|}
+%			\hline
+%			\textbf{Model} & \textbf{\shortstack{Max \\ Sequence Length}} & 
+%			\textbf{\shortstack{Embedding \\ Dimension}} & 
+%			\textbf{\shortstack{Number \\ of Heads}} & 
+%			\textbf{\shortstack{Feedforward \\ Dimension}} & 
+%			\textbf{\shortstack{Batch \\ Size}} & 
+%			\textbf{Epochs} & 
+%			\textbf{\shortstack{Val \\ Loss}} & 
+%			\textbf{\shortstack{Val \\ Accuracy}} \\
+%			\hline
+%			Hybrid+SAS & 50  & 128  & 2 & 128  & 128 & 10 & 4.5721 & 0.1530  \\ \hline
+%			MoL+SAS    & 50  & 128  & 2 & 128  & 128 & 10 & 4.5733 & 0.1526  \\ \hline
+%			Hybrid+SAS & 128 & 256  & 4 & 256  & 128 & 10 & 3.4152 & 0.3680  \\ \hline
+%			MoL+SAS    & 128 & 256  & 4 & 256  & 128 & 10 & 3.4671 & 0.3627  \\ \hline
+%			Hybrid+SAS & 512 & 512  & 4 & 512  & 128 & 10 & 1.0275 & {\bf 0.8350} \\ \hline
+%			MoL+SAS    & 512 & 512  & 4 & 512  & 128 & 10 & 1.0350 & 0.8276  \\ \hline
+%			
+%		\end{tabular}
+%	}
+%	\caption{Results on 1M Movies dataset}
+%	\label{tab:1m_movies}
+%\end{table}
 \begin{table}[ht]
 	\centering
-		\large
-		\renewcommand{\arraystretch}{1.7} % Increase row spacing
-		\resizebox{\textwidth}{!}{  % Scale table to fit width
-			\begin{tabular}{|l|c|c|c|c|c|c|c|c|}
-				\hline
-				\textbf{Model} & \textbf{\shortstack{Max \\ Sequence Length}} & 
-				\textbf{\shortstack{Embedding \\ Dimension}} & 
-				\textbf{\shortstack{Number \\ of Heads}} & 
-				\textbf{\shortstack{Feedforward \\ Dimension}} & 
-				\textbf{\shortstack{Batch \\ Size}} & 
+	\renewcommand{\arraystretch}{1.5} % Balanced row spacing
+	\setlength{\tabcolsep}{0.8em}    % Improved column padding
+	\caption{Performance comparison of Hybrid+SAS and MoL+SAS models on the 1M Movies dataset}
+	\label{tab:1m_movies}
+	\resizebox{\textwidth}{!}{ % Remove if table appears too compressed
+		\begin{tabular}{@{}l*{8}{c}@{}}
+			\toprule
+			\textbf{Model} & 
+			\textbf{\makecell{Max \\ Seq. Len.}} & 
+			\textbf{\makecell{Embed. \\ Dim.}} & 
+			\textbf{Heads} & 
+			\textbf{\makecell{FFN \\ Dim.}} & 
+			\textbf{\makecell{Batch \\ Size}} & 
 			\textbf{Epochs} & 
-				\textbf{\shortstack{Val \\ Loss}} & 
-				\textbf{\shortstack{Val \\ Accuracy}} \\
-				\hline
-				Hybrid+SAS & 50  & 128  & 2 & 128  & 128 & 10 & 4.5721 & 0.1530  \\ \hline
-				MoL+SAS    & 50  & 128  & 2 & 128  & 128 & 10 & 4.5733 & 0.1526  \\ \hline
-				Hybrid+SAS & 128 & 256  & 4 & 256  & 128 & 10 & 3.4152 & 0.3680  \\ \hline
-				MoL+SAS    & 128 & 256  & 4 & 256  & 128 & 10 & 3.4671 & 0.3627  \\ \hline
-				Hybrid+SAS & 512 & 512  & 4 & 512  & 128 & 10 & 1.0275 & {\bf 0.8350} \\ \hline
-				MoL+SAS    & 512 & 512  & 4 & 512  & 128 & 10 & 1.0350 & 0.8276  \\ \hline
-
+			\textbf{\makecell{Val. \\ Loss}} & 
+			\textbf{\makecell{Val. \\ Acc.}} \\
+			\midrule
+			Hybrid+SAS & 50  & 128 & 2 & 128 & 128 & 10 & \textbf{4.5721} & \textbf{0.1530}  \\ 
+			MoL+SAS    & 50  & 128 & 2 & 128 & 128 & 10 & 4.5733 & 0.1526  \\ 
+			\addlinespace[0.2em]
+			Hybrid+SAS & 128 & 256 & 4 & 256 & 128 & 10 & \textbf{3.4152} & \textbf{0.3680}  \\ 
+			MoL+SAS    & 128 & 256 & 4 & 256 & 128 & 10 & 3.4671 & 0.3627  \\
+			\addlinespace[0.2em]
+			Hybrid+SAS & 512 & 512 & 4 & 512 & 128 & 10 & \textbf{1.0275} & \textbf{0.8350} \\ 
+			MoL+SAS    & 512 & 512 & 4 & 512 & 128 & 10 & 1.0350 & 0.8276  \\
+			\bottomrule
 		\end{tabular}
 	}
-		\caption{Results on 1M Movies dataset}
-		\label{tab:1m_movies}
 \end{table}
+
 \newpage
@@
 
-	\newpage
-
-	%%===================================================%%
-	%% For presentation purpose, we have included        %%
-	%% \bigskip command. Please ignore this.             %%
-	%%===================================================%%
-
-	%%===========================================================================================%%
-	%% If you are submitting to one of the Nature Portfolio journals, using the eJP submission   %%
-	%% system, please include the references within the manuscript file itself. You may do this  %%
-	%% by copying the reference list from your .bbl file, paste it into the main manuscript .tex %%
-	%% file, and delete the associated \verb+\bibliography+ commands.                            %%
-	%%===========================================================================================%%
-
-	\bibliography{sn-bibliography}% common bib file
-	%% if required, the content of .bbl file can be included here once bbl is generated
-	%%\input sn-article.bbl
+%
+%\section{Research Methods}
+%
+%\subsection{Part One}
+%
+%Lorem ipsum dolor sit amet, consectetur adipiscing elit. Morbi
+%malesuada, quam in pulvinar varius, metus nunc fermentum urna, id
+%sollicitudin purus odio sit amet enim. Aliquam ullamcorper eu ipsum
+%vel mollis. Curabitur quis dictum nisl. Phasellus vel semper risus, et
+%lacinia dolor. Integer ultricies commodo sem nec semper.
+%
+%\subsection{Part Two}
+%
+%Etiam commodo feugiat nisl pulvinar pellentesque. Etiam auctor sodales
+%ligula, non varius nibh pulvinar semper. Suspendisse nec lectus non
+%ipsum convallis congue hendrerit vitae sapien. Donec at laoreet
+%eros. Vivamus non purus placerat, scelerisque diam eu, cursus
+%ante. Etiam aliquam tortor auctor efficitur mattis.
+%
+%\section{Online Resources}
+%
+%Nam id fermentum dui. Suspendisse sagittis tortor a nulla mollis, in
+%pulvinar ex pretium. Sed interdum orci quis metus euismod, et sagittis
+%enim maximus. Vestibulum gravida massa ut felis suscipit
+%congue. Quisque mattis elit a risus ultrices commodo venenatis eget
+%dui. Etiam sagittis eleifend elementum.
+%
+%Nam interdum magna at lectus dignissim, ac dignissim lorem
+%rhoncus. Maecenas eu arcu ac neque placerat aliquam. Nunc pulvinar
+%massa et mattis lacinia.
+%
+%\section{Reproducibility Checklist for JAIR}
+%
+%Select the answers that apply to your research -- one per item. 
+%
+%\subsection*{All articles:}
+%
+%%\hh{revised for stylistic consistency:}
+%\begin{enumerate}
+%    \item All claims investigated in this work are clearly stated. 
+%    [yes/partially/no]
+%    \item Clear explanations are given how the work reported substantiates the claims. 
+%    [yes/partially/no]
+%    \item Limitations or technical assumptions are stated clearly and explicitly. 
+%    [yes/partially/no]
+%    \item Conceptual outlines and/or pseudo-code descriptions of the AI methods introduced in this work are provided, and important implementation details are discussed. 
+%    [yes/partially/no/NA]
+%    \item 
+%    Motivation is provided for all design choices, including algorithms, implementation choices, parameters, data sets and experimental protocols beyond metrics.
+%    [yes/partially/no]
+%\end{enumerate}
+%
+%\subsection*{Articles containing theoretical contributions:}
+%Does this paper make theoretical contributions? 
+%[yes/no] 
+%
+%If yes, please complete the list below.
+%
+%\begin{enumerate}
+%    \item All assumptions and restrictions are stated clearly and formally. 
+%    [yes/partially/no]
+%    \item All novel claims are stated formally (e.g., in theorem statements). 
+%    [yes/partially/no]
+%    \item Proofs of all non-trivial claims are provided in sufficient detail to permit verification by readers with a reasonable degree of expertise (e.g., that expected from a PhD candidate in the same area of AI).
+[yes/partially/no]
+%    \item
+%    Complex formalism, such as definitions or proofs, is motivated and explained clearly.
+%%hh: was:
+%%Proof sketches or intuitions are given for complex and/or novel results.
+%    [yes/partially/no]
+%    \item 
+%    The use of mathematical notation and formalism serves the purpose of enhancing clarity and precision; gratuitous use of mathematical formalism (i.e., use that does not enhance clarity or precision) is avoided.
+%    [yes/partially/no]
+%    \item 
+%    Appropriate citations are given for all non-trivial theoretical tools and techniques. 
+%    [yes/partially/no]
+%\end{enumerate}
+%
+%\subsection*{Articles reporting on computational experiments:}
+%Does this paper include computational experiments? [yes/no]
+%
+%If yes, please complete the list below.
+%\begin{enumerate}
+%    \item 
+%    All source code required for conducting experiments is included in an online appendix 
+%    or will be made publicly available upon publication of the paper.
+%    The online appendix follows best practices for source code readability and documentation as well as for long-term accessibility.
+%    [yes/partially/no]
+%    \item The source code comes with a license that
+%    allows free usage for reproducibility purposes.
+%    [yes/partially/no]
+%    \item The source code comes with a license that
+%    allows free usage for research purposes in general.
+%    [yes/partially/no]
+%    \item 
+%    Raw, unaggregated data from all experiments is included in an online appendix 
+%    or will be made publicly available upon publication of the paper.
+%    The online appendix follows best practices for long-term accessibility.
+%    [yes/partially/no]
+%    \item The unaggregated data comes with a license that
+%    allows free usage for reproducibility purposes.
+%    [yes/partially/no]
+%    \item The unaggregated data comes with a license that
+%    allows free usage for research purposes in general.
+%    [yes/partially/no]
+%    \item If an algorithm depends on randomness, then the method used for generating random numbers and for setting seeds is described in a way sufficient to allow replication of results. 
+%    [yes/partially/no/NA]
+%    \item The execution environment for experiments, the computing infrastructure (hardware and software) used for running them, is described, including GPU/CPU makes and models; amount of memory (cache and RAM); make and version of operating system; names and versions of relevant software libraries and frameworks.
+%    [yes/partially/no]
+%    \item 
+%    The evaluation metrics used in experiments are clearly explained and their choice is explicitly motivated. 
+%    [yes/partially/no]
+%    \item 
+%    The number of algorithm runs used to compute each result is reported. 
+%    [yes/no]
+%    \item 
+%    Reported results have not been ``cherry-picked'' by silently ignoring unsuccessful or unsatisfactory experiments. 
+%    [yes/partially/no]
+%    \item 
+%    Analysis of results goes beyond single-dimensional summaries of performance (e.g., average, median) to include measures of variation, confidence, or other distributional information. 
+%    [yes/no]
+%    \item 
+%    All (hyper-) parameter settings for 
+%    the algorithms/methods used in experiments have been reported, along with the rationale or method for determining them. 
+%    [yes/partially/no/NA]
+%    \item 
+%    The number and range of (hyper-) parameter settings explored prior to conducting final experiments have been indicated, along with the effort spent on (hyper-) parameter optimisation. 
+%    [yes/partially/no/NA]
+%    \item 
+%    Appropriately chosen statistical hypothesis tests are used to establish statistical significance
+%    in the presence of noise effects.
+%    [yes/partially/no/NA]
+%\end{enumerate}
+%
+%
+%\subsection*{Articles using data sets:}
+%Does this work rely on one or more data sets (possibly obtained from a benchmark generator or similar software artifact)? 
+%[yes/no]
+%
+%If yes, please complete the list below.
+%\begin{enumerate}
+%    \item 
+%    All newly introduced data sets 
+%    are included in an online appendix 
+%    or will be made publicly available upon publication of the paper.
+%    The online appendix follows best practices for long-term accessibility with a license
+%    that allows free usage for research purposes.
+%    [yes/partially/no/NA]
+%    \item The newly introduced data set comes with a license that
+%    allows free usage for reproducibility purposes.
+%    [yes/partially/no]
+%    \item The newly introduced data set comes with a license that
+%    allows free usage for research purposes in general.
+%    [yes/partially/no]
+%    \item All data sets drawn from the literature or other public sources (potentially including authors' own previously published work) are accompanied by appropriate citations.
+%    [yes/no/NA]
+%    \item All data sets drawn from the existing literature (potentially including authors’ own previously published work) are publicly available. [yes/partially/no/NA]
+%    %\item All data sets that are not publicly available are described in detail.
+%    %[yes/partially/no/NA]
+%    \item All new data sets and data sets that are not publicly available are described in detail, including relevant statistics, the data collection process and annotation process if relevant.
+%    [yes/partially/no/NA]
+%    \item 
+%    All methods used for preprocessing, augmenting, batching or splitting data sets (e.g., in the context of hold-out or cross-validation)
+%    are described in detail. [yes/partially/no/NA]
+%\end{enumerate}
+%
+%\subsection*{Explanations on any of the answers above (optional):}
+%
+%[Text here; please keep this brief.]
+%%
+%% The next two lines define the bibliography style to be used, and
+%% the bibliography file.
+\bibliographystyle{ACM-Reference-Format}
+\bibliography{sn-bibliography}
 
 \end{document}
+\endinput
+%%
+%% End of file `sample-acmlarge.tex'.

### 2025-07-18 Paper/Submission3/RAG3.tex
All changes
Last submitted paper is in folder submission3

--- 2025-07-18 Paper/JAIR_Example_Template/JAIR_RAG_Paper.tex
@@
-%% This is a JAIR Example File Compiled by Nicholas Mattei (nsmattei@tulane.edu) 
-%% and Odd Erik Gundersen (odderik@ntnu.no)
-%% and Mykel Kochenderfer (mykel@stanford.edu)
-%% March 29, 2025
-%%
-%% This file is based off the ACM Latex Template https://www.acm.org/publications/proceedings-template
-%% Revision 2.12 (12/28/2024)
-%% 
-%% Please see https://www.jair.org/index.php/jair for more information and submission instructions.
-%%
+% sage_latex_guidelines.tex V1.20, 14 January 2017
 
-%% The first command in your LaTeX source must be the \documentclass
-%% command.
-%%
-%% For submission and review of your manuscript please change the
-%% command to \documentclass[manuscript, screen, review]{jair}.
-%%
-\documentclass[review]{jair}
+\documentclass[Afour,sageh,times]{sagej}
 
-\setcopyright{cc}
-\copyrightyear{2025}
-\acmYear{2025}
-\acmDOI{10.1613/jair.1.xxxxx}
+\usepackage{moreverb,url}
 
-%%
-\JAIRAE{Insert JAIR AE Name}
-\JAIRTrack{Insert JAIR Track Name Here}
-\acmVolume{4}
-\acmArticle{111}
-\acmMonth{8}
-\acmYear{2025}
+\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}
 
-%%
-%% For managing citations, it is recommended to use bibliography
-%% files in BibTeX format.
-%%
-%% You can then either use BibTeX with the ACM-Reference-Format style,
-%% or BibLaTeX with the acmnumeric or acmauthoryear sytles, that include
-%% support for advanced citation of software artefact from the
-%% biblatex-software package, also separately available on CTAN.
-%%
-%% Look at the sample-*-biblatex.tex files for templates showcasing
-%% the biblatex styles.
-%%
+\newcommand\BibTeX{{\rmfamily B\kern-.05em \textsc{i\kern-.025em b}\kern-.08em
+T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
 
-%%
-%% The majority of ACM publications use numbered citations and
-%% references.  The command \citestyle{authoryear} switches to the
-%% "author year" style.
-%%
-%% If you are preparing content for an event
-%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
-%% citations and references.
-%% Uncommenting
-%% the next command will enable that style.
-%%\citestyle{acmauthoryear}
+\def\volumeyear{2016}
 
-%%
-%% end of the preamble, start of the body of the document source.
-\usepackage{graphicx}%
-\usepackage{multirow}%
-%\usepackage{amsfonts, amsmath,amssymb}%
-%%\usepackage{amsthm}%
-%%\usepackage{mathrsfs}%
-%\usepackage[title]{appendix}%
-%\usepackage{xcolor}%
-%\usepackage{textcomp}%
-%\usepackage{manyfoot}%
-%\usepackage{booktabs}%
 \usepackage{algorithm}%
@@
 
-%%
-%% The "title" command has an optional parameter,
-%% allowing the author to define a "short title" to be used in page headers.
-%\title{JAIR Example Template}
-\title[dyRAG.Dynamic K-Selection]{Boosting RAG Efficiency with dyRAG: Dynamic K-Selection for Optimal Retrieval}
+\runninghead{Smith and Wittkopf}
 
-%%
-%% The "author" command and its associated commands are used to define
-%% the authors and their affiliations.
-%% Of note is the shared affiliation of the first two authors, and the
-%% "authornote" and "authornotemark" commands
-%% used to denote shared contribution to the research and/or corresponding author.
-\author{Zoubida Asmaa Boudjenane}
+\title{Boosting RAG Efficiency with dyRAG: Dynamic Candidate Selection for Optimal Retrieval}
 
-\email{asmaa.boudj04@gmail.com}
-\orcid{0000-0002-2037-3694}
-% \orcid{https://orcid.org/0000-0002-2037-3694} % This works too if you want to see the full URL
-\affiliation{%
-  \institution{Computer Science Department, University of Mascara}
-  \city{Mascara}
-  \state{Mascara}
-  \country{Algeria}
-}
+\author{Zoubida Asmaa Boudjenane \affilnum{1} and Mohammed SALEM \affilnum{2}}
 
-\author{Mohammed Salem}
-\orcid{0000-0002-2037-3694}
-\authornote{Corresponding Author.}
+\affiliation{\affilnum{1}Computer Science Department, University of Mascara, Algeria\\
+\affilnum{2}LISYS Laboratory, University of Mascara, Algeria}
+
+\corrauth{Mohammed SALEM, LISYS Lab, University of Mascara, BP 305 Route Mamounia, Mascara, 29000, Algeria.}
+
 \email{salem@univ-mascara.dz}
-\affiliation{%
-  \institution{LISYS Laboratory, University of Mascara}
-  \city{Mascara}
-  \state{Mascara}
-  \country{Algeria}}
 
-%\author{Holger H. Hoos}
-%\orcid{0000-0003-0629-0099}
-%\email{hoos@cs.ubc.ca}
-%\affiliation{%
-%  \institution{University of British Columbia}
-%  \city{Vancouver}
-%  \state{British Columbia}
-%  \country{Canada}
-%}
-
-%\author{Kevin Leyton-Brown}
-%\orcid{0000-0002-7644-5327}
-%\email{kevinlb@cs.ubc.ca}
-%\affiliation{%
-%  \institution{University of British Columbia}
-%  \city{Vancouver}
-%  \state{British Columbia}
-%  \country{Canada}
-%}
-
-%\author{Ben Trovato}
-
-%%
-%% By default, the full list of authors will be used in the page
-%% headers. Often, this list is too long, and will overlap
-%% other information printed in the page headers. This command allows
-%% the author to define a more concise list
-%% of authors' names for this purpose.
-\renewcommand{\shortauthors}{Boudjenane and Salem.}
-
-%%
-%% The abstract is a short summary of the work to be presented in the
-%% article.
 \begin{abstract}
-{\bf Background:} 
 Retrieval-Augmented Generation (RAG) enhances language model responses by incorporating external knowledge.
 However, its effectiveness heavily depends on the quality of the retrieved documents.
-Using a fixed number of retrieved documents (K) may fail to adapt to varying query complexity, often leading to suboptimal retrieval—either including irrelevant documents or missing crucial ones.
-
-    {\bf Objectives:}
+Using a fixed number of retrieved documents K may fail  to adapt to varying query complexity, often leading to suboptimal retrieval either including irrelevant documents or missing crucial ones.
 To address this issue, we propose dyRAG, a hybrid method that dynamically adjusts K based on query characteristics while maintaining computational efficiency.
-Our goal is to improve retrieval performance by maximizing relevant information and minimizing noise.
-
-    {\bf Methods:}
-   We introduce dyRAG, which dynamically adjusts the number of retrieved documents (K) in response to query complexity and dataset characteristics.
-The method balances precision and computational efficiency to optimize RAG systems.
-
-    {\bf Results:}
-    Experimental results show that dyRAG outperforms fixed-K retrieval, offering a more effective solution for optimizing RAG systems.
-It demonstrates greater adaptability across various contexts compared to traditional methods.
-
-    {\bf Conclusions:}
-    dyRAG provides a dynamic and efficient approach to document retrieval in RAG systems, improving performance by tailoring retrieval to query-specific needs.
-This advancement enhances the flexibility and effectiveness of RAG in diverse applications.
+Our method improves retrieval performance by maximizing relevant information and minimizing noise.
+Experimental results show that it outperforms fixed-K retrieval, offering a more effective solution for optimizing RAG systems.
+Compared to traditional methods, it dynamically adjusts based on the query and dataset characteristics, offering greater adaptability across various contexts.
 \end{abstract}
 
-%\begin{abstract}
-%      A clear and well-documented \LaTeX\ document is presented as an
-%  article formatted for publication by ACM in a conference proceedings
-%  or journal publication. Based on the ``acmart'' document class, this
-%  article presents and explains many of the common variations, as well
-%  as many of the formatting elements an author may use in the
-%  preparation of the documentation of their work.
-%\end{abstract}
+\keywords{Retrieval-Augmented Generation (RAG), Dynamic Retrieval (dyRAG), large Llanguage model, 	Query Complexity, Relevant Information}
 
-%% JAIR Note: 
-%% Do not include ACM CCS Concepts or Keywords
+\maketitle
+\section{Introduction}
+Large Language Models (LLMs) have emerged as state-of-the-art artificial intelligence systems that excel at understanding and generating human-like text \cite{Radford2019}, Built on deep learning architectures, particularly transformers \cite{vaswani2017attention},these models showcase impressive adaptability across diverse tasks, such as machine translation \cite{Gu2018SearchEngine}, text generation \cite{liang2024controllabletextgenerationlarge} and question answering \cite{Zhang2023}.
+Notable examples of LLMs include OpenAI GPT series \cite{brown2020language}, Google BERT \cite{devlin2018bert} and T5 \cite{raffel2020exploring}.
+Despite their impressive capabilities, LLMs suffer from hallucination \cite{Radford2019}, often generating confident but incorrect information.
+This is a critical challenge for knowledge-intensive tasks requiring high accuracy \cite{Radford2019}.
+Additionally, their static nature being trained on fixed datasets \cite{Radford2019} limits their ability to update knowledge dynamically, making them prone to outdated or inconsistent responses, especially for time-sensitive information \cite{Mousavi2025}.
 
-%% To be updated by authors.
-\received{20 February 2007}
-\received[revised]{12 March 2009}
-\received[accepted]{5 June 2009}
-
-%%
-%% This command processes the author and affiliation and title
-%% information and builds the first part of the formatted document.
-\maketitle
-
-\section{Introduction}
-Large Language Models (LLMs) have emerged as state-of-the-art artificial intelligence systems that excel at understanding and generating human-like text\cite{Radford2019}, Built on deep learning architectures, particularly transformers \cite{vaswani2017attention},these models showcase impressive adaptability across diverse tasks, such as machine translation\cite{Gu2018SearchEngine}, text generation\cite{liang2024controllabletextgenerationlarge} and question answering\cite{Zhang2023}.
-Notable examples of LLMs include OpenAI GPT series\cite{brown2020language}, Google BERT\cite{devlin2018bert} and T5 \cite{raffel2020exploring}.
-Despite their impressive capabilities, LLMs suffer from hallucination\cite{Radford2019}, often generating confident but incorrect information.
-This is a critical challenge for knowledge-intensive tasks requiring high accuracy\cite{Radford2019}.
-Additionally, their static nature being trained on fixed datasets\cite{Radford2019} limits their ability to update knowledge dynamically, making them prone to outdated or inconsistent responses, especially for time-sensitive information.\cite{Mousavi2025}.
-
-Researchers have explored various approaches to address this limitation, including transfer learning\cite{zhuang2020comprehensive}, fine-tuning \cite{Howard2018} which adapt pre-trained models to specific tasks.
+Researchers have explored various approaches to address this limitation, including transfer learning \cite{zhuang2020comprehensive}, fine-tuning \cite{Howard2018} which adapt pre-trained models to specific tasks.
 Despite their strengths, they are computationally expensive and impractical for real-time knowledge updates.
@@
 For instance, The performance of RAG heavily depends on the quality of the retriever \cite{karpukhin2020dpr}, as irrelevant or noisy documents can degrade the generator output.
-Additionally, a critical challenge is determining the optimal number of retrieved passages K to feed into the generator\cite{lewis2020retrieval}.
+Additionally, a critical challenge is determining the optimal number of retrieved passages K to feed into the generator \cite{lewis2020retrieval}.
 Often resulting in either insufficient or excessive information being retrieved, reduce coherence and increase computational overhead.
@@
 
-In this paper, we investigate the problem of selecting K in RAG and propose a novel approach to dynamically adjust K based on  thresholds and a Mixture of Logits (MoL)\cite{ding2024efficient} scoring mechanism, ensuring efficient and context-sensitive candidate selection while reducing computational costs.
+In this paper, we investigate the problem of selecting K in RAG and propose a novel approach to dynamically adjust K based on  thresholds and a Mixture of Logits (MoL) \cite{ding2024efficient} scoring mechanism, ensuring efficient and context-sensitive candidate selection while reducing computational costs.
 The use of  adaptive thresholds algorithm improves the relevance of retrieved documents, refining the set to include only the most contextually appropriate matches.
@@
 Static Top-K selection in Retrieval :In traditional Retrieval-Augmented Generation (RAG) systems, static K-selection is a widely used method where a fixed number of top-k documents are retrieved based on their relevance scores.
-Sparse retrieval\cite{bai2020spartermlearningtermbasedsparse} methods where documents and queries are represented as high-dimensional, sparse vectors In these method, the "keys" represent documents to be retrieved, and the "values" are the importance or weight of those terms in the document such as query likelihood models\cite{Lafferty2001}, TF-IDF\cite{Robertson1997} and BM25\cite{Zaragoza2009} often rely on this static approach to identify and rank documents.
+Sparse retrieval \cite{bai2020spartermlearningtermbasedsparse} methods where documents and queries are represented as high-dimensional, sparse vectors In these method, the "keys" represent documents to be retrieved, and the "values" are the importance or weight of those terms in the document such as query likelihood models \cite{Lafferty2001}, TF-IDF \cite{Robertson1997} and BM25 \cite{Zaragoza2009} often rely on this static approach to identify and rank documents.
 For instance, if k=5, the system retrieves the top 5 documents deemed most relevant to a query, regardless of the query's complexity or the variability in relevance distribution among documents.
@@
 Dynamic top-K Selection in Retrieval :addresses the limitations of static K by varying the number of retrieved documents based on the specific query or context.
-These approaches typically involve heuristics, machine learning models, or adaptive algorithms to determine the optimal K for each query such as Dynamic Query Cutoff (DQC)\cite{Culpepper2016} This approach uses machine learning models trained on query-document features to predict the optimal K for retrieval, a reinforcement learning-based approach\cite{Zhou2021} where the agent learns to adjust K by maximizing downstream task performance (e.g accurate response generation in RAG) Another example is the Dynamic Selection of K Nearest Neighbors \cite{DynamicKNN2012} where the selection process often depends on the distribution of neighbors, their distances to the query point, or their relevance scores.
+These approaches typically involve heuristics, machine learning models, or adaptive algorithms to determine the optimal K for each query such as Dynamic Query Cutoff (DQC) \cite{Culpepper2016} This approach uses machine learning models trained on query-document features to predict the optimal K for retrieval, a reinforcement learning-based approach \cite{Zhou2021} where the agent learns to adjust K by maximizing downstream task performance (e.g accurate response generation in RAG) Another example is the Dynamic Selection of K Nearest Neighbors \cite{DynamicKNN2012} where the selection process often depends on the distribution of neighbors, their distances to the query point, or their relevance scores.
 However, these methods come with limitations.
@@
 Hybrid Approaches For Top-K selection in Retrieval: combine elements of static and dynamic K-selection to achieve a balance between simplicity and adaptability.
-These systems might use a static K as a baseline but dynamically adjust it such as, the "Blended RAG"\cite{sawarkar2024blendedragimprovingrag} approach enhances retrieval accuracy by combining semantic search techniques with hybrid query strategies, Furthermore STAYKATE (Static-Dynamic Hybrid Selection)\cite{Zhu2024} enhances LLM performance in scientific information extraction by integrating representativeness sampling from active learning with a retrieval-based strategy.
-Another example is DR-RAG (Dynamic Relevance for Retrieval-Augmented Generation)\cite{hei2024drragapplyingdynamicdocument} First, it retrieves an initial set of $k_1$ documents.
+These systems might use a static K as a baseline but dynamically adjust it such as, the "Blended RAG" \cite{sawarkar2024blendedragimprovingrag} approach enhances retrieval accuracy by combining semantic search techniques with hybrid query strategies, Furthermore STAYKATE (Static-Dynamic Hybrid Selection) \cite{Zhu2024} enhances LLM performance in scientific information extraction by integrating representativeness sampling from active learning with a retrieval-based strategy.
+Another example is DR-RAG (Dynamic Relevance for Retrieval-Augmented Generation) \cite{hei2024drragapplyingdynamicdocument} First, it retrieves an initial set of $k_1$ documents.
 Then, a classifier dynamically evaluates their relevance and retrieves additional relevant $K_2$ documents to enhance recall.
@@
 \section{Overview of Retrieval-Augmented Generation }\label{sec3}
-Retrieval-Augmented Generation (RAG) is a method designed to enhance the performance of large language models (LLMs) by integrating external, reliable knowledge sources into their response generation process\cite{awsRAG}.
+Retrieval-Augmented Generation (RAG) is a method designed to enhance the performance of large language models (LLMs) by integrating external, reliable knowledge sources into their response generation process \cite{awsRAG}.
 RAG combines search capabilities with LLM prompting.
-The process involves feeding both the userss query and the relevant information retrieved by a search algorithm into the LLMs prompt\cite{ilin2023advancedrag}, enabling the model to generate answers based on the provided context.
+The process involves feeding both the userss query and the relevant information retrieved by a search algorithm into the LLMs prompt \cite{ilin2023advancedrag}, enabling the model to generate answers based on the provided context.
 \begin{figure}[h]
 	\centering
-	\includegraphics[width=0.9\textwidth]{rag.JPG}
-	\caption{RAG system component\cite{gupta20241}. }\label{rag.PNG}
+	\includegraphics[width=0.9\linewidth]{rag.JPG}
+	\caption{RAG system component \cite{gupta20241}. }\label{rag.PNG}
 \end{figure}
@@
 The retriever's role is to find relevant information from a pre-built data store, while the generator utilizes this retrieved data to produce coherent and contextually appropriate content.
-The overall RAG process operates as follows :
+The overall RAG process operates as follows:
 \subsection{Retrieval in RAG Systems}\label{subsec2}
@@
 Depending on the similarity function used, retrieval methods can be grouped into categories such as sparse retrieval\cite{bai2020spartermlearningtermbasedsparse}which relies on exact term matching techniques like BM25 and TF-IDF, dense retrieval\cite{karpukhin2020dpr}which Uses deep learning models to encode queries and documents into dense vector embeddings, retrieving relevant documents based on semantic similarity rather than exact word matching, and others.
-For the widely used sparse and dense retrieval approaches, the process typically involves two main steps:  \\
-1. Encoding each object into a specific representation (e.g., term-based for sparse retrieval, vector embeddings for dense retrieval).\\
-2. Building an index to organize and efficiently retrieve relevant data during search .
-
-This structured approach ensures that the most relevant information is retrieved quickly and accurately\cite{zhao2024retrievalaugmentedgenerationaigeneratedcontent}.
+For the widely used sparse and dense retrieval approaches, the process typically involves two main steps:
+\begin{itemize}
+\item Encoding each object into a specific representation (e.g., term-based for sparse retrieval, vector embeddings for dense retrieval).
+\item Building an index to organize and efficiently retrieve relevant data during search .
+\end{itemize}
+This structured approach ensures that the most relevant information is retrieved quickly and accurately \cite{zhao2024retrievalaugmentedgenerationaigeneratedcontent}.
 
 \subsection{Generator  in RAG Systems}\label{subsec2}
-In Retrieval-Augmented Generation (RAG) systems, the generator\cite{huang2024surveyretrievalaugmentedtextgeneration} is a key component responsible for crafting the final output.
+In Retrieval-Augmented Generation (RAG) systems, the generator  \cite{huang2024surveyretrievalaugmentedtextgeneration} is a key component responsible for crafting the final output.
 It works by combining the retrieved information with the users input query to produce a well-structured and contextually relevant response.
@@
 
-In various generative tasks, different models are selected based on the specific requirements: Models like BART\cite{lewis2019bart}, GPT\cite{brown2020language}and and T5 \cite{raffel2020exploring} are commonly used for tasks such as text summarization and translation, Vision-Language Models (VLMs) \cite{radford2021learningtransferablevisualmodels} are employed to generate textual descriptions from images, for text-to-Code Generation: Models like OpenAIs Codex \cite{chen2021evaluatinglargelanguagemodels} are designed to translate natural language prompts into executable code.
+In various generative tasks, different models are selected based on the specific requirements: Models like BART \cite{lewis2019bart}, GPT \cite{brown2020language}and and T5 \cite{raffel2020exploring} are commonly used for tasks such as text summarization and translation, Vision-Language Models (VLMs) \cite{radford2021learningtransferablevisualmodels} are employed to generate textual descriptions from images, for text-to-Code Generation: Models like OpenAIs Codex \cite{chen2021evaluatinglargelanguagemodels} are designed to translate natural language prompts into executable code.
 
@@
 	\large  % Increase overall font size
-	\begin{tabular}{|c|p{10cm}|}  % Use p{width} for text wrapping
+	\begin{tabular}{|c|p{5cm}|}  % Use p{width} for text wrapping
 		\hline
@@
 \end{table}
-
 \subsection{Mixture of Logits} 
-
-The Mixture of Logits (MoL)\cite{zhai2023revisiting,ding2024efficient} methode provides a flexible and adaptive mechanism for computing similarity scores between queries and items.
+The Mixture of Logits (MoL) \cite{zhai2023revisiting,ding2024efficient} method provides a flexible and adaptive mechanism for computing similarity scores between queries and items.
 Given a query \( q \) and an item \( x \), MoL assumes that both are mapped to \( P \) groups of low-rank embeddings, denoted as \( f_p(q) \) and \( g_p(x) \), respectively.
 These embeddings are parameterized by neural networks that extract features from the query and item.
-The similarity between \( q \) and \( x \) is computed as a weighted sum of the inner products of these embeddings, where the weights \( \pi_p(q, x) \)\cite{zhai2023revisiting,ding2024efficient} are adaptive gating parameters constrained to sum to 1:
-
+The similarity between \( q \) and \( x \) is computed as a weighted sum of the inner products of these embeddings, where the weights \( \pi_p(q,x)\) \cite{zhai2023revisiting,ding2024efficient} are adaptive gating parameters constrained to sum to 1:
 \begin{equation}
@@
 \end{equation}
-
 This formulation allows MoL to capture complex relationships between queries and items, making it particularly suitable for dynamic candidate selection tasks.
@@
 To efficiently scale MoL for large datasets and hardware-optimized implementations, the formulation is extended by decomposing the dot products into batched outer products of query-side and document-side embeddings.
-This decomposition improves computational efficiency, particularly on accelerators like GPUs, by normalizing the embeddings using the \( l_2 \)-norm:\cite{zhai2023revisiting}
+This decomposition improves computational efficiency, particularly on accelerators like GPUs, by normalizing the embeddings using the \( l_2 \)-norm: \cite{zhai2023revisiting}
 
@@
 \end{equation}
-
 Since embedding normalization can be precomputed, both formulations remain interchangeable in practical applications.
@@
 \end{equation}
-
 \noindent
 2.\textbf{ Initial Candidate Retrieval:} This step involves computing similarity scores between a query representation and item representations for each feature component $p \in P$.  
-
 \begin{equation}
@@
 	\begin{algorithmic}[1]
-
 		\State Set $G \gets \emptyset$     \Comment{Initial candidate set}
@@
 \end{algorithm}
-\newpage
 \subsection{Evaluation metrics in recommendation system}
@@
 These metrics are widely used in sequential recommendation tasks to evaluate the model's ability to rank relevant items higher in a user's preference list:
-
 \begin{itemize}
@@
 In order to measure the effectiveness of our proposed method, we perform a comprehensive evaluation of both the retrieval algorithm and the Mixture-of-Logits (MoL) approach.
-This assessment focuses on the next-item prediction task, a fundamental challenge in recommendation systems\cite{Zhu_2018}\cite{kang2018selfattentivesequentialrecommendation}, where the goal is to predict the most relevant item a user will interact with next based on their past interactions.
+This assessment focuses on the next-item prediction task, a fundamental challenge in recommendation systems \cite{Zhu_2018} \cite{kang2018selfattentivesequentialrecommendation}, where the goal is to predict the most relevant item a user will interact with next based on their past interactions.
 \subsection{Dataset description}
-We conducted experiments using two MovieLens datasets : ML-100K and ML-1M \cite{Harper2015}which are widely used benchmarks for evaluating sequential recommendation models as detailed in Table\ref{tab:movielens}.\\
+We conducted experiments using two MovieLens datasets : ML-100K and ML-1M \cite{Harper2015}which are widely used benchmarks for evaluating sequential recommendation models as detailed in Table\ref{tab:movielens}.
 \begin{table}[h]
@@
 	\renewcommand{\arraystretch}{1.2}
-	\begin{tabular}{|l|p{10cm}|}
+	\begin{tabular}{|l|p{5cm}|}
 		\hline
@@
 
-\setlength{\tabcolsep}{1pt} % Increase column padding
+%\setlength{\tabcolsep}{1pt} % Increase column padding
 %
@@
 
-\begin{table}[ht]
+\begin{table*}[h!]
 	\centering
@@
 	}
-\end{table}
+\end{table*}
 
-\setlength{\tabcolsep}{1pt} % Increase column padding
-
-%\begin{table}[ht]
-%	\centering
-%	\large
-%	\renewcommand{\arraystretch}{1.7} % Increase row spacing
-%	\resizebox{\textwidth}{!}{  % Scale table to fit width
-%		\begin{tabular}{|l|c|c|c|c|c|c|c|c|}
-%			\hline
-%			\textbf{Model} & \textbf{\shortstack{Max \\ Sequence Length}} & 
-%			\textbf{\shortstack{Embedding \\ Dimension}} & 
-%			\textbf{\shortstack{Number \\ of Heads}} & 
-%			\textbf{\shortstack{Feedforward \\ Dimension}} & 
-%			\textbf{\shortstack{Batch \\ Size}} & 
-%			\textbf{Epochs} & 
-%			\textbf{\shortstack{Val \\ Loss}} & 
-%			\textbf{\shortstack{Val \\ Accuracy}} \\
-%			\hline
-%			Hybrid+SAS & 50  & 128  & 2 & 128  & 128 & 10 & 4.5721 & 0.1530  \\ \hline
-%			MoL+SAS    & 50  & 128  & 2 & 128  & 128 & 10 & 4.5733 & 0.1526  \\ \hline
-%			Hybrid+SAS & 128 & 256  & 4 & 256  & 128 & 10 & 3.4152 & 0.3680  \\ \hline
-%			MoL+SAS    & 128 & 256  & 4 & 256  & 128 & 10 & 3.4671 & 0.3627  \\ \hline
-%			Hybrid+SAS & 512 & 512  & 4 & 512  & 128 & 10 & 1.0275 & {\bf 0.8350} \\ \hline
-%			MoL+SAS    & 512 & 512  & 4 & 512  & 128 & 10 & 1.0350 & 0.8276  \\ \hline
+%\setlength{\tabcolsep}{1pt} % Increase column padding
 %
-%		\end{tabular}
-%	}
-%	\caption{Results on 1M Movies dataset}
-%	\label{tab:1m_movies}
-%\end{table}
-\begin{table}[ht]
+%%\begin{table}[ht]
+%%	\centering
+%%	\large
+%%	\renewcommand{\arraystretch}{1.7} % Increase row spacing
+%%	\resizebox{\textwidth}{!}{  % Scale table to fit width
+%%		\begin{tabular}{|l|c|c|c|c|c|c|c|c|}
+%%			\hline
+%%			\textbf{Model} & \textbf{\shortstack{Max \\ Sequence Length}} & 
+%%			\textbf{\shortstack{Embedding \\ Dimension}} & 
+%%			\textbf{\shortstack{Number \\ of Heads}} & 
+%%			\textbf{\shortstack{Feedforward \\ Dimension}} & 
+%%			\textbf{\shortstack{Batch \\ Size}} & 
+%%			\textbf{Epochs} & 
+%%			\textbf{\shortstack{Val \\ Loss}} & 
+%%			\textbf{\shortstack{Val \\ Accuracy}} \\
+%%			\hline
+%%			Hybrid+SAS & 50  & 128  & 2 & 128  & 128 & 10 & 4.5721 & 0.1530  \\ \hline
+%%			MoL+SAS    & 50  & 128  & 2 & 128  & 128 & 10 & 4.5733 & 0.1526  \\ \hline
+%%			Hybrid+SAS & 128 & 256  & 4 & 256  & 128 & 10 & 3.4152 & 0.3680  \\ \hline
+%%			MoL+SAS    & 128 & 256  & 4 & 256  & 128 & 10 & 3.4671 & 0.3627  \\ \hline
+%%			Hybrid+SAS & 512 & 512  & 4 & 512  & 128 & 10 & 1.0275 & {\bf 0.8350} \\ \hline
+%%			MoL+SAS    & 512 & 512  & 4 & 512  & 128 & 10 & 1.0350 & 0.8276  \\ \hline
+%%			
+%%		\end{tabular}
+%%	}
+%%	\caption{Results on 1M Movies dataset}
+%%	\label{tab:1m_movies}
+%%\end{table}
+\begin{table*}[ht]
 	\centering
@@
 	}
-\end{table}
-
-\newpage
+\end{table*}
 Both approaches demonstrate significant performance improvements as model capacity increases, with larger configurations consistently delivering better results.
@@
 	\centering
-	\begin{subfigure}{0.48\textwidth}
+	\begin{subfigure}{0.48\linewidth}
 		\centering
@@
 	\hfill
-	\begin{subfigure}{0.48\textwidth}
+	\begin{subfigure}{0.48\linewidth}
 		\centering
@@
 	\centering
-	\includegraphics[width=0.7\textwidth]{HISTOGRAM.png}
+	\includegraphics[width=0.7\linewidth]{HISTOGRAM.png}
 	\caption{Comparison of Models Across Evaluation Metrics}\label{histo_p}
@@
 The improvements are consistent across both datasets, highlighting the robustness of our approach.\\
-%%=============================================%%
-%% For presentation purpose, we have included  %%
-%% \bigskip command. Please ignore this.       %%
-%%=============================================%%
-\bigskip
-\begin{minipage}{\hsize}%
-	\lstset{frame=single,framexleftmargin=-1pt,framexrightmargin=-17pt,framesep=12pt,linewidth=0.98\textwidth,language=pascal}% Set your language (you can change the language for each code-block optionally)
-	%%% Start your code-block
-
-\end{minipage}
-
-%Here is an example for \verb+\cite{...}+: \cite{bib1}.
+%%%=============================================%%
+%%% For presentation purpose, we have included  %%
+%%% \bigskip command. Please ignore this.       %%
+%%%=============================================%%
+%\bigskip
+%\begin{minipage}{\hsize}%
+%	\lstset{frame=single,framexleftmargin=-1pt,framexrightmargin=-17pt,framesep=12pt,linewidth=0.98\textwidth,language=pascal}% Set your language (you can change the language for each code-block optionally)
+%	%%% Start your code-block
+%	
+%\end{minipage}
+%
+%%Here is an example for \verb+\cite{...}+: \cite{bib1}.
 Another example for %\verb+\citep{...}+: \citep{bib2}.
@@
 (1990) and \verb+\citep{...}+ prints (Jones et al., 1990).
-
-%All cited bib entries are printed at the end of this article: \cite{bib3}, %\cite{bib4}, \cite{bib5}, \cite{bib6}, \cite{bib7}, \cite{bib8}, \cite{bib9}, %\cite{bib10}, \cite{bib11}, \cite{bib12} and \cite{bib13}.
-%%=============================================%%
-%% For presentation purpose, we have included  %%
-%% \bigskip command. Please ignore this.       %%
-%%=============================================%%
-
+%
+%%All cited bib entries are printed at the end of this article: \cite{bib3}, %\cite{bib4}, \cite{bib5}, \cite{bib6}, \cite{bib7}, \cite{bib8}, \cite{bib9}, %\cite{bib10}, \cite{bib11}, \cite{bib12} and \cite{bib13}.
+%%%=============================================%%
+%%% For presentation purpose, we have included  %%
+%%% \bigskip command. Please ignore this.       %%
+%%%=============================================%%
+%
 \section{Conclusion}\label{sec13}
@@
 Future work can explore enhancing our model with learning-based retrieval strategies such as reinforcement learning or attention mechanisms could refine the dynamic selection of K Additionally, testing our approach in real-world scenarios and conducting large-scale evaluations across diverse domains will help validate its robustness and adaptability in various knowledge bases and retrieval settings.
+\begin{dci}
+The authors have no Conflict of Interests to share.
+\end{dci}
+%%Harvard (name/date)
+\bibliographystyle{SageH}
+%%Vancouver (numbered)
+%\bibliographystyle{SageV}
+\bibliography{sn-bibliography}
 
@@
 %% the bibliography file.
-\bibliographystyle{ACM-Reference-Format}
-\bibliography{sn-bibliography}
+
+%
+%\begin{thebibliography}{99}
+%\bibitem[Kopka and Daly(2003)]{R1}
+%Kopka~H and Daly~PW (2003) \textit{A Guide to \LaTeX}, 4th~edn.
+%Addison-Wesley.
+%
+%\bibitem[Lamport(1994)]{R2}
+%Lamport~L (1994) \textit{\LaTeX: a Document Preparation System},
+%2nd~edn. Addison-Wesley.
+%
+%\bibitem[Mittelbach and Goossens(2004)]{R3}
+%Mittelbach~F and Goossens~M (2004) \textit{The \LaTeX\ Companion},
+%2nd~edn. Addison-Wesley.
+%
+%\end{thebibliography}
 
 \end{document}
-\endinput
-%%
-%% End of file `sample-acmlarge.tex'.
